### 2024-01-18 reports/mutation_load.Rmd
add mutation_load report, lower infercnv resources per sample

---
title: "Mutation loads in muscle ageing data from SComatic"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: true
    code_folding: show
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
params:
    do_rerun : TRUE

---

```{css style settings, echo = FALSE}
blockquote {
    padding: 10px 20px;
    margin: 0 0 20px;
    font-size: 14px;
    border-left: 5px solid #eee;
}
```

```{r setup, include = F, message = F, warning = F, class.source = 'fold-hide'}
# Rscript -e "rmarkdown::render('reports/mutation_load.Rmd', output_dir = 'out/mutation_load/')"
knitr::opts_knit$set(root.dir = '../')
knitr::opts_chunk$set(
  warning = FALSE, 
  dpi = 300, 
  message = FALSE,
  cache.path = 'reports/mutation_load_cache/')

# libraries
library(magrittr)
library(ggplot2)
library(gridExtra)
```

```{r plotting_fns, include = F, echo = F}
get_hq_cts <- 
  function(ct_info, min_n_cells_per_celltype_across_ids) {
    ct_info %>% 
      dplyr::group_by(id, lvl, annotation_level1_alias) %>% 
      dplyr::filter(min(n_cells_per_id_per_celltype) >= min_n_cells_per_celltype_across_ids) %>%
      dplyr::ungroup() %>%
      dplyr::select(id, lvl, annotation_level1_alias)
  }

colour_kbl_column <- function(kbl, df, col) {
  kbl %>% 
    kableExtra::column_spec(
      which(colnames(df) == col), 
    color = "white",
    background = kableExtra::spec_color(df[, col, drop = T]))
}

plot_vafs_density <-
  function(som_muts) {
    som_muts %>%
      dplyr::filter(lvl == 'donor_id') %>%
      {split(., .$lvl)} %>%
      purrr::map(function(df) {
       df %>% 
          dplyr::group_by(lvl, id, annotation_level1) %>%
          dplyr::mutate(n = dplyr::n()) %>%
          ggplot(aes(x = VAF, label = n, fill = age_bracket)) +
          geom_density() +
          geom_vline(aes(xintercept = median_VAF_per_id_per_celltype), colour = 'grey') +
          geom_text(aes(x = 0.6, y = Inf), vjust = 1.3) +
          ggh4x::facet_nested(lvl + age_bracket + id ~ annotation_level1, scales = 'free_y') +
          theme_bw() +
          theme(axis.text.x = element_text(angle = -90, hjust = 0),
                axis.text.y = element_blank(),
                strip.text.y.right = element_text(angle = 0, hjust = 0),
                panel.grid.major = element_blank(),
                panel.grid.minor = element_blank())
      }) %>%
      {do.call("grid.arrange", c(., ncol = length(.)))}
  }

plot_vafs_boxplot <-
  function(som_muts) {
    som_muts %>%
      {split(., .$lvl)} %>%
      purrr::map(function(df) {
       df %>% 
          ggplot(aes(x = VAF)) +
          geom_boxplot() +
          ggh4x::facet_nested(annotation_level1_alias ~ lvl + age_bracket + donor_id + id, scales = 'free_y') +
          theme_bw() +
          theme(axis.text.y = element_blank(),
                axis.text.x = element_text(angle = -90, hjust = 0),
                strip.text.y.right = element_text(angle = 0, hjust = 0),
                panel.grid.major = element_blank(),
                panel.grid.minor = element_blank())
      }) %>%
      {do.call("grid.arrange", c(., ncol = length(unique(som_muts$lvl))))}
  }

plot_ct_vs_mut_load_per_id_bar <- 
  function(som_mut_loads) {
    som_mut_loads %>%
    {split(., .$lvl)} %>%
      purrr::map(function(lvl_i) {
        lvl_i %>%
          ggplot(aes(x = annotation_level1, y = som_mut_load_per_Mb,
                     fill = age_bracket)) +
          geom_col() +
          theme_bw() +
          theme(axis.text.x = element_text(angle = 270, hjust = 0, vjust = 0.5, size = 5),
                legend.position = 'none') +
          ggh4x::facet_nested(. ~ lvl + age_bracket + donor_id + id) 
      }) %>%
    {do.call("grid.arrange", c(., ncol = 1))}
  }

plot_ct_vs_mut_load_per_id_bar2 <- 
  function(som_mut_loads) {
    som_mut_loads %>%
    {split(., .$lvl)} %>%
      purrr::map(function(lvl_i) {
        lvl_i %>%
          ggplot(aes(x = annotation_level1, y = som_mut_load_per_Mb, 
                     fill = age_bracket)) +
          geom_col(position = position_dodge2(preserve = "single")) +
          theme_minimal() +
          theme(axis.text.x = element_text(angle = -45, hjust = 0, vjust = 0.5, size = 10)) 
      }) %>%
    {do.call("grid.arrange", c(., ncol = 1))}
  }

plot_celltype_vs_mut_load_per_id_heatmap <- 
  function(som_mut_loads) {
    pre_mat <-
      som_mut_loads %>%
      dplyr::ungroup() %>%
      dplyr::select(som_mut_load_per_Mb, annotation_level1, id, age_bracket, lvl) %>%
      tidyr::pivot_wider(names_from = annotation_level1, values_from = som_mut_load_per_Mb) %>%
      tibble::column_to_rownames('id') 
    mat <- 
      pre_mat %>%
      dplyr::select(-c(age_bracket, lvl)) %>%
      as.matrix()
    annot <-
      pre_mat[, c('lvl', 'age_bracket')]
    pheatmap::pheatmap(
      mat, cluster_rows = F, cluster_cols = F,
      annotation_row = annot)
  }

plot_celltype_vs_mut_load_vs_n_cells_per_id_heatmap <-
  function(som_mut_loads) {
    som_mut_loads %>%
      ggplot(aes(x = id, y = annotation_level1, 
                 colour = som_mut_load_per_Mb, 
                 size = n_cells_per_id_per_celltype)) +
      geom_point() +
      theme_bw() +
      viridis::scale_color_viridis() +
      theme(axis.text.x = element_text(angle = -90, hjust = 0),
            panel.grid = element_blank()) +
      ggh4x::facet_nested(~ lvl + age_bracket, 
                          scales = 'free_x', space = 'free_x')
  }

plot_mut_load_btwn_replicates <- 
  function(som_mut_loads) {
    p_dat <-
      som_mut_loads %>%
      dplyr::filter(lvl == 'sample_id') %>%
      dplyr::group_by(donor_id, age_bracket) %>%
      dplyr::filter(dplyr::n_distinct(id) == 2) %>%
      dplyr::mutate(rep = paste0('rep', as.numeric(as.factor(id)))) %>%
      tidyr::pivot_wider(id_cols = c('donor_id', 'age_bracket', 'annotation_level1'),
                         names_from = rep, values_from = som_mut_load_per_Mb) 
    if (nrow(p_dat) > 0) {
      p_dat %>%
      #dplyr::filter(!is.na(rep1), !is.na(rep2)) %>%
      ggplot(aes(x = rep1, y = rep2, colour = annotation_level1, group = 1)) +
      geom_smooth(method = 'lm') +
      geom_point() + 
      ggh4x::facet_nested(age_bracket + donor_id ~ .) +
      theme_bw() +
      coord_equal()
    }
  }

plot_n_cells_vs_metrics <- 
  function(som_mut_loads) {
    som_mut_loads %>% 
      tidyr::pivot_longer(cols = c('n_callable_sites', 'n_som_muts', 'som_mut_load_per_Mb')) %>%
      ggplot(aes(x = n_cells_per_id_per_celltype, y = value,
                 colour = donor_id, shape = age_bracket)) +
      geom_point() +
      theme_bw() +
      theme(aspect.ratio = 1) +
      facet_wrap(lvl ~ name, scales = 'free_y') 
  }

plot_mutational_signatures <- function(som_muts) {
  # get all possible trios
  all_REF_trios <-
    c('A', 'T', 'G', 'C') %>%
    {tibble::tibble(up = ., REF_fix = ., down = ., ALT_fix = .)} %>%
    tidyr::expand(up, REF_fix, down, ALT_fix) %>%
    dplyr::filter(REF_fix %in% c('C', 'T'), ALT_fix != REF_fix) %>%
    dplyr::transmute(ALT_fix, REF_fix,
                     REF_trio = paste0(up, REF_fix, down),
                     REF_label = paste0(up, '**', REF_fix, '**', down)) %>%
    dplyr::cross_join(
      som_muts %>% 
        dplyr::ungroup() %>%
        dplyr::distinct(lvl, id, age_bracket) 
    )

  # mutation colours
  mut_cols <-
    c('C > A' = '#1493d2',
      'C > G' = '#070707',
      'C > T' = '#bb0f12',
      'T > A' = '#ababa5',
      'T > C' = '#6ab538',
      'T > G' = '#dfa99a')

  # plot
  som_muts %>%
    dplyr::filter(lvl == 'donor_id') %>%
    dplyr::mutate(dplyr::across(
      c('REF', 'ALT', 'Up_context', 'Down_context'), 
      ~ dplyr::case_when(REF %in% c('G', 'A') ~ chartr("ATGC", "TACG", .x),
                         TRUE ~ .x), 
      .names = "{.col}_fix"),
      REF_trio = paste0(substr(Up_context_fix, 5, 5), REF_fix, substr(Down_context_fix, 1, 1))) %>%
    dplyr::group_by(REF_trio, REF_fix, ALT_fix, id, age_bracket) %>%
    dplyr::count() %>%
    dplyr::group_by(id) %>%
    dplyr::mutate(prop = n / sum(n)) %>%
    dplyr::full_join(all_REF_trios) %>%
    dplyr::mutate(mutation = paste(REF_fix, '>', ALT_fix),
                  dplyr::across(c('prop', 'n'),
                                ~ dplyr::case_when(is.na(.x) ~ 0,
                                                   TRUE ~ .x))) %>%
    ggplot(aes(x = REF_label, y = prop, fill = mutation)) +
    geom_col() +
    ggh4x::facet_nested(age_bracket + id ~ mutation, scales = 'free_x') +
    theme_linedraw() +
    theme(axis.title.x = element_blank(),
          axis.text.x = ggtext::element_markdown(angle = 90, vjust = 0.5, family = 'cairo'),
          panel.grid.major.x = element_blank(),
          legend.position = 'none') +
    labs(y = 'relative contribution') +
    scale_fill_manual(values = mut_cols)
}

make_plots <- function(som_muts, som_mut_loads) {

  cat('\n\n**Density plot of somatic mutation VAFs**')
  plot_vafs_density(som_muts)

  cat('\n\n**Boxplot of somatic mutation VAFs**')
  plot_vafs_boxplot(som_muts)

  cat('\n\n**Barplot of somatic mutation load per ID, by level**\n')
  plot_ct_vs_mut_load_per_id_bar(som_mut_loads)

  cat('\n\n**Barplot of somatic mutation load per celltype, by level**\n')
  plot_ct_vs_mut_load_per_id_bar2(som_mut_loads)

  cat('\n\n**Heatmap of somatic mutation load per celltype, by level**\n')
  plot_celltype_vs_mut_load_per_id_heatmap(som_mut_loads)

  cat('\n\n**Heatmap of somatic mutation load per celltype vs n cells, by level**\n')
  print(plot_celltype_vs_mut_load_vs_n_cells_per_id_heatmap(som_mut_loads %>% dplyr::filter(lvl == 'donor_id')))

  cat('\n\n**Dotplot of somatic mutation load per Mb, compared between replicates of each donor**\n')
  print(plot_mut_load_btwn_replicates(som_mut_loads))

  cat('\n\n**Dotplot of n cells per level per celltype vs various metrics**\n')
  plot_n_cells_vs_metrics(som_mut_loads)

}
```

SComatic was run on scRNA-seq rheumatoid arthritis data using a modified version of kp9 and jc48’s [Nextflow implementation](https://github.com/Teichlab/scripts/tree/bd17ed22ec9e669056862d28c5f12a7e5033e68c/genotyping/scomatic/nextflow). 
Their implementation was intended to capture all germline and somatic variants in order to reconstruct tissue phylogeny, using both shared and private mutations for lineage tracing. In order to achieve this, they set `max_cell_types` to 1e6. I set it to 1. I also forced the pipeline to publish the celltype-split BAMs for each patient, which will be used to assess coverage of the driver genes in each cell type. The pipeline can be used with both GEX and ATAC modalities, hence the `*-GEX` suffix on the output directories.

Here is an example of the output files for donor BRI-401:

```
out/scomatic/BRI_401-GEX
├── BRI_401.calling.step2.intersect.tsv
├── BRI_401.calling.step2.pass.tsv
├── BRI_401.calling.step2.tsv
├── BRI_401.coverage_cell_count.per_chromosome.report.tsv
├── BRI_401.coverage_cell_count.report.tsv
├── cell_callable_sites
│   ├── BRI_401.B_cell_plasma_cell.SitesPerCell.tsv
│   ├── BRI_401.Endothelial_cell.SitesPerCell.tsv
│   ├── BRI_401.Myeloid_cell.SitesPerCell.tsv
│   ├── BRI_401.NK.SitesPerCell.tsv
│   ├── BRI_401.Stromal_cell.SitesPerCell.tsv
│   └── BRI_401.T_cell.SitesPerCell.tsv
└── celltype_bams
    ├── BRI_401.B_cell_plasma_cell.bam
    ├── BRI_401.B_cell_plasma_cell.bam.bai
    ├── BRI_401.Endothelial_cell.bam
    ├── BRI_401.Endothelial_cell.bam.bai
    ├── BRI_401.Myeloid_cell.bam
    ├── BRI_401.Myeloid_cell.bam.bai
    ├── BRI_401.NK.bam
    ├── BRI_401.NK.bam.bai
    ├── BRI_401.Stromal_cell.bam
    ├── BRI_401.Stromal_cell.bam.bai
    ├── BRI_401.T_cell.bam
    └── BRI_401.T_cell.bam.bai
```

### Load metadata

```{r sm_and_ct_info, results = F, message = F, class.source = 'fold-hide'}
# sample metadata
sm <-
  '../scomatic_vs_monopogen/out/metadata.tsv' %>%
  readr::read_tsv() %>%
  dplyr::select(-c(irods_bam_file, median_n_counts, median_n_genes, n_cells)) %>%
  dplyr::distinct() %>%
  dplyr::mutate(age_bracket = factor(age_bracket, levels = c('young', 'old')),
                id = ifelse(lvl == 'sample_id', paste0(donor_id, '_', id), id))

# celltype info
ct_info <-
  '../scomatic_vs_monopogen/out/celltypes_w_counts.tsv' %>%
  readr::read_tsv() %>%
  dplyr::mutate(donor_id2 = donor_id) %>%
  tidyr::pivot_longer(
    cols = c('donor_id', 'sample_id'),
    names_to = 'lvl',
    values_to = 'id'
  ) %>%
  dplyr::rename(donor_id = donor_id2) %>%
  dplyr::mutate(id = ifelse(lvl == 'sample_id', paste0(donor_id, '_', id), id)) %>%
  dplyr::group_by(lvl, id, donor_id, annotation_level1, annotation_level1_alias) %>%
  dplyr::summarise(n_cells_per_id_per_celltype = sum(n))

# sce object + coldata
sce <- 
  readRDS('/nfs/team205/vk8/projects/somatic_mut_benchmark/SKM_cells_data2benchmark_all.Rds')
col_data <- 
  sce@colData %>%
  tibble::as_tibble(rownames = 'CB') %>%
  dplyr::rename(sample_id = SampleID, donor_id = DonorID) %>%
  dplyr::mutate(sample_id = paste0(sample_id, '_', donor_id))

# cellranger qc
cr_qc <-
  '../scomatic_vs_monopogen/data/cellranger_web_summaries/counts.tsv' %>%
  readr::read_tsv()

# celltypes of interest
cts_oi <- c('MuSC', 'ArtEC',  'CapEC', 'VenEC', 'SMC', 'Pericyte',  'FB', 'Macrophage')
cts_pal <-
  ct_info %>%
  dplyr::group_by(annotation_level1) %>%
  dplyr::summarise(n = sum(n_cells_per_id_per_celltype)) %>%
  dplyr::arrange(-n) %>%
  dplyr::mutate(oi = ifelse(annotation_level1 %in% cts_oi, 'oi', 'noi')) %>%
  {split(.$annotation_level1, .$oi)}
cts_pal$oi <- rev(RColorBrewer::brewer.pal(length(cts_oi), 'Dark2')) %>% setNames(cts_pal$oi)
cts_pal$oi[1] <- '#0370b0'
cts_pal$noi <- rep('darkgrey', length(cts_pal$noi)) %>% setNames(cts_pal$noi)
cts_pal <- c(cts_pal$oi, cts_pal$noi)
```

#### Donor-level metadata

```{r donor_sm_df, echo = F}
sm %>%
  dplyr::filter(lvl == 'donor_id') %>%
  knitr::kable("html") %>% 
  kableExtra::kable_styling("striped")
```

#### Sample-level metadata

```{r sample_sm_df, echo = F}
sm %>%
  dplyr::filter(lvl == 'sample_id') %>%
  knitr::kable("html") %>% 
  kableExtra::kable_styling("striped") 
```

#### Cellranger QC

```{r cr_qc, echo = F}
cr_qc %>%
  knitr::kable("html") %>% 
  kableExtra::kable_styling("striped", full_width = F) 
```

#### Celltype-level counts

```{r ct_df, echo = F}
ct_info %>%
  knitr::kable("html") %>% 
  kableExtra::kable_styling("striped", full_width = F) %>% 
  kableExtra::scroll_box(width = "100%", height = "300px")
```

```{r plot_cts_vs_ids, eval = F}
p_dat <- 
  ct_info %>%
  dplyr::left_join(sm) %>%
  dplyr::filter(lvl == 'donor_id') %>%
  dplyr::group_by(annotation_level1) %>%
  dplyr::mutate(
    n_cells_per_celltype = sum(n_cells_per_id_per_celltype)) %>%
  dplyr::group_by(id) %>%
  dplyr::mutate(
    n_cells_per_id = sum(n_cells_per_id_per_celltype)) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    id = forcats::fct_reorder(id, n_cells_per_id),
    annotation_level1_lab = 
      ifelse(
        annotation_level1 %in% cts_oi,
        paste0('**', annotation_level1, '**'),
        annotation_level1) %>%
      forcats::fct_reorder(-n_cells_per_celltype),
    annotation_level1 = forcats::fct_reorder(annotation_level1, -n_cells_per_celltype),
    dplyr::across(
      .cols = dplyr::starts_with('n_'), 
      .fns = ~ prettyNum(.x, big.mark = ",", scientific = FALSE),
      .names = '{.col}_label'))

p_ids_x_cts <-
  p_dat %>%
  ggplot(aes(x = annotation_level1_lab, y = id, 
             colour = annotation_level1)) +
  geom_point(aes(size = n_cells_per_id_per_celltype, 
                 alpha = n_cells_per_id_per_celltype)) +
  scale_alpha(range = c(0.5, 1)) +
  scale_y_discrete(position = 'right')  +
  scale_x_discrete(position = 'bottom', guide = guide_axis(angle = -90)) +
  scale_size_continuous(range = c(1, 8)) +
  scale_colour_manual(guide = 'none', values = cts_pal) +
  theme_linedraw() +
  theme(legend.position = 'left', legend.title = element_blank(),
        axis.text.x = ggtext::element_markdown(angle = 90, vjust = 0.5),
        axis.title = element_blank(),
        strip.placement = 'outside') +
  facet_grid(age_bracket ~ ., scales = 'free_y', space = 'free_y', 
             switch = 'y')

p_ids <-
  p_dat %>%
  dplyr::mutate(annotation_level1 = factor(annotation_level1, levels = rev(names(cts_pal)))) %>%
  ggplot(aes(x = id, y = n_cells_per_id_per_celltype,
             fill = annotation_level1)) +
  geom_col() +
  geom_text(aes(x = id, y = n_cells_per_id, label = n_cells_per_id_label),
            size = 3.5,
            hjust = -0.1, vjust = 0.5) +
  coord_flip() +
  scale_y_continuous(
    expand = c(0, 0), 
    limits = c(0, 1.4 * max(p_dat$n_cells_per_id))) +
  scale_fill_manual(values = cts_pal) +
  theme_void() +
  theme(legend.position = 'none', 
        strip.text.y = element_blank()) +
  facet_grid(age_bracket ~ ., scales = 'free_y', space = 'free_y')

p_cts <-
  p_dat %>%
  dplyr::distinct(annotation_level1, 
                  n_cells_per_celltype, 
                  n_cells_per_celltype_label) %>%
  ggplot(aes(x = annotation_level1, 
             y = n_cells_per_celltype, 
             fill = annotation_level1)) +
  geom_col() +
  geom_text(aes(x = annotation_level1, y = n_cells_per_celltype, 
                label = n_cells_per_celltype_label),
            size = 5,
            hjust = 0, vjust = -0.1, angle = 45) +
  scale_y_continuous(
    expand = c(0, 0), 
    limits = c(0, 1.2 * max(p_dat$n_cells_per_celltype))) +
  scale_fill_manual(values = cts_pal) +
  theme_void() +
  theme(legend.position = 'none')

# plots
cowplot::plot_grid(p_ids_x_cts, p_ids, ncol = 2, align = 'h',
                   rel_widths = c(1, .3))
p_cts
```

#### Quality metrics

```{r qc_plots, fig.height = 8}
p_dat <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::left_join(sm %>% dplyr::mutate(DonorID = donor_id)) %>%
  dplyr::mutate(DonorID = forcats::fct_reorder(
                  paste0(DonorID, '\n(', X10X_version, ')'), age)) %>%
  dplyr::group_by(DonorID) %>%
  dplyr::mutate(n_cells = dplyr::n()) %>%
  dplyr::ungroup() %>%
  tidyr::pivot_longer(cols = c('age', 'n_counts', 'n_genes', 'n_cells')) %>%
  dplyr::mutate(
    name = gsub('_', ' ', name),
    name = ifelse(name %in% c('n cells', 'age'), name, 
                  paste0('log10(', name, ')')) %>%
      factor(levels = c('age', 'n cells', 'log10(n counts)', 'log10(n genes)')),
    value = ifelse(name %in% c('n cells', 'age'), value, log10(value)))

p_dat %>%
  ggplot(aes(x = DonorID, y = value, fill = age_bracket)) +
  geom_violin(data = p_dat %>% dplyr::filter(!(name %in% c('age', 'n cells'))),
              aes(fill = age_bracket, colour = age_bracket)) +
  geom_jitter(data = p_dat %>% dplyr::filter(!(name %in% c('age', 'n cells'))),
              height = 0, alpha = 0.02, size = 0.005) +
  geom_col(data = p_dat %>% 
             dplyr::filter(name %in% c('age', 'n cells')) %>%
             dplyr::distinct(DonorID, name, value, age_bracket)) +
  facet_grid(name ~ age_bracket, scales = 'free', space = 'free_x') +
  theme_linedraw() +
  theme(axis.title.y = element_blank(),
        legend.position = 'none') +
  xlab('Donor ID\n(10X version)')
```

### Load called mutations

```{r load_som_muts, eval = F}
scom <- xfun::cache_rds({

  # initiate lists
  scom <- list(
    all_muts = list(),
    som_muts = list(),
    coverages = list(),
    sc_muts = list(),
    annovar = list()
  )

  # load somatic mutations and coverages for each ID/lvl
  c('donor_id', 'sample_id') %>%
    purrr::walk(function(lvl) {

      dir <- paste0('out/', lvl)

      # initiate sublists
      scom$all_muts[[lvl]]  <<- list()
      scom$som_muts[[lvl]]  <<- list()
      scom$coverages[[lvl]] <<- list()
      scom$sc_muts[[lvl]]   <<- list()

      list.files(dir, pattern = '-GEX$') %>%
        purrr::walk(function(subdir) {

          id <- gsub('-GEX', '', subdir)
          file_prefix <- paste0(dir, '/', subdir, '/', id)
          col_suffix <- ifelse(lvl == 'sample_id', 
                               sub('.*?_', '', id),
                               id) %>% {paste0('_', .)}

          message(paste(lvl, id))

          # get all called muts
          scom$all_muts[[lvl]][[id]] <<-
            paste0(file_prefix, '.calling.step2.pass.tsv') %>%
            readr::read_tsv(comment = '##') %>%
            # separate longer the Cell_types column and pivot longer the
            # calling columns and then separate wider the INFO 
            # -> produces one row per variant per celltype
            tidyr::pivot_longer(dplyr::ends_with(col_suffix), names_to = 'ct', values_to = 'call') %>%
            tidyr::separate_rows(Cell_types, VAF, sep = ',') %>%
            tidyr::separate(
              'call', 
              into = paste0('ct_', c('DP', 'NC', 'CC', 'BC', 'BQ', 'BCf', 'BCr')),
              sep = '\\|',  remove = F, convert = T) %>%
            readr::type_convert() %>%
            dplyr::filter(ct == Cell_types) %>%
            # get celltype names
            dplyr::mutate(
              annotation_level1_alias = gsub(col_suffix, '', Cell_types))

          # get called somatic mutations (exclusive to celltypes)
          scom$som_muts[[lvl]][[id]] <<-
            scom$all_muts[[lvl]][[id]] %>%
            # get vars shared by <= max_n_celltypes
            dplyr::group_by(`#CHROM`, Start, End, REF, ALT) %>%
            dplyr::filter(dplyr::n_distinct(annotation_level1_alias) == 1) %>%
            # get median VAF per celltype
            dplyr::group_by(annotation_level1_alias) %>%
            dplyr::mutate(median_VAF_per_id_per_celltype = median(VAF)) %>%
            dplyr::ungroup()

          # get callable sites per celltype
          scom$coverages[[lvl]][[id]] <<-
            paste0(file_prefix, '.coverage_cell_count.per_chromosome.report.tsv') %>%
            readr::read_tsv() %>%
            # get celltype names
            dplyr::mutate(
              annotation_level1_alias = gsub(col_suffix, '', Cell_types))

          # get annovar gene annotations
          scom$annovar[[lvl]][[id]] <<-
            paste0(dir, '/', subdir, '/annovar/', 
                   id, '.variants.annovar.hg38_multianno.csv') %>%
            readr::read_csv()

          # get somatic mutations at single cell resolution
          scom$sc_muts[[lvl]][[id]] <<-
            paste0(dir, '/', subdir, '-genotypes') %>%
            list.files(full.names = T) %>%
            purrr::map(function(file) {
              # message(file)
              sc_df <-
                readr::read_tsv(file, ) %>%
                dplyr::filter(Cell_type_expected == Cell_type_observed,
                              ALT_expected == Base_observed) %>%
                # get celltype names
                dplyr::mutate(
                  annotation_level1_alias = gsub(
                    paste0('_', id), '', Cell_type_observed))

              # get callable sites per cell
              ct <- unique(sc_df$Cell_type_observed)
              if (length(ct) != 1) {
                message(paste(length(ct), 'celltypes in the single cell genotypes file! Should be 1!'))
                message(paste('File', file, 'will be skipped!'))
                sc_df <- tibble::tibble()
              } else {
                sc_coverage <- 
                  paste0(dir, '/', subdir, '/cell_callable_sites/', 
                         id, '.', ct, '.SitesPerCell.tsv') %>%
                  readr::read_csv()
                # add single cell-level coverage
                sc_df <- dplyr::left_join(sc_df, sc_coverage, by = 'CB')
              }

              # return
              sc_df

            }) %>%
            dplyr::bind_rows()

        })
    })

  # add sample and celltype metadata
  scom <-
    scom %>% 
    purrr::map(function(df_ls_ls) {
      df_ls_ls %>%
        purrr::map(function(df_ls) {
          df_ls[sapply(df_ls, function(x) dim(x)[1]) > 0] %>%
            dplyr::bind_rows(.id = 'id')
        }) %>%
        dplyr::bind_rows(.id = 'lvl') %>%
        # # add celltype info
        dplyr::left_join(ct_info) %>%
        # # add ID info
        dplyr::left_join(sm) %>%
        # add grouping structure
        dplyr::group_by(# sample info
                        lvl, id, donor_id, sex, age_bracket, age, X10X_version, n_cells_per_id,
                        # celltype info
                        annotation_level1_alias, annotation_level1, n_cells_per_id_per_celltype)
    })

  # return
  scom

}, file = 'scom.rds', rerun = do_rerun)

# assign dfs globally
all_muts <- xfun::cache_rds({scom$all_muts}, file = 'all_muts.rds', rerun = do_rerun)
som_muts <- xfun::cache_rds({scom$som_muts}, file = 'som_muts.rds', rerun = do_rerun)
coverages <- xfun::cache_rds({scom$coverages}, file = 'coverages.rds', rerun = do_rerun)
sc_muts <- xfun::cache_rds({scom$sc_muts}, file = 'sc_muts.rds', rerun = do_rerun)
annovar <- xfun::cache_rds({scom$annovar}, file = 'annovar.rds', rerun = do_rerun)
# list2env(scom, globalenv())
```

```{r load_up}
all_muts <- readRDS('reports/mutation_load_cache/all_muts.rds')
som_muts <- readRDS('reports/mutation_load_cache/som_muts.rds')
coverages <- readRDS('reports/mutation_load_cache/coverages.rds')
sc_muts <- readRDS('reports/mutation_load_cache/sc_muts.rds')
annovar <- readRDS('reports/mutation_load_cache/annovar.rds')
```

`*.calling.step2.pass.tsv` represents the somatic mutation calls at celltype resolution. These files look like this:

```{r pass_df, echo = F}
head(som_muts) %>%
  knitr::kable("html") %>% 
  kableExtra::kable_styling("striped") %>% 
  kableExtra::scroll_box(width = "100%", height = "300px")
```

The columns have the following definitions:

```{r pass_df_cols, echo = F}
'out/donor_id/339C-GEX/339C.calling.step2.pass.tsv' %>% 
  readr::read_delim(delim = ',Description=', skip = 1, col_names = F) %>%
  dplyr::filter(grepl('##INFO', X1)) %>%
  dplyr::transmute(
    col = gsub('##INFO=', '', X1),
    definition = gsub('>', '', X2)
  ) %>%
  knitr::kable("html") %>% 
  kableExtra::kable_styling("striped")
```

`*.coverage_cell_count.report.tsv` files look like this:

```{r cov_df, echo = F}
head(coverages, 10) %>%
  knitr::kable("html") %>% 
  kableExtra::kable_styling("striped")
```

The columns have the following definitions:

```{r cov_df_cols, echo = F}
tibble::tribble(
  ~col, ~definition,
  'Cell_types', 'celltype (Z)',
  'Cov', 'coverage (n)',
  'DP', 'number of sites in celltype Z with at least n reads',
  'NC', 'number of sites in celltype Z in at least n cells'
) %>%
  knitr::kable("html") %>% 
  kableExtra::kable_styling("striped")
```

`*.single_cell_genotypes.tsv` files look like this:

```{r sc_df, echo = F}
head(sc_muts) %>%
  knitr::kable("html") %>% 
  kableExtra::kable_styling("striped") %>% 
  kableExtra::scroll_box(width = "100%", height = "300px")
```

### Calculate somatic mutation loads within celltypes

Somatic mutation load was calculated according to [issue #23](https://github.com/cortes-ciriano-lab/SComatic/issues/23) on the SComatic GitHub.

> Regarding the "callable sites" question, it is important to take into account that there are two types of values when we speak about coverage (Cov column):
>
> 1. one based on the number of cells with at least one read at a given position (NC column)
>
> 2. another based on the depth or number of reads at each site (DP column).
>
> To clarify this concept and the file format, it is much easier to understand it with one example:
>
> In the B_memory showed in your attached figure, when Cov == 5:
>
> * DP shows the number of sites with 5 reads = 472874
>
> * NC shows the number of sites with 5 cells = 2242427
>
> You have this site counting value for each coverage (up to 150 by default). By looking at these values, you can get the number of callable sites based on the minimum coverage (Cov) that you want. For instance, if you want to get the callable sites with at least 10 cells, you should sum the column NC for all rows in the cell type with Cov >= 10.
>
> In our manuscript, we computed the mutation load at cell-type resolution by using a minimum Cov >= 5 and the next formula:
>
> `(# somatic mutations in the cell type Z) / (# callable sites in the cell type Z)`

`# somatic mutations in the cell type Z` can be calculated by "counting the number of rows in the file sample.calling.step2.pass.tsv for each cell type without any more filtering".

So, for celltype *Z* of ID *X*:

*somatic mutation load in Z = (n somatic mutations in Z) / (n callable sites in Z)*

* n somatic mutations in Z = `X.calling.step2.pass.tsv %>% count(Cell_types)`
* n callable sites in celltype Z = `X.coverage_cell_count.report.tsv %>% filter(NC > min_n_cells_per_site) %>% sum(NC per Z)`

**Steps:**

1. Get n somatic mutations per celltype per ID. Somatic mutations must be filtered to exclusive mutations within each celltype. Because SComatic was run with `max_cell_types = 100000`, the output will also contain germline and developmental (mosaic) mutations, shared between celltypes. We are interested in celltype-specific somatic mutations, so must remove these. This can be done by filtering the `Cell_types` column to only variants reported in one celltype. The instances of each celltype in the `Cell_types` column can then be counted to get n somatic mutations.
2. Get n callable sites per celltype per ID by summing the `NC` column of the coverage report.
3. Calculate mutational load per celltype per ID as `n_som_muts` / `n_callable_sites`.

This will be done at different cut-offs for coverage depth, number of cells, and label granularity.

```{r, echo = F}
tibble::tribble(
  ~arg, ~definition, ~default,
  'max_n_celltypes', 'Maximum n celltypes sharing the variant for it to be called as a somatic mutation.', '1',
  'min_n_cells', 'Minimum n cells in which a variant is detected for it to be called as a somatic mutation.', '5',
  'min_coverage_depth', 'Minimum n reads at a site at which a variant is detected for it to be called as a somatic mutation', '5',
  'min_n_cells_per_celltype_across_ids', 'Minimum number of cells for a celltype in all IDs for it to be interrogated for somatic mutations. If any ID has less than this number of cells in a celltype, the celltype is not interrogated. This is to ensure that only well-represented celltypes are included in comparisons.', '1',
  'min_n_callable_sites_per_celltype', 'Minimum number of callable sites for a cell type in an ID for it to be included.', '500,000',
  'label', 'The labelling level at which to group cells into celltypes. annotation_level0 is coarsest, annotation_level2 is finest.', 'annotation_level1'
) %>%
  knitr::kable("html") %>% 
  kableExtra::kable_styling("striped") 
```

```{r get_som_mut_loads}
get_som_mut_loads <- 
  function(
    som_muts,
    coverages,
    ct_info,
    min_n_cells = 5,
    min_coverage_depth = 5,
    min_n_cells_per_celltype_across_ids = 1,
    min_n_callable_sites_per_celltype = 500000,
    remove_XY = T,
    label = 'annotation_level1' # annotation_level2 = finer, annotation_level0 = coarser
  ) {
    # min_n_cells=5; min_coverage_depth=5;min_n_cells_per_celltype_across_ids=1;min_n_callable_sites_per_celltype=500000;remove_XY=T

    # check that no celltype names contain commas
    if(any(grepl(',', ct_info$annotation_level1))) stop("Commas in the celltype names!")

    # 1. get n somatic mutations
    som_mut <-
      som_muts %>%
      # i. filter out variants...
      #    - in cell types with insufficient coverage 
      #    - in cell types insufficient cells
      dplyr::filter(ct_DP >= min_coverage_depth,
                    ct_NC >= min_n_cells) %>%
      # ii. optionally filter out variants on the sex chromosomes
      { if (remove_XY == T) dplyr::filter(., !(`#CHROM` %in% c('X', 'Y'))) else . } %>%
      # ii. count mutations per celltype
      dplyr::count(name = 'n_som_muts')

    # 2. get n callable sites per celltype
    coverage <-
      coverages %>%
      # i. sum n callable sites per celltype 
      #    (with coverage >= min_n_cells, on autosomes)
      dplyr::filter(Cov >= min_n_cells) %>%
      { if(remove_XY == T) dplyr::filter(.,  !(CHROM %in% c('X', 'Y'))) else . } %>%
      dplyr::summarise(n_callable_sites = sum(NC)) %>%
      # ii. filter n callable sites >= min_n_callable_sites_per_celltype
      dplyr::filter(n_callable_sites >= min_n_callable_sites_per_celltype) %>%
      dplyr::group_by(lvl, id)

    # 3. calculate somatic mutation load
    som_mut_loads <-
      # i. join n somatic mutations and coverage by celltype
      dplyr::inner_join(som_mut, coverage) %>%
      # ii. calculate som mut load and som mut rate:
      #       if no callable sites ~ NA
      #       if callable sites ~ n_som_muts / n_callable_sites
      dplyr::mutate(
        som_mut_load = ifelse(
          n_callable_sites == 0, NA,
          n_som_muts / n_callable_sites),
        som_mut_load_per_Mb = som_mut_load * 1e6,
        som_mut_load_per_genome = som_mut_load * 3.2e9,
        som_mut_load_per_genome_per_yr = som_mut_load_per_genome / age) %>%
      # iii. filter out celltypes with insufficient cells in any one ID
      dplyr::inner_join(
        get_hq_cts(ct_info, min_n_cells_per_celltype_across_ids))

    # return
    som_mut_loads

    }
```

The function is applied to the somatic mutations at different cutoff stringencies.

```{r}
som_mut_loads <-
  get_som_mut_loads(
    som_muts, coverages, ct_info,
    min_n_cells = 5,
    min_coverage_depth = 5,
    min_n_cells_per_celltype_across_ids = 50,
    min_n_callable_sites_per_celltype = 500000
  )

som_mut_loads %>% 
    dplyr::filter(lvl == 'sample_id', annotation_level1 %in% cts_oi) %>% 
    ggplot(aes(y = som_mut_load_per_Mb, x = age_bracket, fill = age_bracket)) + 
    geom_boxplot() + 
    geom_point(alpha = 0.5, position = 'dodge') + 
    facet_grid(~ annotation_level1)

# save 
som_mut_loads %>%
  readr::write_tsv('out/som_mut_loads.tsv')
```

```{r eval = F, echo = F}
df <- 
  som_mut_loads %>%
  dplyr::arrange(-som_mut_load_per_Mb)  %>%
  dplyr::ungroup() %>%
  dplyr::select(
    lvl, id, annotation_level1_alias,
    n_callable_sites, n_som_muts, som_mut_load_per_Mb, n_cells_per_id_per_celltype
  )
df %>%
  kableExtra::kbl() %>%
  kableExtra::kable_styling("striped") %>% 
  colour_kbl_column(df, 'n_callable_sites') %>%
  colour_kbl_column(df, 'n_som_muts') %>%
  colour_kbl_column(df, 'som_mut_load_per_Mb') %>%
  colour_kbl_column(df, 'n_cells_per_id_per_celltype') %>% 
  kableExtra::scroll_box(width = "100%", height = "400px")
```

### Plots

```{r plots_final, results = 'asis', echo = F}
make_plots(som_muts, som_mut_loads)
```

### Plots of celltypes of interest

We are primarily interested in 8 celltypes:

* `r paste(cts_oi, collapse = '\n\n* ')`

```{r plots_cts_oi, results = 'asis', echo = F}
make_plots(
  som_muts %>% dplyr::filter(lvl == 'donor_id', annotation_level1_alias %in% cts_oi), 
  som_mut_loads %>% dplyr::filter(lvl == 'donor_id', annotation_level1_alias %in% cts_oi))
```

### Mutational signatures

```{r plot_mutational_signatures, echo = F, fig.width = 10, fig.height = 10, eval = F}
som_muts %>%
  dplyr::filter(lvl == 'donor_id') %>%
  plot_mutational_signatures()
```

### Mutation load at single cell resolution

How do we calculate the mutation load of every individual cell?

This question was asked and answered in [issue #30](https://github.com/cortes-ciriano-lab/SComatic/issues/30) on the SComatic GitHub:

> The basic strategy would be:
>
> 1. Compute how many rows per cell accomplish the `ALT_expected == Base_observed` and `Cell_type_expected == Cell_type_observed`. Basically, the number of mutations per cell.
> 2. It is essential to consider the number of callable sites per cell, as it will affect the number of mutations detected. To perform this correction, you will need to compute the number of callable sites per cell using `scripts/GetCallableSites/GetAllCallableSites.py`. You can use these callable sites to compute for example the mutation load per cell and MB (# Mutations per cell / # Callable sites per cell).
> 3. Plot the resulting values. I would ignore those cells with a very low number of callable sites. Generally, you will see an enrichment of cells at 0. This is due to the low number of callable sites per cell in this type of approaches (scRNA-seq) and a low mutation load (depending on the cancer or cell type).

```{r sc_som_mut_loads}
sc_som_mut_loads <- xfun::cache_rds({
  sc_muts %>%
    # only sites in the callable sites at cell type level
    dplyr::inner_join(
      all_muts %>%
        dplyr::ungroup() %>%
        dplyr::select(lvl, id, `#CHROM`, Start, End, REF, ALT, annotation_level1_alias)
    ) %>%
    dplyr::group_by(CB, SitesPerCell, .add = T) %>%
    dplyr::count(name = 'n_som_muts') %>%
    dplyr::mutate(som_mut_load = n_som_muts / SitesPerCell,
                  som_mut_load_per_Mb = som_mut_load * 1e6) 
}, file = 'sc_som_mut_loads.rds')
```

```{r}
GeomSplitViolin <- 
  ggproto("GeomSplitViolin", GeomViolin, draw_group = function(self, data, ..., draw_quantiles = NULL) {
  data <- transform(data, xminv = x - violinwidth * (x - xmin), xmaxv = x + violinwidth * (xmax - x))
  grp <- data[1, "group"]
  newdata <- plyr::arrange(transform(data, x = if (grp %% 2 == 1) xminv else xmaxv), if (grp %% 2 == 1) y else -y)
  newdata <- rbind(newdata[1, ], newdata, newdata[nrow(newdata), ], newdata[1, ])
  newdata[c(1, nrow(newdata) - 1, nrow(newdata)), "x"] <- round(newdata[1, "x"])

  if (length(draw_quantiles) > 0 & !scales::zero_range(range(data$y))) {
    stopifnot(all(draw_quantiles >= 0), all(draw_quantiles <=
      1))
    quantiles <- ggplot2:::create_quantile_segment_frame(data, draw_quantiles)
    aesthetics <- data[rep(1, nrow(quantiles)), setdiff(names(data), c("x", "y")), drop = FALSE]
    aesthetics$alpha <- rep(1, nrow(quantiles))
    both <- cbind(quantiles, aesthetics)
    quantile_grob <- GeomPath$draw_panel(both, ...)
    ggplot2:::ggname("geom_split_violin", grid::grobTree(GeomPolygon$draw_panel(newdata, ...), quantile_grob))
  }
  else {
    ggplot2:::ggname("geom_split_violin", GeomPolygon$draw_panel(newdata, ...))
  }
})
geom_split_violin <- function(mapping = NULL, data = NULL, stat = "ydensity", position = "identity", ..., 
                              draw_quantiles = NULL, trim = TRUE, scale = "area", na.rm = FALSE, 
                              show.legend = NA, inherit.aes = TRUE) {
  layer(data = data, mapping = mapping, stat = stat, geom = GeomSplitViolin, 
        position = position, show.legend = show.legend, inherit.aes = inherit.aes, 
        params = list(trim = trim, scale = scale, draw_quantiles = draw_quantiles, na.rm = na.rm, ...))
}
sc_som_mut_loads %>%
  dplyr::left_join(sm %>% dplyr::select(id, X10X_version)) %>%
  dplyr::filter(annotation_level1 %in% cts_oi) %>%
  ggplot(aes(x = annotation_level1, 
             y = som_mut_load_per_Mb, 
             fill = age_bracket)) + 
  geom_split_violin() +
  scale_x_discrete(guide = guide_axis(angle = -90)) +
  theme_bw() +
  facet_wrap(~ X10X_version)
```

We can look at the extreme cells to see how their somatic mutation loads are being derived.

```{r echo = F}
df <- 
  sc_som_mut_loads %>%
  dplyr::arrange(-som_mut_load_per_Mb)  %>%
  head(100) %>%
  dplyr::ungroup() %>%
  dplyr::select(
    lvl, id, annotation_level1_alias,
    n_som_muts, SitesPerCell, som_mut_load_per_Mb, n_cells_per_id_per_celltype
  )
df %>%
  kableExtra::kbl() %>%
  kableExtra::kable_styling("striped") %>% 
  colour_kbl_column(df, 'SitesPerCell') %>%
  colour_kbl_column(df, 'n_som_muts') %>%
  colour_kbl_column(df, 'som_mut_load_per_Mb') %>%
  colour_kbl_column(df, 'n_cells_per_id_per_celltype') %>% 
  kableExtra::scroll_box(width = "100%", height = "400px")
```

```{r plot_sc_som_mut_loads, echo = F, fig.width = 10}
sc_som_mut_loads %>% 
  dplyr::group_by(lvl, id, annotation_level1) %>% 
  dplyr::mutate(median_som_mut_load_per_Mb = median(som_mut_load_per_Mb)) %>%
  dplyr::filter(lvl == 'donor_id', annotation_level1_alias %in% cts_oi) %>%
  ggplot(aes(x = "", y = som_mut_load_per_Mb)) +
  geom_violin() +
  geom_jitter(alpha = 0.2, size = 0.2, height = 0,
              aes(colour = SitesPerCell)) +
  geom_hline(aes(yintercept = median_som_mut_load_per_Mb)) +
  theme(axis.text.x = element_text(angle = 270, hjust = 0, vjust = -0.5)) +
  scale_colour_viridis_b() +
  coord_flip() +
  ggh4x::facet_nested(age_bracket + id ~ annotation_level1) 
```

```{r plot_sc_som_mut_load_vs_variables}
sc_som_mut_loads %>%
  ggplot(aes(x = SitesPerCell, y = som_mut_load_per_Mb, 
             colour = donor_id)) +
  geom_point() +
  facet_grid(~ lvl)
```

```{r}
cb_x_som_mut <-
  tibble::tibble(CB = colnames(sce)) %>%
  dplyr::left_join(
    sc_som_mut_loads %>%
      dplyr::filter(lvl == 'donor_id') %>%
      dplyr::ungroup() %>%
      dplyr::transmute(
        CB = CB %>% stringi::stri_replace_last_fixed('_', '-'), 
        som_mut_load_per_Mb)) %>%
  tibble::column_to_rownames('CB')
sce$som_mut_load_per_Mb <- 
  cb_x_som_mut$som_mut_load_per_Mb
scater::plotReducedDim(sce, 'UMAP', colour_by = 'som_mut_load_per_Mb')
```

### Building a GLM

```{r glm}
p1 <-
  som_mut_loads %>%
  dplyr::left_join(sm) %>% 
  dplyr::filter(lvl == 'sample_id', annotation_level1 %in% cts_oi) %>% 
  ggplot(aes(x = n_cells_per_id_per_celltype, y = n_som_muts, colour = X10X_version, shape = age_bracket)) + 
  geom_point()
p2 <-
  som_mut_loads %>%
  dplyr::left_join(sm) %>% 
  dplyr::filter(lvl == 'sample_id', annotation_level1 %in% cts_oi) %>% 
  ggplot(aes(x = n_callable_sites, y = n_som_muts, colour = X10X_version, shape = age_bracket)) + 
  geom_point()
p3 <-
  som_mut_loads %>%
  dplyr::left_join(sm) %>% 
  dplyr::filter(lvl == 'sample_id', annotation_level1 %in% cts_oi) %>% 
  ggplot(aes(x = n_callable_sites, y = n_som_muts, colour = X10X_version, shape = age_bracket)) + 
  geom_point()
som_mut_loads %>%
  dplyr::left_join(sm) %>% 
  dplyr::inner_join(
    sce@colData %>% 
      tibble::as_tibble() %>% 
      dplyr::transmute(annotation_level1, id = paste0(DonorID, '_', SampleID), n_counts) %>%
      dplyr::group_by(id, annotation_level1) %>%
      dplyr::summarise(avg_n_counts = mean(n_counts))) %>%
  dplyr::filter(lvl == 'sample_id', annotation_level1 %in% cts_oi) %>% 
  ggplot(aes(x = avg_n_counts, y = som_mut_load_per_Mb, colour = X10X_version, shape = age_bracket)) + 
  geom_point()
som_mut_loads %>%
    dplyr::left_join(sm) %>% 
    dplyr::inner_join(
        sce@colData %>% 
            tibble::as_tibble() %>% 
            dplyr::transmute(annotation_level1, id = paste0(DonorID, '_', SampleID), n_counts) %>%
            dplyr::group_by(id, annotation_level1) %>%
            dplyr::summarise(avg_n_counts = mean(n_counts))) %>%
    dplyr::filter(lvl == 'sample_id', annotation_level1 %in% cts_oi, X10X_version == "3'v2") %>% 
    ggplot(aes(x = avg_n_counts, y = som_mut_load_per_Mb, colour = X10X_version, shape = age_bracket)) + 
    geom_point()

# n_callable_sites > 5e6
som_mut_loads %>%
    dplyr::left_join(sm) %>% 
    dplyr::filter(lvl == 'sample_id', annotation_level1 %in% cts_oi, X10X_version == "3'v2", n_callable_sites > 5e6) %>% 
    ggplot(aes(x = n_callable_sites, y = som_mut_load_per_Mb, colour = age_bracket, shape = age_bracket)) + 
    geom_point()

glm_dat <-
  som_mut_loads %>%
  dplyr::filter(annotation_level1 %in% cts_oi) %>%
  dplyr::left_join(sm) %>%
  dplyr::inner_join(
        sce@colData %>% 
            tibble::as_tibble() %>% 
            dplyr::transmute(annotation_level1, id = paste0(DonorID, '_', SampleID), n_counts) %>%
            dplyr::group_by(id, annotation_level1) %>%
            dplyr::summarise(avg_n_counts = mean(n_counts))) %>%
  dplyr::transmute(
    n_som_muts,
    som_mut_load_per_Mb,
    age_bracket = factor(age_bracket, levels = c('young', 'old')),
    n_cells_per_id_per_celltype,
    n_callable_sites,
    avg_n_counts,
    X10X_version)
glm_obj <-
  glm(
  n_som_muts ~
    # donor-level effects
    #age_bracket + annotation_level1 +
    age_bracket * annotation_level1 + 
    annotation_level1 + 
    avg_n_counts,
   # n_cells_per_id_per_celltype + 
    # cell type-level effects
    # n_callable_sites +
    # technical effects
    #X10X_version, 
  family = 'poisson',
  data = glm_dat
)
summary(glm_obj)
```

```{r fig.height = 8}
annovar %>%
  # get somatic muts only
  dplyr::inner_join(
    som_muts %>% dplyr::select(Chr = `#CHROM`, Start, End, Ref = REF, Alt = ALT)) %>%
  # get celltypes of interest
  dplyr::filter(annotation_level1 %in% cts_oi) %>%
  # get counts as fraction of all muts
  dplyr::mutate(n = dplyr::n(), `genomic region` = Func.refGene) %>%
  # get counts overall
  dplyr::group_by(`genomic region`) %>%
  dplyr::mutate(n = dplyr::n()) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(`genomic region` = forcats::fct_reorder(`genomic region`, -n)) %>%
  ggplot(aes(x = `genomic region`, fill = age_bracket)) +
  geom_bar() +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5),
        legend.position = 'none',
        panel.grid = element_blank()) +
  facet_grid(annotation_level1 ~ age_bracket)
```

```{r, fig.height = 4}
# deleterious mutations
annovar %>%
  # get somatic muts only
  dplyr::inner_join(
    som_muts %>% 
      dplyr::select(Chr = `#CHROM`, Start, End, Ref = REF, Alt = ALT)) %>% 
  # classify deleterious vs non deleterious
  dplyr::mutate(
    `deleterious?` = 
      ifelse((is.na(MetaLR_pred) & is.na(MetaSVM_pred)), 'unknown',
             ifelse((MetaLR_pred == "T" | MetaSVM_pred == 'T'), 'deleterious',
             'not deleterious'))) %>%
  dplyr::group_by(`deleterious?`, .add = T) %>%
  ggplot(aes(x = donor_id, fill = `deleterious?`)) + 
  geom_bar() +
  theme_linedraw() +
  theme(panel.grid = element_blank()) +
  scale_fill_manual(values = c('deleterious' = '#c2625b',
                               'not deleterious' = '#72b86e',
                               'unknown' = 'darkgrey')) +
  facet_grid(~ age_bracket, scales = 'free_x', space = 'free_x')
```

### Final plots

```{r boxplot, echo = F, fig.width = 7, fig.height = 3}
# boxplot som_mut_load_per_Mb vs age
som_mut_loads %>%
  dplyr::filter(lvl == 'sample_id') %>%
  ggplot(aes(x = age, y = som_mut_load_per_Mb, group = donor_id)) +
  geom_boxplot(outlier.alpha = 0) +
  geom_jitter(height = 0, size = 0.1, aes(colour = age_bracket)) +
  theme_linedraw() +
  theme(panel.grid.major.x  = element_blank(),
        panel.grid.minor.x = element_blank())
```

```{r sc_and_ct_levels, fig.height = 8}
# metrics at cell type and single cell level
p_dat <-
  som_mut_loads %>%
  tidyr::pivot_longer(
    cols = c('n_callable_sites', 
             'som_mut_load_per_Mb', 
             'n_som_muts')) %>%
  dplyr::mutate(level = 'cell type-level') %>%
  dplyr::bind_rows(
    sc_som_mut_loads %>% 
      dplyr::transmute(som_mut_load_per_Mb,
                       n_callable_sites = SitesPerCell,
                       n_som_muts = n_som_muts,
                       level = 'single cell-level') %>%
      tidyr::pivot_longer(
        cols = c('som_mut_load_per_Mb', 
                 'n_callable_sites',
                 'n_som_muts'))) %>%
  dplyr::filter(lvl == 'sample_id', annotation_level1 %in% cts_oi) %>%
  dplyr::group_by(annotation_level1) %>%
  dplyr::mutate(name = factor(gsub('_', ' ', name), 
                              levels = c('n callable sites', 'n som muts', 'som mut load per Mb')))

p_dat %>%
  ggplot(aes(x = age_bracket, y = value)) +
  geom_boxplot(data = p_dat %>% dplyr::filter(level == 'cell type-level'),
               outlier.alpha = 0) + 
  geom_jitter(data = p_dat %>% dplyr::filter(level == 'cell type-level'),
              aes(colour = age_bracket), height = 0, width = 0.3) + 
  geom_violin(data = p_dat %>% dplyr::filter(level == 'single cell-level'),
              draw_quantiles = 0.5) +
  geom_jitter(data = p_dat %>% dplyr::filter(level == 'single cell-level'),
              aes(colour = age_bracket), height = 0, width = 0.3, size = 0.1) + 
  theme_linedraw() +
  ggh4x::facet_nested(level + name ~ annotation_level1, scales = 'free_y') +
  labs(y = '', x = 'age bracket') + 
  theme(legend.position = 'none')
```

```{r metrics_vs_corrections}
# somatic mutation load vs n cells in the cell type
# somatic mutation load vs average n counts
# somatic mutation load vs n callable sites
p_ls <-
  som_mut_loads %>%
  # add qc metrics from sce
  dplyr::left_join(
    col_data %>%
      dplyr::group_by(sample_id, annotation_level1) %>%
      dplyr::summarise(dplyr::across(.cols = dplyr::starts_with('n_'),
                                     .fns = mean, .names = 'avg_{.col}'))
  ) %>%
  dplyr::rename(n_cells = n_cells_per_id_per_celltype) %>%
  dplyr::filter(lvl == 'sample_id') %>% 
  tidyr::pivot_longer(cols = c('n_callable_sites', 
                               'n_cells', 
                               'avg_n_counts'), 
                      names_to = 'x_var', values_to = 'x_value') %>%
  tidyr::pivot_longer(cols = c('n_som_muts', 'som_mut_load_per_Mb'),
                      names_to = 'y_var', values_to = 'y_value') %>%
  dplyr::arrange(y_var, x_var) %>%
  dplyr::mutate(split = paste0(y_var, '_vs_', x_var)) %>%
  dplyr::group_by(split) %>%
  split(., .$split) %>%
  purrr::map(function(df) {
    p <-
      df %>%
      ggplot(aes(x = x_value, y = y_value,
                 colour = donor_id, shape = age_bracket)) +
      geom_point() +
      theme_bw() +
      theme(aspect.ratio = 1) +
      labs(x = gsub('_', ' ', unique(df$x_var)), 
           y = gsub('_', ' ', unique(df$y_var)))
    p_legend <<- cowplot::get_legend(p)
    p + theme(legend.position = 'none')
  }) 
p_ls %>%
  patchwork::wrap_plots() 
patchwork::wrap_elements(p_legend)
```

```{r final_vafs, fig.height = 7}
som_muts %>%
  dplyr::filter(annotation_level1 %in% cts_oi, lvl == 'donor_id') %>%
  dplyr::group_by(id, annotation_level1) %>%
  dplyr::mutate(n = dplyr::n()) %>%
  ggplot(aes(x = VAF, label = n, fill = age_bracket)) +
  geom_density() +
  geom_vline(aes(xintercept = median_VAF_per_id_per_celltype), colour = 'grey') +
  geom_text(aes(x = 0.6, y = Inf), vjust = 1.3) +
  ggh4x::facet_nested(age_bracket + id ~ annotation_level1, scales = 'free_y') +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = -90, hjust = 0),
        axis.text.y = element_blank(),
        strip.text.y.right = element_text(angle = 0, hjust = 0),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = 'none')
```

```{r final_vafs_superimposed, fig.height = 7}
som_muts %>%
  dplyr::filter(annotation_level1 %in% cts_oi, lvl == 'donor_id') %>%
  dplyr::group_by(annotation_level1) %>%
  dplyr::mutate(n = dplyr::n(),
                median_VAF_per_celltype = median(VAF)) %>%
  ggplot(aes(x = VAF, label = n, fill = age_bracket, group = id)) +
  geom_density(alpha = 0.2) +
  geom_vline(aes(xintercept = median_VAF_per_celltype)) +
  geom_text(aes(x = 0.6, y = Inf), vjust = 1.3) +
  ggh4x::facet_grid2(annotation_level1 ~ age_bracket, 
                     scales = 'free_y', independent = "y",
                strip = ggh4x::strip_vanilla(size = "variable")) +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = -90, vjust = 0.5),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size = 5),
        axis.ticks.y = element_blank(),
        strip.text.y.right = element_text(angle = 0, hjust = 0),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = 'none')
```

### 2024-01-18 src/02_run_infercnv.R

@@ -1,23 +1,0 @@
-#!/bin/bash
-# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/ ; . ~/.bashrc ; mamba activate jupy ; /software/team205/bin/jsub lsf -q week -n infercnv -c 1 -m 2g -l log "bash src/02_run_infercnv.sh" | bsub
-
-set -eo pipefail
-
-# wd
-wd=/lustre/scratch125/casm/team268im/at31/RA_som_mut/
-cd $wd
-mkdir -p work/ out/infercnv/ log/
-
-# conda activate an environment which has docopt installed so jsub works
-. ~/.bashrc
-mamba activate jupy
-
-# run infercnv
-/software/team205/nextflow-23.04.1-all run nextflow/infercnv.nf \
-    --output_dir out/infercnv \
-    --mappings data/infercnv/mappings.csv \
-    --gene_order_file data/infercnv/hg38_gencode_v27.txt \
-    -c /nfs/team205/kp9/nextflow/scomatic/LSF.config  \
-    -w work/ \
-    -resume
-    

### 2024-01-19 scomatic/reports/scomatic.Rmd
restructure infercnv and scomatic runs into separate subdirectories

---
title: "Mutation loads in rheumatoid arthritis data from SComatic"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: true
    code_folding: show
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
params:
    do_rerun: TRUE
    celltype_counts: NULL
    mappings: NULL
---

```{css style settings, echo = FALSE}
blockquote {
    padding: 10px 20px;
    margin: 0 0 20px;
    font-size: 14px;
    border-left: 5px solid #eee;
}
```

```{r setup, include = F, message = F, warning = F, class.source = 'fold-hide'}
# params <- list(celltype_counts = 'data/scomatic/celltype_counts.tsv', do_rerun = T)
knitr::opts_knit$set(root.dir = '../')
knitr::opts_chunk$set(
  warning = FALSE, 
  dpi = 300, 
  message = FALSE,
  cache.path = 'reports/mutation_load_cache/')

# libraries
library(magrittr)
library(ggplot2)
library(gridExtra)
```

SComatic was run on scRNA-seq rheumatoid arthritis data using a modified version of kp9 and jc48’s [Nextflow implementation](https://github.com/Teichlab/scripts/tree/bd17ed22ec9e669056862d28c5f12a7e5033e68c/genotyping/scomatic/nextflow). 
Their implementation was intended to capture all germline and somatic variants in order to reconstruct tissue phylogeny, using both shared and private mutations for lineage tracing. In order to achieve this, they set `max_cell_types` to 1e6. I set it to 1. I also forced the pipeline to publish the celltype-split BAMs for each patient, which will be used to assess coverage of the driver genes in each cell type. The pipeline can be used with both GEX and ATAC modalities, hence the `*-GEX` suffix on the output directories.

Here is an example of the output files for donor BRI-401:

```
out/scomatic/BRI_401-GEX
├── BRI_401.calling.step2.intersect.tsv
├── BRI_401.calling.step2.pass.tsv
├── BRI_401.calling.step2.tsv
├── BRI_401.coverage_cell_count.per_chromosome.report.tsv
├── BRI_401.coverage_cell_count.report.tsv
├── cell_callable_sites
│   ├── BRI_401.B_cell_plasma_cell.SitesPerCell.tsv
│   ├── BRI_401.Endothelial_cell.SitesPerCell.tsv
│   ├── BRI_401.Myeloid_cell.SitesPerCell.tsv
│   ├── BRI_401.NK.SitesPerCell.tsv
│   ├── BRI_401.Stromal_cell.SitesPerCell.tsv
│   └── BRI_401.T_cell.SitesPerCell.tsv
└── celltype_bams
    ├── BRI_401.B_cell_plasma_cell.bam
    ├── BRI_401.B_cell_plasma_cell.bam.bai
    ├── BRI_401.Endothelial_cell.bam
    ├── BRI_401.Endothelial_cell.bam.bai
    ├── BRI_401.Myeloid_cell.bam
    ├── BRI_401.Myeloid_cell.bam.bai
    ├── BRI_401.NK.bam
    ├── BRI_401.NK.bam.bai
    ├── BRI_401.Stromal_cell.bam
    ├── BRI_401.Stromal_cell.bam.bai
    ├── BRI_401.T_cell.bam
    └── BRI_401.T_cell.bam.bai
```

### Sample metadata

```{r md}
ct_counts <-
    params$celltype_counts %>%
    readr::read_tsv()
mappings <-
    params$mappings %>%
    readr::read_csv()
```

### Celltype counts

```{r plot_cts_vs_ids, results = F, message = F}
p_dat <-
    ct_counts %>%
    dplyr::group_by(celltype) %>%
    dplyr::mutate(
        n_cells_per_celltype = sum(n_cells_per_id_per_celltype)) %>%
    dplyr::group_by(id) %>%
    dplyr::mutate(
        n_cells_per_id = sum(n_cells_per_id_per_celltype)) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(
        id = forcats::fct_reorder(id, n_cells_per_id),
        celltype = forcats::fct_reorder(celltype, -n_cells_per_celltype),
        dplyr::across(
          .cols = dplyr::starts_with('n_'), 
          .fns = ~ prettyNum(.x, big.mark = ",", scientific = FALSE),
          .names = '{.col}_label'))

p_ids_x_cts <-
  p_dat %>%
  ggplot(aes(x = id, y = celltype, 
             colour = celltype)) +
  geom_point(aes(size = n_cells_per_id_per_celltype, 
                 alpha = n_cells_per_id_per_celltype)) +
  scale_alpha(range = c(0.5, 1)) +
  scale_y_discrete(position = 'right')  +
  scale_x_discrete(position = 'bottom', guide = guide_axis(angle = -90)) +
  scale_size_continuous(range = c(1, 8)) +
  theme_linedraw() +
  theme(legend.position = 'left', legend.title = element_blank(),
        axis.text.x = ggtext::element_markdown(angle = 90, vjust = 0.5),
        axis.title = element_blank(),
        strip.placement = 'outside') 
```

### PASS/FAIL criteria

```{r pass_crit}
prefilter <-
    mappings$id %>%
    purrr::set_names() %>% 
    purrr::map(function(id) {
        step2 <- 
            paste0(params$output_dir, id, '-GEX') %>%
            list.files(pattern = '.calling.step2.tsv', full.names = T) 
        if (length(step2)) { readr::read_tsv(step2, comment = '##') }
    }) %>%
    dplyr::bind_rows(.id = 'id')
prefilter %>%
    dplyr::mutate(FILTER = FILTER == 'PASS') %>%
    dplyr::group_by(id) %>%
    dplyr::mutate(
        n_muts = dplyr::n(),
        id = forcats::fct_reorder(id, -n_muts)) %>%
    ggplot(aes(x = id, fill = FILTER)) +
    geom_bar() +
    theme(axis.text.x = element_text(angle = -90))
```

### Load called mutations

```{r load_som_muts}
scom <- xfun::cache_rds({

  # initiate lists
  scom <- list(
    all_muts = list(),
    som_muts = list(),
    coverages = list(),
    sc_muts = list(),
    annovar = list()
  )

  # load somatic mutations and coverages for each ID
  dir <- params$output_dir
  list.files(dir, pattern = '-GEX$') %>%
    purrr::walk(function(subdir) {

      id <- gsub('-GEX', '', subdir)
      file_prefix <- paste0(dir, '/', subdir, '/', id)
      message(id)

      # get all called muts
      scom$all_muts[[id]] <<-
        paste0(file_prefix, '.calling.step2.pass.tsv') %>%
        readr::read_tsv(comment = '##') %>%
        # separate longer the Cell_types column and pivot longer the
        # calling columns and then separate wider the INFO 
        # -> produces one row per variant per celltype
        tidyr::pivot_longer(!c(`#CHROM`:INFO), names_to = 'ct', values_to = 'call') %>%
        tidyr::separate_rows(Cell_types, VAF, sep = ',') %>%
        tidyr::separate(
          'call', 
          into = paste0('ct_', c('DP', 'NC', 'CC', 'BC', 'BQ', 'BCf', 'BCr')),
          sep = '\\|',  remove = F, convert = T) %>%
        readr::type_convert() %>%
        dplyr::filter(ct == Cell_types) %>%
        # get celltype names
        dplyr::mutate(
          annotation_level1_alias = gsub(col_suffix, '', Cell_types))

      # get called somatic mutations (exclusive to celltypes)
      scom$som_muts[[id]] <<-
        scom$all_muts[[id]] %>%
        # get vars shared by <= max_n_celltypes
        dplyr::group_by(`#CHROM`, Start, End, REF, ALT) %>%
        dplyr::filter(dplyr::n_distinct(annotation_level1_alias) == 1) %>%
        # get median VAF per celltype
        dplyr::group_by(annotation_level1_alias) %>%
        dplyr::mutate(median_VAF_per_id_per_celltype = median(VAF)) %>%
        dplyr::ungroup()

      # get callable sites per celltype
      scom$coverages[[id]] <<-
        paste0(file_prefix, '.coverage_cell_count.per_chromosome.report.tsv') %>%
        readr::read_tsv() %>%
        # get celltype names
        dplyr::mutate(
          annotation_level1_alias = gsub(col_suffix, '', Cell_types))

      # get annovar gene annotations
      scom$annovar[[id]] <<-
        paste0(dir, '/', subdir, '/annovar/', 
               id, '.variants.annovar.hg38_multianno.csv') %>%
        readr::read_csv()

      # get somatic mutations at single cell resolution
      scom$sc_muts[[id]] <<-
        paste0(dir, '/', subdir, '-genotypes') %>%
        list.files(full.names = T) %>%
        purrr::map(function(file) {
          # message(file)
          sc_df <-
            readr::read_tsv(file, ) %>%
            dplyr::filter(Cell_type_expected == Cell_type_observed,
                          ALT_expected == Base_observed) %>%
            # get celltype names
            dplyr::mutate(
              annotation_level1_alias = gsub(
                paste0('_', id), '', Cell_type_observed))

          # get callable sites per cell
          ct <- unique(sc_df$Cell_type_observed)
          if (length(ct) != 1) {
            message(paste(length(ct), 'celltypes in the single cell genotypes file! Should be 1!'))
            message(paste('File', file, 'will be skipped!'))
            sc_df <- tibble::tibble()
          } else {
            sc_coverage <- 
              paste0(dir, '/', subdir, '/cell_callable_sites/', 
                     id, '.', ct, '.SitesPerCell.tsv') %>%
              readr::read_csv()
            # add single cell-level coverage
            sc_df <- dplyr::left_join(sc_df, sc_coverage, by = 'CB')
          }

          # return
          sc_df

        }) %>%
        dplyr::bind_rows()

    })

  # add sample and celltype metadata
  scom <-
    scom %>% 
    purrr::map(function(df_ls) {
      df_ls[sapply(df_ls, function(x) dim(x)[1]) > 0] %>%
        dplyr::bind_rows(.id = 'id')
    }) %>%
        # # add celltype info
        dplyr::left_join(ct_info) %>%
        # # add ID info
        dplyr::left_join(sm) %>%
        # add grouping structure
        dplyr::group_by(# sample info
                        id, donor_id, sex, age_bracket, age, X10X_version, n_cells_per_id,
                        # celltype info
                        annotation_level1_alias, annotation_level1, n_cells_per_id_per_celltype)
    })

  # return
  scom

}, file = 'scom.rds', rerun = params$do_rerun)

# assign dfs globally
all_muts <- xfun::cache_rds({scom$all_muts}, file = 'all_muts.rds', rerun = do_rerun)
som_muts <- xfun::cache_rds({scom$som_muts}, file = 'som_muts.rds', rerun = do_rerun)
coverages <- xfun::cache_rds({scom$coverages}, file = 'coverages.rds', rerun = do_rerun)
sc_muts <- xfun::cache_rds({scom$sc_muts}, file = 'sc_muts.rds', rerun = do_rerun)
annovar <- xfun::cache_rds({scom$annovar}, file = 'annovar.rds', rerun = do_rerun)
```

### 2024-01-24 scomatic/reports/driver_coverage.Rmd
build up driver coverage rmd report, fix sample IDs issue in coverage calculations

---
title: "Mutation loads in rheumatoid arthritis data from SComatic"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: true
    code_folding: show
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
params:
    rerun: true
---

```{r setup, include = F, message = F, warning = F, class.source = 'fold-hide'}
# wd <- ifelse(Sys.info()['nodename'] == "mib119104s", '~/Volumes/', '') %>% paste0('/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/') ; setwd(wd)
# rmarkdown::render('reports/driver_coverage.Rmd', output_file = 'driver_coverage.html', output_dir = 'reports/', params = list(rerun = T))

knitr::opts_knit$set(root.dir = '../')
knitr::opts_chunk$set(
  warning = FALSE, 
  dpi = 300, 
  message = FALSE,
  cache.path = 'reports/driver_coverage_cache/')

# libraries
library(magrittr)
library(ggplot2)
library(gridExtra)

# dev.off() if there is a device
dev_off_if <- function() {if (length(dev.list()) != 0) { dev.off() }}
```

### Drivers

Genes of interest ('drivers') in rheumatoid arthritis are loaded. Their coverage in the dataset, across samples and cell types, will be checked.

```{r drivers, warning=F, message=F}
drivers <-
  readr::read_tsv('data/driver_genes/driver_gene_coords_for_coverage.tsv')
knitr::kable(drivers)
```

### Calculate coverage, by sample and by sample-x-celltype

We use `samtools depth` to get the coverage of each site in the genes of interest, both across each sample and within each cell type in each sample.

```{bash cov, eval=F}
# activate conda env
mamba activate jupy

# dirs
input_dir=/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/cellranger_output
ct_input_dir=out/Zhang2023/default_thresholds/
output_dir=out/Zhang2023/coverage/
drivers=data/driver_genes/driver_gene_coords_for_coverage.tsv
mkdir -p $output_dir

# iterate over each bam file and pileup specific regions
(
  echo -e 'chr\tpos\tdepth\tgene\tsample' ;
  for bam in $(find $input_dir -iname 'possorted_genome_bam.bam'); do
  	sample=$(echo $bam | cut -d "/" -f 10)
    while read -r gene coords ; do
      samtools depth -a -r $coords -Q 30 $bam |
      awk -v sample=$sample -v gene=$gene '{print $0"\t"gene"\t"sample}'
    done < <(sed 1d $drivers) ;
  done
) | cat > $output_dir/drivers_samtools_depth.tsv

# iterate over celltype-specific bam files and pileup specific regions
(
  echo -e 'chr\tpos\tdepth\tgene\tsample\tcelltype' ;
  for sample_dir in $ct_input_dir/*-GEX ; do
    sample=$(basename $sample_dir | sed 's/\-GEX$//g')
    for ct_bam in $sample_dir/celltype_bams/*.bam ; do
      ct=$(basename $ct_bam | cut -f2 -d.)
      while read -r gene coords ; do
        samtools depth -a -r $coords -Q 30 $ct_bam |
        awk -v sample=$sample -v gene=$gene -v ct=$ct \
        '{print $0"\t"gene"\t"sample"\t"ct}'
      done < <(sed 1d $drivers) ;
    done
  done
) | cat > $output_dir/drivers_samtools_depth_by_celltype.tsv
```

We read in the results at both levels.

```{r read_cov}
cov <- xfun::cache_rds({
  cov_files <-
    list.files(
      'out/Zhang2023/coverage/', 
      pattern = 'drivers_samtools_depth.*.tsv',
      full.names = T) 
  cov <-
    cov_files %>%
    setNames(cov_files %>% basename() %>% tools::file_path_sans_ext()) %>%
    purrr::map(readr::read_tsv) %>%
    dplyr::bind_rows(.id = 'lvl') %>%
    dplyr::mutate(chr = as.character(chr))
  cov
}, file = 'cov.rds', rerun = params$rerun)
```

### Annotate exons and introns

We use the reference object from DNDSCV to delimit intronic and exonic regions of the genes.

```{r features}
# annotate gene features
cov <- xfun::cache_rds({

  # load rda to get ref_cds object from DNDSCV
  refcds_38 <- '/lustre/scratch125/casm/team268im/fa8/119/DNDSCV_COVARIATES/refcds_GRCh38-GencodeV18+Appris.rda'
  load(refcds_38)

  # set gene names
  ref_cds <-
    RefCDS %>%
    setNames(RefCDS %>% purrr::map(~ .x$gene_name) %>% unlist())

  # annotate exons
  exons <-
    ref_cds[c(drivers$gene)] %>%
    purrr::map(~ .x$intervals_cds %>% 
                 tibble::as_tibble() %>%
                 dplyr::transmute(chr = .x$chr, start = V1, end = V2) %>%
                 dplyr::mutate(exon = paste0('exon', dplyr::row_number()))) %>%
    dplyr::bind_rows(.id = 'gene') %>%
    dplyr::mutate(pos = purrr::map2(start, end, ~ .x:.y)) %>%
    tidyr::unnest(cols = 'pos') %>%
    dplyr::select(-start, -end)
  cov <-
    cov %>%
    dplyr::left_join(exons) %>%
    dplyr::mutate(exon = ifelse(is.na(exon), 'intron', exon),
                  feature = ifelse(exon == 'intron', 'intron', 'exon'))

  # return
  cov

}, file = 'cov_w_features.rds', rerun = params$rerun)
```

```{r plots}
# get exon coords for plotting
exon_coords <- xfun::cache_rds({
  exon_coords <-
    cov %>%
    dplyr::filter(feature == 'exon') %>%
    dplyr::group_by(gene, exon) %>%
    dplyr::summarise(start = min(pos), 
                     end = max(pos))
  exon_coords
}, file = 'exon_coords.rds', rerun = params$rerun)

# function: plot all
plot_all <- function(p_df, p_exon_coords, info) {
  ggplot() +
    geom_rect(
      data = p_exon_coords,
      aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf),
      colour = 'lightblue', fill = 'lightblue'
    ) +
    geom_col(
      data = p_df,
      aes(x = pos, y = depth, fill = feature)) +
    theme_classic() +
    scale_fill_manual(values = c(exon = 'darkblue', intron = 'darkgrey')) +
    labs(title = paste(info, '(all)'))
}

# function: plot exons 
plot_exons <- function(p_df, title) {
  p_df %>%
    # get only exonic sequence
    dplyr::filter(feature == 'exon') %>%
    # fix exon ordering
    dplyr::mutate(
      exon = exon %>% forcats::fct_reorder(exon %>% gsub('exon', '', .) %>% as.numeric())
    ) %>%
    ggplot(aes(x = pos, y = depth)) +
    geom_col(fill = 'darkblue') +
    facet_grid(. ~ exon, space = 'free_x', scales = 'free_x') +
    theme_classic() +
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          panel.spacing = unit(.05, "lines"),
          panel.border = element_rect(fill = NA)) +
    labs(title = paste(info, '(exons)'))
}

# create list of plots by gene x sample
p <- xfun::cache_rds({
  # initiate plot list
  p <- list()

  cov %>% 
    dplyr::group_by(gene, sample) %>%
    dplyr::group_split() %>%
    purrr::map(function(df_i) {

      # for testing:
      # cov %>% dplyr::group_by(gene, sample) %>% dplyr::group_split() %>% {.[[1]]} -> df_i

      # get info
      i <- 
        df_i %>%
        dplyr::distinct(gene, sample)
      info <- t(i[1,]) %>% paste(collapse = '.')
      p[[info]] <<- list()

      list(by_sample = list(lvl = 'drivers_samtools_depth', title_suffix = ''),
           by_celltype = list(lvl = 'drivers_samtools_depth', title_suffix = '_by_celltype')) %>%
        purrr::map(function(x){

          # plot all
          p_all <-
            plot_all(
              p_df = df_i %>% dplyr::filter(lvl == x$lvl),
              p_exon_coords = exon_coords %>% dplyr::filter(gene == i$gene),
              paste0(info, x$title_suffix)
            )

          # plot exons only
          p_all <<-
            plot_exons(
              p_df = df_i %>% dplyr::filter(lvl == x$lvl),
              paste0(info, x$title_suffix)
            )

          # facet grid for celltypes
          if (title_suffix == '_by_celltype') {
            p_all <- p_all + facet_grid(~ celltype)
            p_exons <- p_exons + facet_grid(~ celltype)
          }

          # assign globally
          p[[info]][[paste0('all', title_suffix)]] <<- p_all
          p[[info]][[paste0('exons', title_suffix)]] <<- p_exons

        })

    })

  # return 
  p

}, file = 'plots.rds', rerun = params$rerun)
```

```{r knit_plots, echo = F, warning = F, message = F, results = 'asis'}
chunks <- c()

for (info in names(p)) {
  chunks <- c(chunks, paste0('\n####', info))

  knit_plot <- c(
    '',
    paste0('```{r ', info, ', class.source = "fold-hide", warning = F, message = F, ',
           'fig.dim = c(20, ', 3 * length(p[[info]]), ')}'),
    paste0('do.call("grid.arrange", c(p[["', info, '"]], ncol = 1))')
    '```',
    '')
  knitted_plot <- knitr::knit_child(text = knit_plot, envir = environment(), quiet = T)
  dev_off_if()

  chunks <- c(chunks, knitted_plot)
}
dev_off_if()

# evaluate chunks
cat(unlist(chunks), sep = '\n')
```

### 2024-01-24 infercnv/src/00_preprocess.R

@@ -20,7 +20,7 @@ list.files(
     dplyr::transmute(
         sample_id,
         raw_counts_matrix = paste0(processed_dir, 'raw_mRNA_count_matrix.rds'),
-        annotations = paste0(wd, '/data/annotations/', sample_id, '.tsv'),
+        annotations = paste0(wd, '/data/Zhang2023/annotations/', sample_id, '.tsv'),
         id = sample_id) %>%
     readr::write_csv('data/Zhang2023/mappings.csv')
 
@@ -36,8 +36,8 @@ dat %>%
         df %>%
             dplyr::select(cell, cell_type) %>%
             readr::write_tsv(
-            paste0('data/Zhang2023/annotations/', sample, '.tsv'),
-            col_names = F
+              paste0('data/Zhang2023/annotations/', sample, '.tsv'),
+              col_names = F
             )
     })
 

### 2024-01-24 infercnv/src/01_run_infercnv.sh

@@ -13,7 +13,7 @@ mkdir -p work/ out/ log/
 mamba activate jupy
 
 # run infercnv
-/software/team205/nextflow-23.04.1-all run infercnv.nf \
+/software/team205/nextflow-23.04.1-all run nextflow/infercnv.nf \
     -c config/Zhang2023.config \
     -c config/infercnv.config \
     -c /nfs/team205/kp9/nextflow/scomatic/LSF.config  \

### 2024-01-24 scomatic/src/02_run_scomatic_low_thresholds.sh

@@ -1,5 +1,5 @@
 #!/bin/bash
-# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/ ; . ~/.bashrc ; mamba activate jupy ; /software/team205/bin/jsub lsf -q week -n scomatic -c 1 -m 2g -l log "bash src/01_run_scomatic.sh" | bsub
+# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/ ; . ~/.bashrc ; mamba activate jupy ; /software/team205/bin/jsub lsf -q week -n scomatic_lo -c 1 -m 2g -l log "bash src/02_run_scomatic_low_thresholds.sh" | bsub
 
 set -eo pipefail
 
@@ -18,9 +18,9 @@ mamba activate jupy
     --projectDir $PWD \
     -params-file $config_dir/GEX.json \
     -c config/Zhang2023.config \
+    -c config/low_thresholds.config \
     -c $config_dir/LSF.config \
     -w work/ \
-    --output_dir out/Zhang2023/default_thresholds/ \
     --location local \
     -resume
 
@@ -30,8 +30,8 @@ mamba activate jupy
     --projectDir $PWD \
     -params-file $config_dir/GEX.json \
     -c config/Zhang2023.config \
+    -c config/low_thresholds.config \
     -c $config_dir/LSF.config \
     -w work/ \
-    --output_dir out/Zhang2023/default_thresholds/ \
     --location local \
     -resume

### 2024-01-24 scomatic/src/03_get_driver_coverage.sh

@@ -1,0 +1,12 @@
+#!/bin/bash
+# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/ ; . ~/.bashrc ; mamba activate jupy ; /software/team205/bin/jsub lsf -q week -n driver_coverage -c 1 -m 10g -l log "bash src/03_knit_driver_coverage.sh" | bsub
+
+# wd
+wd=/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/
+cd $wd
+
+# conda activate an environment which has docopt installed so jsub works
+. ~/.bashrc
+mamba activate jupy
+
+# knit

### 2024-01-24 scomatic/src/03_get_driver_coverage.sh
rename knit driver coverage

@@ -1,12 +1,0 @@
-#!/bin/bash
-# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/ ; . ~/.bashrc ; mamba activate jupy ; /software/team205/bin/jsub lsf -q week -n driver_coverage -c 1 -m 10g -l log "bash src/03_knit_driver_coverage.sh" | bsub
-
-# wd
-wd=/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/
-cd $wd
-
-# conda activate an environment which has docopt installed so jsub works
-. ~/.bashrc
-mamba activate jupy
-
-# knit

### 2024-01-26 scomatic/reports/driver_coverage.Rmd
fix infercnv checkifexists issue, update driver coverage report, add knitting src

@@
   for bam in $(find $input_dir -iname 'possorted_genome_bam.bam'); do
-  	sample=$(echo $bam | cut -d "/" -f 10)
+  	sample=$(echo $bam | cut -d "/" -f 10 | sed 's/\-/_/g')
     while read -r gene coords ; do
@@
     cov %>%
-    dplyr::left_join(exons) %>%
+    dplyr::left_join(exons, by = c('chr', 'pos', 'gene')) %>%
     dplyr::mutate(exon = ifelse(is.na(exon), 'intron', exon),
@@
 # function: plot all
-plot_all <- function(p_df, p_exon_coords, info) {
+plot_all <- function(p_df, p_exon_coords, p_title) {
   ggplot() +
@@
     theme_classic() +
+    theme(panel.border = element_rect(fill = NA),
+          panel.spacing = unit(.05, 'lines')) +
     scale_fill_manual(values = c(exon = 'darkblue', intron = 'darkgrey')) +
-    labs(title = paste(info, '(all)'))
+    labs(title = paste(p_title, '(all)')) +
+    facet_grid(celltype ~ ., space = 'free_x', scales = 'free_x') 
 }
@@
 # function: plot exons 
-plot_exons <- function(p_df, title) {
+plot_exons <- function(p_df, p_title) {
   p_df %>%
@@
     ) %>%
-    ggplot(aes(x = pos, y = depth)) +
+    ggplot(aes(x = pos, y = depth, group = celltype)) +
     geom_col(fill = 'darkblue') +
-    facet_grid(. ~ exon, space = 'free_x', scales = 'free_x') +
     theme_classic() +
@@
           axis.ticks.x = element_blank(),
-          panel.spacing = unit(.05, "lines"),
+          panel.spacing = unit(.05, 'lines'),
           panel.border = element_rect(fill = NA)) +
-    labs(title = paste(info, '(exons)'))
+    labs(title = paste(p_title, '(exons)')) +
+    facet_grid(celltype ~ exon, space = 'free_x', scales = 'free_x') 
 }
@@
 p <- xfun::cache_rds({
+
   # initiate plot list
   p <- list()
+  n_cts <- list()
 
@@
 
+      # get number of celltypes for plot dims
+      p[[info]]$n_cts <<- 
+        df_i %>% 
+        dplyr::filter(!is.na(celltype)) %>% 
+        dplyr::pull(celltype) %>% 
+        dplyr::n_distinct()
+
+      # plot by sample and by celltype
       list(by_sample = list(lvl = 'drivers_samtools_depth', title_suffix = ''),
-           by_celltype = list(lvl = 'drivers_samtools_depth', title_suffix = '_by_celltype')) %>%
+           by_celltype = list(lvl = 'drivers_samtools_depth_by_celltype', title_suffix = '_by_celltype')) %>%
         purrr::map(function(x){
@@
           # plot all
-          p_all <-
+          p[[info]][[paste0('all', x$title_suffix)]] <<-
             plot_all(
@@
               p_exon_coords = exon_coords %>% dplyr::filter(gene == i$gene),
-              paste0(info, x$title_suffix)
+              p_title = paste0(info, x$title_suffix)
             )
@@
           # plot exons only
-          p_all <<-
+          p[[info]][[paste0('exons', x$title_suffix)]] <<-
             plot_exons(
               p_df = df_i %>% dplyr::filter(lvl == x$lvl),
-              paste0(info, x$title_suffix)
+              p_title = paste0(info, x$title_suffix)
             )
-
-          # facet grid for celltypes
-          if (title_suffix == '_by_celltype') {
-            p_all <- p_all + facet_grid(~ celltype)
-            p_exons <- p_exons + facet_grid(~ celltype)
-          }
-
-          # assign globally
-          p[[info]][[paste0('all', title_suffix)]] <<- p_all
-          p[[info]][[paste0('exons', title_suffix)]] <<- p_exons
 
@@
   chunks <- c(chunks, paste0('\n####', info))
+  message(info)
 
@@
     paste0('```{r ', info, ', class.source = "fold-hide", warning = F, message = F, ',
-           'fig.dim = c(20, ', 3 * length(p[[info]]), ')}'),
-    paste0('do.call("grid.arrange", c(p[["', info, '"]], ncol = 1))')
+           'fig.dim = c(20, ', 4 + 4 * p[[info]]$n_cts, ')}'),
+    paste0('p_i <- p[["', info, '"]]'),
+    paste0('grid.arrange(p_i$all, p_i$all_by_celltype, p_i$exons, p_i$exons_by_celltype, ncol = 1, ',
+           'heights = c(2, p_i$n_cts * 2, 2, p_i$n_cts * 2))'),
     '```',
@@
 ```
+
+```{r mean_coverage}
+cov %>%
+  dplyr::filter(lvl == 'drivers_samtools_depth') %>%
+  dplyr::group_by(sample, gene, feature) %>%
+  dplyr::summarise(mean_depth = mean(depth)) %>%
+  ggplot(aes(x = sample, y = mean_depth, fill = feature)) +
+  geom_col(position = 'dodge') +
+  facet_grid(gene ~ .) +
+  theme_classic()
+```

### 2024-01-26 infercnv/src/00_preprocess.R

@@ -2,7 +2,9 @@
 library(magrittr)
 
 # dirs
-wd <- '/lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/'
+wd <- 
+  ifelse(Sys.info()['nodename'] == "mib119104s", '~/Volumes/', '') %>%
+  paste0('/lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/') 
 setwd(wd)
 
 # Zhang 2023 ---
@@ -11,35 +13,50 @@ cellranger_dir <- paste0(dat_dir, 'cellranger_output/')
 processed_dir <- paste0(dat_dir, 'processed_output/')
 dir.create('data/Zhang2023/annotations/', recursive = T)
 
-# mappings CSV (sample_id,raw_counts_matrix,annotations,id) ----
-list.files(
-        cellranger_dir,
-        pattern = '^BRI\\-',
-        include.dirs = T) %>%
-    tibble::enframe(value = 'sample_id') %>%
-    dplyr::transmute(
-        sample_id,
-        raw_counts_matrix = paste0(processed_dir, 'raw_mRNA_count_matrix.rds'),
-        annotations = paste0(wd, '/data/Zhang2023/annotations/', sample_id, '.tsv'),
-        id = sample_id) %>%
-    readr::write_csv('data/Zhang2023/mappings.csv')
-
 # annotations TSVs (cell celltype) ----
-dat <-
-    readRDS(paste0(processed_dir, 'all_cells_reference.rds'))$meta_data
-dat %>%
-    dplyr::select(cell, cell_type, sample) %>%
-    dplyr::group_by(sample) %>%
-    dplyr::group_split() %>%
-    purrr::walk(function(df) {
-        sample <- unique(df$sample)
-        df %>%
-            dplyr::select(cell, cell_type) %>%
-            readr::write_tsv(
-              paste0('data/Zhang2023/annotations/', sample, '.tsv'),
-              col_names = F
-            )
-    })
+annotations <-
+  readRDS(paste0(processed_dir, 'all_cells_reference.rds'))$meta_data %>%
+  dplyr::select(cell, cell_type, sample_id = sample) 
+annotations %>%
+  dplyr::group_by(sample_id) %>%
+  dplyr::group_split() %>%
+  purrr::walk(function(df) {
+      sample_id <- unique(df$sample_id)
+      df %>%
+          dplyr::select(cell, cell_type) %>%
+          readr::write_tsv(
[…]

### 2024-01-26 infercnv/src/01_run_infercnv.sh

@@ -17,6 +17,5 @@ mamba activate jupy
     -c config/Zhang2023.config \
     -c config/infercnv.config \
     -c /nfs/team205/kp9/nextflow/scomatic/LSF.config  \
-    -w work/ \
-    -resume
+    -w work/ # -resume
     

### 2024-01-26 scomatic/src/03_knit_driver_coverage.sh

@@ -1,0 +1,13 @@
+#!/bin/bash
+# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/ ; . ~/.bashrc ; mamba activate jupy ; /software/team205/bin/jsub lsf -q week -n driver_coverage -c 1 -m 20g -l log "bash src/03_knit_driver_coverage.sh" | bsub
+
+# wd
+wd=/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/
+cd $wd
+
+# conda activate an environment which has docopt installed so jsub works
+. ~/.bashrc
+mamba activate jupy
+
+# knit
+Rscript -e "rmarkdown::render('reports/driver_coverage.Rmd', output_file = 'driver_coverage.html', output_dir = 'reports/', params = list(rerun = F))"

### 2024-01-30 scomatic/reports/driver_coverage.Rmd
add driver preprocessing script to find exons, will eventually add to driver_coverage.Rmd, fix infercnv output pattern matching

@@
 
-Genes of interest ('drivers') in rheumatoid arthritis are loaded. Their coverage in the dataset, across samples and cell types, will be checked.
+Genes of interest in rheumatoid arthritis (hereafter referred to as 'drivers') are defined as genes that are found to undergo significant selection (q-value < 0.2) according to the whole exome and targeted NanoSeq data.
+The coverage of these sites in the dataset, across samples and celltypes, will be checked.
 
-```{r drivers, warning=F, message=F}
+First, we read in the dNdS results and filter by q-value.
+
+```{r read_dnds, warning = F, message = F}
+# read in dnds results
+muts <-
+  'data/dnds_nanoseq/' %>%
+  list.files(pattern = 'nanoseq') %>%
+  purrr::map(function(assay) {
+    list.files(paste0('data/dnds_nanoseq/', assay), 
+               pattern = '_dnds_sel_cv.tsv', 
+               full.names = T) %>%
+      purrr::map(function(file) {
+        file %>%
+          readr::read_tsv(show_col_types = F) %>%
+          dplyr::mutate(
+            assay = assay,
+            celltype = file %>% 
+              basename() %>% 
+              gsub('_dnds_sel_cv.tsv', '', .))
+      }) %>%
+      dplyr::bind_rows()
+  }) %>%
+  dplyr::bind_rows() %>%
+  dplyr::select(assay, celltype, everything())
+
+# get genes with strong evidence (q-value)
 drivers <-
-  readr::read_tsv('data/driver_genes/driver_gene_coords_for_coverage.tsv')
-knitr::kable(drivers)
+  muts %>%
+  dplyr::filter(qglobal_cv < 0.2) %>%
+  dplyr::rename(gene = gene_name)
+
+# manually write aliases for genes that are not matched
+# HIST1H3B -> H3C2
+# HIST1H1E -> H1-4
+# CRIPAK   -> This record has been withdrawn by NCBI, after discussions with 
+#             CCDS collaborators. It was decided that this locus is not an 
+#             independent gene.
+drivers <-
+  drivers %>%
+  dplyr::mutate(
+    gene_fixed = dplyr::case_when(
+      gene == 'HIST1H3B' ~ 'H3C2',
+      gene == 'HIST1H1E' ~ 'H1-4',
+      TRUE ~ gene
+    )
+  )
+```
+
+```{r plot_drivers, class.source = "fold-hide", warning = F, message = F}
+drivers %>%
+  dplyr::distinct(celltype, gene, assay, qglobal_cv) %>%
+  dplyr::mutate(neglog10q = -log10(qglobal_cv)) %>%
+  ggplot(aes(x = tidytext::reorder_within(gene, -neglog10q, celltype), 
+             y = neglog10q, fill = celltype)) +
+  geom_col(position = 'dodge') +
+  theme_classic() +
+  theme(axis.text.x = element_text(angle = -90, hjust = 0)) +
+  facet_grid(assay ~ celltype, scales = 'free_x', space = 'free_x') +
+  tidytext::scale_x_reordered() +
+  labs(x = 'gene', y = '-log10(p)')
+```
+
+```{r drivers_df, echo = F}
+drivers %>%
+  knitr::kable("html") %>% 
+  kableExtra::kable_styling("striped", full_width = F) %>% 
+  kableExtra::scroll_box(width = "100%", height = "300px")
+```
+
+In order to calculate their coverage in the BAMs, we must get their positions.
+
+```{r get_driver_pos}
+# load biomart
+library(biomaRt)
+
+# load ref
+ensembl <- useEnsembl(biomart = 'ensembl',
+                      dataset = 'hsapiens_gene_ensembl')
+
+# get positions
+drivers_pos <-
+  getBM(attributes = c('hgnc_symbol', 'chromosome_name', 'start_position', 'end_position'),
+        filters = 'hgnc_symbol', values = unique(drivers$gene),
+        mart = ensembl) %>%
+  # fix colnames
+  dplyr::mutate(gene = hgnc_symbol, 
+                chr = chromosome_name, 
+                start = start_position,
+                end = end_position) %>%
+  # remove weird chromosomes
+  dplyr::filter(chr %in% c(as.character(1:22, 'X', 'Y')))
+
+# save positions
+drivers_pos %>% 
+  dplyr::group_by(gene) %>%
+  dplyr::summarise(coords = 
+                     paste0(unique(chromosome_name), ':', 
+                            min(start_position), '-', 
+                            max(end_position))) %>%
+  readr::write_tsv('data/driver_genes/driver_gene_coords_for_coverage.tsv')
 ```
@@
 
-We use `samtools depth` to get the coverage of each site in the genes of interest, both across each sample and within each cell type in each sample.
+We use `samtools depth` to get the coverage of each site in the drivers, both across each sample and within each cell type in each sample.
 
-```{bash cov, eval=F}
+```{bash cov, eval = T}
 # activate conda env

### 2024-01-30 infercnv/src/01_run_infercnv.sh

@@ -17,6 +17,5 @@ mamba activate jupy
     -c config/Zhang2023.config \
     -c config/infercnv.config \
     -c /nfs/team205/kp9/nextflow/scomatic/LSF.config  \
-    -w work/ \
-    -resume
+    -w work/ #-resume
     

### 2024-01-30 scomatic/src/00_preprocess_BAMs.R

@@ -2,7 +2,10 @@
 library(magrittr)
 
 # dirs
-setwd('/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/')
+wd <- 
+  ifelse(Sys.info()['nodename'] == "mib119104s", '~/Volumes/', '') %>% 
+  paste0('/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/') 
+setwd(wd)
 dat_dir <- '/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/'
 cellranger_dir <- paste0(dat_dir, 'cellranger_output/')
 processed_dir <- paste0(dat_dir, 'processed_output/')

### 2024-01-30 scomatic/src/00_preprocess_drivers.R

@@ -1,0 +1,124 @@
+# libraries
+library(magrittr)
+library(ggplot2)
+library(biomaRt)
+library(org.Hs.eg.db)
+
+# dirs
+wd <- 
+  ifelse(Sys.info()['nodename'] == "mib119104s", '~/Volumes/', '') %>% 
+  paste0('/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/') 
+setwd(wd)
+
+# load ref
+ensembl <- useEnsembl(biomart = 'ensembl',
+                      dataset = 'hsapiens_gene_ensembl')
+
+# read in dnds results
+muts <-
+  'data/dnds_nanoseq/' %>%
+  list.files(pattern = 'nanoseq') %>%
+  purrr::map(function(assay) {
+    list.files(paste0('data/dnds_nanoseq/', assay), 
+               pattern = '_dnds_sel_cv.tsv', 
+               full.names = T) %>%
+      purrr::map(function(file) {
+        file %>%
+          readr::read_tsv(show_col_types = F) %>%
+          dplyr::mutate(
+            assay = assay,
+            celltype = file %>% 
+              basename() %>% 
+              gsub('_dnds_sel_cv.tsv', '', .))
+      }) %>%
+      dplyr::bind_rows()
+  }) %>%
+  dplyr::bind_rows() %>%
+  dplyr::select(assay, celltype, everything())
+
+# get genes with strong evidence
+drivers <-
+  muts %>%
+  dplyr::filter(qglobal_cv < 0.2) %>%
+  dplyr::rename(gene = gene_name)
+
+# get positions
+drivers_pos <-
+  getBM(attributes = c('hgnc_symbol', 'chromosome_name', 'start_position', 'end_position'),
+        filters = 'hgnc_symbol', values = unique(drivers$gene),
+        mart = ensembl)
+
+# manually get aliases for genes that are not matched
+unique(drivers$gene[!drivers$gene %in% drivers_pos$hgnc_symbol])
+# HIST1H3B -> H3C2
+# HIST1H1E -> H1-4
+# CRIPAK   -> This record has been withdrawn by NCBI, after discussions with 
+#             CCDS collaborators. It was decided that this locus is not an 
+#             independent gene.
+drivers <-
+  drivers %>%
+  dplyr::mutate(
+    gene_fixed = dplyr::case_when(
+      gene == 'HIST1H3B' ~ 'H3C2',
[…]

### 2024-01-30 scomatic/src/03_knit_driver_coverage.sh

@@ -10,4 +10,4 @@ cd $wd
 mamba activate jupy
 
 # knit
-Rscript -e "rmarkdown::render('reports/driver_coverage.Rmd', output_file = 'driver_coverage.html', output_dir = 'reports/', params = list(rerun = F))"
+Rscript -e "rmarkdown::render('reports/driver_coverage.Rmd', output_file = 'driver_coverage.html', output_dir = 'reports/', params = list(rerun = T)"

### 2024-01-31 infercnv/reports/infercnv.Rmd
add infercnv rmd to show all plots, changes to driver coverage rmd

---
title: "Copy number variants in rheumatoid arthritis data from inferCNV"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: true
    code_folding: show
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
params:
    rerun: true
---

```{r setup, include = F, message = F, warning = F, class.source = 'fold-hide'}
# wd <- ifelse(Sys.info()['nodename'] == "mib119104s", '~/Volumes/', '') |> paste0('/lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/') ; setwd(wd)
# rmarkdown::render('reports/infercnv.Rmd', output_file = 'driver_coverage.html', output_dir = 'reports/', params = list(rerun = T))

knitr::opts_knit$set(root.dir = '../')
knitr::opts_chunk$set(
  warning = FALSE, 
  dpi = 300, 
  message = FALSE,
  cache.path = 'reports/infercnv_cache/')

# libraries
library(magrittr)
library(ggplot2)
library(gridExtra)
library(slickR)

# dev.off() if there is a device
dev_off_if <- function() {if (length(dev.list()) != 0) { dev.off() }}
```

inferCNV was run on the scRNA-seq data to call copy number variants in synovial cells. No reference group was used, so the average profile of all cells was used in each patient as the baseline.

```{r, results = 'asis'}
pngs <-
  list.files('out/Zhang2023/', pattern = '^BRI') %>% head %>%
  purrr::set_names() %>%
  purrr::map(function(sample_id) {
    list.files(paste0('out/Zhang2023/', sample_id), pattern = '.png', 
                     recursive = T, full.names = T) %>%
      {setNames(., basename(.))}
  })

chunks <- c()

for (sample_id in names(pngs)) {

  chunks <- c(chunks, paste0('\n#### ', sample_id))

  # for (i in names(pngs[[sample_id]])) {
  #   
  #   chunks <- c(chunks, paste0('\n##### ', i))

    knit_plot <- c(
      '',
      '```{r, warning = F, message = F, echo = F}',
      paste0('slickR(obj = pngs[["', sample_id, '"]]) %synch% ',
             '( slickR(names(pngs[["', sample_id, '"]]), slideType = "p") + settings(arrows = F) )'),
      '```',
      ''
    )
    knitted_plot <- knitr::knit_child(text = knit_plot, envir = environment(), quiet = T)
    dev_off_if()

  }

  knit_plot <- c(
    '',
    '```{r, warning = F, message = F, echo = F}',
    paste0('slickR(obj = pngs[["', sample_id, '"]]) %synch% ',
           '( slickR(names(pngs[["', sample_id, '"]]), slideType = "p") + settings(arrows = F) )'),
    '```',
    ''
  )
  knitted_plot <- knitr::knit_child(text = knit_plot, envir = environment(), quiet = T)
  dev_off_if()

  chunks <- c(chunks, knitted_plot)

}

cat(unlist(chunks), sep = '\n')
```

### 2024-01-31 scomatic/reports/driver_coverage.Rmd
add infercnv rmd to show all plots, changes to driver coverage rmd

@@
     fig_width: 8
-    keep_md: true
+    keep_md: false
     code_folding: show
@@
 ```{r setup, include = F, message = F, warning = F, class.source = 'fold-hide'}
-# wd <- ifelse(Sys.info()['nodename'] == "mib119104s", '~/Volumes/', '') %>% paste0('/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/') ; setwd(wd)
-# rmarkdown::render('reports/driver_coverage.Rmd', output_file = 'driver_coverage.html', output_dir = 'reports/', params = list(rerun = T))
+# wd <- ifelse(Sys.info()['nodename'] == "mib119104s", '~/Volumes/', '') |> paste0('/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/') ; setwd(wd)
+# rmarkdown::render('reports/driver_coverage.Rmd', output_file = 'driver_coverage.html', output_dir = 'reports/', params = list(rerun = F))
 
@@
 
-```{r plot_drivers, class.source = "fold-hide", warning = F, message = F}
+```{r plot_drivers, class.source = "fold-hide", warning = F, message = F, fig.width = 10}
 drivers %>%
@@
 
-```{r get_driver_pos}
-# load biomart
-library(biomaRt)
+```{r get_drivers_pos}
+drivers_pos <- xfun::cache_rds({
 
   # load ref
-ensembl <- useEnsembl(biomart = 'ensembl',
-                      dataset = 'hsapiens_gene_ensembl')
+  library(biomaRt)
+  ensembl <- useEnsembl(biomart = 'ensembl', dataset = 'hsapiens_gene_ensembl')
 
@@
           mart = ensembl) %>%
+    tibble::as_tibble() %>% 
     # fix colnames
@@
     dplyr::filter(chr %in% c(as.character(1:22, 'X', 'Y')))
+
+  # return
+  drivers_pos
+
+}, file = 'drivers_pos.rds')
 

### 2024-01-31 infercnv/src/02_knit_infercnv.sh

@@ -1,0 +1,13 @@
+#!/bin/bash
+# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/ ; . ~/.bashrc ; mamba activate jupy ; /software/team205/bin/jsub lsf -q week -n knit_infercnv -c 1 -m 20g -l log "bash src/02_knit_infercnv.sh" | bsub
+
+# wd
+wd=/lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/
+cd $wd
+
+# conda activate an environment which has docopt installed so jsub works
+. ~/.bashrc
+mamba activate jupy
+
+# knit
+Rscript -e "rmarkdown::render('reports/infercnv.Rmd', output_file = 'infercnv.html', output_dir = 'reports/', params = list(rerun = F))"

### 2024-01-31 scomatic/src/03_knit_driver_coverage.sh

@@ -1,5 +1,5 @@
 #!/bin/bash
-# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/ ; . ~/.bashrc ; mamba activate jupy ; /software/team205/bin/jsub lsf -q week -n driver_coverage -c 1 -m 20g -l log "bash src/03_knit_driver_coverage.sh" | bsub
+# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/ ; . ~/.bashrc ; mamba activate jupy ; /software/team205/bin/jsub lsf -q week -n knit_driver_coverage -c 1 -m 20g -l log "bash src/03_knit_driver_coverage.sh" | bsub
 
 # wd
 wd=/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/
@@ -10,4 +10,4 @@ cd $wd
 mamba activate jupy
 
 # knit
-Rscript -e "rmarkdown::render('reports/driver_coverage.Rmd', output_file = 'driver_coverage.html', output_dir = 'reports/', params = list(rerun = T)"
+Rscript -e "rmarkdown::render('reports/driver_coverage.Rmd', output_file = 'driver_coverage.html', output_dir = 'reports/', params = list(rerun = F))"

### 2024-02-01 scomatic/src/03_knit_driver_coverage.sh
add infercnv run wo immune

@@ -1,5 +1,5 @@
 #!/bin/bash
-# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/ ; . ~/.bashrc ; mamba activate jupy ; /software/team205/bin/jsub lsf -q week -n knit_driver_coverage -c 1 -m 20g -l log "bash src/03_knit_driver_coverage.sh" | bsub
+# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/ ; . ~/.bashrc ; mamba activate jupy ; /software/team205/bin/jsub lsf -q week -n knit_driver_coverage -c 1 -m 30g -l log "bash src/03_knit_driver_coverage.sh" | bsub
 
 # wd
 wd=/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/

### 2024-02-01 scomatic/reports/driver_coverage.Rmd
move driver wrangling from Rmd to Rscript

--- 2024-01-26 scomatic/reports/driver_coverage.Rmd
@@
 
-### Drivers
-
-Genes of interest ('drivers') in rheumatoid arthritis are loaded. Their coverage in the dataset, across samples and cell types, will be checked.
-
-```{r drivers, warning=F, message=F}
-drivers <-
-  readr::read_tsv('data/driver_genes/driver_gene_coords_for_coverage.tsv')
-knitr::kable(drivers)
-```
-
 ### Calculate coverage, by sample and by sample-x-celltype
 
-We use `samtools depth` to get the coverage of each site in the genes of interest, both across each sample and within each cell type in each sample.
+We use `samtools depth` to get the coverage of each site in the drivers, both across each sample and within each cell type in each sample.
 
-```{bash cov, eval=F}
+```{bash cov, eval = T}
 # activate conda env

### 2024-02-01 scomatic/src/00_preprocess_drivers.R

@@ -10,9 +10,11 @@ wd <-
   paste0('/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/') 
 setwd(wd)
 
-# load ref
-ensembl <- useEnsembl(biomart = 'ensembl',
-                      dataset = 'hsapiens_gene_ensembl')
+# Genes of interest in rheumatoid arthritis (hereafter referred to as 'drivers') 
+# are defined as genes that are found to undergo significant selection (q-value 
+# < 0.2) according to the whole exome and targeted NanoSeq data. The coverage of 
+# these sites in the dataset, across samples and celltypes, will be checked.
+# First, we read in the dNdS results and filter by q-value.
 
 # read in dnds results
 muts <-
@@ -36,20 +38,13 @@ muts <-
   dplyr::bind_rows() %>%
   dplyr::select(assay, celltype, everything())
 
-# get genes with strong evidence
+# get genes with strong evidence (q-value)
 drivers <-
   muts %>%
   dplyr::filter(qglobal_cv < 0.2) %>%
   dplyr::rename(gene = gene_name)
 
-# get positions
-drivers_pos <-
-  getBM(attributes = c('hgnc_symbol', 'chromosome_name', 'start_position', 'end_position'),
-        filters = 'hgnc_symbol', values = unique(drivers$gene),
-        mart = ensembl)
-
-# manually get aliases for genes that are not matched
-unique(drivers$gene[!drivers$gene %in% drivers_pos$hgnc_symbol])
+# manually write aliases for genes that are not matched
 # HIST1H3B -> H3C2
 # HIST1H1E -> H1-4
 # CRIPAK   -> This record has been withdrawn by NCBI, after discussions with 
@@ -65,46 +60,7 @@ drivers <-
     )
   )
 
-# get positions again
-drivers_pos <-
-  getBM(attributes = c('hgnc_symbol', 'ensembl_transcript_id',
-                       'chromosome_name', 'start_position', 'end_position'),
-        filters = 'hgnc_symbol', values = unique(drivers$gene_fixed),
-        mart = ensembl)
-
-# check if it was fixed
-unique(drivers$gene_fixed[!drivers$gene_fixed %in% drivers_pos$hgnc_symbol])
-
-# get exon positions
-drivers_exons <-
-  getBM(attributes = c('ensembl_transcript_id', 'exon_chrom_start', 'exon_chrom_end'),
[…]

### 2024-02-01 scomatic/reports/driver_coverage.Rmd
add nextflow wrapper for driver_coverage

@@
     rerun: true
+    gene: NULL
+    coords: NULL
 ---
@@
 
-We use `samtools depth` to get the coverage of each site in the drivers, both across each sample and within each cell type in each sample.
+We use `samtools depth` to get the coverage of each site in the drivers, both across each sample and within each celltype in each sample.
 

### 2024-02-01 scomatic/src/03_knit_driver_coverage.sh

@@ -11,3 +11,19 @@ mamba activate jupy
 
 # knit
 Rscript -e "rmarkdown::render('reports/driver_coverage.Rmd', output_file = 'driver_coverage.html', output_dir = 'reports/', params = list(rerun = F))"
+
+
+# run scomatic
+(
+  echo 'id,bam,out_dir' ;
+  cat data/Zhang2023/mappings.csv |
+  sed 1d | head -3 | cut -d, -f 1,2 |
+  awk -F'/' '{print $0",/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/out/Zhang2023/coverage/"$10"/"}' ;
+) | cat > data/test_mappings.csv
+
+/software/team205/nextflow-23.04.1-all run nextflow/driver_coverage.nf \
+    --mappings data/test_mappings.csv \
+    --drivers /lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/data/driver_genes/driver_gene_coords_for_coverage.tsv \
+    --location local \
+    -w work/ \
+    -resume

### 2024-02-05 infercnv/reports/infercnv.Rmd
Recover repository after accidental deletion of .git directory

@@
 pngs <-
-  list.files('out/Zhang2023/', pattern = '^BRI') %>% head %>%
+  list.files('out/Zhang2023/', pattern = '^BRI') %>%
   purrr::set_names() %>%
   purrr::map(function(sample_id) {
+    lis <-
       list.files(paste0('out/Zhang2023/', sample_id), pattern = '.png', 
                  recursive = T, full.names = T) %>%
-      {setNames(., basename(.))}
+      {setNames(., gsub('infercnv.', '', tools::file_path_sans_ext(basename(.))))}
+
+    # reorder
+    c(lis['infercnv'], lis[names(lis) != 'infercnv'])
   })
@@
 
-  chunks <- c(chunks, paste0('\n#### ', sample_id))
+  chunks <- c(chunks, paste0('\n### ', sample_id))
 
-  # for (i in names(pngs[[sample_id]])) {
-  #   
-  #   chunks <- c(chunks, paste0('\n##### ', i))
+  for (i in names(pngs[[sample_id]])) {
+
+    chunks <- c(chunks, paste0('\n#### ', i))
 
@@
       '```{r, warning = F, message = F, echo = F}',
-      paste0('slickR(obj = pngs[["', sample_id, '"]]) %synch% ',
-             '( slickR(names(pngs[["', sample_id, '"]]), slideType = "p") + settings(arrows = F) )'),
-      '```',
-      ''
-    )
-    knitted_plot <- knitr::knit_child(text = knit_plot, envir = environment(), quiet = T)
-    dev_off_if()
-
-  }
-
-  knit_plot <- c(
-    '',
-    '```{r, warning = F, message = F, echo = F}',
-    paste0('slickR(obj = pngs[["', sample_id, '"]]) %synch% ',
-           '( slickR(names(pngs[["', sample_id, '"]]), slideType = "p") + settings(arrows = F) )'),
+      paste0('knitr::include_graphics("../', pngs[[sample_id]][[i]], '")'),
       '```',
@@
   }
+}
 

### 2024-02-05 scomatic/reports/driver_coverage.Rmd
Recover repository after accidental deletion of .git directory

--- 2024-02-01 scomatic/reports/driver_coverage.Rmd
@@
 ---
-title: "Mutation loads in rheumatoid arthritis data from SComatic"
+title: "Driver coverage"
 author: "Alexandra Tidd"
@@
 params:
+  id: NULL
+  depths: NULL
+  drivers: NULL
+  ref_cds: NULL
+  cache_dir: './driver_coverage_cache/'
   rerun: true
@@
 
-knitr::opts_knit$set(root.dir = '../')
+knitr::opts_knit$set(root.dir = './')
 knitr::opts_chunk$set(
@@
   message = FALSE,
-  cache.path = 'reports/driver_coverage_cache/')
+  cache.path = params$cache_dir)
 
@@
 
-### Calculate coverage, by sample and by sample-x-celltype
-
-We use `samtools depth` to get the coverage of each site in the drivers, both across each sample and within each cell type in each sample.
-
-```{bash cov, eval = T}
-# activate conda env
-mamba activate jupy
-
-# dirs
-input_dir=/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/cellranger_output
-ct_input_dir=out/Zhang2023/default_thresholds/
-output_dir=out/Zhang2023/coverage/
-drivers=data/driver_genes/driver_gene_coords_for_coverage.tsv
-mkdir -p $output_dir
-
-# iterate over each bam file and pileup specific regions
-(
-  echo -e 'chr\tpos\tdepth\tgene\tsample' ;
-  for bam in $(find $input_dir -iname 'possorted_genome_bam.bam'); do
-  	sample=$(echo $bam | cut -d "/" -f 10 | sed 's/\-/_/g')
-    while read -r gene coords ; do
-      samtools depth -a -r $coords -Q 30 $bam |
-      awk -v sample=$sample -v gene=$gene '{print $0"\t"gene"\t"sample}'
-    done < <(sed 1d $drivers) ;
-  done
-) | cat > $output_dir/drivers_samtools_depth.tsv
-
-# iterate over celltype-specific bam files and pileup specific regions
-(
-  echo -e 'chr\tpos\tdepth\tgene\tsample\tcelltype' ;
-  for sample_dir in $ct_input_dir/*-GEX ; do
-    sample=$(basename $sample_dir | sed 's/\-GEX$//g')
-    for ct_bam in $sample_dir/celltype_bams/*.bam ; do
-      ct=$(basename $ct_bam | cut -f2 -d.)
-      while read -r gene coords ; do
-        samtools depth -a -r $coords -Q 30 $ct_bam |
-        awk -v sample=$sample -v gene=$gene -v ct=$ct \
-        '{print $0"\t"gene"\t"sample"\t"ct}'
-      done < <(sed 1d $drivers) ;
-    done
-  done
-) | cat > $output_dir/drivers_samtools_depth_by_celltype.tsv
-```
-
-We read in the results at both levels.
+We read in drivers and their depths.
 
 ```{r read_cov}
+# read coverage
 cov <- xfun::cache_rds({
-  cov_files <-
-    list.files(
-      'out/Zhang2023/coverage/', 
-      pattern = 'drivers_samtools_depth.*.tsv',
-      full.names = T) 
   cov <- 
-    cov_files %>%
-    setNames(cov_files %>% basename() %>% tools::file_path_sans_ext()) %>%
-    purrr::map(readr::read_tsv) %>%
-    dplyr::bind_rows(.id = 'lvl') %>%
+    readr::read_tsv('depths.tsv') %>% 
     dplyr::mutate(chr = as.character(chr))
@@
 }, file = 'cov.rds', rerun = params$rerun)
+
+# read drivers
+drivers <- readr::read_tsv(params$drivers)
 ```
@@
 ```{r features}
+# get ref_cds object from DNDSCV
+ref_cds <- 
+  readRDS(params$ref_cds)
+
 # annotate gene features
@@
 
-  # load rda to get ref_cds object from DNDSCV
-  refcds_38 <- '/lustre/scratch125/casm/team268im/fa8/119/DNDSCV_COVARIATES/refcds_GRCh38-GencodeV18+Appris.rda'
-  load(refcds_38)
+  # get all positions in each gene
+  gene_spans <-
+    ref_cds[c(drivers$gene)] %>%
+    purrr::map(function(i) {
+      tibble::tibble(
+        chr = as.character(i$chr),
+        start = i$intervals_cds[, 1],
+        end = i$intervals_cds[, 2]) %>%
+        dplyr::mutate(exon = paste0('exon', dplyr::row_number()))
+    })
 
-  # set gene names
-  ref_cds <-
-    RefCDS %>%
-    setNames(RefCDS %>% purrr::map(~ .x$gene_name) %>% unlist())
+  # drop those not found
+  gene_spans[is.na(names(gene_spans))] <- NULL
 
@@
   exons <-
-    ref_cds[c(drivers$gene)] %>%
-    purrr::map(~ .x$intervals_cds %>% 
-                 tibble::as_tibble() %>%
-                 dplyr::transmute(chr = .x$chr, start = V1, end = V2) %>%
-                 dplyr::mutate(exon = paste0('exon', dplyr::row_number()))) %>%
+    gene_spans %>%
     dplyr::bind_rows(.id = 'gene') %>%
@@
   cov %>% 
-    dplyr::group_by(gene, sample) %>%
+    dplyr::group_by(gene) %>%
     dplyr::group_split() %>%
@@
       # for testing:
-      # cov %>% dplyr::group_by(gene, sample) %>% dplyr::group_split() %>% {.[[1]]} -> df_i
+      # cov %>% dplyr::group_by(gene) %>% dplyr::group_split() %>% {.[[1]]} -> df_i
 
       # get info
-      i <- 
-        df_i %>%
-        dplyr::distinct(gene, sample)
-      info <- t(i[1,]) %>% paste(collapse = '.')
-      p[[info]] <<- list()
+      gene_i <- unique(df_i$gene)
+      p[[gene_i]] <<- list()
 
       # get number of celltypes for plot dims
-      p[[info]]$n_cts <<- 
+      p[[gene_i]]$n_cts <<- 
         df_i %>% 
@@
       # plot by sample and by celltype
-      list(by_sample = list(lvl = 'drivers_samtools_depth', title_suffix = ''),
-           by_celltype = list(lvl = 'drivers_samtools_depth_by_celltype', title_suffix = '_by_celltype')) %>%
-        purrr::map(function(x){
+      df_i %>%
+        dplyr::mutate(suffix = dplyr::case_when(is.na(celltype) ~ '',
+                                                TRUE ~ '_by_celltype')) %>%
+        {split(., .$suffix)} %>%
+        purrr::map(function(lvl) {
+
+          # get suffix
+          p_suffix <- unique(lvl$suffix)
 
           # plot all
-          p[[info]][[paste0('all', x$title_suffix)]] <<-
+          p[[gene_i]][[paste0('all', p_suffix)]] <<-
             plot_all(
-              p_df = df_i %>% dplyr::filter(lvl == x$lvl),
-              p_exon_coords = exon_coords %>% dplyr::filter(gene == i$gene),
-              p_title = paste0(info, x$title_suffix)
+              p_df = lvl,
+              p_exon_coords = exon_coords %>% dplyr::filter(gene == gene_i),
+              p_title = paste0(gene_i, p_suffix)
             )
 
-          # plot exons only
-          p[[info]][[paste0('exons', x$title_suffix)]] <<-
+          # plot exons
+          p[[gene_i]][[paste0('exons', p_suffix)]] <<-
             plot_exons(
-              p_df = df_i %>% dplyr::filter(lvl == x$lvl),
-              p_title = paste0(info, x$title_suffix)
+              p_df = lvl,
+              p_title = paste0(gene_i, p_suffix)
             )
-
         })
-
     })

### 2024-02-05 infercnv/src/00_preprocess.R

@@ -1,0 +1,116 @@
+# libraries
+library(magrittr)
+
+# dirs
+wd <- 
+  ifelse(Sys.info()['nodename'] == "mib119104s", '~/Volumes/', '') |>
+  paste0('/lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/') 
+setwd(wd)
+
+# Zhang 2023 ---
+dat_dir <- '/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/'
+cellranger_dir <- paste0(dat_dir, 'cellranger_output/')
+processed_dir <- paste0(dat_dir, 'processed_output/')
+
+# annotations TSVs (cell celltype) ----
+
+# all celltypes
+dir.create('data/Zhang2023/annotations/', recursive = T)
+annotations <-
+  readRDS(paste0(processed_dir, 'all_cells_reference.rds'))$meta_data %>%
+  dplyr::select(cell, celltype = cell_type, sample_id = sample) 
+annotations %>%
+  {split(., .$sample_id)} %>%
+  purrr::walk2(.x = names(.), .y = ., function(sample_id, df) {
+      df %>%
+          dplyr::select(cell, celltype) %>%
+          readr::write_tsv(
+            paste0('data/Zhang2023/annotations/', sample_id, '.tsv'),
+            col_names = F
+          )
+  })
+
+# wo immune celltypes
+dir.create('data/Zhang2023/annotations_wo_immune/', recursive = T)
+immune_cts <- c('NK', 'B cell/plasma cell', 'Myeloid cell', 'T cell')
+annotations %>%
+  dplyr::filter(!(celltype %in% immune_cts)) %>%
+  {split(., .$sample_id)} %>%
+  purrr::walk2(.x = names(.), .y = ., function(sample_id, df) {
+    df %>%
+      dplyr::select(cell, celltype) %>%
+      readr::write_tsv(
+        paste0('data/Zhang2023/annotations_wo_immune/', sample_id, '.tsv'),
+        col_names = F
+      )
+  })
+
+# mappings CSV (sample_id,raw_counts_matrix,annotations,id) ----
+
+# all celltypes
+mappings <-
+  list.files(
+    cellranger_dir,
+    pattern = '^BRI\\-',
+    include.dirs = T) %>%
+  tibble::tibble(sample_id = .)  %>%
+  # write file paths
+  dplyr::transmute(
+    sample_id,
[…]

### 2024-02-05 infercnv/src/01_run_infercnv.sh

@@ -1,0 +1,29 @@
+#!/bin/bash
+# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/ ; . ~/.bashrc ; mamba activate jupy ; /software/team205/bin/jsub lsf -q week -n infercnv -c 1 -m 2g -l log "bash src/01_run_infercnv.sh" | bsub
+
+set -eo pipefail
+
+# wd
+wd=/lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/
+cd $wd
+mkdir -p work/ out/ log/
+
+# conda activate an environment which has docopt installed so jsub works
+. ~/.bashrc
+mamba activate jupy
+
+# # run infercnv
+# /software/team205/nextflow-23.04.1-all run nextflow/infercnv.nf \
+#     -c config/Zhang2023.config \
+#     -c config/infercnv.config \
+#     -c /nfs/team205/kp9/nextflow/scomatic/LSF.config  \
+#     -w work/ \
+#     -resume
+    
+# run infercnv wo immune
+/software/team205/nextflow-23.04.1-all run nextflow/infercnv.nf \
+    -c config/Zhang2023_wo_immune.config \
+    -c config/infercnv.config \
+    -c /nfs/team205/kp9/nextflow/scomatic/LSF.config  \
+    -w work/ \
+    -resume

### 2024-02-05 scomatic/src/00_check_bams.sh

@@ -1,0 +1,5 @@
+#!/bin/bash
+
+java -jar ~/bin/picard.jar ValidateSamFile \
+      I=/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/cellranger_output/BRI-419/possorted_genome_bam.bam \
+      MODE=SUMMARY

### 2024-02-05 scomatic/src/00_preprocess_BAMs.R

@@ -1,0 +1,51 @@
+# libraries
+library(magrittr)
+
+# dirs
+wd <- 
+  ifelse(Sys.info()['nodename'] == "mib119104s", '~/Volumes/', '') %>% 
+  paste0('/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/') 
+setwd(wd)
+dat_dir <- '/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/'
+cellranger_dir <- paste0(dat_dir, 'cellranger_output/')
+processed_dir <- paste0(dat_dir, 'processed_output/')
+dir.create('data/')
+
+# Zhang 2023 ----
+# mappings CSV (sample_id,bam_file,id) -----
+dir.create('data/Zhang2023/')
+list.files(
+        cellranger_dir,
+        pattern = '^BRI\\-',
+        include.dirs = T) %>%
+    tibble::enframe(value = 'path_id') %>%
+    dplyr::transmute(
+        sample_id = gsub('-', '_', path_id),
+        bam_file = paste0(cellranger_dir, path_id, '/possorted_genome_bam.bam'),
+        id = gsub('-', '_', path_id)) %>%
+    readr::write_csv('data/Zhang2023/mappings.csv')
+
+# celltypes TSV (Index Cell_type) -----
+dat <-
+    readRDS(paste0(processed_dir, 'all_cells_reference.rds'))$meta_data
+aliases <-
+    dat %>%
+    dplyr::transmute(
+        sample_id = sample %>% gsub('-', '_', .),
+        # fix indices - must be formatted SAMPLE_BARCODE-1
+        cell = cell %>% gsub('-', '_', .) %>% paste0('-1'),
+        # fix cell types - only letters, numbers and underscores
+        celltype = cell_type,
+        celltype_alias = cell_type %>% gsub(' |/', '_', .))
+aliases %>%
+    dplyr::select(Index = cell, Cell_type = celltype_alias) %>%
+    readr::write_tsv('data/Zhang2023/celltypes.tsv')
+
+# celltype counts
+dat %>%
+    dplyr::select(sample_id = sample,
+                  id = sample, 
+                  celltype = cell_type) %>%
+    dplyr::group_by(sample_id, id, celltype) %>%
+    dplyr::count(name = 'n_cells_per_id_per_celltype') %>%
+    readr::write_tsv('data/Zhang2023/celltype_counts.tsv')

### 2024-02-05 scomatic/src/00_preprocess_drivers.R

@@ -1,0 +1,118 @@
+# libraries
+library(magrittr)
+library(ggplot2)
+library(biomaRt)
+library(org.Hs.eg.db)
+
+# dirs
+wd <- 
+  ifelse(Sys.info()['nodename'] == "mib119104s", '~/Volumes/', '') %>% 
+  paste0('/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/') 
+setwd(wd)
+
+# Genes of interest in rheumatoid arthritis (hereafter referred to as 'drivers') 
+# are defined as genes that are found to undergo significant selection (q-value 
+# < 0.2) according to the whole exome and targeted NanoSeq data. The coverage of 
+# these sites in the dataset, across samples and celltypes, will be checked.
+# First, we read in the dNdS results and filter by q-value.
+
+# read in dnds results
+muts <-
+  'data/dnds_nanoseq/' %>%
+  list.files(pattern = 'nanoseq') %>%
+  purrr::map(function(assay) {
+    list.files(paste0('data/dnds_nanoseq/', assay), 
+               pattern = '_dnds_sel_cv.tsv', 
+               full.names = T) %>%
+      purrr::map(function(file) {
+        file %>%
+          readr::read_tsv(show_col_types = F) %>%
+          dplyr::mutate(
+            assay = assay,
+            celltype = file %>% 
+              basename() %>% 
+              gsub('_dnds_sel_cv.tsv', '', .))
+      }) %>%
+      dplyr::bind_rows()
+  }) %>%
+  dplyr::bind_rows() %>%
+  dplyr::select(assay, celltype, everything())
+
+# get genes with strong evidence (q-value)
+drivers <-
+  muts %>%
+  dplyr::filter(qglobal_cv < 0.2) %>%
+  dplyr::rename(gene = gene_name)
+
+# manually write aliases for genes that are not matched
+# HIST1H3B -> H3C2
+# HIST1H1E -> H1-4
+# CRIPAK   -> This record has been withdrawn by NCBI, after discussions with 
+#             CCDS collaborators. It was decided that this locus is not an 
+#             independent gene.
+drivers <-
+  drivers %>%
+  dplyr::mutate(
+    gene_fixed = dplyr::case_when(
+      gene == 'HIST1H3B' ~ 'H3C2',
+      gene == 'HIST1H1E' ~ 'H1-4',
+      TRUE ~ gene
+    )
+  )
+
[…]

### 2024-02-05 scomatic/src/01_run_scomatic.sh

@@ -1,0 +1,37 @@
+#!/bin/bash
+# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/ ; . ~/.bashrc ; mamba activate jupy ; /software/team205/bin/jsub lsf -q week -n scomatic -c 1 -m 2g -l log "bash src/01_run_scomatic.sh" | bsub
+
+set -eo pipefail
+
+# wd
+wd=/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/
+cd $wd
+config_dir=/nfs/team205/kp9/nextflow/scomatic/
+mkdir -p work/ out/Zhang2023/default_thresholds/ log/
+
+# conda activate an environment which has docopt installed so jsub works
+. ~/.bashrc
+mamba activate jupy
+
+# run scomatic
+/software/team205/nextflow-23.04.1-all run nextflow/scomatic.nf \
+    --projectDir $PWD \
+    -params-file $config_dir/GEX.json \
+    -c config/Zhang2023.config \
+    -c $config_dir/LSF.config \
+    -w work/ \
+    --output_dir out/Zhang2023/default_thresholds/ \
+    --location local \
+    -resume
+
+# get single cell genotypes
+/software/team205/nextflow-23.04.1-all run nextflow/scomatic.nf \
+    -entry genotypes \
+    --projectDir $PWD \
+    -params-file $config_dir/GEX.json \
+    -c config/Zhang2023.config \
+    -c $config_dir/LSF.config \
+    -w work/ \
+    --output_dir out/Zhang2023/default_thresholds/ \
+    --location local \
+    -resume

### 2024-02-05 scomatic/src/02_run_scomatic_low_thresholds.sh

@@ -1,0 +1,39 @@
+#!/bin/bash
+# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/ ; . ~/.bashrc ; mamba activate jupy ; /software/team205/bin/jsub lsf -q week -n scomatic_lo -c 1 -m 2g -l log "bash src/02_run_scomatic_low_thresholds.sh" | bsub
+
+set -eo pipefail
+
+# wd
+wd=/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/
+cd $wd
+config_dir=/nfs/team205/kp9/nextflow/scomatic/
+mkdir -p work/ out/Zhang2023/default_thresholds/ log/
+
+# conda activate an environment which has docopt installed so jsub works
+. ~/.bashrc
+mamba activate jupy
+
+# run scomatic
+/software/team205/nextflow-23.04.1-all run nextflow/scomatic.nf \
+    --projectDir $PWD \
+    --min_dp 1 \
+    --output_dir '/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/out/Zhang2023/low_thresholds/' \
+    -params-file $config_dir/GEX.json \
+    -c config/Zhang2023.config \
+    -c $config_dir/LSF.config \
+    -w work/ \
+    --location local \
+    -resume
+
+# get single cell genotypes
+/software/team205/nextflow-23.04.1-all run nextflow/scomatic.nf \
+    -entry genotypes \
+    --projectDir $PWD \
+    --min_dp 1 \
+    --output_dir '/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/out/Zhang2023/low_thresholds/' \
+    -params-file $config_dir/GEX.json \
+    -c config/Zhang2023.config \
+    -c $config_dir/LSF.config \
+    -w work/ \
+    --location local \
+    -resume

### 2024-02-05 scomatic/src/03_knit_driver_coverage.sh

@@ -1,0 +1,45 @@
+#!/bin/bash
+# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/ ; . ~/.bashrc ; mamba activate jupy ; /software/team205/bin/jsub lsf -q week -n knit_driver_coverage -c 1 -m 30g -l log "bash src/03_knit_driver_coverage.sh" | bsub
+
+# wd
+wd=/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/
+cd $wd
+
+# conda activate an environment which has docopt installed so jsub works
+. ~/.bashrc
+mamba activate jupy
+
+# # knit
+# Rscript -e "rmarkdown::render('reports/driver_coverage.Rmd', output_file = 'driver_coverage.html', output_dir = 'reports/', params = list(rerun = F))"
+
+# create mappings
+bams_dir=/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/cellranger_output/
+ct_bams_dir=/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/out/Zhang2023/default_thresholds/
+out_dir=/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/out/Zhang2023/coverage/
+(
+  echo 'id,bam,celltype' ;
+  while read -r id ; do
+    echo "$id,$bams_dir/$id/possorted_genome_bam.bam,NA" ;
+    for file in $ct_bams_dir/${id/-/_}/celltype_bams/*.bam ; do
+      ct=$(basename $file | cut -d. -f2)
+      echo "$id,$file,$ct" ;
+    done
+  done < \
+  <(sed 1d /lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/data/Zhang2023/mappings.csv | cut -d, -f1 | sort -u)
+) | cat > data/Zhang2023/coverage_mappings.csv
+
+# test mappings
+cat data/Zhang2023/coverage_mappings.csv | head -10 > data/Zhang2023/coverage_mappings_test.csv
+
+# run
+/software/team205/nextflow-23.04.1-all run nextflow/driver_coverage.nf \
+    --mappings data/Zhang2023/coverage_mappings_test.csv \
+    --drivers /lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/data/driver_genes/driver_gene_coords_for_coverage.tsv \
[…]

### 2024-02-05 scomatic/reports/driver_coverage.Rmd
fix knitting location

@@
 
-knitr::opts_knit$set(root.dir = './')
+#knitr::opts_knit$set(root.dir = '../')
 knitr::opts_chunk$set(
@@
 for (info in names(p)) {
-  chunks <- c(chunks, paste0('\n####', info))
+  chunks <- c(chunks, paste0('\n#### ', info))
   message(info)

### 2024-02-05 scomatic/src/03_knit_driver_coverage.sh

@@ -33,10 +33,9 @@ cat data/Zhang2023/coverage_mappings.csv | head -10 > data/Zhang2023/coverage_ma
 
 # run
 /software/team205/nextflow-23.04.1-all run nextflow/driver_coverage.nf \
-    --mappings data/Zhang2023/coverage_mappings_test.csv \
+    --mappings data/Zhang2023/coverage_mappings.csv \
     --drivers /lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/data/driver_genes/driver_gene_coords_for_coverage.tsv \
     --ref_cds /lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/data/dndscv/ref_cds.rds \
     --out_dir /lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/out/Zhang2023/coverage/ \
     --location local \
-    -w work/ \
-    -resume
+    -w work/ 

### 2024-02-06 infercnv/reports/infercnv.Rmd
fix infercnv metadata propagation, change to standard mappings and annotations, add help nf functions

@@
 
-knitr::opts_knit$set(root.dir = '../')
+# chunk options
 knitr::opts_chunk$set(
@@
   dpi = 300, 
-  message = FALSE,
-  cache.path = 'reports/infercnv_cache/')
+  message = FALSE)
 
@@
 pngs <-
-  list.files('out/Zhang2023/', pattern = '^BRI') %>%
+  list.files(pattern = '^BRI', include.dirs = T) %>%
   purrr::set_names() %>%
-  purrr::map(function(sample_id) {
+  purrr::map(function(id) {
     lis <-
-      list.files(paste0('out/Zhang2023/', sample_id), pattern = '.png', 
-                 recursive = T, full.names = T) %>%
+      list.files(id, pattern = '.png', recursive = T, full.names = T) %>%
       {setNames(., gsub('infercnv.', '', tools::file_path_sans_ext(basename(.))))}
 
-    # reorder
+    # reorder (final plot first)
     c(lis['infercnv'], lis[names(lis) != 'infercnv'])
@@
 
-for (sample_id in names(pngs)) {
+for (id in names(pngs)) {
 
-  chunks <- c(chunks, paste0('\n### ', sample_id))
+  chunks <- c(chunks, paste0('\n### ', id))
 
-  for (i in names(pngs[[sample_id]])) {
+  for (i in names(pngs[[id]])) {
 
@@
       '```{r, warning = F, message = F, echo = F}',
-      paste0('knitr::include_graphics("../', pngs[[sample_id]][[i]], '")'),
+      paste0('knitr::include_graphics("../', pngs[[id]][[i]], '")'),
       '```',

### 2024-02-06 infercnv/src/00_preprocess.R

@@ -13,115 +13,60 @@ cellranger_dir <- paste0(dat_dir, 'cellranger_output/')
 processed_dir <- paste0(dat_dir, 'processed_output/')
 
 # annotations TSVs (cell celltype) ----
-
-# load annotations
-annotations <-
-  readRDS(paste0(processed_dir, 'all_cells_reference.rds'))$meta_data %>%
+readRDS(paste0(processed_dir, 'all_cells_reference.rds'))$meta_data %>%
   dplyr::select(cell, 
                 celltype = cell_type, 
                 cluster = cluster_name, 
-                sample_id = sample) 
+                id = sample) %>%
+  readr::write_tsv('data/Zhang2023/annotations.tsv')
 
-# all celltypes
-dir.create('data/Zhang2023/annotations/celltypes/', recursive = T)
-annotations %>%
-  {split(., .$sample_id)} %>%
-  purrr::walk2(.x = names(.), .y = ., function(sample_id, df) {
-      df %>%
-          dplyr::select(cell, celltype) %>%
-          readr::write_tsv(
-            paste0('data/Zhang2023/annotations/celltypes/', sample_id, '.tsv'),
-            col_names = F
-          )
-  })
+# mappings CSV (id,raw_counts_matrix) ----
+files <-
+  list.files(
+    cellranger_dir,
+    pattern = '^BRI\\-',
+    include.dirs = T) %>%
+  tibble::tibble(id = .)  %>%
+  # write file paths
+  dplyr::transmute(
+    id,
+    raw_counts_matrix = paste0(processed_dir, 'raw_mRNA_count_matrix.rds')) 
 
-# wo immune celltypes
-dir.create('data/Zhang2023/annotations/celltypes_wo_immune/', recursive = T)
-immune_cts <- c('NK', 'B cell/plasma cell', 'Myeloid cell', 'T cell')
-annotations %>%
-  dplyr::filter(!(celltype %in% immune_cts)) %>%
-  {split(., .$sample_id)} %>%
-  purrr::walk2(.x = names(.), .y = ., function(sample_id, df) {
-    df %>%
-      dplyr::select(cell, celltype) %>%
-      readr::write_tsv(
-        paste0('data/Zhang2023/annotations/celltypes_wo_immune/', sample_id, '.tsv'),
-        col_names = F
-      )
-  })
+# check for missing files
+missing <- 
+  files %>%
[…]

### 2024-02-06 infercnv/src/01_run_infercnv.sh

@@ -8,37 +8,29 @@ wd=/lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/
 cd $wd
 mkdir -p work/ out/ log/
 
-# annotation level
-lvl=clusters
-
 # conda activate an environment which has docopt installed so jsub works
 . ~/.bashrc
 mamba activate jupy
 
-# # run infercnv on celltypes
-# /software/team205/nextflow-23.04.1-all run nextflow/infercnv.nf \
-#   --out_dir $wd/out/Zhang2023/$lvl/ \
-#   --mappings $wd/data/Zhang2023/mappings/$lvl.csv \
-#   -c config/infercnv.config \
-#   -c /nfs/team205/kp9/nextflow/scomatic/LSF.config  \
-#   -w work/ \
-#   -resume
-    
-# # run infercnv on celltypes wo immune
-# /software/team205/nextflow-23.04.1-all run nextflow/infercnv.nf \
-#   --out_dir $wd/out/Zhang2023/$lvl/ \
-#   --mappings $wd/data/Zhang2023/mappings/$lvl.csv \
-#   -c config/infercnv.config \
-#   -c /nfs/team205/kp9/nextflow/scomatic/LSF.config  \
-#   -w work/ \
-#   -resume
-    
-# run infercnv on clusters
+# run infercnv on celltypes
 /software/team205/nextflow-23.04.1-all run nextflow/infercnv.nf \
-  --out_dir $wd/out/Zhang2023/$lvl/ \
-  --mappings $wd/data/Zhang2023/mappings/$lvl.csv \
+  --out_dir $wd/out/Zhang2023/by_celltype/ \
+  --annotations $wd/data/Zhang2023/annotations.tsv \
+  --annotation_col 'celltype' \
+  --mappings $wd/data/Zhang2023/mappings.csv \
   -c config/infercnv.config \
   -c /nfs/team205/kp9/nextflow/scomatic/LSF.config  \
   -w work/ \
   -resume
-
+    
+# run infercnv on clusters 
+/software/team205/nextflow-23.04.1-all run nextflow/infercnv.nf \
+  --out_dir $wd/out/Zhang2023/by_cluster/ \
+  --annotations $wd/data/Zhang2023/annotations.tsv \
+  --annotation_col 'cluster' \
+  --mappings $wd/data/Zhang2023/mappings.csv \
+  -c config/infercnv.config \
[…]

### 2024-02-07 infercnv/src/01_run_infercnv.sh
attempt to add knit_report module to infercnv

@@ -1,5 +1,5 @@
 #!/bin/bash
-# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/ ; . ~/.bashrc ; mamba activate jupy ; /software/team205/bin/jsub lsf -q week -n infercnv -c 1 -m 2g -l log "bash src/01_run_infercnv.sh" | bsub
+# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/ ; . ~/.bashrc ; mamba activate jupy ; /software/team205/bin/jsub lsf -q week -n infercnv_clusters -c 1 -m 2g -l log "bash src/01_run_infercnv.sh" | bsub
 
 set -eo pipefail
 
@@ -12,25 +12,34 @@ mkdir -p work/ out/ log/
 . ~/.bashrc
 mamba activate jupy
 
-# run infercnv on celltypes
+# # run infercnv on celltypes
+# /software/team205/nextflow-23.04.1-all run nextflow/infercnv.nf \
+#   --out_dir $wd/out/Zhang2023/by_celltype/ \
+#   --mappings $wd/data/Zhang2023/mappings.csv \
+#   --annotations $wd/data/Zhang2023/annotations.tsv \
+#   --annotation_col 'celltype' \
+#   -c config/infercnv.config \
+#   -c /nfs/team205/kp9/nextflow/scomatic/LSF.config  \
+#   -w work/ \
+#   -resume
+    
+# run infercnv on clusters
 /software/team205/nextflow-23.04.1-all run nextflow/infercnv.nf \
-  --out_dir $wd/out/Zhang2023/by_celltype/ \
+  --out_dir $wd/out/Zhang2023/by_cluster/ \
   --mappings $wd/data/Zhang2023/mappings.csv \
+  --annotations $wd/data/Zhang2023/annotations.tsv \
+  --annotation_col 'cluster' \
+  -c config/infercnv.config \
+  -c /nfs/team205/kp9/nextflow/scomatic/LSF.config  \
+  -w work/
+#   -resume
+    
+# run test
+/software/team205/nextflow-23.04.1-all run nextflow/infercnv.nf \
+  --out_dir $wd/out/test/ \
+  --mappings <((head -1 $wd/data/Zhang2023/mappings.csv; grep 'BRI-527\|BRI-581' $wd/data/Zhang2023/mappings.csv)| cat) \
   --annotations $wd/data/Zhang2023/annotations.tsv \
[…]

### 2024-02-07 scomatic/reports/driver_coverage.Rmd
fix infercnv metadata propagation, change to standard mappings and annotations, add help nf functions
update driver coverage functions

@@
 
-#knitr::opts_knit$set(root.dir = '../')
 knitr::opts_chunk$set(
@@
 
+  # only drivers in ref_cds
+  avail_drivers <-
+    drivers %>%
+    dplyr::filter(gene %in% names(ref_cds))
+
   # get all positions in each gene
   gene_spans <-
-    ref_cds[c(drivers$gene)] %>%
+    ref_cds[c(avail_drivers$gene)] %>%
     purrr::map(function(i) {
@@
     })
-
-  # drop those not found
-  gene_spans[is.na(names(gene_spans))] <- NULL
 
@@
 
-```{r knit_plots, echo = F, warning = F, message = F, results = 'asis'}
+```{r knit_plots, echo = F, message = F, results = 'asis', warning = T, error = T}
 chunks <- c()
@@
 
-```{r mean_coverage}
+```{r mean_coverage, eval = F}
 cov %>%

### 2024-02-07 scomatic/src/03_knit_driver_coverage.sh

@@ -12,28 +12,28 @@ mamba activate jupy
 # # knit
 # Rscript -e "rmarkdown::render('reports/driver_coverage.Rmd', output_file = 'driver_coverage.html', output_dir = 'reports/', params = list(rerun = F))"
 
-# create mappings 
-bams_dir=/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/cellranger_output/
-ct_bams_dir=/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/out/Zhang2023/default_thresholds/
-out_dir=/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/out/Zhang2023/coverage/
-(
-  echo 'id,bam,celltype' ;
-  while read -r id ; do
-    # check if ct bams directory exists
-    if [ -d $ct_bams_dir/${id/-/_}/ ] ; then 
-      ct_bams=($(ls $ct_bams_dir/${id/-/_}/celltype_bams/*.bam))
-      # check if scomatic successfully ran
-      if [ ${#ct_bams[*]} -gt 0 ] ; then
-        echo "$id,$bams_dir/$id/possorted_genome_bam.bam,NA" ;
-        for file in ${ct_bams[@]} ; do
-          ct=$(basename $file | cut -d. -f2)
-          echo "$id,$file,$ct" ;
-        done
-      fi
-    fi
-  done < \
-  <(sed 1d /lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/data/Zhang2023/mappings.csv | cut -d, -f1 | sort -u)
-) | cat > data/Zhang2023/coverage_mappings.csv
+# # create mappings 
+# bams_dir=/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/cellranger_output/
+# ct_bams_dir=/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/out/Zhang2023/default_thresholds/
+# out_dir=/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/out/Zhang2023/coverage/
+# (
+#   echo 'id,bam,celltype' ;
+#   while read -r id ; do
+#     # check if ct bams directory exists
+#     if [ -d $ct_bams_dir/${id/-/_}/ ] ; then 
+#       ct_bams=($(ls $ct_bams_dir/${id/-/_}/celltype_bams/*.bam))
[…]

### 2024-02-07 infercnv/reports/infercnv.Rmd
change knit infercnv back to manual integration of all pngs

@@
     fig_width: 8
-    keep_md: true
-    code_folding: show
+    keep_md: false
+    code_folding: hide
     toc: true
@@
 params:
+  infercnv_dir: NULL
   rerun: true
@@
 # chunk options
+knitr::opts_knit$set(root.dir = '../')
 knitr::opts_chunk$set(
@@
 pngs <-
-  list.files(pattern = '^BRI', include.dirs = T) %>%
+  list.files(params$infercnv_dir,
+             pattern = '^BRI', 
+             include.dirs = T) %>%
   purrr::set_names() %>%
   purrr::map(function(id) {
+    # collect pngs
     lis <-
-      list.files(id, pattern = '.png', recursive = T, full.names = T) %>%
+      list.files(file.path(params$infercnv_dir, id), 
+                 pattern = '.png', 
+                 recursive = T, 
+                 full.names = T) %>%
       {setNames(., gsub('infercnv.', '', tools::file_path_sans_ext(basename(.))))}

### 2024-02-07 infercnv/src/01_run_infercnv.sh

@@ -34,12 +34,12 @@ mamba activate jupy
   -w work/
 #   -resume
     
-# run test
-/software/team205/nextflow-23.04.1-all run nextflow/infercnv.nf \
-  --out_dir $wd/out/test/ \
-  --mappings <((head -1 $wd/data/Zhang2023/mappings.csv; grep 'BRI-527\|BRI-581' $wd/data/Zhang2023/mappings.csv)| cat) \
-  --annotations $wd/data/Zhang2023/annotations.tsv \
-  --annotation_col 'celltype' \
-  -c config/infercnv.config \
-  -c /nfs/team205/kp9/nextflow/scomatic/LSF.config  \
-  -w work/test/
+# # run test
+# /software/team205/nextflow-23.04.1-all run nextflow/infercnv.nf \
+#   --out_dir $wd/out/test/ \
+#   --mappings <((head -1 $wd/data/Zhang2023/mappings.csv; grep 'BRI-527\|BRI-581' $wd/data/Zhang2023/mappings.csv)| cat) \
+#   --annotations $wd/data/Zhang2023/annotations.tsv \
+#   --annotation_col 'celltype' \
+#   -c config/infercnv.config \
+#   -c /nfs/team205/kp9/nextflow/scomatic/LSF.config  \
+#   -w work/test/

### 2024-02-07 infercnv/src/02_knit_infercnv.sh

@@ -10,4 +10,4 @@ cd $wd
 mamba activate jupy
 
 # knit
-Rscript -e "rmarkdown::render('reports/infercnv.Rmd', output_file = 'infercnv.html', output_dir = 'reports/', params = list(rerun = F))"
+Rscript -e "rmarkdown::render('reports/infercnv.Rmd', params = list(infercnv_dir = 'out/Zhang2023/by_celltype/', rerun = F), output_file = 'infercnv.html', output_dir = 'out/Zhang2023/by_celltype/')"

### 2024-02-07 infercnv/reports/gains_and_losses.Rmd
add gains_and_losses report file

---
title: "Gains and losses in rheumatoid arthritis data from inferCNV"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: false
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
params:
  infercnv_dir: NULL
  rerun: true
---

```{r setup, include = F, message = F, warning = F, class.source = 'fold-hide'}
# wd <- ifelse(Sys.info()['nodename'] == "mib119104s", '~/Volumes/', '') |> paste0('/lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/') ; setwd(wd)
# rmarkdown::render('reports/infercnv.Rmd', output_file = 'driver_coverage.html', output_dir = 'reports/', params = list(rerun = T))

# chunk options
knitr::opts_knit$set(root.dir = '../')
knitr::opts_chunk$set(
  warning = FALSE, 
  dpi = 300, 
  message = FALSE)

# libraries
library(magrittr)
library(ggplot2)
library(gridExtra)

# dev.off() if there is a device
dev_off_if <- function() {if (length(dev.list()) != 0) { dev.off() }}
```

inferCNV was run on the scRNA-seq data to call copy number variants in synovial cells. No reference group was used, so the average profile of all cells was used in each patient as the baseline.

First, we load the CITE-seq object and extract the metadata and UMAP embeddings.

```{r load_obj, results = 'asis'}
# infercnv output dir
infercnv_dir <- 'out/Zhang2023/by_celltype/'

# load object
obj <- 
  readRDS('/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/processed_output/all_cells_reference.rds')

# get umap embeddings
umap <-
  obj$umap$embedding %>%
  tibble::as_tibble() %>%
  dplyr::bind_cols(
    # add metadata
    obj$meta_data %>%
      tibble::as_tibble() %>%
      dplyr::mutate(id = sample)) 
```

Next, we add the CNV metadata from the inferCNV run. We must wrangle it from wide- to long-format.

```{r infercnv_output}
# get cnv metadata
cnvs <- 
  list.files(
    infercnv_dir,
    pattern = 'map_metadata_from_infercnv.txt',
    recursive = T) %>%
  purrr::set_names(., dirname(.)) %>%
  purrr::map(function(file) {
    file.path(infercnv_dir, file) %>% 
      read.table(sep = '\t') %>% 
      tibble::as_tibble(rownames = 'cell')
    }) %>%
  dplyr::bind_rows(.id = 'id') %>%
  # convert from wide to long format
  tidyr::pivot_longer(
    cols = tidyselect::starts_with(c('has_', 'proportion_', 'top_'))
  ) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(
    chr = name %>%
      gsub('.*_', '', .) %>%
      {forcats::fct_reorder(., gsub('chr', '', .) %>% as.numeric())},
    event = sub(paste0(chr, '_'), '', name),
    measure = sub(paste0(chr, '_', event), '', name)) %>%
  dplyr::ungroup()

# combine umap and cnvs
umap <-
  umap %>%
  # add cnv metadata
  dplyr::inner_join(cnvs, by = c('id', 'cell')) 
```

The `inferCNV` output file `map_metadata_from_infercnv.txt` contains columns with different CNV measures at whole-chromosome level per cell, with rownames denoting cells. The columns have the following values...

```{r echo = F, results = 'asis'}
tibble::tribble(
  ~name, ~`type - value`,
  '`has_*`', 'binary - whether any CNV / loss / duplication is found within the given chr',
  '`proportion_*`', 'proportion - proportion of genes that are part of any CNV / loss / duplication within the given chr',
  '`proportion_scaled_*`', 'proportion - same as proportion_*, but taking into account whether genes that are duplicated / lost are a single copy or two copies, as a weight',
  '`top_loss_*`', 'the top CNVs that affect the most genes'
) %>%
  knitr::kable()
```

We will define whole chromosome events as those affecting >90% of genes in a given chromosome.

```{r call_ploidy}

```

Now, we will plot gain and loss events by chromosome

```{r}
# function: plot gains and losses
plot_gains_and_losses <- 
  function(umap, annot) {
    umap %>%
      dplyr::rename(annotation = tidyselect::all_of(annot)) %>%
      dplyr::filter(measure == 'has', event != 'cnv') %>%
      dplyr::group_by(id, chr, event, annotation) %>%
      dplyr::summarise(n_cells = dplyr::n(),
                       n_events = sum(value),
                       prop_of_cells = n_events / n_cells) %>%
      ggplot(aes(x = id, y = chr, alpha = prop_of_cells, fill = event)) +
      geom_tile() +
      theme_classic() +
      theme(panel.spacing = unit(0, "lines")) +
      scale_fill_manual(values = c('loss' = scales::muted('blue'),
                                   'dupli' = scales::muted('red'))) +
      scale_alpha(range = c(0, 1)) +
      facet_grid(event ~ annotation)
  }

plot_gains_and_losses(umap, 'cell_type')
plot_gains_and_losses(umap, 'cluster_name')

# plot chr7 gain on umap
umap %>%
  dplyr::filter(measure == 'has', event == 'dupli', chr == 'chr7') %>%
  ggplot(aes(x = UMAP1, y = UMAP2, colour = cell_type, alpha = has_dupli_chr7)) +
  geom_point(size = 0.1) +
  theme_classic() +
  coord_fixed()
```

### 2024-02-07 scomatic/reports/driver_coverage.Rmd
add gains_and_losses report file

@@
   id: NULL
-  depths: NULL
   drivers: NULL
@@
   cov <-
-    readr::read_tsv('depths.tsv') %>% 
+    list.files(pattern = '_depths.tsv') %>% 
+    purrr::set_names(., gsub('_depths.tsv', '', .)) %>%
+    purrr::map(function(file) {
+      file %>%
+        readr::read_tsv() %>%
         dplyr::mutate(chr = as.character(chr))
+    }) %>%
+    dplyr::bind_rows(.id = 'id')
   cov
@@
   cov %>% 
-    dplyr::group_by(gene) %>%
-    dplyr::group_split() %>%
+    {split(., .$id)} %>%
+    purrr::map(function(df_id) {
+
+      # get id, inititate sublist
+      id_i <- unique(df_id$id)
+      p[[id_i]] <<- list()
+
+      # go by gene
+      df_id %>%
+        {split(., .$gene)} %>%
         purrr::map(function(df_i) {
 
-      # for testing:
-      # cov %>% dplyr::group_by(gene) %>% dplyr::group_split() %>% {.[[1]]} -> df_i
-
-      # get info
-      gene_i <- unique(df_i$gene)
-      p[[gene_i]] <<- list()
+          # get gene, initiate sublist
+          gene_i <- unique(df_gene$gene)
+          p[[id_i]][[gene_i]] <<- list()
 
           # get number of celltypes for plot dims
-      p[[gene_i]]$n_cts <<- 
+          p[[id_i]][[gene_i]]$n_cts <<- 
             df_i %>% 
@@
               # plot all
-          p[[gene_i]][[paste0('all', p_suffix)]] <<-
+              p[[id_i]][[gene_i]][[paste0('all', p_suffix)]] <<-
                 plot_all(
                   p_df = lvl,
-              p_exon_coords = exon_coords %>% dplyr::filter(gene == gene_i),
-              p_title = paste0(gene_i, p_suffix)
+                  p_exon_coords = exon_coords %>% dplyr::filter(gene == i),
+                  p_title = paste0(i, p_suffix)
                 )
@@
               # plot exons
-          p[[gene_i]][[paste0('exons', p_suffix)]] <<-
+              p[[id_i]][[gene_i]][[paste0('exons', p_suffix)]] <<-
                 plot_exons(
                   p_df = lvl,
-              p_title = paste0(gene_i, p_suffix)
+                  p_title = paste0(i, p_suffix)
                 )
+            })
         })

### 2024-02-07 infercnv/src/03_plot_gains_and_losses.R

@@ -1,3 +1,4 @@
+# libraries
 library(magrittr)
 library(ggplot2)
 

### 2024-02-08 infercnv/reports/gains_and_losses.Rmd
write output to infercnv output dir to take advantage of cacheing

@@
   infercnv_dir: NULL
+  cache_dir: NULL
   rerun: true
@@
   dpi = 300, 
-  message = FALSE)
+  message = FALSE,
+  cache.path = params$cache_dir)
 
@@
   readRDS('/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/processed_output/all_cells_reference.rds')
+```
+
+Next, we add the CNV metadata from the inferCNV run to the UMAP embeddings. We must wrangle it from wide- to long-format.
+
+```{r infercnv_output}
+cnvs <- xfun::cache_rds({
 
@@
         dplyr::mutate(id = sample))
-```
 
-Next, we add the CNV metadata from the inferCNV run. We must wrangle it from wide- to long-format.
-
-```{r infercnv_output}
   # get cnv metadata
@@
       cols = tidyselect::starts_with(c('has_', 'proportion_', 'top_'))
-  ) %>%
+    )
+
+  # get chr / event / name
+  names <-
+    cnvs %>%
+    dplyr::distinct(name) %>%
     dplyr::rowwise() %>%
     dplyr::mutate(
-    chr = name %>%
-      gsub('.*_', '', .) %>%
-      {forcats::fct_reorder(., gsub('chr', '', .) %>% as.numeric())},
-    event = sub(paste0(chr, '_'), '', name),
-    measure = sub(paste0(chr, '_', event), '', name)) %>%
+      chr = gsub('.*_', '', name),
+      event = gsub(paste0('_', chr), '', name) %>% sub('.*_', '', .),
+      measure = sub(paste0('_', event, '_', chr), '', name),
+      chr = dplyr::case_when(
+        measure == 'top' ~ paste0('chr', chr), TRUE ~ chr) %>% 
+        {forcats::fct_reorder(., gsub('chr', '', .) %>% as.numeric())}) %>%
     dplyr::ungroup()
 
-# combine umap and cnvs
-umap <-
+  # combine umap, names, and cnvs
+  cnvs <-
     umap %>%
     # add cnv metadata
-  dplyr::inner_join(cnvs, by = c('id', 'cell')) 
+    dplyr::inner_join(cnvs, by = c('id', 'cell')) %>%
+    # add names
+    dplyr::left_join(names, by = 'name')
+
+  # return
+  cnvs
+
+}, file = 'cnvs.rds')
 ```
@@
 
-We will define whole chromosome events as those affecting >90% of genes in a given chromosome.
-
-```{r call_ploidy}
-
-```
-
 Now, we will plot gain and loss events by chromosome
@@
 plot_gains_and_losses <- 
-  function(umap, annot) {
-    umap %>%
+  function(cnvs, annot) {
+    cnvs %>%
       dplyr::rename(annotation = tidyselect::all_of(annot)) %>%
@@
 
-plot_gains_and_losses(umap, 'cell_type')
-plot_gains_and_losses(umap, 'cluster_name')
+plot_gains_and_losses(cnvs, 'cell_type')
+plot_gains_and_losses(cnvs, 'cluster_name')
 
-# plot chr7 gain on umap
-umap %>%
-  dplyr::filter(measure == 'has', event == 'dupli', chr == 'chr7') %>%
-  ggplot(aes(x = UMAP1, y = UMAP2, colour = cell_type, alpha = has_dupli_chr7)) +
+# plot celltype
+cnvs %>%
+  dplyr::filter(measure == 'proportion', event == 'dupli', chr == 'chr7') %>%
+  ggplot(aes(x = UMAP1, y = UMAP2, colour = cell_type)) +
   geom_point(size = 0.1) +
@@
   coord_fixed()
+
+# plot subclusters on umap
+cnvs %>%
+  dplyr::filter(measure == 'proportion', event == 'dupli', chr == 'chr7') %>%
+  ggplot(aes(x = UMAP1, y = UMAP2, colour = subcluster)) +
+  geom_point(size = 0.1) +
+  theme_classic() +
+  coord_fixed()
+
+# plot chr7 gain on umap
+cnvs %>%
+  dplyr::filter(measure == 'proportion', event == 'dupli', chr == 'chr7') %>%
+  ggplot(aes(x = UMAP1, y = UMAP2, colour = value)) +
+  geom_point(size = 0.1) +
+  theme_classic() +
+  scale_colour_viridis() +
+  coord_fixed()
 ```
+
+#### chr7 gain
+
+We are most interested in the chr7 gain events in fibroblasts.
+
+We will define cells with chr7 gain as those affecting >90% of genes.
+
+```{r chr7_prop_dist}
+p_dat <-
+  cnvs %>%
+  dplyr::filter(measure == 'proportion', event == 'dupli', chr == 'chr7') %>%
+  dplyr::group_by(id, cluster_name, cell_type) %>%
+  dplyr::mutate(median_proportion = median(value)) %>%
+  dplyr::ungroup()
+
+p_dat %>%
+  ggplot(aes(x = value)) + 
+  geom_density() +
+  geom_vline(data = p_dat, aes(xintercept = median_proportion)) +
+  ggh4x::facet_nested(~ subcluster + cell_type)
+```
+
+##### Pseudobulk
+
+We pseudobulk chr7+ cells and perform differential expression analysis.
+
+```{r pseudobulk}
+
+```

### 2024-02-08 infercnv/src/01_run_infercnv.sh

@@ -1,5 +1,5 @@
 #!/bin/bash
-# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/ ; . ~/.bashrc ; mamba activate jupy ; /software/team205/bin/jsub lsf -q week -n infercnv_clusters -c 1 -m 2g -l log "bash src/01_run_infercnv.sh" | bsub
+# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/ ; . ~/.bashrc ; mamba activate jupy ; /software/team205/bin/jsub lsf -q week -n infercnv -c 1 -m 2g -l log "bash src/01_run_infercnv.sh" | bsub
 
 set -eo pipefail
 
@@ -23,17 +23,31 @@ mamba activate jupy
 #   -w work/ \
 #   -resume
     
-# run infercnv on clusters
+# # run infercnv on clusters
+# /software/team205/nextflow-23.04.1-all run nextflow/infercnv.nf \
+#   --out_dir $wd/out/Zhang2023/by_cluster/ \
+#   --mappings $wd/data/Zhang2023/mappings.csv \
+#   --annotations $wd/data/Zhang2023/annotations.tsv \
+#   --annotation_col 'cluster' \
+#   -c config/infercnv.config \
+#   -c /nfs/team205/kp9/nextflow/scomatic/LSF.config  \
+#   -w work/ \
+#   -resume
+
+# run infercnv on stromal clusters
+cat $wd/data/Zhang2023/annotations.tsv |
+awk -F'\t' '{if ((NR == 1) || ($2 == "Stromal cell")) {print} }' \
+> $wd/data/Zhang2023/stromal_annotations.tsv
 /software/team205/nextflow-23.04.1-all run nextflow/infercnv.nf \
-  --out_dir $wd/out/Zhang2023/by_cluster/ \
+  --out_dir $wd/out/Zhang2023/by_stromal_cluster/ \
   --mappings $wd/data/Zhang2023/mappings.csv \
-  --annotations $wd/data/Zhang2023/annotations.tsv \
+  --annotations $wd/data/Zhang2023/stromal_annotations.tsv \
   --annotation_col 'cluster' \
   -c config/infercnv.config \
   -c /nfs/team205/kp9/nextflow/scomatic/LSF.config  \
-  -w work/
-#   -resume
-    
+  -w work/ \
+  -resume
+
 # # run test
[…]

### 2024-02-08 infercnv/src/02_knit_infercnv.sh

@@ -4,11 +4,19 @@
 # wd
 wd=/lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/
 cd $wd
-mkdir out/Zhang2023/by_celltype/summary/
 
 # conda activate an environment which has docopt installed so jsub works
 . ~/.bashrc
 mamba activate jupy
 
-# knit
+# knit by_celltype
+mkdir out/Zhang2023/by_celltype/summary/
 Rscript -e "rmarkdown::render('reports/infercnv.Rmd', params = list(infercnv_dir = 'out/Zhang2023/by_celltype/', rerun = F), output_file = 'infercnv.html', output_dir = 'out/Zhang2023/by_celltype/summary/')"
+
+# knit by_cluster
+mkdir out/Zhang2023/by_cluster/summary/
+Rscript -e "rmarkdown::render('reports/infercnv.Rmd', params = list(infercnv_dir = 'out/Zhang2023/by_cluster/', rerun = F), output_file = 'infercnv.html', output_dir = 'out/Zhang2023/by_cluster/summary/')"
+
+# knit by_stromal_cluster
+mkdir out/Zhang2023/by_stromal_cluster/summary/
+Rscript -e "rmarkdown::render('reports/infercnv.Rmd', params = list(infercnv_dir = 'out/Zhang2023/by_stromal_cluster/', rerun = F), output_file = 'infercnv.html', output_dir = 'out/Zhang2023/by_stromal_cluster/summary/')"

### 2024-02-08 infercnv/src/03_knit_gains_and_losses.sh

@@ -1,14 +1,14 @@
 #!/bin/bash
-# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/ ; . ~/.bashrc ; mamba activate jupy ; /software/team205/bin/jsub lsf -q week -n knit_infercnv -c 1 -m 20g -l log "bash src/02_knit_infercnv.sh" | bsub
+# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/ ; . ~/.bashrc ; mamba activate jupy ; /software/team205/bin/jsub lsf -q week -n knit_gains_and_losses -c 1 -m 20g -l log "bash src/03_knit_gains_and_losses.sh" | bsub
 
 # wd
 wd=/lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/
 cd $wd
-mkdir out/Zhang2023/by_celltype/summary/
 
 # conda activate an environment which has docopt installed so jsub works
 . ~/.bashrc
 mamba activate jupy
 
-# knit
-Rscript -e "rmarkdown::render('reports/infercnv.Rmd', params = list(infercnv_dir = 'out/Zhang2023/by_celltype/', rerun = F), output_file = 'infercnv.html', output_dir = 'out/Zhang2023/by_celltype/summary/')"
+# knit by_celltype
+mkdir -p out/Zhang2023/by_celltype/summary/
+Rscript -e "rmarkdown::render('reports/gains_and_losses.Rmd', params = list(infercnv_dir = 'out/Zhang2023/by_celltype/', cache_dir = 'out/Zhang2023/by_celltype/summary/gains_and_losses_cache/', rerun = F), output_file = 'gains_and_losses.html', output_dir = 'out/Zhang2023/by_celltype/summary/')"

### 2024-02-09 infercnv/reports/gains_and_losses.Rmd
delete gains_and_losses

@@
       dplyr::rename(annotation = tidyselect::all_of(annot)) %>%
-      dplyr::filter(measure == 'has', event != 'cnv') %>%
+      dplyr::filter(measure == 'proportion', event != 'cnv') %>%
       dplyr::group_by(id, chr, event, annotation) %>%
-      dplyr::summarise(n_cells = dplyr::n(),
-                       n_events = sum(value),
-                       prop_of_cells = n_events / n_cells) %>%
-      ggplot(aes(x = id, y = chr, alpha = prop_of_cells, fill = event)) +
+      dplyr::summarise(proportion = mean(value)) %>%
+      ggplot(aes(x = chr, y = id, alpha = proportion, fill = event)) +
       geom_tile() +
       theme_classic() +
-      theme(panel.spacing = unit(0, "lines")) +
+      theme(panel.spacing = unit(0, "lines"),
+            axis.text.y = element_text(size = 3),
+            axis.text.x = element_text(angle = 270, hjust = 0, vjust = 0.5)) +
       scale_fill_manual(values = c('loss' = scales::muted('blue'),
@@
       scale_alpha(range = c(0, 1)) +
-      facet_grid(event ~ annotation)
+      facet_grid(annotation ~ event)
   }
 
+pdf('test.pdf', height = 20)
 plot_gains_and_losses(cnvs, 'cell_type')
 plot_gains_and_losses(cnvs, 'cluster_name')
+dev.off()
 
+pdf('umaps.pdf', height = 15)
+
+list(
   # plot celltype
@@
     ggplot(aes(x = UMAP1, y = UMAP2, colour = cell_type)) +
-  geom_point(size = 0.1) +
+    geom_point(size = 0.05) +
     theme_classic() +
-  coord_fixed()
-
-# plot subclusters on umap
-cnvs %>%
-  dplyr::filter(measure == 'proportion', event == 'dupli', chr == 'chr7') %>%
-  ggplot(aes(x = UMAP1, y = UMAP2, colour = subcluster)) +
-  geom_point(size = 0.1) +
-  theme_classic() +
-  coord_fixed()
-
+    coord_fixed(),
   # plot chr7 gain on umap
@@
     dplyr::filter(measure == 'proportion', event == 'dupli', chr == 'chr7') %>%
+    dplyr::arrange(value) %>%
     ggplot(aes(x = UMAP1, y = UMAP2, colour = value)) +
-  geom_point(size = 0.1) +
+    geom_point(size = 0.05) +
     theme_classic() +
-  scale_colour_viridis() +
-  coord_fixed()
+    scale_colour_viridis_c() +
+    coord_fixed() +
+    ggtitle('proportion of genes duplicated on chr7')
+) %>%
+  patchwork::wrap_plots(ncol = 1)
+
+dev.off()
 ```

### 2024-02-09 infercnv/src/01_run_infercnv.sh

@@ -1,5 +1,5 @@
 #!/bin/bash
-# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/ ; . ~/.bashrc ; mamba activate jupy ; /software/team205/bin/jsub lsf -q week -n infercnv -c 1 -m 2g -l log "bash src/01_run_infercnv.sh" | bsub
+# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/ ; . ~/.bashrc ; mamba activate jupy ; ~/bin/jsub lsf -q week -n infercnv -c 1 -m 2g -l log "bash src/01_run_infercnv.sh" | bsub
 
 set -eo pipefail
 
@@ -12,27 +12,27 @@ mkdir -p work/ out/ log/
 . ~/.bashrc
 mamba activate jupy
 
-# # run infercnv on celltypes
-# /software/team205/nextflow-23.04.1-all run nextflow/infercnv.nf \
-#   --out_dir $wd/out/Zhang2023/by_celltype/ \
-#   --mappings $wd/data/Zhang2023/mappings.csv \
-#   --annotations $wd/data/Zhang2023/annotations.tsv \
-#   --annotation_col 'celltype' \
-#   -c config/infercnv.config \
-#   -c /nfs/team205/kp9/nextflow/scomatic/LSF.config  \
-#   -w work/ \
-#   -resume
+# run infercnv on celltypes
+/nextflow-23.04.1-all run nextflow/infercnv.nf \
+  --out_dir $wd/out/Zhang2023/by_celltype/ \
+  --mappings $wd/data/Zhang2023/mappings.csv \
+  --annotations $wd/data/Zhang2023/annotations.tsv \
+  --annotation_col 'celltype' \
+  -c config/infercnv.config \
+  -c /nfs/team205/kp9/nextflow/scomatic/LSF.config  \
+  -w work/ \
+  -resume
     
-# # run infercnv on clusters
-# /software/team205/nextflow-23.04.1-all run nextflow/infercnv.nf \
-#   --out_dir $wd/out/Zhang2023/by_cluster/ \
-#   --mappings $wd/data/Zhang2023/mappings.csv \
-#   --annotations $wd/data/Zhang2023/annotations.tsv \
-#   --annotation_col 'cluster' \
-#   -c config/infercnv.config \
-#   -c /nfs/team205/kp9/nextflow/scomatic/LSF.config  \
-#   -w work/ \
-#   -resume
[…]

### 2024-02-09 infercnv/src/03_plot_gains_and_losses.R

@@ -1,92 +1,0 @@
-# libraries
-library(magrittr)
-library(ggplot2)
-
-obj <- readRDS('/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/processed_output/all_cells_reference.rds')
-
-cnvs <- 
-  list.files(
-    'out/Zhang2023/by_celltype/',
-    pattern = 'map_metadata_from_infercnv.txt',
-    recursive = T
-  ) %>%
-  purrr::set_names(., dirname(.)) %>%
-  purrr::map(function(file) {
-    paste0('out/Zhang2023/by_celltype/', file) %>% 
-      read.table(sep = '\t') %>% 
-      tibble::as_tibble(rownames = 'cell')
-    }) %>%
-  dplyr::bind_rows(.id = 'id')
-
-metadata <-
-  obj$meta_data %>%
-  tibble::as_tibble() %>%
-  dplyr::mutate(id = sample)
-
-umap <-
-  obj$umap$embedding %>%
-  tibble::as_tibble() %>%
-  dplyr::bind_cols(metadata) %>%
-  dplyr::left_join(cnvs, by = c('id', 'cell'))
-
-long_umap <-
-  umap %>%
-  dplyr::filter(!is.na(top_loss_1)) %>%
-  tidyr::pivot_longer(
-    cols = tidyselect::starts_with(c('has_', 'proportion_', 'top_'))
-  ) %>%
-  dplyr::rowwise() %>%
-  dplyr::mutate(
-    chr = name %>%
-      gsub('.*_', '', .) %>%
-      {forcats::fct_reorder(., gsub('chr', '', .) %>% as.numeric())},
-    event = sub(paste0(chr, '_'), '', name),
-    measure = sub(paste0(chr, '_', event), '', name))
-  
-# plot gains and losses
-plot_gains_and_losses <- 
-  function(long_umap, annot) {
-    long_umap %>%
-      dplyr::rename(annotation = tidyselect::all_of(annot)) %>%
-      dplyr::filter(measure == 'has', event != 'cnv') %>%
-      dplyr::group_by(id, chr, event, annotation) %>%
-      dplyr::summarise(n_cells = dplyr::n(),
-                       n_events = sum(value),
-                       prop_of_cells = n_events / n_cells) %>%
-      ggplot(aes(x = id, y = chr, alpha = prop_of_cells, fill = event)) +
-      geom_tile() +
-      theme_classic() +
[…]

### 2024-02-09 infercnv/reports/gains_and_losses.Rmd
fix infercnv.nf bug - used single quotes instead of doubles so variables were not taken; convert Zhang object to sce

--- 2024-02-08 infercnv/reports/gains_and_losses.Rmd
@@
 
+# libraries
+library(magrittr)
+library(ggplot2)
+library(gridExtra)
+library(Matrix)
+library(SummarizedExperiment)
+library(SingleCellExperiment)
+
+# dev.off() if there is a device
+dev_off_if <- function() {if (length(dev.list()) != 0) { dev.off() }}
+
 # chunk options
@@
   cache.path = params$cache_dir) 
-
-# libraries
-library(magrittr)
-library(ggplot2)
-library(gridExtra)
-
-# dev.off() if there is a device
-dev_off_if <- function() {if (length(dev.list()) != 0) { dev.off() }}
 ```
@@
 
-First, we load the CITE-seq object and extract the metadata and UMAP embeddings.
+First, we load the CITE-seq object. We convert it to a SingleCellExperiment.
 
 ```{r load_obj, results = 'asis'}
-# infercnv output dir
+# dirs
+data_dir <- '/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/processed_output/'
 infercnv_dir <- 'out/Zhang2023/by_celltype/'
+
+# convert to sce
+sce <- xfun::cache_rds({
 
@@
   obj <- 
-  readRDS('/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/processed_output/all_cells_reference.rds')
+    readRDS(file.path(data_dir, 'all_cells_reference.rds'))
+
+  # load rna data
+  assays <-
+    list(
+      counts = 
+        readRDS(file.path(data_dir, 'raw_mRNA_count_matrix.rds')),
+      logcounts = 
+        readRDS(file.path(data_dir, 'qc_mRNA_314011cells_log_normalized_matrix.rds')))
+
+  # get cell order from counts
+  cell_order <- colnames(assays$logcounts)
+
+  # filter and reorder to cells that pass qc from counts
+  assays$counts <- 
+    assays$counts[rownames(assays$logcounts), 
+                  cell_order]
+
+  # generate coldata
+  colData <-
+    obj$meta_data[cell_order, ]
+
+  # populate sce
+  sce <- 
+    SingleCellExperiment::SingleCellExperiment(
+      assays = assays,
+      colData = colData
+    )
+
+  # name, reorder, and add umap embeddings
+  umap <- obj$umap$embedding
+  rownames(umap) <- rownames(obj$meta_data)
+  umap <- umap[cell_order, ]
+  reducedDim(sce, 'UMAP_uwot') <- umap
+
+  # return
+  sce
+
+}, file = 'sce.rds')
+
+sce <- xfun::cache_rds({
+
+  # get cnv metadata, reorder, left join to colData
+  cnvs <- 
+    list.files(
+      infercnv_dir,
+      pattern = 'map_metadata_from_infercnv.txt',
+      recursive = T) %>%
+    purrr::set_names(., dirname(.)) %>%
+    purrr::map(function(file) {
+      file.path(infercnv_dir, file) %>% 
+        read.table(sep = '\t')
+      }) %>%
+    dplyr::bind_rows()
+
+  # left join cnvs to coldata, reorder, restore rownames
+  colData <- merge(colData, cnvs, by = 'row.names', all.x = T)
+  rownames(colData) <- colData[, 1]
+  colData <- colData[, -1]
+  colData <- colData[colnames(sce), ]
+
+  # add to sce
+  sce@colData <- colData
+
+  # return
+  sce
+
+}, file = 'sce_w_infercnv.rds')
 ```
@@
 cnvs <- xfun::cache_rds({
-
-  # get umap embeddings
-  umap <- 
-    obj$umap$embedding %>%
-    tibble::as_tibble() %>%
-    dplyr::bind_cols(
-      # add metadata
-      obj$meta_data %>%
-        tibble::as_tibble() %>%
-        dplyr::mutate(id = sample))
 
@@
   cnvs <-
-    umap %>%
-    # add cnv metadata
-    dplyr::inner_join(cnvs, by = c('id', 'cell')) %>%
+    cnvs %>%
     # add names
@@
 
-```{r}
+```{r plot_gains_and_losses, fig.height = 20}
 # function: plot gains and losses
@@
       scale_alpha(range = c(0, 1)) +
-      facet_grid(event ~ annotation)
+      facet_grid(annotation ~ event)
   }
@@
 plot_gains_and_losses(cnvs, 'cluster_name')
+```
 
+```{r plot_umaps, fig.height = 15}
+list(
   # plot celltype
@@
     dplyr::filter(measure == 'proportion', event == 'dupli', chr == 'chr7') %>%
+    dplyr::arrange(value) %>%
     ggplot(aes(x = UMAP1, y = UMAP2, colour = value)) +
-  geom_point(size = 0.1) +
+    geom_point(size = 0.05) +
     theme_classic() +
-  scale_colour_viridis() +
-  coord_fixed()
+    scale_colour_viridis_c() +
+    coord_fixed() +
+    ggtitle('proportion of genes duplicated on chr7')
+) %>%
+  patchwork::wrap_plots(ncol = 1)
 ```
@@
   ggh4x::facet_nested(~ subcluster + cell_type)
+
+plot_prop_chr7_gain <- function(p_dat, position_arg) {
+  p_dat %>%
+    ggplot() +
+    geom_bar(aes(x = cell_type, fill = value, group = value),
+             position = position_arg) +
+    theme_classic() +
+    scale_fill_viridis_b(n.breaks = 9) 
+}
+
+plot_prop_chr7_gain(p_dat, position_stack(reverse = T))
+plot_prop_chr7_gain(p_dat, position_fill(reverse = T))
+p_dat %>%
+  {split(., .$id)} %>%
+  purrr::map2(names(.), ., function(id, df) {
+    plot_prop_chr7_gain(df, position_fill(reverse = T)) +
+      ggtitle(paste0('chr7 gain count - ', id))
+    plot_prop_chr7_gain(df, position_stack(reverse = T)) +
+      ggtitle(paste0('chr7 gain proportion - ', id))
+  })
 ```
@@
 We pseudobulk chr7+ cells and perform differential expression analysis.
+In order to do this, we use the `glmGamPoi` package in R.
+This analysis follows [this tutorial](https://bioconductor.org/packages/devel/bioc/vignettes/glmGamPoi/inst/doc/pseudobulk.html).
 
 ```{r pseudobulk}
+pseudobulk <- xfun::cache_rds({
 
+  # pseudobulk the sce object by chr7 status
+  reduced_sce <-
+    glmGamPoi::pseudobulk(
+      sce, 
+      group_by = vars()
+    )
+
+}, file = 'pseudobulk.rds')
 ```

### 2024-02-09 infercnv/src/01_run_infercnv.sh

@@ -13,40 +13,40 @@ mkdir -p work/ out/ log/
 mamba activate jupy
 
 # run infercnv on celltypes
-/nextflow-23.04.1-all run nextflow/infercnv.nf \
+/software/team205//nextflow-23.04.1-all run nextflow/infercnv.nf \
   --out_dir $wd/out/Zhang2023/by_celltype/ \
   --mappings $wd/data/Zhang2023/mappings.csv \
   --annotations $wd/data/Zhang2023/annotations.tsv \
   --annotation_col 'celltype' \
   -c config/infercnv.config \
   -c /nfs/team205/kp9/nextflow/scomatic/LSF.config  \
-  -w work/ \
-  -resume
+  -w work/ #\
+  #-resume
     
-# run infercnv on clusters
-/software/team205/nextflow-23.04.1-all run nextflow/infercnv.nf \
-  --out_dir $wd/out/Zhang2023/by_cluster/ \
-  --mappings $wd/data/Zhang2023/mappings.csv \
-  --annotations $wd/data/Zhang2023/annotations.tsv \
-  --annotation_col 'cluster' \
-  -c config/infercnv.config \
-  -c /nfs/team205/kp9/nextflow/scomatic/LSF.config  \
-  -w work/ \
-  -resume
-
-# run infercnv on stromal clusters
-cat $wd/data/Zhang2023/annotations.tsv |
-awk -F'\t' '{if ((NR == 1) || ($2 == "Stromal cell")) {print} }' \
-> $wd/data/Zhang2023/stromal_annotations.tsv
-/software/team205/nextflow-23.04.1-all run nextflow/infercnv.nf \
-  --out_dir $wd/out/Zhang2023/by_stromal_cluster/ \
-  --mappings $wd/data/Zhang2023/mappings.csv \
-  --annotations $wd/data/Zhang2023/stromal_annotations.tsv \
-  --annotation_col 'cluster' \
-  -c config/infercnv.config \
-  -c /nfs/team205/kp9/nextflow/scomatic/LSF.config  \
-  -w work/ \
-  -resume
+# # run infercnv on clusters
+# /software/team205/nextflow-23.04.1-all run nextflow/infercnv.nf \
+#   --out_dir $wd/out/Zhang2023/by_cluster/ \
+#   --mappings $wd/data/Zhang2023/mappings.csv \
+#   --annotations $wd/data/Zhang2023/annotations.tsv \
[…]

### 2024-02-09 infercnv/reports/gains_and_losses.Rmd
use dittoseq for plotting umaps

@@
 
-First, we load the CITE-seq object. We convert it to a SingleCellExperiment.
+First, we load the CITE-seq object. We convert it to a SingleCellExperiment object.
 
@@
 }, file = 'sce.rds')
+```
 
+Next, we add the CNV metadata from the inferCNV run to the object.
+
+```{r get_infercnv}
 sce <- xfun::cache_rds({
@@
 
-Next, we add the CNV metadata from the inferCNV run to the UMAP embeddings. We must wrangle it from wide- to long-format.
-
-```{r infercnv_output}
-cnvs <- xfun::cache_rds({
-
-  # get cnv metadata
-  cnvs <- 
-    list.files(
-      infercnv_dir,
-      pattern = 'map_metadata_from_infercnv.txt',
-      recursive = T) %>%
-    purrr::set_names(., dirname(.)) %>%
-    purrr::map(function(file) {
-      file.path(infercnv_dir, file) %>% 
-        read.table(sep = '\t') %>% 
-        tibble::as_tibble(rownames = 'cell')
-      }) %>%
-    dplyr::bind_rows(.id = 'id') %>%
-    # convert from wide to long format
-    tidyr::pivot_longer(
-      cols = tidyselect::starts_with(c('has_', 'proportion_', 'top_'))
-    )
-
-  # get chr / event / name
-  names <-
-    cnvs %>%
-    dplyr::distinct(name) %>%
-    dplyr::rowwise() %>%
-    dplyr::mutate(
-      chr = gsub('.*_', '', name),
-      event = gsub(paste0('_', chr), '', name) %>% sub('.*_', '', .),
-      measure = sub(paste0('_', event, '_', chr), '', name),
-      chr = dplyr::case_when(
-        measure == 'top' ~ paste0('chr', chr), TRUE ~ chr) %>% 
-        {forcats::fct_reorder(., gsub('chr', '', .) %>% as.numeric())}) %>%
-    dplyr::ungroup()
-
-  # combine umap, names, and cnvs
-  cnvs <-
-    cnvs %>%
-    # add names
-    dplyr::left_join(names, by = 'name')
-
-  # return
-  cnvs
-
-}, file = 'cnvs.rds')
-```
-
 The `inferCNV` output file `map_metadata_from_infercnv.txt` contains columns with different CNV measures at whole-chromosome level per cell, with rownames denoting cells.
@@
 
-Now, we will plot gain and loss events by chromosome
+First, we plot UMAPs by ID and celltype.
+
+```{r plot_umaps, fig.height = 10}
+dittoSeq::dittoDimPlot(sce, 'cell_type', size = 0.2) +
+  coord_fixed()
+```
+
+```{r plot_umaps_by_id, fig.height = 30}
+dittoSeq::dittoDimPlot(sce, 'cell_type', size = 0.2, 
+                       split.by = 'sample') +
+  coord_fixed()
+```
+
+Then, we visualise chr7 and chr5 gains.
+
+```{r plot_chr_gains}
+dittoSeq::dittoDimPlot(
+  sce, 'proportion_dupli_chr7', size = 0.2) +
+  coord_fixed()
+dittoSeq::dittoDimPlot(
+  sce, 'proportion_dupli_chr7', size = 0.2, split.by = 'sample') +
+  coord_fixed()
+dittoSeq::dittoDimPlot(
+  sce, 'proportion_dupli_chr5', size = 0.2) +
+  coord_fixed()
+dittoSeq::dittoDimPlot(
+  sce, 'proportion_dupli_chr5', size = 0.2, split.by = 'sample') +
+  coord_fixed()
+```
+
+Now, we will plot gain and loss events by chromosome.
 
 ```{r plot_gains_and_losses, fig.height = 20}
+# mean proportion loss / gain per chromosome per celltype per id
+p_dat <-
+  sce@colData %>%
+  tibble::as_tibble() %>%
+  dplyr::group_by(sample, cell_type) %>%
+  dplyr::summarise(
+    dplyr::across(dplyr::starts_with('proportion_'), ~ mean(.x, na.rm = T))) %>%
+  dplyr::ungroup()
+
+# split gain and loss
+p_dat <-
+  c('dupli', 'loss') %>% 
+  purrr::set_names() %>%
+  purrr::map(function(direction) {
+    list(
+      matrix = 
+        p_dat %>% 
+        dplyr::select(dplyr::contains(paste0('proportion_', direction))) %>%
+        dplyr::filter(!is.na(paste0('proportion_', direction, '_chr1'))) %>%
+        as.matrix(),
+      annot = p_dat %>% dplyr::select(sample, cell_type)
+    )
+  })
+
+# plot heatmaps
+pheatmap::pheatmap(p_dat$dupli$matrix, cluster_cols = F, cluster_rows = F)
+pheatmap::pheatmap(sce@colData[,grepl('proportion_dupli', colnames(sce@colData))])
+
 # function: plot gains and losses
@@
 
-We are most interested in the chr7 gain events in fibroblasts.
-
-We will define cells with chr7 gain as those affecting >90% of genes.
+We are most interested in the chr7 gain events in fibroblasts. We visualise their distribution across the dataset.
 
@@
 p_dat <-
-  cnvs %>%
-  dplyr::filter(measure == 'proportion', event == 'dupli', chr == 'chr7') %>%
-  dplyr::group_by(id, cluster_name, cell_type) %>%
-  dplyr::mutate(median_proportion = median(value)) %>%
-  dplyr::ungroup()
-
-p_dat %>%
-  ggplot(aes(x = value)) + 
-  geom_density() +
-  geom_vline(data = p_dat, aes(xintercept = median_proportion)) +
-  ggh4x::facet_nested(~ subcluster + cell_type)
+  sce@colData %>%
+  tibble::as_tibble() %>%
+  dplyr::filter(!is.na(proportion_dupli_chr7))
 
@@
     ggplot() +
-    geom_bar(aes(x = cell_type, fill = value, group = value),
+    geom_bar(aes(x = cell_type, 
+                 fill = proportion_dupli_chr7, 
+                 group = proportion_dupli_chr7),
              position = position_arg) +
+    scale_x_discrete(guide = guide_axis(angle = -45)) +
     theme_classic() +
@@
 
-plot_prop_chr7_gain(p_dat, position_stack(reverse = T))
-plot_prop_chr7_gain(p_dat, position_fill(reverse = T))
+plot_prop_chr7_gain(p_dat, position_stack(reverse = T)) +
+  ggtitle('chr7 gain count - all cells')
+plot_prop_chr7_gain(p_dat, position_fill(reverse = T)) +
+  ggtitle('chr7 gain proportion - all cells')
 p_dat %>%
@@
 
-```{r pseudobulk}
+We will define cells with chr7 gain as those affecting >90% of genes.
+
+```{r pseudobulk_chr7_gain}
 pseudobulk <- xfun::cache_rds({
+
+  # we assign chr7 status as >90% of genes affected by a duplication
+  sce@colData$chr7_gain <-
+    as.numeric(sce@colData$proportion_dupli_chr7 > 0.9)
 
@@
   reduced_sce <-
-    glmGamPoi::pseudobulk(
+    glmGamPoi::pseudobulk_sce(
       sce, 
-      group_by = vars()
+      group_by = vars(chr7_gain, cell_type)
     )

### 2024-02-12 infercnv/reports/gains_and_losses.Rmd
reorganise gains_and_losses report

@@
 
-First, we load the CITE-seq object. We convert it to a SingleCellExperiment object.
+First, we load the CITE-seq object. We convert it to a SingleCellExperiment object. Next, we add the CNV metadata from the inferCNV run to the object.
 
@@
   # generate coldata
-  colData <-
+  col_data <-
     obj$meta_data[cell_order, ]
@@
       assays = assays,
-      colData = colData
+      colData = col_data
     )
@@
 
-  # return
-  sce
-
-}, file = 'sce.rds')
-```
-
-Next, we add the CNV metadata from the inferCNV run to the object.
-
-```{r get_infercnv}
-sce <- xfun::cache_rds({
-
-  # get cnv metadata, reorder, left join to colData
+  # get cnv metadata
   cnvs <- 
@@
   # left join cnvs to coldata, reorder, restore rownames
-  colData <- merge(colData, cnvs, by = 'row.names', all.x = T)
-  rownames(colData) <- colData[, 1]
-  colData <- colData[, -1]
-  colData <- colData[colnames(sce), ]
+  col_data <- merge(col_data, cnvs, by = 'row.names', all.x = T)
+  rownames(col_data) <- col_data[, 1]
+  col_data <- 
+    col_data[colnames(sce), -1] %>%
+    DataFrame()
 
   # add to sce
-  sce@colData <- colData
+  sce@colData <- col_data
 
@@
 
-}, file = 'sce_w_infercnv.rds')
+}, file = 'sce.rds')
 ```
@@
   purrr::map(function(direction) {
+    p_dat_i <-
+      p_dat %>% 
+        dplyr::filter(!is.na(paste0('proportion_', direction, '_chr1')))
     list(
       matrix = 
-        p_dat %>% 
+        p_dat_i %>%
         dplyr::select(dplyr::contains(paste0('proportion_', direction))) %>%
-        dplyr::filter(!is.na(paste0('proportion_', direction, '_chr1'))) %>%
         as.matrix(),
-      annot = p_dat %>% dplyr::select(sample, cell_type)
+      annot = 
+        p_dat_i %>% 
+        dplyr::select(sample, cell_type)
     )
@@
 pheatmap::pheatmap(sce@colData[,grepl('proportion_dupli', colnames(sce@colData))])
-
-# function: plot gains and losses
-plot_gains_and_losses <- 
-  function(cnvs, annot) {
-    cnvs %>%
-      dplyr::rename(annotation = tidyselect::all_of(annot)) %>%
-      dplyr::filter(measure == 'proportion', event != 'cnv') %>%
-      dplyr::group_by(id, chr, event, annotation) %>%
-      dplyr::summarise(proportion = mean(value)) %>%
-      ggplot(aes(x = chr, y = id, alpha = proportion, fill = event)) +
-      geom_tile() +
-      theme_classic() +
-      theme(panel.spacing = unit(0, "lines"),
-            axis.text.y = element_text(size = 3),
-            axis.text.x = element_text(angle = 270, hjust = 0, vjust = 0.5)) +
-      scale_fill_manual(values = c('loss' = scales::muted('blue'),
-                                   'dupli' = scales::muted('red'))) +
-      scale_alpha(range = c(0, 1)) +
-      facet_grid(annotation ~ event)
-  }
-
-plot_gains_and_losses(cnvs, 'cell_type')
-plot_gains_and_losses(cnvs, 'cluster_name')
-```
-
-```{r plot_umaps, fig.height = 15}
-list(
-  # plot celltype
-  cnvs %>%
-    dplyr::filter(measure == 'proportion', event == 'dupli', chr == 'chr7') %>%
-    ggplot(aes(x = UMAP1, y = UMAP2, colour = cell_type)) +
-    geom_point(size = 0.05) +
-    theme_classic() +
-    coord_fixed(),
-  # plot chr7 gain on umap
-  cnvs %>%
-    dplyr::filter(measure == 'proportion', event == 'dupli', chr == 'chr7') %>%
-    dplyr::arrange(value) %>%
-    ggplot(aes(x = UMAP1, y = UMAP2, colour = value)) +
-    geom_point(size = 0.05) +
-    theme_classic() +
-    scale_colour_viridis_c() +
-    coord_fixed() +
-    ggtitle('proportion of genes duplicated on chr7')
-) %>%
-  patchwork::wrap_plots(ncol = 1)
 ```
@@
 
-We will define cells with chr7 gain as those affecting >90% of genes.
+We will define cells with chr7 gain as those affecting >90% of genes. We will pseudobulk by patient, celltype, and chr7 ploidy to produce a reduced object.
 
 ```{r pseudobulk_chr7_gain}
-pseudobulk <- xfun::cache_rds({
+reduced_sce <- xfun::cache_rds({
 
@@
 
-  # pseudobulk the sce object by chr7 status
+  # pseudobulk the sce object by patient, celltype, and chr7 ploidy
   reduced_sce <-
@@
       sce, 
-      group_by = vars(chr7_gain, cell_type)
+      group_by = vars(sample, cell_type, chr7_gain),
+      n_cells = n()
     )
 
-}, file = 'pseudobulk.rds')
+  # remove NAs
+  reduced_sce <- reduced_sce[, !is.na(reduced_sce$cell_type)]
+
+  # return
+  reduced_sce
+
+}, file = 'reduced_sce.rds')
 ```
+
+With the reduced data, we can conduct differential expression analysis, treating the data like bulk RNA-seq data.
+
+```{r deseq, message = T}
+# use DESeq2's size factor calculation procedure
+fit <-
+  glmGamPoi::glm_gp(
+    reduced_sce,
+    design = ~ chr7_gain * cell_type + sample,
+    size_factor = 'ratio',
+    verbose = T
+  )
+res <-
+  glmGamPoi::test_de(
+    fit, 
+    contrast = 
+      cond(cell_type = 'Stromal cell', chr7_gain = 1) -
+      cond(cell_type = 'Stromal cell', chr7_gain = 0)
+  )
+```
+
+```{r plot_de}
+res %>%
+  ggplot(aes(x = lfc, y = log10(pval))) +
+  geom_point(aes(colour = adj_pval < 0.01), size = 0.5) +
+  t
+```

### 2024-02-12 infercnv/src/03_knit_gains_and_losses.sh

@@ -1,5 +1,5 @@
 #!/bin/bash
-# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/ ; . ~/.bashrc ; mamba activate jupy ; /software/team205/bin/jsub lsf -q week -n knit_gains_and_losses -c 1 -m 20g -l log "bash src/03_knit_gains_and_losses.sh" | bsub
+# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/ ; . ~/.bashrc ; mamba activate jupy ; /software/team205/bin/jsub lsf -q week -n knit_gains_and_losses -c 1 -m 50g -l log "bash src/03_knit_gains_and_losses.sh" | bsub
 
 # wd
 wd=/lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/

### 2024-02-12 infercnv/reports/gains_and_losses.Rmd
add hclust ceiling to pheatmap

@@
 library(SingleCellExperiment)
+library(glmGamPoi)
 
@@
 
-}, file = 'sce.rds')
+}, file = 'sce.rds', rerun = params$rerun)
 ```
@@
 
-```{r plot_umaps_by_id, fig.height = 30}
-dittoSeq::dittoDimPlot(sce, 'cell_type', size = 0.2, 
-                       split.by = 'sample') +
-  coord_fixed()
-```
-
-Then, we visualise chr7 and chr5 gains.
+Then, we visualise chr7 gains on the UMAP.
 
@@
 dittoSeq::dittoDimPlot(
-  sce, 'proportion_dupli_chr7', size = 0.2) +
-  coord_fixed()
-dittoSeq::dittoDimPlot(
-  sce, 'proportion_dupli_chr7', size = 0.2, split.by = 'sample') +
-  coord_fixed()
-dittoSeq::dittoDimPlot(
-  sce, 'proportion_dupli_chr5', size = 0.2) +
-  coord_fixed()
-dittoSeq::dittoDimPlot(
-  sce, 'proportion_dupli_chr5', size = 0.2, split.by = 'sample') +
+  sce, 'proportion_dupli_chr7', size = 0.1) +
   coord_fixed()
@@
 
-```{r plot_gains_and_losses, fig.height = 20}
+First, we plot single cell-level CNVs by direction and by celltype.
+
+```{r plot_gains_and_losses_by_ct, fig.height = 20, eval = F}
+# at single-cell level
+unique(sce$cell_type) %>%
+  purrr::walk(function(ct) {
+    c('dupli', 'loss') %>%
+      purrr::walk(function(direction) {
+        p_dat <-
+          sce@colData[!is.na(sce$proportion_cnv_chr1) & sce$cell_type == ct, ]
+        p_mat <-
+          p_dat[, grepl(paste0('proportion_', direction), colnames(p_dat))]
+        p_annot <- 
+          p_dat[, c('sample'), drop = F] %>%
+          as.data.frame()
+        p <-
+          pheatmap::pheatmap(
+            p_mat, 
+            show_rownames = F, 
+            annotation_row = p_annot,
+            # hclust error: size cannot be NA nor exceed 65536
+            cluster_rows = ifelse(nrow(p_mat) > 65536, F, T),
+            legend = F,
+            main = paste0('celltype: ', ct, '\ndirection: ', direction))
+        print(p)
+      })
+  })
+```
+
+We also plot sample-x-celltype level events (mean-summarised).
+
+```{r plot_gains_and_losses_by_id}
 # mean proportion loss / gain per chromosome per celltype per id
@@
   dplyr::group_by(sample, cell_type) %>%
+  dplyr::filter(!is.na(proportion_cnv_chr1)) %>%
   dplyr::summarise(
     dplyr::across(dplyr::starts_with('proportion_'), ~ mean(.x, na.rm = T))) %>%
-  dplyr::ungroup()
+  dplyr::ungroup() %>%
+  dplyr::mutate(sample_x_ct = paste0(sample, '_', cell_type)) %>%
+  tibble::column_to_rownames('sample_x_ct')
 
@@
   c('dupli', 'loss') %>% 
-  purrr::set_names() %>%
   purrr::map(function(direction) {
-    p_dat_i <-
-      p_dat %>% 
-        dplyr::filter(!is.na(paste0('proportion_', direction, '_chr1')))
-    list(
-      matrix = 
-        p_dat_i %>%
-        dplyr::select(dplyr::contains(paste0('proportion_', direction))) %>%
-        as.matrix(),
-      annot = 
-        p_dat_i %>% 
-        dplyr::select(sample, cell_type)
-    )
+    p_mat <-
+      p_dat[, grepl(paste0('proportion_', direction), colnames(p_dat))]
+    p_annot <- 
+      p_dat[, c('sample', 'cell_type')] %>%
+      as.data.frame()
+    pheatmap::pheatmap(
+      p_mat, 
+      show_rownames = F, 
+      annotation_row = p_annot,
+      main = paste0('direction: ', direction))
   })
-
-# plot heatmaps
-pheatmap::pheatmap(p_dat$dupli$matrix, cluster_cols = F, cluster_rows = F)
-pheatmap::pheatmap(sce@colData[,grepl('proportion_dupli', colnames(sce@colData))])
 ```
@@
 ```{r chr7_prop_dist}
+# prep data
 p_dat <-
@@
   tibble::as_tibble() %>%
-  dplyr::filter(!is.na(proportion_dupli_chr7))
+  dplyr::filter(!is.na(proportion_dupli_chr7)) %>%
+  dplyr::group_by(sample) %>%
+  dplyr::mutate(n_cells = dplyr::n())
 
-plot_prop_chr7_gain <- function(p_dat, position_arg) {
+# define functions
+plot_prop_chr7_gain <- function(p_dat, title) {
+  plot_func <- function(p_dat, position_arg) {
     p_dat %>%
@@
       theme_classic() +
-    scale_fill_viridis_b(n.breaks = 9) 
+      scale_fill_viridis_b(n.breaks = 9, limits = c(0, 1)) 
   }
 
-plot_prop_chr7_gain(p_dat, position_stack(reverse = T)) +
-  ggtitle('chr7 gain count - all cells')
-plot_prop_chr7_gain(p_dat, position_fill(reverse = T)) +
-  ggtitle('chr7 gain proportion - all cells')
+  # plot both count (stack) and proportion (fill), annotate n cells
+  list(
+    plot_func(p_dat, position_fill(reverse = T)) +
+      ggtitle(title) +
+      theme(axis.text.x = element_blank(),
+            axis.title.x = element_blank()) +
+      ylab('count'),
+    plot_func(p_dat, position_stack(reverse = T)) +
+      ylab('proportion')) %>%
+    patchwork::wrap_plots(ncol = 1) 
+}
+
+# plot, across all and by patient
+plot_prop_chr7_gain(p_dat, 'chr7 gain - all cells')
 p_dat %>%
-  {split(., .$id)} %>%
-  purrr::map2(names(.), ., function(id, df) {
-    plot_prop_chr7_gain(df, position_fill(reverse = T)) +
-      ggtitle(paste0('chr7 gain count - ', id))
-    plot_prop_chr7_gain(df, position_stack(reverse = T)) +
-      ggtitle(paste0('chr7 gain proportion - ', id))
+  {split(., .$sample)} %>%
+  purrr::walk2(names(.), ., function(id, df) {
+    df %>%
+    plot_prop_chr7_gain(
+      paste0('chr7 gains - ', id, ' (', unique(df$n_cells), ' cells)')) %>%
+      print()
   })
@@
 In order to do this, we use the `glmGamPoi` package in R.
-This analysis follows [this tutorial](https://bioconductor.org/packages/devel/bioc/vignettes/glmGamPoi/inst/doc/pseudobulk.html).
+This analysis follows a [tutorial](https://bioconductor.org/packages/devel/bioc/vignettes/glmGamPoi/inst/doc/pseudobulk.html) and the [glmGamPoi GitHub](https://github.com/const-ae/glmGamPoi).
 
@@
 
-  # pseudobulk the sce object by patient, celltype, and chr7 ploidy
+  # pseudobulk the sce object by patient and chr7 ploidy
+  # subset to stromal cells
   reduced_sce <-
-    glmGamPoi::pseudobulk_sce(
-      sce, 
-      group_by = vars(sample, cell_type, chr7_gain),
-      n_cells = n()
+    glmGamPoi::pseudobulk(
+      sce[, sce$cell_type == 'Stromal cell'], 
+      group_by = vars(sample, chr7_gain)
     )
 
-  # remove NAs
-  reduced_sce <- reduced_sce[, !is.na(reduced_sce$cell_type)]
+  # remove NA's
+  reduced_sce <-
+    reduced_sce[, !is.na(reduced_sce$sample) & !is.na(reduced_sce$chr7_gain)]
 
@@
 
-}, file = 'reduced_sce.rds')
+}, file = 'reduced_sce.rds', rerun = params$rerun)
 ```
@@
 
+We will identify which genes in stromal cells are changed most by chr7 gain. We do this by fitting one GLM model for each gene, with the chr7 status and the patient as factors.
+
 ```{r deseq, message = T}
-# use DESeq2's size factor calculation procedure
+# fit one GLM model for each gene
+fit <- xfun::cache_rds({
   fit <-
@@
     reduced_sce,
-    design = ~ chr7_gain * cell_type + sample,
-    size_factor = 'ratio',
+    design = ~ chr7_gain + sample,
     verbose = T
   )
-res <-
+  fit
+}, file = 'fit.rds', rerun = params$rerun)
+
+# show model output
+summary(fit)
+```
+
+Next, we test for differential gene expression.
+
+```{r test_ge}
+# get differential expression
+de_res <-
   glmGamPoi::test_de(
@@
     contrast = 
-      cond(cell_type = 'Stromal cell', chr7_gain = 1) -
-      cond(cell_type = 'Stromal cell', chr7_gain = 0)
+      cond(chr7_gain = 1) -
+      cond(chr7_gain = 0)
   )
@@
 ```{r plot_de}
-res %>%
-  ggplot(aes(x = lfc, y = log10(pval))) +
-  geom_point(aes(colour = adj_pval < 0.01), size = 0.5) +
-  t
+# minimum accepted adj_pval
+min_p <- 0.05
+
+# define up/downregulation
+de_res <-
+  de_res %>% 
+  tibble::as_tibble() %>%
+  dplyr::mutate(
+    direction = dplyr::case_when(
+      lfc < -1.5 & adj_pval < min_p ~ 'downregulated',
+      lfc > 1.5 & adj_pval < min_p ~ 'upregulated',
+      TRUE ~ NA
+    )
+  )
+
+# plot
+de_res %>%
+  ggplot(aes(x = lfc, y = -log10(adj_pval), colour = direction)) +
+  geom_point(size = 0.5) +
+  geom_hline(yintercept = -log10(min_p), linetype = 'dashed') +
+  geom_vline(xintercept = c(-1.5, 1.5), linetype = 'dashed') +
+  ggrepel::geom_text_repel(
+    data = 
+      de_res %>% 
+      dplyr::filter(!is.na(direction)) %>%
+      dplyr::group_by(direction) %>%
+      dplyr::filter(rank(adj_pval) <= 10 | rank(lfc, ties.method = 'min') <= 10),
+    aes(label = name),
+    min.segment.length = 0, nudge_x = 0.5) +
+  theme_classic() +
+  scale_colour_manual(values = c('upregulated' = 'darkred',
+                                 'downregulated' = 'darkblue'), 
+                      na.value = 'darkgrey')
 ```
+
+These are the most strongly downregulated genes:
+
+```{r get_downreg_degs}
+de_res %>%
+  tibble::as_tibble() %>% 
+  dplyr::filter(adj_pval < 0.01) %>%
+  dplyr::arrange(-lfc, adj_pval) %>%
+  head(10) %>%
+  knitr::kable()
+```
+
+These are the most strongly upregulated genes:
+
+```{r get_upreg_degs}
+de_res %>%
+  tibble::as_tibble() %>% 
+  dplyr::filter(adj_pval < 0.01) %>%
+  dplyr::arrange(lfc, adj_pval) %>%
+  head(10) %>%
+  knitr::kable()
+```

### 2024-02-12 infercnv/src/01_run_infercnv.sh

@@ -12,16 +12,16 @@ mkdir -p work/ out/ log/
 . ~/.bashrc
 mamba activate jupy
 
-# run infercnv on celltypes
-/software/team205//nextflow-23.04.1-all run nextflow/infercnv.nf \
-  --out_dir $wd/out/Zhang2023/by_celltype/ \
-  --mappings $wd/data/Zhang2023/mappings.csv \
-  --annotations $wd/data/Zhang2023/annotations.tsv \
-  --annotation_col 'celltype' \
-  -c config/infercnv.config \
-  -c /nfs/team205/kp9/nextflow/scomatic/LSF.config  \
-  -w work/ #\
-  #-resume
+# # run infercnv on celltypes
+# /software/team205//nextflow-23.04.1-all run nextflow/infercnv.nf \
+#   --out_dir $wd/out/Zhang2023/by_celltype/ \
+#   --mappings $wd/data/Zhang2023/mappings.csv \
+#   --annotations $wd/data/Zhang2023/annotations.tsv \
+#   --annotation_col 'celltype' \
+#   -c config/infercnv.config \
+#   -c /nfs/team205/kp9/nextflow/scomatic/LSF.config  \
+#   -w work/ \
+#   -resume
     
 # # run infercnv on clusters
 # /software/team205/nextflow-23.04.1-all run nextflow/infercnv.nf \
@@ -33,20 +33,20 @@ mamba activate jupy
 #   -c /nfs/team205/kp9/nextflow/scomatic/LSF.config  \
 #   -w work/ \
 #   -resume
-# 
-# # run infercnv on stromal clusters
-# cat $wd/data/Zhang2023/annotations.tsv |
-# awk -F'\t' '{if ((NR == 1) || ($2 == "Stromal cell")) {print} }' \
-# > $wd/data/Zhang2023/stromal_annotations.tsv
-# /software/team205/nextflow-23.04.1-all run nextflow/infercnv.nf \
-#   --out_dir $wd/out/Zhang2023/by_stromal_cluster/ \
-#   --mappings $wd/data/Zhang2023/mappings.csv \
-#   --annotations $wd/data/Zhang2023/stromal_annotations.tsv \
-#   --annotation_col 'cluster' \
-#   -c config/infercnv.config \
-#   -c /nfs/team205/kp9/nextflow/scomatic/LSF.config  \
-#   -w work/ \
-#   -resume
+
+# run infercnv on stromal clusters
[…]

### 2024-02-13 infercnv/reports/gains_and_losses.Rmd
implement infercnv_dir parameter in gains_and_losses

@@
 
+# Pre-processing
+
 inferCNV was run on the scRNA-seq data to call copy number variants in synovial cells. No reference group was used, so the average profile of all cells was used in each patient as the baseline.
@@
 data_dir <- '/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/processed_output/'
-infercnv_dir <- 'out/Zhang2023/by_celltype/'
+infercnv_dir <- params$infercnv_dir
 
-# convert to sce
+# load gene coords
+coords <-
+  readr::read_tsv('data/hg38_gencode_v27.txt', col_names = c('name', 'chr', 'start', 'stop'))
+
+# convert obj to sce
 sce <- xfun::cache_rds({
@@
 
-```{r plot_umaps, fig.height = 10}
+```{r plot_umaps}
 dittoSeq::dittoDimPlot(sce, 'cell_type', size = 0.2) +
@@
 
-```{r plot_gains_and_losses_by_ct, fig.height = 20, eval = F}
+```{r plot_gains_and_losses_by_ct}
 # at single-cell level
 unique(sce$cell_type) %>%
-  purrr::walk(function(ct) {
+  purrr::map(function(ct) {
     c('dupli', 'loss') %>%
-      purrr::walk(function(direction) {
+      purrr::map(function(direction) {
         p_dat <-
@@
           as.data.frame()
-        p <-
         pheatmap::pheatmap(
@@
           legend = F,
-            main = paste0('celltype: ', ct, '\ndirection: ', direction))
-        print(p)
+          main = paste0('celltype: ', ct, '\ndirection: ', direction)) %>%
+          print()
       })
@@
       annotation_row = p_annot,
-      main = paste0('direction: ', direction))
+      main = paste0('direction: ', direction)) %>%
+      print()
   })
@@
 
-#### chr7 gain
+# chr7 gain
 
@@
 
-##### Pseudobulk
+# Pseudobulking and DGEA
 
@@
 
-  # we assign chr7 status as >90% of genes affected by a duplication
+  # we assign chr7 status as 
+  # >80% of genes affected by a duplication -> chr7+
+  # <20% of genes affected by a duplication -> chr7 WT
   sce@colData$chr7_gain <-
-    as.numeric(sce@colData$proportion_dupli_chr7 > 0.9)
+    ifelse(sce$proportion_dupli_chr7 > 0.8,
+           1, ifelse(sce$proportion_dupli_chr7 < 0.2, 0, NA))
 
@@
     glmGamPoi::pseudobulk(
-      sce[, sce$cell_type == 'Stromal cell'], 
+      sce[, sce$cell_type == 'Stromal cell' & !is.na(sce$chr7_gain)], 
       group_by = vars(sample, chr7_gain)
@@
 ```{r test_ge}
+# minimum accepted adj_pval
+min_p <- 0.05
+
 # get differential expression
+de_res <- xfun::cache_rds({
+
+  # test differences
   de_res <-
@@
     )
-```
 
-```{r plot_de}
-# minimum accepted adj_pval
-min_p <- 0.05
-
-# define up/downregulation
+  # define significant up/downregulation and position
   de_res <-
@@
         lfc > 1.5 & adj_pval < min_p ~ 'upregulated',
-      TRUE ~ NA
-    )
-  )
+        TRUE ~ NA)) %>%
+    dplyr::left_join(coords)
 
+  de_res
+}, file = 'de_res.rds', rerun = params$rerun)
+```
+
+```{r plot_de}
 # plot
 de_res %>%
-  ggplot(aes(x = lfc, y = -log10(adj_pval), colour = direction)) +
-  geom_point(size = 0.5) +
+  ggplot(aes(x = lfc, y = -log10(adj_pval), colour = direction, 
+             size = chr == 'chr7', shape = chr == 'chr7')) +
+  geom_point() +
   geom_hline(yintercept = -log10(min_p), linetype = 'dashed') +
+  geom_vline(xintercept = log2(1.5), linetype = 'dashed', colour = 'purple') +
   geom_vline(xintercept = c(-1.5, 1.5), linetype = 'dashed') +
@@
       dplyr::filter(rank(adj_pval) <= 10 | rank(lfc, ties.method = 'min') <= 10),
-    aes(label = name),
-    min.segment.length = 0, nudge_x = 0.5) +
+    aes(label = name, ),
+    min.segment.length = 0, nudge_x = 0.7, show.legend = F) +
   theme_classic() +
@@
                                  'downregulated' = 'darkblue'), 
-                      na.value = 'darkgrey')
+                      na.value = 'darkgrey') +
+  scale_size_manual(values = c('TRUE' = 3, 'FALSE' = 2), na.value = 2) +
+  scale_shape_manual(values = c('TRUE' = 55, 'FALSE' = 16), na.value = 16)
 ```
 
-These are the most strongly downregulated genes:
+These are the most strongly upregulated genes:
 
@@
 
-These are the most strongly upregulated genes:
+These are the most strongly downregulated genes:
 

### 2024-02-13 infercnv/src/03_knit_gains_and_losses.sh

@@ -9,6 +9,10 @@ cd $wd
 . ~/.bashrc
 mamba activate jupy
 
-# knit by_celltype
-mkdir -p out/Zhang2023/by_celltype/summary/
-Rscript -e "rmarkdown::render('reports/gains_and_losses.Rmd', params = list(infercnv_dir = 'out/Zhang2023/by_celltype/', cache_dir = 'out/Zhang2023/by_celltype/summary/gains_and_losses_cache/', rerun = F), output_file = 'gains_and_losses.html', output_dir = 'out/Zhang2023/by_celltype/summary/')"
+# # knit by_celltype
+# mkdir -p out/Zhang2023/by_celltype/summary/
+# Rscript -e "rmarkdown::render('reports/gains_and_losses.Rmd', params = list(infercnv_dir = 'out/Zhang2023/by_celltype/', cache_dir = 'out/Zhang2023/by_celltype/summary/gains_and_losses_cache/', rerun = F), output_file = 'gains_and_losses.html', output_dir = 'out/Zhang2023/by_celltype/summary/')"
+
+# knit by_stromal_cluster
+mkdir -p out/Zhang2023/by_stromal_cluster/summary/
+Rscript -e "rmarkdown::render('reports/gains_and_losses.Rmd', params = list(infercnv_dir = 'out/Zhang2023/by_stromal_cluster/', cache_dir = 'out/Zhang2023/by_stromal_cluster/summary/gains_and_losses_cache/', rerun = F), output_file = 'gains_and_losses.html', output_dir = 'out/Zhang2023/by_stromal_cluster/summary/')"

### 2024-02-14 infercnv/reports/gains_and_losses.Rmd
looking for sub-chromosome-level copy number events, add nebulosa density plotting for umap, change gains_and_losses pseudobulk to all cells

@@
     fig_width: 8
-    keep_md: false
+    keep_md: true
     code_folding: hide
@@
 
-# Pre-processing
+# Loading the inferCNV outputs
 
@@
 
-First, we load the CITE-seq object. We convert it to a SingleCellExperiment object. Next, we add the CNV metadata from the inferCNV run to the object.
+First, we load the CITE-seq object. We convert it to a SingleCellExperiment object. Then we add the CNV metadata from the inferCNV run to the object.
 
@@
 data_dir <- '/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/processed_output/'
-infercnv_dir <- params$infercnv_dir
 
@@
 sce <- xfun::cache_rds({
+
+  # load CTAP and metadata
+  ctap <-
+    readr::read_tsv('data/Zhang2023/CTAP_donor_mapping.tsv') %>%
+    dplyr::select(sample = donor, CTAP)
+  md <- 
+    readr::read_tsv('data/Zhang2023/SYNAPSE_METADATA_MANIFEST.tsv') %>%
+    dplyr::distinct()
 
@@
     list.files(
-      infercnv_dir,
+      params$infercnv_dir,
       pattern = 'map_metadata_from_infercnv.txt',
@@
     purrr::map(function(file) {
-      file.path(infercnv_dir, file) %>% 
+      file.path(params$infercnv_dir, file) %>% 
         read.table(sep = '\t')
       }) %>%
-    dplyr::bind_rows()
+    dplyr::bind_rows() %>%
+    dplyr::mutate(infercnv = T)
 
   # left join cnvs to coldata, reorder, restore rownames
-  col_data <- merge(col_data, cnvs, by = 'row.names', all.x = T)
+  col_data <- merge(col_data, cnvs, by = 'row.names', all.x = T, sort = F)
   rownames(col_data) <- col_data[, 1]
@@
 }, file = 'sce.rds', rerun = params$rerun)
+
+ # remove NA's for inferCNV analysis
+ sce <- sce[, !is.na(sce$proportion_cnv_chr1)]
 ```
+
+## `map_metadata_from_infercnv.txt`
 
@@
 
-First, we plot UMAPs by ID and celltype.
+## HMM regional predictions
+
+We also load the CNV regions prediced by HMM. According to [their wiki](https://github.com/broadinstitute/infercnv/wiki/infercnv-i6-HMM-type#infercnv-i6-hmm)...
+
+As well as aneuploidy, we want to look for smaller recurrent structural changes, including focal-/arm-level events.
+
+> The inferCNV i6 HMM is a six-state CNV model that predicts the following CNV levels:
+>
+> * State 1: 0x: complete loss
+> * State 2: 0.5x: loss of one copy
+> * State 3: 1x: neutral
+> * State 4: 1.5x: addition of one copy
+> * State 5: 2x: addition of two copies
+> * State 6: 3x: essentially a placeholder for >2x copies but modeled as 3x.
+
+```{r get_hmm}
+cnv_regions <- xfun::cache_rds({
+  cnv_regions <- 
+    list.files(
+      params$infercnv_dir,
+      pattern = 'HMM_CNV_predictions.HMMi6.leiden.hmm_mode-subclusters.Pnorm_0.5.pred_cnv_regions.dat',
+      recursive = T) %>%
+    purrr::set_names(., dirname(.)) %>%
+    purrr::map(function(file) {
+      file.path(params$infercnv_dir, file) %>% 
+        readr::read_tsv()
+      }) %>%
+    dplyr::bind_rows(.id = 'sample') %>%
+    # refactor chromosomes
+    dplyr::mutate(
+      chr = chr %>% {forcats::fct_reorder(., gsub('chr', '', .) %>% as.numeric())},
+      direction = dplyr::case_when(
+        state < 3 ~ 'loss',
+        state > 3 ~ 'dupli',
+        TRUE ~ 'neutral'),
+      region = gsub('.*region_', '', cnv_name) %>% as.numeric())
+}, file = 'cnv_regions', rerun = params$rerun)
+```
+
+```{r plot_cnv_regions_pred}
+pdf('test.pdf', height = 25)
+cnv_regions %>%
+  dplyr::filter(direction != 'neutral') %>% 
+  ggplot(aes(x = region)) +
+  geom_bar(position = 'stack') +
+  facet_wrap(~ chr, ncol = 1, scales = 'free') +
+  theme_classic() +
+  scale_x_discrete(guide = guide_axis(angle = 90)) +
+  theme(legend.position = 'none')
+dev.off()
+```
+
+## UMAPs
+
+First, we plot UMAPs by celltype.
 
@@
 
-Then, we visualise chr7 gains on the UMAP.
+Then, we visualise gains and loss proportions for each chromosome on the UMAP.
 
-```{r plot_chr_gains}
-dittoSeq::dittoDimPlot(
-  sce, 'proportion_dupli_chr7', size = 0.1) +
-  coord_fixed()
+```{r plot_chr_gains, fig.width = 10, fig.height = 5}
+pdf('test.pdf')
+paste0('chr', 1:22) %>% head(2) %>%
+  purrr::map(function(chr) {
+    sce %>%
+      Nebulosa::plot_density(
+        paste('proportion', c('dupli', 'loss'), chr, sep = '_'),
+        size = 0.5) 
+  })
 ```
 
-Now, we will plot gain and loss events by chromosome.
+## Heatmaps
+
+Now, we will plot gain and loss proportions by chromosome.
 
@@
 # at single-cell level
-unique(sce$cell_type) %>%
+sce$cell_type[!is.na(sce$infercnv)] %>%
+  unique() %>%
   purrr::map(function(ct) {
@@
           sce@colData[!is.na(sce$proportion_cnv_chr1) & sce$cell_type == ct, ]
+        if (nrow(p_dat) > 0) {
           p_mat <-
@@
             print()
+        }
       })
@@
 
-# chr7 gain
+## chr7 gain
 
@@
     glmGamPoi::pseudobulk(
-      sce[, sce$cell_type == 'Stromal cell' & !is.na(sce$chr7_gain)], 
-      group_by = vars(sample, chr7_gain)
+      sce[, !is.na(sce$sample) & 
+            !is.na(sce$chr7_gain) &
+            !is.na(sce$cell_type)], 
+      group_by = vars(sample, cell_type, chr7_gain)
     )
-
-  # remove NA's
-  reduced_sce <-
-    reduced_sce[, !is.na(reduced_sce$sample) & !is.na(reduced_sce$chr7_gain)]
 
@@
     reduced_sce,
-    design = ~ chr7_gain + sample,
+    design = ~ chr7_gain + sample + cell_type,
     verbose = T
@@
       contrast = 
-        cond(chr7_gain = 1) -
-        cond(chr7_gain = 0)
+        cond(chr7_gain = 1, cell_type = 'Stromal cells') -
+        cond(chr7_gain = 0, cell_type = 'Stromal cells')
     )

### 2024-02-14 infercnv/src/02_knit_infercnv.sh

@@ -9,13 +9,13 @@ cd $wd
 . ~/.bashrc
 mamba activate jupy
 
-# knit by_celltype
-mkdir out/Zhang2023/by_celltype/summary/
-Rscript -e "rmarkdown::render('reports/infercnv.Rmd', params = list(infercnv_dir = 'out/Zhang2023/by_celltype/', rerun = F), output_file = 'infercnv.html', output_dir = 'out/Zhang2023/by_celltype/summary/')"
-
-# knit by_cluster
-mkdir out/Zhang2023/by_cluster/summary/
-Rscript -e "rmarkdown::render('reports/infercnv.Rmd', params = list(infercnv_dir = 'out/Zhang2023/by_cluster/', rerun = F), output_file = 'infercnv.html', output_dir = 'out/Zhang2023/by_cluster/summary/')"
+# # knit by_celltype
+# mkdir out/Zhang2023/by_celltype/summary/
+# Rscript -e "rmarkdown::render('reports/infercnv.Rmd', params = list(infercnv_dir = 'out/Zhang2023/by_celltype/', rerun = F), output_file = 'infercnv.html', output_dir = 'out/Zhang2023/by_celltype/summary/')"
+# 
+# # knit by_cluster
+# mkdir out/Zhang2023/by_cluster/summary/
+# Rscript -e "rmarkdown::render('reports/infercnv.Rmd', params = list(infercnv_dir = 'out/Zhang2023/by_cluster/', rerun = F), output_file = 'infercnv.html', output_dir = 'out/Zhang2023/by_cluster/summary/')"
 
 # knit by_stromal_cluster
 mkdir out/Zhang2023/by_stromal_cluster/summary/

### 2024-02-14 infercnv/src/03_knit_gains_and_losses.sh

@@ -1,5 +1,5 @@
 #!/bin/bash
-# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/ ; . ~/.bashrc ; mamba activate jupy ; /software/team205/bin/jsub lsf -q week -n knit_gains_and_losses -c 1 -m 50g -l log "bash src/03_knit_gains_and_losses.sh" | bsub
+# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/ ; . ~/.bashrc ; mamba activate jupy ; /software/team205/bin/jsub lsf -q week -n knit_gains_and_losses -c 1 -m 70g -l log "bash src/03_knit_gains_and_losses.sh" | bsub
 
 # wd
 wd=/lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/
@@ -9,10 +9,10 @@ cd $wd
 . ~/.bashrc
 mamba activate jupy
 
-# # knit by_celltype
-# mkdir -p out/Zhang2023/by_celltype/summary/
-# Rscript -e "rmarkdown::render('reports/gains_and_losses.Rmd', params = list(infercnv_dir = 'out/Zhang2023/by_celltype/', cache_dir = 'out/Zhang2023/by_celltype/summary/gains_and_losses_cache/', rerun = F), output_file = 'gains_and_losses.html', output_dir = 'out/Zhang2023/by_celltype/summary/')"
+# knit by_celltype
+mkdir -p out/Zhang2023/by_celltype/summary/
+Rscript -e "rmarkdown::render('reports/gains_and_losses.Rmd', params = list(infercnv_dir = 'out/Zhang2023/by_celltype/', cache_dir = 'out/Zhang2023/by_celltype/summary/gains_and_losses_cache/', rerun = F), output_file = 'gains_and_losses.html', output_dir = 'out/Zhang2023/by_celltype/summary/')"
 
-# knit by_stromal_cluster
-mkdir -p out/Zhang2023/by_stromal_cluster/summary/
-Rscript -e "rmarkdown::render('reports/gains_and_losses.Rmd', params = list(infercnv_dir = 'out/Zhang2023/by_stromal_cluster/', cache_dir = 'out/Zhang2023/by_stromal_cluster/summary/gains_and_losses_cache/', rerun = F), output_file = 'gains_and_losses.html', output_dir = 'out/Zhang2023/by_stromal_cluster/summary/')"
+# # knit by_stromal_cluster
[…]

### 2024-02-16 infercnv/reports/gains_and_losses.Rmd
updating gains_and_losses

@@
         TRUE ~ 'neutral'),
-      region = gsub('.*region_', '', cnv_name) %>% as.numeric())
+      ploidy = dplyr::case_when(
+        state == 1 ~ '0x',
+        state == 2 ~ '0.5x',
+        state == 3 ~ '1x',
+        state == 4 ~ '1.5x',
+        state == 5 ~ '2x',
+        state == 6 ~ '>2x',
+        TRUE ~ NA
+        ) %>% forcats::fct_reorder(state),
+      region = gsub('.*region_', '', cnv_name) %>% as.numeric(),
+      cell_type = cell_group_name %>% gsub('\\..*', '', .))
+  cnv_regions$midpoint <-
+    (cnv_regions$start + cnv_regions$end) / 2
+  cnv_regions
 }, file = 'cnv_regions', rerun = params$rerun)
@@
 
-```{r plot_cnv_regions_pred}
-pdf('test.pdf', height = 25)
-cnv_regions %>%
-  dplyr::filter(direction != 'neutral') %>% 
-  ggplot(aes(x = region)) +
-  geom_bar(position = 'stack') +
-  facet_wrap(~ chr, ncol = 1, scales = 'free') +
-  theme_classic() +
-  scale_x_discrete(guide = guide_axis(angle = 90)) +
-  theme(legend.position = 'none')
-dev.off()
+```{r plot_cnv_regions_pred, fig.height = 20, fig.width = 10}
+p_dat %>%
+  ggplot(aes(x = pos, fill = ploidy, group = ploidy)) +
+  geom_density(data = p_dat %>% dplyr::filter(direction == 'dupli'),
+               aes(y = ..density..), 
+               position = position_stack(), 
+               colour = NA) +
+  geom_density(data = p_dat %>% dplyr::filter(direction == 'loss'),
+               aes(y = -..density..), 
+               position = position_stack(reverse = T), 
+               colour = NA) +
+  ggh4x::facet_grid2(chr ~ cell_type, scales = 'free', independent = 'x') +
+  theme_bw() +
+  theme(axis.text.x = element_blank()) +
+  scale_fill_manual(
+    values = c('0x' = '#00008B', '0.5x' = '#b4b0ff', '1' = 'white', 
+               '1.5x' = '#ffc2c2', '2x' = '#d64242', '>2x' = '#8B0000')) 
 ```
@@
 ```{r plot_chr_gains, fig.width = 10, fig.height = 5}
-pdf('test.pdf')
 paste0('chr', 1:22) %>% head(2) %>%
@@
         paste('proportion', c('dupli', 'loss'), chr, sep = '_'),
-        size = 0.5) 
+        size = 0.5) %>%
+      print()
   })
@@
       contrast = 
-        cond(chr7_gain = 1, cell_type = 'Stromal cells') -
-        cond(chr7_gain = 0, cell_type = 'Stromal cells')
+        cond(chr7_gain = 1, cell_type = 'Stromal cell') -
+        cond(chr7_gain = 0, cell_type = 'Stromal cell')
     )
@@
 ```
+
+Next, we plot the top differentially expressed genes onto the plot of locally amplified / lost regions within chr7 to see where they fall.
+
+```{r plot_des_on_regions}
+
+```

### 2024-02-16 infercnv/src/03_knit_gains_and_losses.sh

@@ -11,7 +11,7 @@ mamba activate jupy
 
 # knit by_celltype
 mkdir -p out/Zhang2023/by_celltype/summary/
-Rscript -e "rmarkdown::render('reports/gains_and_losses.Rmd', params = list(infercnv_dir = 'out/Zhang2023/by_celltype/', cache_dir = 'out/Zhang2023/by_celltype/summary/gains_and_losses_cache/', rerun = F), output_file = 'gains_and_losses.html', output_dir = 'out/Zhang2023/by_celltype/summary/')"
+Rscript -e "rmarkdown::render('reports/gains_and_losses.Rmd', params = list(infercnv_dir = 'out/Zhang2023/by_celltype/', cache_dir = 'out/Zhang2023/by_celltype/summary/gains_and_losses_cache/', rerun = T), output_file = 'gains_and_losses.html', output_dir = 'out/Zhang2023/by_celltype/summary/')"
 
 # # knit by_stromal_cluster
 # mkdir -p out/Zhang2023/by_stromal_cluster/summary/

### 2024-02-16 infercnv/reports/gains_and_losses.Rmd
move sce wrangling out of rmd, make gains_and_losses.rmd accept different cell_group_name values

--- 2024-02-14 infercnv/reports/gains_and_losses.Rmd
@@
 params:
+  sce: NULL
+  cell_group_name: NULL
   infercnv_dir: NULL
@@
 
-```{r load_obj, results = 'asis'}
-# dirs
-data_dir <- '/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/processed_output/'
-
+```{r load_sce, results = 'asis'}
 # load gene coords
@@
 
-# convert obj to sce
+# load CTAP and metadata
+ctap <-
+  readr::read_tsv('data/Zhang2023/syn26710600/CTAP_donor_mapping.tsv') %>%
+  dplyr::select(sample = donor, CTAP)
+md <- 
+  readr::read_tsv('data/Zhang2023/syn26710600/SYNAPSE_METADATA_MANIFEST.tsv') %>%
+  dplyr::distinct()
+
+# load sce, add inferCNV 
 sce <- xfun::cache_rds({
 
-  # load CTAP and metadata
-  ctap <-
-    readr::read_tsv('data/Zhang2023/CTAP_donor_mapping.tsv') %>%
-    dplyr::select(sample = donor, CTAP)
-  md <- 
-    readr::read_tsv('data/Zhang2023/SYNAPSE_METADATA_MANIFEST.tsv') %>%
-    dplyr::distinct()
-
-  # load object
-  obj <- 
-    readRDS(file.path(data_dir, 'all_cells_reference.rds'))
-
-  # load rna data
-  assays <-
-    list(
-      counts = 
-        readRDS(file.path(data_dir, 'raw_mRNA_count_matrix.rds')),
-      logcounts = 
-        readRDS(file.path(data_dir, 'qc_mRNA_314011cells_log_normalized_matrix.rds')))
-
-  # get cell order from counts
-  cell_order <- colnames(assays$logcounts)
-
-  # filter and reorder to cells that pass qc from counts
-  assays$counts <- 
-    assays$counts[rownames(assays$logcounts), 
-                  cell_order]
-
-  # generate coldata
-  col_data <-
-    obj$meta_data[cell_order, ]
-
-  # populate sce
-  sce <- 
-    SingleCellExperiment::SingleCellExperiment(
-      assays = assays,
-      colData = col_data
-    )
-
-  # name, reorder, and add umap embeddings
-  umap <- obj$umap$embedding
-  rownames(umap) <- rownames(obj$meta_data)
-  umap <- umap[cell_order, ]
-  reducedDim(sce, 'UMAP_uwot') <- umap
+  # load sce
+  sce <- readRDS(params$sce)
 
@@
   # left join cnvs to coldata, reorder, restore rownames
-  col_data <- merge(col_data, cnvs, by = 'row.names', all.x = T, sort = F)
+  col_data <- merge(sce@colData, cnvs, by = 'row.names', all.x = T, sort = F)
   rownames(col_data) <- col_data[, 1]
@@
 sce <- sce[, !is.na(sce$proportion_cnv_chr1)]
+
+ # assign cell group
+sce@colData$cell_group <-
+  sce@colData[, params$cell_group_name]
 ```
@@
 cnv_regions <- xfun::cache_rds({
+
+  # get HMM predictions
   cnv_regions <- 
@@
       file.path(params$infercnv_dir, file) %>% 
-        readr::read_tsv()
+        readr::read_tsv(show_col_types = F)
       }) %>%
@@
         TRUE ~ 'neutral'),
-      region = gsub('.*region_', '', cnv_name) %>% as.numeric())
-}, file = 'cnv_regions', rerun = params$rerun)
+      ploidy = dplyr::case_when(
+        state == 1 ~ '0x',
+        state == 2 ~ '0.5x',
+        state == 3 ~ '1x',
+        state == 4 ~ '1.5x',
+        state == 5 ~ '2x',
+        state == 6 ~ '>2x',
+        TRUE ~ NA
+        ) %>% forcats::fct_reorder(state),
+      region = gsub('.*region_', '', cnv_name) %>% as.numeric(),
+      cell_group = cell_group_name %>% gsub('\\..*', '', .))
+
+  # calculate region midpoints for plotting
+  cnv_regions$midpoint <-
+    (cnv_regions$start + cnv_regions$end) / 2
+
+  # return
+  cnv_regions
+
+}, file = 'cnv_regions.rds', rerun = params$rerun)
 ```
 
-```{r plot_cnv_regions_pred}
-pdf('test.pdf', height = 25)
+```{r plot_cnv_regions_pred, fig.height = 20, fig.width = 10}
 cnv_regions %>%
-  dplyr::filter(direction != 'neutral') %>% 
-  ggplot(aes(x = region)) +
-  geom_bar(position = 'stack') +
-  facet_wrap(~ chr, ncol = 1, scales = 'free') +
-  theme_classic() +
-  scale_x_discrete(guide = guide_axis(angle = 90)) +
-  theme(legend.position = 'none')
-dev.off()
+  ggplot(aes(x = midpoint, fill = ploidy, group = ploidy)) +
+  geom_density(data = cnv_regions %>% dplyr::filter(direction == 'dupli'),
+               aes(y = ..density..), 
+               position = position_stack(), 
+               colour = NA) +
+  geom_density(data = cnv_regions %>% dplyr::filter(direction == 'loss'),
+               aes(y = -..density..), 
+               position = position_stack(reverse = T), 
+               colour = NA) +
+  ggh4x::facet_grid2(chr ~ cell_group, scales = 'free', independent = 'x') +
+  theme_bw() +
+  theme(axis.text.x = element_blank()) +
+  scale_fill_manual(
+    values = c('0x' = '#00008B', '0.5x' = '#b4b0ff', '1' = 'white', 
+               '1.5x' = '#ffc2c2', '2x' = '#d64242', '>2x' = '#8B0000')) 
 ```
@@
 
-First, we plot UMAPs by celltype.
+First, we plot UMAPs by cell grouping.
 
 ```{r plot_umaps}
-dittoSeq::dittoDimPlot(sce, 'cell_type', size = 0.2) +
+dittoSeq::dittoDimPlot(sce, params$cell_group_name, size = 0.2) +
   coord_fixed()
@@
 
-Then, we visualise gains and loss proportions for each chromosome on the UMAP.
+Then, we visualise gains and loss proportions for each chromosome on the UMAP. The `Nebulosa` package is used to show the values as densities.
 
-```{r plot_chr_gains, fig.width = 10, fig.height = 5}
-pdf('test.pdf')
-paste0('chr', 1:22) %>% head(2) %>%
+```{r plot_chr_gains, fig.width = 10, fig.height = 4, results = 'hide'}
+paste0('chr', 1:22) %>% 
   purrr::map(function(chr) {
+    events <-
+      paste('proportion', c('dupli', 'loss'), chr, sep = '_')
     sce %>%
-      Nebulosa::plot_density(
-        paste('proportion', c('dupli', 'loss'), chr, sep = '_'),
-        size = 0.5) 
+      Nebulosa::plot_density(events, size = 0.5) %>%
+      print()
+    reducedDim(sce, 'UMAP_uwot') %>%
+      cbind(sce@colData[, events]) %>%
+      tibble::as_tibble() %>%
+      tidyr::pivot_longer(3:4) %>%
+      dplyr::arrange(value) %>%
+      ggplot(aes(x = UMAP1, y = UMAP2, colour = value)) + 
+      geom_point(size = 0.1) +
+      coord_fixed() +
+      theme_classic() +
+      scale_color_viridis_c() +
+      facet_grid(~ name) %>%
+      print()
   })
@@
 
-```{r plot_gains_and_losses_by_ct}
+```{r plot_gains_and_losses_by_ct, results = 'hide'}
 # at single-cell level
-sce$cell_type[!is.na(sce$infercnv)] %>%
+sce@colData[, params$cell_group_name] %>%
   unique() %>%
-  purrr::map(function(ct) {
+  purrr::map(function(cgn) {
     c('dupli', 'loss') %>%
@@
         p_dat <-
-          sce@colData[!is.na(sce$proportion_cnv_chr1) & sce$cell_type == ct, ]
+          sce@colData[sce@colData[, params$cell_group_name] == cgn, ]
         if (nrow(p_dat) > 0) {
@@
             # hclust error: size cannot be NA nor exceed 65536
-            cluster_rows = ifelse(nrow(p_mat) > 65536, F, T),
+            cluster_rows = ((nrow(p_mat) < 65536) == T),
             legend = F,
-            main = paste0('celltype: ', ct, '\ndirection: ', direction)) %>%
+            main = paste0(params$cell_group_name, ': ', cgn, '\ndirection: ', direction)) %>%
             print()
@@
 
-```{r plot_gains_and_losses_by_id}
+```{r plot_gains_and_losses_by_id, fig.width = 12}
 # mean proportion loss / gain per chromosome per celltype per id
@@
   tibble::as_tibble() %>%
-  dplyr::group_by(sample, cell_type) %>%
-  dplyr::filter(!is.na(proportion_cnv_chr1)) %>%
+  dplyr::mutate(cell_group = get(params$cell_group_name)) %>%
+  dplyr::group_by(sample, cell_group) %>%
   dplyr::summarise(
@@
   dplyr::ungroup() %>%
-  dplyr::mutate(sample_x_ct = paste0(sample, '_', cell_type)) %>%
-  tibble::column_to_rownames('sample_x_ct')
+  dplyr::mutate(sample_x_cg = paste0(sample, '_', cell_group)) %>%
+  tibble::column_to_rownames('sample_x_cg')
 
 # split gain and loss
-p_dat <-
 c('dupli', 'loss') %>% 
@@
     p_annot <- 
-      p_dat[, c('sample', 'cell_type')] %>%
+      p_dat[, c('sample', 'cell_group')] %>%
       as.data.frame()
@@
   tibble::as_tibble() %>%
-  dplyr::filter(!is.na(proportion_dupli_chr7)) %>%
   dplyr::group_by(sample) %>%
-  dplyr::mutate(n_cells = dplyr::n())
+  dplyr::mutate(cell_group = get(params$cell_group_name),
+                n_cells = dplyr::n())
 
@@
       ggplot() +
-      geom_bar(aes(x = cell_type, 
+      geom_bar(aes(x = cell_group, 
                    fill = proportion_dupli_chr7, 
@@
 
-We will define cells with chr7 gain as those affecting >90% of genes. We will pseudobulk by patient, celltype, and chr7 ploidy to produce a reduced object.
+We will define chr7+ cells as those with >80% of genes affected by duplication and chr7 WT cells as those with <20% of genes affected.
+We will pseudobulk by patient, cell group, and chr7 ploidy to produce a reduced object.
 
-```{r pseudobulk_chr7_gain}
+```{r chr7_gain, class.source = 'fold-show'}
+sce$chr7_gain <-
+  ifelse(sce$proportion_dupli_chr7 > 0.8, 1, 
+         ifelse(sce$proportion_dupli_chr7 < 0.2, 0, 
+                NA))
+```
+
+```{r chr7_gain_prop, result = 'asis'}
+dat <-
+  tibble::tibble(chr7_gain_value = sce$chr7_gain,
+               cell_group = sce$cell_group) %>%
+  dplyr::mutate(
+    chr7_gain_status = dplyr::case_when(
+      is.na(chr7_gain_value) ~ 'unknown/intermediate',
+      chr7_gain_value == 0 ~ 'WT',
+      chr7_gain_value == 1 ~ 'gain')
+  )
+
+cat('Counts and proportions of chr7 gain, all cells')
+dat %>%
+  dplyr::mutate(n_cells = dplyr::n()) %>%
+  dplyr::group_by(chr7_gain_status, n_cells) %>%
+  dplyr::summarise(n = dplyr::n(),
+                   `%` = 100 * dplyr::n() / unique(n_cells)) %>%
+  knitr::kable()
+
+cat('Counts and proportions of chr7 gain, by cell group')
+dat %>%
+  dplyr::group_by(cell_group) %>%
+  dplyr::mutate(n_cells = dplyr::n()) %>%
+  dplyr::group_by(cell_group, chr7_gain_status, n_cells) %>%
+  dplyr::summarise(n = dplyr::n(),
+                   `%` = 100 * dplyr::n() / unique(n_cells)) %>%
+  knitr::kable()
+```
+
+```{r pseudobulk_chr7_gain, class.source = 'fold-show'}
 reduced_sce <- xfun::cache_rds({
 
-  # we assign chr7 status as 
-  # >80% of genes affected by a duplication -> chr7+
-  # <20% of genes affected by a duplication -> chr7 WT
-  sce@colData$chr7_gain <-
-    ifelse(sce$proportion_dupli_chr7 > 0.8,
-           1, ifelse(sce$proportion_dupli_chr7 < 0.2, 0, NA))
-
-  # pseudobulk the sce object by patient and chr7 ploidy
-  # subset to stromal cells
+  # pseudobulk the sce object by patient, cell group and chr7 ploidy
   reduced_sce <-
@@
             !is.na(sce$chr7_gain) &
-            !is.na(sce$cell_type)], 
-      group_by = vars(sample, cell_type, chr7_gain)
+            !is.na(sce$cell_group)], 
+      group_by = vars(sample, cell_group, chr7_gain)
     )
@@
 
-We will identify which genes in stromal cells are changed most by chr7 gain. We do this by fitting one GLM model for each gene, with the chr7 status and the patient as factors.
+We will identify which genes in stromal cells are changed most by chr7 gain. We do this by fitting one GLM model for each gene, with the chr7 status, cell group and patient as factors.
 
@@
       reduced_sce,
-    design = ~ chr7_gain + sample + cell_type,
+      design = ~ chr7_gain + sample + cell_group,
       verbose = T
@@
       contrast = 
-        cond(chr7_gain = 1, cell_type = 'Stromal cells') -
-        cond(chr7_gain = 0, cell_type = 'Stromal cells')
+        cond(chr7_gain = 1) -
+        cond(chr7_gain = 0)
     )
@@
   geom_hline(yintercept = -log10(min_p), linetype = 'dashed') +
-  geom_vline(xintercept = log2(1.5), linetype = 'dashed', colour = 'purple') +
+  geom_vline(xintercept = c(log2(1.5), log2(2)), linetype = 'dashed', colour = 'purple') +
   geom_vline(xintercept = c(-1.5, 1.5), linetype = 'dashed') +

### 2024-02-16 infercnv/src/00_preprocess.R

@@ -50,6 +50,7 @@ files %>%
     paste0('data/Zhang2023/mappings.csv')
   )
 
+######################################
 # # Zhang 2019 ---
 # # mp34 previously ran inferCNV on RA samples from the Zhang 2019 paper. 
 # # Here we will run this smaller dataset again as a test.

### 2024-02-16 infercnv/src/00_preprocess_sce.R

@@ -1,0 +1,80 @@
+# Zhang et al. 2023 performed CITE-seq on rheumatoid arthritis samples.
+# The processed scRNA-seq output from this was saved to various files.
+# Here I will construct these outputs into SingleCellExperiment objects.
+
+# libraries
+library(magrittr)
+library(SummarizedExperiment)
+library(SingleCellExperiment)
+
+# dirs
+wd <- 
+  ifelse(Sys.info()['nodename'] == "mib119104s", '~/Volumes/', '') |>
+  paste0('/lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/') 
+setwd(wd)
+data_dir <- '/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/processed_output/'
+out_dir <- 'data/Zhang2023/sce/'
+dir.create(out_dir)
+
+# function: convert Zhang 2023 object to sce
+get_sce <- function(obj, assays) {
+  
+  # get cell order from metadata
+  cell_order <- obj$meta_data$cell
+  
+  # filter and reorder to cells that pass qc from counts
+  assays_pass <-
+    assays %>%
+    purrr::map(~ .x[rownames(assays$logcounts), cell_order])
+  
+  # populate sce
+  sce <- 
+    SingleCellExperiment::SingleCellExperiment(
+      assays = assays_pass,
+      colData = obj$meta_data
+    )
+  
+  # name, reorder, and add umap embeddings
+  reducedDim(sce, 'UMAP_uwot') <- obj$umap$embedding
+  
+  # return
+  sce
+}
+
+# all cells ----
+
+# load object
+obj <- 
+  readRDS(file.path(data_dir, 'all_cells_reference.rds'))
+
+# load rna data
+assays <-
+  list(
+    counts = 
+      readRDS(file.path(data_dir, 'raw_mRNA_count_matrix.rds')),
+    logcounts = 
+      readRDS(file.path(data_dir, 'qc_mRNA_314011cells_log_normalized_matrix.rds')))
+
+# generate sce
+sce <-
+  get_sce(obj, assays)
+
+# save
+saveRDS(sce, 'data/Zhang2023/sce/sce.rds')
+
+# stromal cells ----
+
+# load obj
+obj <-
+  readRDS(file.path(data_dir, 'fibroblast_reference.rds'))
+
[…]

### 2024-02-16 infercnv/src/03_knit_gains_and_losses.sh

@@ -9,10 +9,10 @@ cd $wd
 . ~/.bashrc
 mamba activate jupy
 
-# knit by_celltype
-mkdir -p out/Zhang2023/by_celltype/summary/
-Rscript -e "rmarkdown::render('reports/gains_and_losses.Rmd', params = list(infercnv_dir = 'out/Zhang2023/by_celltype/', cache_dir = 'out/Zhang2023/by_celltype/summary/gains_and_losses_cache/', rerun = T), output_file = 'gains_and_losses.html', output_dir = 'out/Zhang2023/by_celltype/summary/')"
+# # knit by_celltype
+# mkdir -p out/Zhang2023/by_celltype/summary/
+# Rscript -e "rmarkdown::render('reports/gains_and_losses.Rmd', params = list(sce = 'data/Zhang2023/sce/sce.rds', infercnv_dir = 'out/Zhang2023/by_celltype/', cache_dir = 'out/Zhang2023/by_celltype/summary/gains_and_losses_cache/', rerun = T), output_file = 'gains_and_losses.html', output_dir = 'out/Zhang2023/by_celltype/summary/')"
 
-# # knit by_stromal_cluster
-# mkdir -p out/Zhang2023/by_stromal_cluster/summary/
-# Rscript -e "rmarkdown::render('reports/gains_and_losses.Rmd', params = list(infercnv_dir = 'out/Zhang2023/by_stromal_cluster/', cache_dir = 'out/Zhang2023/by_stromal_cluster/summary/gains_and_losses_cache/', rerun = F), output_file = 'gains_and_losses.html', output_dir = 'out/Zhang2023/by_stromal_cluster/summary/')"
+# knit by_stromal_cluster
+mkdir -p out/Zhang2023/by_stromal_cluster/summary/
+Rscript -e "rmarkdown::render('reports/gains_and_losses.Rmd', params = list(sce = 'data/Zhang2023/sce/Stromal_cell_sce.rds', cell_group_name = 'cluster_name', infercnv_dir = 'out/Zhang2023/by_stromal_cluster/', cache_dir = 'out/Zhang2023/by_stromal_cluster/summary/gains_and_losses_cache/', rerun = F), output_file = 'gains_and_losses.html', output_dir = 'out/Zhang2023/by_stromal_cluster/summary/')"

### 2024-02-16 infercnv/reports/gains_and_losses.Rmd
add run that runs all patients together for stromal clusters

@@
   # pseudobulk the sce object by patient, cell group and chr7 ploidy
+  # (exclude intermediate chr7 values)
   reduced_sce <-

### 2024-02-16 infercnv/src/01_run_infercnv.sh

@@ -34,14 +34,28 @@ mamba activate jupy
 #   -w work/ \
 #   -resume
 
-# run infercnv on stromal clusters
-cat $wd/data/Zhang2023/annotations.tsv |
-awk -F'\t' '{if ((NR == 1) || ($2 == "Stromal cell")) {print} }' \
-> $wd/data/Zhang2023/stromal_annotations.tsv
+# # run infercnv on stromal clusters
+# cat $wd/data/Zhang2023/annotations.tsv |
+# awk -F'\t' '{if ((NR == 1) || ($2 == "Stromal cell")) {print} }' \
+# > $wd/data/Zhang2023/stromal_annotations.tsv
+# /software/team205/nextflow-23.04.1-all run nextflow/infercnv.nf \
+#   --out_dir $wd/out/Zhang2023/by_stromal_cluster/ \
+#   --mappings $wd/data/Zhang2023/mappings.csv \
+#   --annotations $wd/data/Zhang2023/stromal_annotations.tsv \
+#   --annotation_col 'cluster' \
+#   -c config/infercnv.config \
+#   -c /nfs/team205/kp9/nextflow/scomatic/LSF.config  \
+#   -w work/ \
+#   -resume
+  
+# run infercnv on stromal clusters, running all individuals together
+cat $wd/data/Zhang2023/stromal_annotations.tsv |
+awk -F'\t' 'BEGIN{OFS="\t";} NR == 1 {print} ; NR > 1 {print $1,$2"_"$4,$2"_"$4,"all"}' \
+> $wd/data/Zhang2023/stromal_annotations_run_all.tsv
 /software/team205/nextflow-23.04.1-all run nextflow/infercnv.nf \
-  --out_dir $wd/out/Zhang2023/by_stromal_cluster/ \
+  --out_dir $wd/out/Zhang2023/by_stromal_cluster_run_all/ \
   --mappings $wd/data/Zhang2023/mappings.csv \
-  --annotations $wd/data/Zhang2023/stromal_annotations.tsv \
+  --annotations $wd/data/Zhang2023/stromal_annotations_run_all.tsv \
   --annotation_col 'cluster' \
   -c config/infercnv.config \
   -c /nfs/team205/kp9/nextflow/scomatic/LSF.config  \

### 2024-02-20 infercnv/reports/gains_and_losses.Rmd
add test for markers to gains_and_losses

@@
 
-```{r load_sce, results = 'asis'}
+```{r load_sce, results = 'asis', class.source = 'fold-show'}
 # load gene coords
@@
 ```
-
-## `map_metadata_from_infercnv.txt`
 
@@
 
+## Cell type compositions
+
+```{r ct_comp, fig.width = 20, fig.height = 7}
+p_dat <-
+  sce@colData %>%
+  tibble::as_tibble() %>%
+  dplyr::mutate(
+    sample = forcats::fct_infreq(sample),
+    cell_group = forcats::fct_infreq(cell_group)) %>%
+  dplyr::add_count(sample, cell_group, name = 'n_sample_x_cell_group')
+p_base <-
+  p_dat %>%
+  ggplot(aes(fill = cell_group, colour = cell_group)) 
+p1 <- 
+  p_base +
+  geom_point(aes(x = sample, y = cell_group, 
+                 alpha = n_sample_x_cell_group, size = n_sample_x_cell_group)) +
+  theme_classic() +
+  theme(legend.position = 'none') +
+  scale_x_discrete(guide = guide_axis(angle = -90))
+p2 <- 
+  p_base +
+  geom_bar(aes(x = sample), position = 'stack') +
+  theme_void() +
+  theme(legend.position = 'none') 
+p3 <-
+  p_base +
+  geom_bar(aes(y = cell_group), position = 'stack') +
+  theme_void() +
+  theme(legend.position = 'none')
+design <-
+'BB#
+AAC'
+patchwork::wrap_plots(A = p1, B = p2, C = p3, design = design)
+```
+
 ## UMAPs
@@
 
-Then, we visualise gains and loss proportions for each chromosome on the UMAP. The `Nebulosa` package is used to show the values as densities.
+Then, we visualise gains and loss proportions for each chromosome on the UMAP.
 
@@
       paste('proportion', c('dupli', 'loss'), chr, sep = '_')
-    sce %>%
-      Nebulosa::plot_density(events, size = 0.5) %>%
-      print()
     reducedDim(sce, 'UMAP_uwot') %>%
@@
 ```{r plot_de}
-# plot
 de_res %>%
@@
 ```
+
+# Cell group markers
+
+We convert the `SingleCellExperiment` object to a `Seurat` object in order to find the top markers of each cell group.
+
+```{r sce_to_seu}
+seu <- xfun::cache_rds({
+  seu <- Seurat::as.Seurat(sce)
+  Seurat::Idents(seu) <- seu$cell_group
+  seu <- Seurat::ScaleData(seu)
+  seu
+}, file = 'seu.rds', rerun = params$rerun)
+```
+
+Now, we find all differentially expressed genes for each cell group.
+
+```{r markers}
+markers <- xfun::cache_rds({
+  markers <-
+    Seurat::FindAllMarkers(seu)
+  markers
+}, file = 'markers.rds', rerun = params$rerun)
+```
+
+```{r inspect_markers, fig.height = 10, fig.width = 10}
+top2 <-
+  markers %>%
+  tibble::as_tibble() %>% 
+  dplyr::filter(p_val_adj < 0.01, avg_log2FC > 1.5) %>% 
+  dplyr::group_by(cluster) %>% 
+  dplyr::slice_max(order_by = avg_log2FC, n = 2)
+unique(top2$cluster) %>%
+  purrr::walk(function(cg) {
+    genes <- 
+      top2 %>% 
+      dplyr::filter(cluster == cg) %>% 
+      dplyr::pull(gene)
+    list(
+      Seurat::VlnPlot(seu, features = genes),
+      Seurat::FeaturePlot(seu, features = genes, label = T) + coord_fixed()) %>%
+      patchwork::wrap_plots(nrow = 2) +
+      patchwork::plot_annotation(
+        title = paste0(params$cell_group_name, ': ', cg)) %>%
+      print()
+  })
+```
+
+```{r marker_heatmap, fig.height = 10, fig.width = 10}
+top10 <-
+  markers %>%
+  dplyr::group_by(cluster) %>%
+  dplyr::filter(avg_log2FC > 1) %>%
+  dplyr::slice_max(order_by = avg_log2FC, n = 100) %>%
+  dplyr::ungroup()
+Seurat::DoHeatmap(seu, features = top10$gene) + 
+  Seurat::NoLegend()
+```

### 2024-02-20 infercnv/src/01_run_infercnv.sh

@@ -49,12 +49,18 @@ mamba activate jupy
 #   -resume
   
 # run infercnv on stromal clusters, running all individuals together
+
 cat $wd/data/Zhang2023/stromal_annotations.tsv |
 awk -F'\t' 'BEGIN{OFS="\t";} NR == 1 {print} ; NR > 1 {print $1,$2"_"$4,$2"_"$4,"all"}' \
 > $wd/data/Zhang2023/stromal_annotations_run_all.tsv
+
+cat $wd/data/Zhang2023/mappings.csv |
+awk -F',' 'BEGIN{OFS=",";} NR == 1 {print} ; NR == 2 {print "all",$2}' \
+> $wd/data/Zhang2023/stromal_mappings_run_all.csv
+
 /software/team205/nextflow-23.04.1-all run nextflow/infercnv.nf \
   --out_dir $wd/out/Zhang2023/by_stromal_cluster_run_all/ \
-  --mappings $wd/data/Zhang2023/mappings.csv \
+  --mappings $wd/data/Zhang2023/stromal_mappings_run_all.csv \
   --annotations $wd/data/Zhang2023/stromal_annotations_run_all.tsv \
   --annotation_col 'cluster' \
   -c config/infercnv.config \

### 2024-02-20 infercnv/src/03_knit_gains_and_losses.sh

@@ -9,10 +9,11 @@ cd $wd
 . ~/.bashrc
 mamba activate jupy
 
-# # knit by_celltype
-# mkdir -p out/Zhang2023/by_celltype/summary/
-# Rscript -e "rmarkdown::render('reports/gains_and_losses.Rmd', params = list(sce = 'data/Zhang2023/sce/sce.rds', infercnv_dir = 'out/Zhang2023/by_celltype/', cache_dir = 'out/Zhang2023/by_celltype/summary/gains_and_losses_cache/', rerun = T), output_file = 'gains_and_losses.html', output_dir = 'out/Zhang2023/by_celltype/summary/')"
+# knit by_celltype
+mkdir -p out/Zhang2023/by_celltype/summary/
+Rscript -e "rmarkdown::render('reports/gains_and_losses.Rmd', params = list(sce = 'data/Zhang2023/sce/sce.rds', cell_group_name = 'cell_type', infercnv_dir = 'out/Zhang2023/by_celltype/', cache_dir = 'out/Zhang2023/by_celltype/summary/gains_and_losses_cache/', rerun = F), output_file = 'gains_and_losses.html', output_dir = 'out/Zhang2023/by_celltype/summary/')"
 
 # knit by_stromal_cluster
 mkdir -p out/Zhang2023/by_stromal_cluster/summary/
 Rscript -e "rmarkdown::render('reports/gains_and_losses.Rmd', params = list(sce = 'data/Zhang2023/sce/Stromal_cell_sce.rds', cell_group_name = 'cluster_name', infercnv_dir = 'out/Zhang2023/by_stromal_cluster/', cache_dir = 'out/Zhang2023/by_stromal_cluster/summary/gains_and_losses_cache/', rerun = F), output_file = 'gains_and_losses.html', output_dir = 'out/Zhang2023/by_stromal_cluster/summary/')"
+

### 2024-02-22 infercnv/reports/gains_and_losses.Rmd
update

@@
   dplyr::group_by(sample) %>%
-  dplyr::mutate(cell_group = get(params$cell_group_name),
-                n_cells = dplyr::n())
+  dplyr::mutate(n_cells = dplyr::n())
 
@@
 
+## Assigning whole chromosome gains
+
+Based on the observed distribution of `proportion_dupli_chr7` values and comparing this with the `inferCNV` heatmaps, whole chromosome gains are defined as `proportion_dupli_chr7 > 0.8`.
+chr7 WT is defined as `proportion_dupli_chr7 < 0.5`.
+
+```{r assign_gains, class.source = 'fold-show'}
+sce <- xfun::cache_rds({
+
+  # apply cut-offs for whole chr gains and losses
+  paste0('chr', 1:22) %>%
+    purrr::walk(function(chr) {
+      chr_dupli <-
+        sce@colData[, paste0("proportion_dupli_", chr)]
+      sce@colData[, paste0(chr, '_gain')] <<-
+        ifelse(chr_dupli > 0.8, 1, ifelse(chr_dupli < 0.5, 0, NA))
+      chr_loss <-
+        sce@colData[, paste0("proportion_loss_", chr)]
+      sce@colData[, paste0(chr, '_loss')] <<-
+        ifelse(chr_loss > 0.8, 1, ifelse(chr_loss < 0.5, 0, NA))
+    })
+
+  # return
+  sce
+
+}, file = 'sce_w_gains.rds', rerun = params$rerun)
+```
+
+We also can observe the rate coincidence of chr7 and other gains. The LCM-WES data suggested a nested relationship in which there were chr5+ subclones within chr7+ populations. We look for this here.
+
+```{r chr7_and_chr5}
+events <-
+  sce@colData %>%
+  tibble::as_tibble() %>%
+  dplyr::select(sample, cell, cell_group, 
+                dplyr::ends_with('_gain'), 
+                dplyr::ends_with('_loss')) %>%
+  tidyr::pivot_longer(-c('sample', 'cell', 'cell_group')) %>%
+  dplyr::group_by(sample, cell, cell_group) %>%
+  # remove cells in which no calls made
+  dplyr::filter(!all(is.na(value))) %>%
+  dplyr::mutate(
+    event = dplyr::case_when(
+      all(value == 0 | is.na(value)) ~ 'WT',
+      value == 1 ~ name,
+      TRUE ~ NA)) %>%
+  dplyr::summarise(
+    events = paste(sort(unique(na.omit(event))), collapse = ','),
+    n_gains = sum(value, na.rm = T))
+
+p_dat <-
+  events %>%
+  dplyr::ungroup() %>%
+  dplyr::add_count(events) %>%
+  dplyr::mutate(prop = n / dplyr::n()) %>%
+  dplyr::filter(events != 'WT') %>%
+  dplyr::mutate(
+    events = events %>% 
+      gsub('chr', '', .) %>% 
+      gsub('_gain', '+', .) %>%
+      gsub('_loss', '+', .),
+    events = forcats::fct_infreq(events),
+    label1 = scales::label_comma()(n), 
+    label = dplyr::case_when(
+      n < 10 ~ label1,
+      TRUE ~ paste0(label1, ' (', round(prop*100, 2), '%)')
+    ))
+p_dat %>%
+  ggplot(aes(x = events, fill = cell_group, label = label)) +
+  geom_bar() +
+  geom_text(aes(y = n), hjust = 0, vjust = -0.1, angle = 45, size = 2) +
+  scale_x_discrete(guide = guide_axis(angle = -90)) +
+  ylim(c(0, max(p_dat$n) * 1.1)) +
+  theme_classic()
+p_dat %>%
+  ggplot(aes(x = forcats::fct_infreq(gains), fill = sample, label = label)) +
+  geom_bar() +
+  geom_text(aes(y = n), hjust = 0, vjust = -0.1, angle = 45, size = 2) +
+  scale_x_discrete(guide = guide_axis(angle = -90)) +
+  ylim(c(0, max(p_dat$n) * 1.1)) +
+  theme_classic()
+```
+
 # Pseudobulking and DGEA
@@
 
-```{r chr7_gain, class.source = 'fold-show'}
-sce$chr7_gain <-
-  ifelse(sce$proportion_dupli_chr7 > 0.8, 1, 
-         ifelse(sce$proportion_dupli_chr7 < 0.2, 0, 
-                NA))
-```
-
-```{r chr7_gain_prop, result = 'asis'}
+```{r chr7_gain_prop, results = 'asis'}
 dat <-
@@
 
+  # return
   de_res
+
 }, file = 'de_res.rds', rerun = params$rerun)
@@
 
+```{r exit}
+knitr::knit_exit()
+```
+
 ```{r sce_to_seu}
+
 seu <- xfun::cache_rds({

### 2024-02-22 infercnv/src/00_preprocess_sce.R

@@ -78,3 +78,12 @@ sce <-
 
 # save
 saveRDS(sce, 'data/Zhang2023/sce/Stromal_cell_sce.rds')
+
+# fibroblasts ----
+
+# subset to fibroblasts only
+sce <-
+  sce[, sce$cluster_name != 'Mu-0: Mural']
+
+# save
+saveRDS(sce, 'data/Zhang2023/sce/Fibroblast_sce.rds')

### 2024-02-22 infercnv/src/01_run_infercnv.sh

@@ -63,6 +63,8 @@ awk -F',' 'BEGIN{OFS=",";} NR == 1 {print} ; NR == 2 {print "all",$2}' \
   --mappings $wd/data/Zhang2023/stromal_mappings_run_all.csv \
   --annotations $wd/data/Zhang2023/stromal_annotations_run_all.tsv \
   --annotation_col 'cluster' \
+  --analysis_mode 'samples' \
+  --cluster_by_groups 'FALSE' \
   -c config/infercnv.config \
   -c /nfs/team205/kp9/nextflow/scomatic/LSF.config  \
   -w work/ \

### 2024-02-22 infercnv/src/03_knit_gains_and_losses.sh

@@ -1,5 +1,5 @@
 #!/bin/bash
-# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/ ; . ~/.bashrc ; mamba activate jupy ; /software/team205/bin/jsub lsf -q week -n knit_gains_and_losses -c 1 -m 70g -l log "bash src/03_knit_gains_and_losses.sh" | bsub
+# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/ ; . ~/.bashrc ; mamba activate jupy ; /software/team205/bin/jsub lsf -q week -n knit_gains_and_losses -c 1 -m 100g -l log "bash src/03_knit_gains_and_losses.sh" | bsub
 
 # wd
 wd=/lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/
@@ -9,11 +9,15 @@ cd $wd
 . ~/.bashrc
 mamba activate jupy
 
-# knit by_celltype
-mkdir -p out/Zhang2023/by_celltype/summary/
-Rscript -e "rmarkdown::render('reports/gains_and_losses.Rmd', params = list(sce = 'data/Zhang2023/sce/sce.rds', cell_group_name = 'cell_type', infercnv_dir = 'out/Zhang2023/by_celltype/', cache_dir = 'out/Zhang2023/by_celltype/summary/gains_and_losses_cache/', rerun = F), output_file = 'gains_and_losses.html', output_dir = 'out/Zhang2023/by_celltype/summary/')"
+# knit by_fibroblast_cluster
+mkdir -p out/Zhang2023/by_fibroblast_cluster/summary/
+Rscript -e "rmarkdown::render('reports/gains_and_losses.Rmd', params = list(sce = 'data/Zhang2023/sce/Fibroblast_sce.rds', cell_group_name = 'cluster_name', infercnv_dir = 'out/Zhang2023/by_stromal_cluster/', cache_dir = 'out/Zhang2023/by_fibroblast_cluster/summary/gains_and_losses_cache/', rerun = F), output_file = 'gains_and_losses.html', output_dir = 'out/Zhang2023/by_fibroblast_cluster/summary/')"
 
 # knit by_stromal_cluster
 mkdir -p out/Zhang2023/by_stromal_cluster/summary/
[…]

### 2024-02-23 scomatic/reports/driver_coverage.Rmd
parameterise scomatic nextflow script, add header keys for reading mappings.csv, introduce STEP1 entry

@@
 
+# function: plot stacked exon coverage
+plot_exons_stacked <- function(p_df, p_title) {
+  p_df %>% 
+    # get only exonic sequence
+    dplyr::filter(feature == 'exon') %>%
+    # fix exon ordering
+    dplyr::mutate(
+      exon = exon %>% gsub('exon', '', .) %>% as.numeric(),
+      facets = ifelse(is.na(celltype), 'all', 'by celltype')) %>%
+    ggplot(aes(x = pos, y = depth, group = celltype, fill = celltype)) +
+    geom_col(position = 'stack') +
+    theme_classic() +
+    theme(axis.text.x = element_blank(),
+          axis.ticks.x = element_blank(),
+          panel.spacing = unit(.05, 'lines'),
+          panel.border = element_rect(fill = NA)) +
+    labs(title = paste(p_title, '(exons)')) +
+    facet_grid(facets ~ exon, space = 'free_x', scales = 'free_x') 
+}
+
 # create list of plots by gene x sample
@@
                 )
+
+              # plot exons stacked
+              p[[id_i]][[gene_i]][[paste0('exons_stacked', p_suffix)]] <<-
+                plot_exons_stacked(
+                  p_df = lvl,
+                  p_title = paste0(i, p_suffix)
+                )
             })

### 2024-02-23 scomatic/src/01_run_scomatic.sh

@@ -7,11 +7,29 @@ set -eo pipefail
 wd=/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/
 cd $wd
 config_dir=/nfs/team205/kp9/nextflow/scomatic/
-mkdir -p work/ out/Zhang2023/default_thresholds/ log/
+mkdir -p work/ log/
 
 # conda activate an environment which has docopt installed so jsub works
 . ~/.bashrc
 mamba activate jupy
+
+# use scomatic to get celltype bams for driver coverage
+/software/team205/nextflow-23.04.1-all run nextflow/scomatic.nf \
+    -entry STEP1 \
+    --max_nM = 1000000 \
+    --max_NH = 1000000 \
+    --min_MQ = 0 \
+    --n_trim = 0 \
+    --min_dp = 0 \
+    --min_cc = 0 \
+    --projectDir $PWD \
+    -params-file $config_dir/GEX.json \
+    -c config/Zhang2023.config \
+    -c $config_dir/LSF.config \
+    -w work/ \
+    --output_dir out/Zhang2023/no_thresholds/ \
+    --location local \
+    -resume
 
 # run scomatic
 /software/team205/nextflow-23.04.1-all run nextflow/scomatic.nf \

### 2024-02-23 scomatic/src/03_knit_driver_coverage.sh

@@ -42,7 +42,6 @@ mamba activate jupy
 /software/team205/nextflow-23.04.1-all run nextflow/driver_coverage.nf \
     --mappings data/Zhang2023/coverage_mappings.csv \
     --drivers /lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/data/driver_genes/driver_gene_coords_for_coverage.tsv \
-    --ref_cds /lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/data/dndscv/ref_cds.rds \
     --out_dir /lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/out/Zhang2023/coverage/ \
     -c /nfs/team205/kp9/nextflow/scomatic/LSF.config  \
     --location local \

### 2024-02-23 infercnv/reports/gains_and_losses.Rmd
check for truncated bams in mappings before submitting to nextflow

@@
 p_dat %>%
-  ggplot(aes(x = forcats::fct_infreq(gains), fill = sample, label = label)) +
+  ggplot(aes(x = forcats::fct_infreq(events), fill = sample, label = label)) +
   geom_bar() +

### 2024-02-23 infercnv/src/01_run_infercnv.sh

@@ -51,7 +51,7 @@ mamba activate jupy
 # run infercnv on stromal clusters, running all individuals together
 
 cat $wd/data/Zhang2023/stromal_annotations.tsv |
-awk -F'\t' 'BEGIN{OFS="\t";} NR == 1 {print} ; NR > 1 {print $1,$2"_"$4,$2"_"$4,"all"}' \
+awk -F'\t' 'BEGIN{OFS="\t";} NR == 1 {print} ; NR > 1 {print $1,$2,$3,"all"}' \
 > $wd/data/Zhang2023/stromal_annotations_run_all.tsv
 
 cat $wd/data/Zhang2023/mappings.csv |

### 2024-02-23 infercnv/src/03_knit_gains_and_losses.sh

@@ -11,13 +11,13 @@ mamba activate jupy
 
 # knit by_fibroblast_cluster
 mkdir -p out/Zhang2023/by_fibroblast_cluster/summary/
-Rscript -e "rmarkdown::render('reports/gains_and_losses.Rmd', params = list(sce = 'data/Zhang2023/sce/Fibroblast_sce.rds', cell_group_name = 'cluster_name', infercnv_dir = 'out/Zhang2023/by_stromal_cluster/', cache_dir = 'out/Zhang2023/by_fibroblast_cluster/summary/gains_and_losses_cache/', rerun = F), output_file = 'gains_and_losses.html', output_dir = 'out/Zhang2023/by_fibroblast_cluster/summary/')"
+Rscript -e "rmarkdown::render('reports/gains_and_losses.Rmd', params = list(sce = 'data/Zhang2023/sce/Fibroblast_sce.rds', cell_group_name = 'cluster_name', infercnv_dir = 'out/Zhang2023/by_stromal_cluster/', cache_dir = 'out/Zhang2023/by_fibroblast_cluster/summary/gains_and_losses_cache/', rerun = T), output_file = 'gains_and_losses.html', output_dir = 'out/Zhang2023/by_fibroblast_cluster/summary/')"
 
 # knit by_stromal_cluster
 mkdir -p out/Zhang2023/by_stromal_cluster/summary/
-Rscript -e "rmarkdown::render('reports/gains_and_losses.Rmd', params = list(sce = 'data/Zhang2023/sce/Stromal_cell_sce.rds', cell_group_name = 'cluster_name', infercnv_dir = 'out/Zhang2023/by_stromal_cluster/', cache_dir = 'out/Zhang2023/by_stromal_cluster/summary/gains_and_losses_cache/', rerun = F), output_file = 'gains_and_losses.html', output_dir = 'out/Zhang2023/by_stromal_cluster/summary/')"
+Rscript -e "rmarkdown::render('reports/gains_and_losses.Rmd', params = list(sce = 'data/Zhang2023/sce/Stromal_cell_sce.rds', cell_group_name = 'cluster_name', infercnv_dir = 'out/Zhang2023/by_stromal_cluster/', cache_dir = 'out/Zhang2023/by_stromal_cluster/summary/gains_and_losses_cache/', rerun = T), output_file = 'gains_and_losses.html', output_dir = 'out/Zhang2023/by_stromal_cluster/summary/')"
 
[…]

### 2024-02-23 scomatic/src/00_check_bams.sh

@@ -8,8 +8,18 @@ cd $wd
 . ~/.bashrc
 mamba activate jupy
 
-for bam in /lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/cellranger_output/*/possorted_genome_bam.bam ; do
-  echo $bam
-  java -jar ~/bin/picard.jar ValidateSamFile I=$bam MODE=SUMMARY 2>&1 
-  echo
-done > out/Zhang2023/check_bams.txt
+# get truncated bams
+(
+  echo -e 'bam_file\tmessage' ;
+  for i in /lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/cellranger_output/BRI-*/possorted_genome_bam.bam ; do 
+    echo -n $i
+    samtools view -h $i | head -1 
+  done 2>&1 | grep '\[' | sed 's/\[/\t\[/g' ;
+) | cat > data/Zhang2023/truncated_bams.tsv 
+
+
+# for bam in /lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/cellranger_output/*/possorted_genome_bam.bam ; do
+#   echo $bam
+#   java -jar ~/bin/picard.jar ValidateSamFile I=$bam MODE=SUMMARY 2>&1 
+#   echo
+# done > out/Zhang2023/check_bams.txt

### 2024-02-23 scomatic/src/00_preprocess_BAMs.R

@@ -14,16 +14,29 @@ dir.create('data/')
 # Zhang 2023 ----
 # mappings CSV (sample_id,bam_file,id) -----
 dir.create('data/Zhang2023/')
-list.files(
+mappings <-
+  list.files(
       cellranger_dir,
       pattern = 'possorted_genome_bam.bam$',
       recursive = T, full.names = T) %>%
   tibble::tibble(bam_file = .) %>%
   # get parent directory (= id)
   dplyr::mutate(
+    bam_file = normalizePath(bam_file),
     sample_id = bam_file %>% dirname() %>% basename() %>% gsub('-', '_', .),
     id = sample_id
-  ) %>%
+  ) 
+
+# check for truncation
+truncated_bams <- 
+  readr::read_tsv('data/Zhang2023/truncated_bams.tsv')
+
+# filter out
+mappings <-
+  mappings %>%
+  dplyr::filter(!bam_file %in% truncated_bams$bam_file)
+
+mappings %>%
     readr::write_csv('data/Zhang2023/mappings.csv')
 
 # celltypes TSV (Index Cell_type) -----

### 2024-02-23 scomatic/src/01_run_scomatic.sh

@@ -14,22 +14,23 @@ mkdir -p work/ log/
 mamba activate jupy
 
 # use scomatic to get celltype bams for driver coverage
+# set all thresholds such that no cells are removedand we get full unfiltered
+# celltype-specific bams
 /software/team205/nextflow-23.04.1-all run nextflow/scomatic.nf \
     -entry STEP1 \
-    --max_nM = 1000000 \
-    --max_NH = 1000000 \
-    --min_MQ = 0 \
-    --n_trim = 0 \
-    --min_dp = 0 \
-    --min_cc = 0 \
+    --max_nM 1000000 \
+    --max_NH 1000000 \
+    --min_MQ 0 \
+    --n_trim 0 \
+    --min_dp 0 \
+    --min_cc 0 \
     --projectDir $PWD \
     -params-file $config_dir/GEX.json \
     -c config/Zhang2023.config \
     -c $config_dir/LSF.config \
     -w work/ \
     --output_dir out/Zhang2023/no_thresholds/ \
-    --location local \
-    -resume
+    --location local
 
 # run scomatic
 /software/team205/nextflow-23.04.1-all run nextflow/scomatic.nf \

### 2024-02-28 infercnv/reports/gains_and_losses.Rmd
introducing genvisr to driver coverage plotting

--- 2024-02-20 infercnv/reports/gains_and_losses.Rmd
@@
   rerun: true
+  max_prop_for_wt: 0.5
+  min_prop_for_cnv: 0.8
 ---
@@
                colour = NA) +
-  ggh4x::facet_grid2(chr ~ cell_group, scales = 'free', independent = 'x') +
+  ggh4x::facet_grid2(chr ~ cell_group, scales = 'free_x', independent = 'x') +
   theme_bw() +
@@
 
+## Assigning whole chromosome-level events
+
+Based on the observed distribution of `proportion_dupli_chr7` values and comparing this with the `inferCNV` heatmaps, whole chromosome gains are defined as `proportion_dupli_chr*` > `r params$min_prop_for_cnv`.
+Diploidy is defined as `proportion_dupli_chr*` < `r params$max_prop_for_wt`.
+
+```{r assign_gains, class.source = 'fold-show'}
+sce <- xfun::cache_rds({
+
+  # apply cut-offs for whole chr gains and losses
+  paste0('chr', 1:22) %>%
+    purrr::walk(function(chr) {
+      chr_dupli <-
+        sce@colData[, paste0("proportion_dupli_", chr)]
+      sce@colData[, paste0(chr, '_gain')] <<-
+        ifelse(chr_dupli > params$min_prop_for_cnv, 1, 
+               ifelse(chr_dupli < params$max_prop_for_wt, 0, 
+                      NA))
+      chr_loss <-
+        sce@colData[, paste0("proportion_loss_", chr)]
+      sce@colData[, paste0(chr, '_loss')] <<-
+        ifelse(chr_loss > params$min_prop_for_cnv, 1, 
+               ifelse(chr_loss < params$max_prop_for_wt, 0, 
+                      NA))
+    })
+
+  # return
+  sce
+
+}, file = 'sce_w_gains.rds', rerun = params$rerun)
+```
+
+We also can observe the rate coincidence of chr7 and other gains. The LCM-WES data suggested a nested relationship in which there were chr5+ subclones within chr7+ populations. We look for this here.
+
+```{r chr7_and_chr5}
+events <-
+  sce@colData %>%
+  tibble::as_tibble() %>%
+  dplyr::select(sample, cell, cell_group, 
+                dplyr::ends_with('_gain'), 
+                dplyr::ends_with('_loss')) %>%
+  tidyr::pivot_longer(-c('sample', 'cell', 'cell_group')) %>%
+  dplyr::group_by(sample, cell, cell_group) %>%
+  # remove cells in which no calls made
+  dplyr::filter(!all(is.na(value))) %>%
+  dplyr::mutate(
+    event = dplyr::case_when(
+      all(value == 0 | is.na(value)) ~ 'WT',
+      value == 1 ~ name,
+      TRUE ~ NA)) %>%
+  dplyr::summarise(
+    events = paste(sort(unique(na.omit(event))), collapse = ','),
+    n_events = sum(value, na.rm = T))
+
+p_dat <-
+  events %>%
+  dplyr::ungroup() %>%
+  dplyr::add_count(events) %>%
+  dplyr::mutate(prop = n / dplyr::n()) %>%
+  # ignore events that happen in <10 cells
+  dplyr::filter(events != 'WT', n >= 10) %>%
+  dplyr::mutate(
+    events = events %>% 
+      gsub('chr', '', .) %>% 
+      gsub('_gain', '+', .) %>%
+      gsub('_loss', '+', .),
+    events = forcats::fct_infreq(events),
+    label1 = scales::label_comma()(n), 
+    label = dplyr::case_when(
+      n < 10 ~ label1,
+      TRUE ~ paste0(label1, ' (', round(prop*100, 2), '%)')
+    ))
+p_dat %>%
+  ggplot(aes(x = events, fill = cell_group, label = label)) +
+  geom_bar() +
+  geom_text(aes(y = n), hjust = 0, vjust = -0.1, angle = 45, size = 2) +
+  scale_x_discrete(guide = guide_axis(angle = -90)) +
+  ylim(c(0, max(p_dat$n) * 1.1)) +
+  theme_classic() +
+  theme(legend.position = 'bottom')
+p_dat %>%
+  ggplot(aes(x = forcats::fct_infreq(events), fill = sample, label = label)) +
+  geom_bar() +
+  geom_text(aes(y = n), hjust = 0, vjust = -0.1, angle = 45, size = 2) +
+  scale_x_discrete(guide = guide_axis(angle = -90)) +
+  ylim(c(0, max(p_dat$n) * 1.1)) +
+  theme_classic() +
+  theme(legend.position = 'none')
+```
+
 # Pseudobulking and DGEA
@@
 
-We will define chr7+ cells as those with >80% of genes affected by duplication and chr7 WT cells as those with <20% of genes affected.
 We will pseudobulk by patient, cell group, and chr7 ploidy to produce a reduced object.
 
-```{r chr7_gain, class.source = 'fold-show'}
-sce$chr7_gain <-
-  ifelse(sce$proportion_dupli_chr7 > 0.8, 1, 
-         ifelse(sce$proportion_dupli_chr7 < 0.2, 0, 
-                NA))
-```
-
-```{r chr7_gain_prop, result = 'asis'}
+```{r chr7_gain_prop, results = 'asis'}
 dat <-

### 2024-02-28 scomatic/reports/driver_coverage.Rmd
introducing genvisr to driver coverage plotting

@@
 params:
-  id: NULL
   drivers: NULL
@@
 library(gridExtra)
+library(GenomicRanges)
+library(BSgenome)
 
@@
 
-We read in drivers and their depths.
+We read in drivers and their coverage by sample / celltype.
 
 ```{r read_cov}
+# read drivers
+drivers <- readr::read_tsv(params$drivers)
+
 # read coverage
 cov <- xfun::cache_rds({
+
   cov <-
-    list.files(pattern = '_depths.tsv') %>% 
-    purrr::set_names(., gsub('_depths.tsv', '', .)) %>%
+    list.files(pattern = '_coverage.tsv') %>% 
+    purrr::set_names(., gsub('_coverage.tsv', '', .)) %>%
     purrr::map(function(file) {
-      file %>%
-        readr::read_tsv() %>%
-        dplyr::mutate(chr = as.character(chr))
+
+      # read  
+      df <- 
+        readr::read_tsv(file, show_col_types = F) %>%
+        dplyr::rename(all = `NA`)
+
+      # reformat for GenVisR
+      df %>%
+        dplyr::select(-c(chr, pos, gene)) %>%
+        colnames() %>%
+        purrr::set_names(., .) %>%
+        purrr::map(function(ct) {
+          df %>%
+            dplyr::transmute(
+              gene,
+              chromosome = paste0('chr', chr), 
+              end = pos, 
+              cov = get(ct))
+
           }) %>%
-    dplyr::bind_rows(.id = 'id')
+        dplyr::bind_rows(.id = 'celltype')
+
+    }) %>%
+    dplyr::bind_rows(.id = 'id') %>%
+    split(., .$gene)
+
+  # return
   cov
+
 }, file = 'cov.rds', rerun = params$rerun)
+```
 
-# read drivers
-drivers <- readr::read_tsv(params$drivers)
+We also load the `BSgenome` and `TxDb` objects.
+
+```{r bsgenome}
+# BSgenome
+library("BSgenome.Hsapiens.UCSC.hg38")
+genome <- BSgenome.Hsapiens.UCSC.hg38
+
+# TxDb
+library("TxDb.Hsapiens.UCSC.hg38.knownGene")
+txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene
+```
+
+Then we plot the coverage.
+
+```{r gencov}
+plot_cov <- function(cov, gene, var, expand_features_of_interest = T) {
+
+  # split by variable(s)
+  x <-
+    cov[[gene]] %>%
+    dplyr::select(chromosome, end, cov) %>%
+    split(., cov[[gene]][, var])
+
+  # make granges obj
+  gr <-
+    GRanges(
+      seqnames = as.character(unique(cov[[gene]]$chromosome)),
+      ranges = IRanges(start = min(cov[[gene]]$end) - 500,
+                       end = max(cov[[gene]]$end) + 500))
+
+  # check what to transform
+  transform <- c('Intron', 'CDS', 'UTR')
+  base <- c(10, 2, 2)
+  if (expand_features_of_interest == T) { base <- transform <- NA }
+
+  # plot
+  GenVisR::genCov(
+    x = x,
+    txdb = txdb, 
+    gr = gr, 
+    genome = genome,
+    cov_plotType = 'line',
+    cov_plotLayer = ggplot2::theme(axis.title.y = ggplot2::element_blank()),
+    label_txtSize = 4,
+    lab2plot_ratio = c(2, length(x)),
+    gene_labelTranscriptSize = 2,
+    gene_name = gene,
+    base = base,
+    transform = transform)
+
+}
+
+plot_cov(cov, 'ARID2', 'celltype') 
+plot_cov(cov, 'ARID2', 'celltype', F)
+plot_cov(cov, 'ARID2', 'id')
+plot_cov(cov, 'ARID2', 'id', F)
+```
+
+```{r exit}
+knitr::knit_exit()
 ```

### 2024-02-28 infercnv/src/03_knit_gains_and_losses.sh

@@ -9,15 +9,19 @@ cd $wd
 . ~/.bashrc
 mamba activate jupy
 
-# knit by_fibroblast_cluster
-mkdir -p out/Zhang2023/by_fibroblast_cluster/summary/
-Rscript -e "rmarkdown::render('reports/gains_and_losses.Rmd', params = list(sce = 'data/Zhang2023/sce/Fibroblast_sce.rds', cell_group_name = 'cluster_name', infercnv_dir = 'out/Zhang2023/by_stromal_cluster/', cache_dir = 'out/Zhang2023/by_fibroblast_cluster/summary/gains_and_losses_cache/', rerun = T), output_file = 'gains_and_losses.html', output_dir = 'out/Zhang2023/by_fibroblast_cluster/summary/')"
+# # knit by_fibroblast_cluster, < 0.5 | > 0.8
+# mkdir -p out/Zhang2023/by_fibroblast_cluster/summary/
+# Rscript -e "rmarkdown::render('reports/gains_and_losses.Rmd', params = list(sce = 'data/Zhang2023/sce/Fibroblast_sce.rds', cell_group_name = 'cluster_name', infercnv_dir = 'out/Zhang2023/by_stromal_cluster/', cache_dir = 'out/Zhang2023/by_fibroblast_cluster/summary/gains_and_losses_cache/', rerun = F), output_file = 'gains_and_losses.html', output_dir = 'out/Zhang2023/by_fibroblast_cluster/summary/')"
 
-# knit by_stromal_cluster
-mkdir -p out/Zhang2023/by_stromal_cluster/summary/
-Rscript -e "rmarkdown::render('reports/gains_and_losses.Rmd', params = list(sce = 'data/Zhang2023/sce/Stromal_cell_sce.rds', cell_group_name = 'cluster_name', infercnv_dir = 'out/Zhang2023/by_stromal_cluster/', cache_dir = 'out/Zhang2023/by_stromal_cluster/summary/gains_and_losses_cache/', rerun = T), output_file = 'gains_and_losses.html', output_dir = 'out/Zhang2023/by_stromal_cluster/summary/')"
+# knit by_fibroblast_cluster, < 0.2 | > 0.8
+mkdir -p out/Zhang2023/by_fibroblast_cluster_0.2/summary/
[…]

### 2024-02-28 scomatic/src/01_run_scomatic.sh

@@ -14,10 +14,12 @@ mkdir -p work/ log/
 mamba activate jupy
 
 # use scomatic to get celltype bams for driver coverage
-# set all thresholds such that no cells are removedand we get full unfiltered
+# set all thresholds such that no cells are removed and we get full unfiltered
 # celltype-specific bams
 /software/team205/nextflow-23.04.1-all run nextflow/scomatic.nf \
     -entry STEP1 \
+    --publish_celltype_bams true \
+    --output_dir out/Zhang2023/coverage/ \
     --max_nM 1000000 \
     --max_NH 1000000 \
     --min_MQ 0 \
@@ -29,7 +31,6 @@ mamba activate jupy
     -c config/Zhang2023.config \
     -c $config_dir/LSF.config \
     -w work/ \
-    --output_dir out/Zhang2023/no_thresholds/ \
     --location local
 
 # # run scomatic

### 2024-02-28 scomatic/src/03_knit_driver_coverage.sh

@@ -9,41 +9,38 @@ cd $wd
 . ~/.bashrc
 mamba activate jupy
 
-# # knit
-# Rscript -e "rmarkdown::render('reports/driver_coverage.Rmd', output_file = 'driver_coverage.html', output_dir = 'reports/', params = list(rerun = F))"
-
-# # create mappings
-# bams_dir=/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/cellranger_output/
-# ct_bams_dir=/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/out/Zhang2023/no_thresholds/
-# out_dir=/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/out/Zhang2023/coverage/
-# (
-#   echo 'id,bam,celltype' ;
-#   while read -r id ; do
-#     # check if ct bams directory exists
-#     if [ -d $ct_bams_dir/${id/-/_}/ ] ; then
-#       ct_bams=($(ls $ct_bams_dir/${id/-/_}/celltype_bams/*.bam))
-#       # check if scomatic successfully ran
-#       if [ ${#ct_bams[*]} -gt 0 ] ; then
-#         echo "$id,$bams_dir/$id/possorted_genome_bam.bam,NA" ;
-#         for file in ${ct_bams[@]} ; do
-#           ct=$(basename $file | cut -d. -f2)
-#           echo "$id,$file,$ct" ;
-#         done
-#       fi
-#     fi
-#   done < \
-#   <(sed 1d /lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/data/Zhang2023/mappings.csv | cut -d, -f1 | sort -u)
-# ) | cat > data/Zhang2023/coverage_mappings.csv
+# create mappings
+mappings=data/Zhang2023/mappings.csv
+bams_dir=/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/cellranger_output/
+ct_bams_dir=/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/out/Zhang2023/coverage/
+(
+  echo 'id,bam,celltype' ;
+  while read -r id ; do
+    # check if ct bams directory exists
+    if [ -d $ct_bams_dir/$id/ ] ; then
+      ct_bams=($(ls $ct_bams_dir/$id/celltype_bams/*.bam))
+      # check if scomatic successfully ran
[…]

### 2024-02-28 scomatic/reports/targeted_mutation_calling.Rmd
beginning scripting for manual interrogation of driver mutations

---
title: "Targeted interrogation of mutational hotspots in rheumatoid arthritis"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: false
    code_folding: show
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
params:
  hotspots: NULL
  ref_cds: NULL
  cache_dir: './driver_coverage_cache/'
  rerun: true
---

```{r setup, include = F, message = F, warning = F, class.source = 'fold-hide'}
# wd <- ifelse(Sys.info()['nodename'] == "mib119104s", '~/Volumes/', '') |> paste0('/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/') ; setwd(wd)
# rmarkdown::render('reports/driver_coverage.Rmd', output_file = 'driver_coverage.html', output_dir = 'reports/', params = list(rerun = F))

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  dpi = 300, 
  cache.path = params$cache_dir)

# libraries
library(magrittr)
library(ggplot2)
library(GenomicRanges)

# dev.off() if there is a device
dev_off_if <- function() {if (length(dev.list()) != 0) { dev.off() }}
```

# Mutational calls in drivers from the LCM-WES data

```{r mut_hotspots}
#params<-list(drivers = 'data/driver_genes/driver_gene_coords_for_coverage.bed',hotspots='/lustre/scratch125/casm/team268im/lh22/synovium_nanoseq/ex_t_dndsout/exome/All_dnds_annot_muts.tsv')
drivers <-
  readr::read_tsv(params$drivers, col_names = c('chr', 'start', 'end', 'gene')) %>%
  dplyr::mutate(chr = gsub('chr', '', chr))
hotspots <-
  readr::read_tsv(params$hotspots)
gr <-
  GRanges(seqnames = drivers$chr,
          ranges = IRanges(start = drivers$start,
                           end = drivers$end))
```

# BAMs from snRNA-seq data

We read in reads in the driver regions

```{r scanbam}
bam_path <-
  '/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/out/Zhang2023/coverage/BRI_401/celltype_bams/BRI_401.B_cell_plasma_cell.bam'
bam_file <- Rsamtools::BamFile(bam_path)
bam_params <- 
  Rsamtools::ScanBamParam(which = gr, what = Rsamtools::scanBamWhat())
bam <- Rsamtools::scanBam(bam_file, param = bam_params)
```

### 2024-03-05 scomatic/reports/driver_coverage.Rmd
tinkering with subset_bams in scomatic.nf

@@
   drivers: NULL
+  mutations: NULL
   ref_cds: NULL
@@
 }, file = 'cov.rds', rerun = params$rerun)
+
+# read in driver mutations from LCM-WES
+muts <-
+  readr::read_tsv(params$mutations)
 ```
 
-We also load the `BSgenome` and `TxDb` objects.
+# GenVisR plots
+
+We load the `BSgenome` and `TxDb` objects.
 
@@
 ```{r gencov}
+# function: use gencov
 plot_cov <- function(cov, gene, var, expand_features_of_interest = T) {
@@
 
-plot_cov(cov, 'ARID2', 'celltype') 
-plot_cov(cov, 'ARID2', 'celltype', F)
-plot_cov(cov, 'ARID2', 'id')
-plot_cov(cov, 'ARID2', 'id', F)
+# plot
+names(cov) %>%
+  purrr::walk(~ plot_cov(cov, .x, 'celltype') %>% print())
+names(cov) %>%
+  purrr::walk(~ plot_cov(cov, .x, 'celltype', F) %>% print())
+names(cov) %>%
+  purrr::walk(~ plot_cov(cov, .x, 'id') %>% print())
+names(cov) %>%
+  purrr::walk(~ plot_cov(cov, .x, 'id', F) %>% print())  
 ```
 
-```{r exit}
-knitr::knit_exit()
-```
+# My plots
 
-### Annotate exons and introns
+## Annotate exons and introns
 
@@
 ```
+
+## Plot driver coverages
 
@@
 
+## Mean_coverage
+
+```{r exit}
+knitr::knit_exit()
+```
+
 ```{r mean_coverage, eval = F}

### 2024-03-05 scomatic/reports/targeted_mutation_calling.Rmd
tinkering with subset_bams in scomatic.nf

@@
 params:
+  drivers: NULL
   hotspots: NULL
-  ref_cds: NULL
-  cache_dir: './driver_coverage_cache/'
+  bam_mappings: NULL
+  cache_dir: './target_mutation_calling_cache/'
   rerun: true
@@
 
+# chunk options
+knitr::opts_knit$set(root.dir = '../')
 knitr::opts_chunk$set(
@@
 ```{r mut_hotspots}
-#params<-list(drivers = 'data/driver_genes/driver_gene_coords_for_coverage.bed',hotspots='/lustre/scratch125/casm/team268im/lh22/synovium_nanoseq/ex_t_dndsout/exome/All_dnds_annot_muts.tsv')
+# read in bam mappings, get celltype-split bams only
+mappings <-
+  readr::read_csv(params$bam_mappings) %>%
+  dplyr::filter(!is.na(celltype))
+
+# read in drivers
 drivers <-
@@
   dplyr::mutate(chr = gsub('chr', '', chr))
+
+# read in mutations in drivers
 hotspots <-
-  readr::read_tsv(params$hotspots)
-gr <-
-  GRanges(seqnames = drivers$chr,
-          ranges = IRanges(start = drivers$start,
-                           end = drivers$end))
+  readr::read_tsv(params$hotspots) %>%
+  dplyr::mutate(mut_pos = pos,
+                chr = gsub('chr', '', chr)) %>%
+  dplyr::group_by(chr, mut_pos, gene, ref_cod) %>%
+  # summarise each position
+  dplyr::summarise(n = dplyr::n(),
+                   id = paste(id, collapse = ','),
+                   mut_cod = paste(mut_cod, collapse = ','),
+                   impact = paste(impact, collapse = ',')) %>% 
+  dplyr::group_by(dplyr::across(everything())) %>%
+  # extend pileup window 20bp either side of mutated position
+  dplyr::reframe(pos = (mut_pos - 20):(mut_pos + 20))
+
+# save as a bed
+hotspots %>%
+  dplyr::transmute(chr = paste0('chr', chr), start = pos - 1, end = pos) %>%
+  readr::write_tsv(
+    'data/driver_genes/lcm_wes_mutations_20bp_window.bed',
+    col_names = F)
+
+# get ref allele at each position
+refs <- xfun::cache_rds({
+  refs <-
+    system('bedtools getfasta -fi /lustre/scratch125/casm/team268im/al28/bed_ref/GRCh38_full_analysis_set_plus_decoy_hla_genome.fa -bed data/driver_genes/lcm_wes_mutations_20bp_window.bed -tab', intern = T) %>%
+    paste(collapse = '\n') %>%
+    readr::read_tsv(col_names = c('pos', 'ref')) %>%
+    tidyr::separate_wider_delim(cols = pos, delim = stringr::regex(':|-'), names = c('chr', 'start', 'end')) %>%
+    type.convert(as.is = T) %>%
+    dplyr::transmute(chr = gsub('chr', '', chr), pos = end, ref) %>%
+    dplyr::distinct()
+  refs
+}, file = 'refs.rds', rerun = params$rerun)
 ```
 
-# BAMs from snRNA-seq data
+# Pileup of BAMs from snRNA-seq data
 
-We read in reads in the driver regions
+```{r pileup}
+# bam2r filtering arguments
+min_phred <- 30
+sam_flag <- 3844
+mapq <- 25
+nucleotides <- c("A", "T", "C", "G", "-", "INS", "a", "t", "c", "g", "_", "ins")
 
-```{r scanbam}
-bam_path <-
-  '/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/out/Zhang2023/coverage/BRI_401/celltype_bams/BRI_401.B_cell_plasma_cell.bam'
-bam_file <- Rsamtools::BamFile(bam_path)
-bam_params <- 
-  Rsamtools::ScanBamParam(which = gr, what = Rsamtools::scanBamWhat())
-bam <- Rsamtools::scanBam(bam_file, param = bam_params)
+# list of id - ct - bam
+bams <-
+  mappings %>%
+  {split(., .$id)} %>%
+  purrr::map(function(df) {split(df$bam, df$celltype)})
+
+# get bases at sites
+sites <- xfun::cache_rds({
+
+  sites <-
+    bams %>% 
+    purrr::map(function(id) {
+      id %>% 
+        purrr::map(function(ct_bam) {
+
+          hotspots %>%
+            purrr::pmap(function(chr, pos, mut_pos, gene, ...) {
+              deepSNV::bam2R(
+                ct_bam,
+                chr = chr,
+                start = pos,
+                stop = pos,
+                q = min_phred,
+                mask = sam_flag,
+                mq = mapq) %>%
+                tibble::as_tibble() %>%
+                dplyr::mutate(chr = chr, pos = pos, mut_pos = mut_pos, gene = gene) %>%
+                dplyr::select(chr, pos, mut_pos, gene, dplyr::all_of(nucleotides))
+            }) %>%
+            dplyr::bind_rows() %>%
+            dplyr::rowwise() %>%
+            dplyr::mutate(
+              total = sum(dplyr::across(dplyr::all_of(nucleotides)))) %>%
+            dplyr::relocate(dplyr::all_of(nucleotides), .after = dplyr::last_col())
+
+        }) %>% dplyr::bind_rows(.id = 'celltype')
+    }) %>% dplyr::bind_rows(.id = 'id')
+
+  # return
+  sites
+
+}, file = 'sites.rds', rerun = params$rerun)
+
+# add refs, compare
+muts <-
+  sites %>%
+  dplyr::left_join(refs) %>%
+  tidyr::pivot_longer(cols = dplyr::all_of(nucleotides),
+                      values_to = 'nt_n_reads', names_to = 'nt') %>%
+  # # ignore indels for now
+  # dplyr::filter(!nt %in% c('-', '_', 'INS', 'ins')) %>%
+  # get number of ref and non-ref alleles at each position
+  dplyr::mutate(n_ref = ifelse(ref == toupper(nt), nt_n_reads, 0),
+                n_non_ref = ifelse(ref != toupper(nt), nt_n_reads, 0)) %>%
+  dplyr::group_by(id, celltype, chr, pos, mut_pos, gene, total, ref) %>%
+  dplyr::summarise(n_ref = sum(n_ref), n_non_ref = sum(n_non_ref)) %>%
+  dplyr::mutate(vaf = n_non_ref / total)
+
+# plot positions
+unique(sites$gene) %>%
+  purrr::walk(function(i) {
+    p <-
+      sites %>%
+      dplyr::filter(gene == i) %>%
+      dplyr::left_join(refs, by = c('chr', 'pos')) %>%
+      tidyr::pivot_longer(cols = dplyr::all_of(nucleotides),
+                          values_to = 'nt_n_reads', names_to = 'nt') %>%
+      dplyr::mutate(
+        strandless_nt = 
+          ifelse(nt %in% c('-', '_'), 'DEL', toupper(nt)),
+        is_ref = ifelse(strandless_nt == ref, 'ref', 'alt'),
+        strandless_nt = paste0(strandless_nt, '_', is_ref) %>% 
+          forcats::fct_relevel(unique(c(paste0(topupper(nucleotides), '_ref'),
+                                        paste0(toupper(nucleotides), '_alt'))))) %>%
+      ggplot(aes(x = pos, y = nt_n_reads, fill = strandless_nt)) + 
+      geom_vline(aes(xintercept = mut_pos)) +
+      geom_col(position = 'stack') +
+      theme_bw() +
+      theme(axis.text.x = element_blank()) +
+      ggh4x::facet_grid2(celltype ~ mut_pos, scales = 'free_x') +
+      scale_fill_manual(values = c(A_ref = 'darkgreen', A_alt = 'lightgreen',
+                                   C_ref = 'darkblue', C_alt = 'lightblue',
+                                   G_ref = 'darkorange3', G_alt = 'peachpuff',
+                                   T_ref = 'darkred', T_alt = 'salmon',
+                                   INS_alt = 'black', DEL_alt = 'grey')) +
+      ggtitle(i)
+    print(p)
+    })
 ```

### 2024-03-05 scomatic/src/00_preprocess_drivers.R

@@ -94,10 +94,10 @@ drivers_pos <-
   # remove weird chromosomes
   dplyr::filter(chr %in% c(as.character(1:22, 'X', 'Y')))
 
-# save bed
+# save driver bed
 drivers_pos %>%
   dplyr::transmute(
-    paste0('chr', chr), start, end, gene
+    chr, start, end, gene
   ) %>%
   readr::write_tsv('data/driver_genes/driver_gene_coords_for_coverage.bed',
                    col_names = F)
@@ -134,7 +134,7 @@ hotspots <-
   readr::read_tsv() %>%
   dplyr::mutate(chr = paste0('chr', chr)) %>%
   dplyr::rename(pos_GRCh37 = pos) %>%
-  dplyr::mutate(driver = gene %in% drivers_pos$gene)
+  dplyr::filter(gene %in% drivers_pos$gene)
 
 # load liftOver output (GRCh37 -> GRCh38 in UCSC LiftOver)
 lo <-
@@ -148,8 +148,8 @@ lo <-
 # lift over mutations (3 positions are lost because they were deleted in the new build)
 hotspots <-
   hotspots %>%
-  dplyr::left_join(lo) %>%
-  dplyr::select(sampleID, chr, pos, everything())
+  dplyr::inner_join(lo) %>%
+  dplyr::select(id = sampleID, chr, pos, everything())
 
 # save
 hotspots %>%

### 2024-03-05 scomatic/src/02_run_scomatic_low_thresholds.sh

@@ -13,7 +13,7 @@ mkdir -p work/ out/Zhang2023/default_thresholds/ log/
 . ~/.bashrc
 mamba activate jupy
 
-# run scomatic
+# run scomatic on drivers
 /software/team205/nextflow-23.04.1-all run nextflow/scomatic.nf \
     --projectDir $PWD \
     --max_nM 1000000 \
@@ -23,23 +23,25 @@ mamba activate jupy
     --min_dp 1 \
     --min_cc 1 \
     --max_cell_types 1000000 \
-    --output_dir '/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/out/Zhang2023/no_thresholds/' \
+    --output_dir $wd/out/Zhang2023/no_thresholds/ \
+    --cleanup false \
     -params-file $config_dir/GEX.json \
     -c config/Zhang2023.config \
-    -c $config_dir/LSF.config \
+    -c config/LSF.config \
     -w work/ \
     --location local \
     -resume
+#    --subset_bed $wd/data/driver_genes/driver_gene_coords_for_coverage.bed \
 
-# get single cell genotypes
-/software/team205/nextflow-23.04.1-all run nextflow/scomatic.nf \
-    -entry genotypes \
-    --projectDir $PWD \
-    --min_dp 1 \
-    --output_dir '/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/out/Zhang2023/no_thresholds/' \
-    -params-file $config_dir/GEX.json \
-    -c config/Zhang2023.config \
-    -c $config_dir/LSF.config \
-    -w work/ \
-    --location local \
-    -resume
+# # get single cell genotypes
+# /software/team205/nextflow-23.04.1-all run nextflow/scomatic.nf \
+#     -entry genotypes \
+#     --projectDir $PWD \
+#     --min_dp 1 \
+#     --output_dir '/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/out/Zhang2023/no_thresholds/' \
+#     -params-file $config_dir/GEX.json \
+#     -c config/Zhang2023.config \
+#     -c $config_dir/LSF.config \
+#     -w work/ \
+#     --location local \
+#     -resume

### 2024-03-05 scomatic/src/03_knit_driver_coverage.sh

@@ -36,13 +36,19 @@ ct_bams_dir=/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/out/Zhang
 /software/team205/nextflow-23.04.1-all run nextflow/driver_coverage.nf \
     --mappings ${wd}out/Zhang2023/coverage/mappings.csv \
     --drivers ${wd}data/driver_genes/driver_gene_coords_for_coverage.tsv \
+    --mutations ${wd}data/driver_genes/lcm_wes_mutations.tsv \
     --out_dir ${wd}out/Zhang2023/coverage/ \
     --min_MQ 0 \
     -c /nfs/team205/kp9/nextflow/scomatic/LSF.config  \
     --location local \
-    -w work/ \
+    -w test/work/ \
     -resume
-    
+
+# knit target_mutation_calling.Rmd
+mkdir out/Zhang2023/coverage/summary/
+Rscript -e "rmarkdown::render('reports/targeted_mutation_calling.Rmd', params = list(drivers = 'data/driver_genes/driver_gene_coords_for_coverage.bed', hotspots = 'data/driver_genes/lcm_wes_mutations.tsv', bam_mappings = 'out/Zhang2023/coverage/mappings.csv', rerun = F, cache_dir = 'out/Zhang2023/coverage/summary/targeted_mutation_calling_cache/'), output_file = 'targeted_mutation_calling.html', output_dir = 'out/Zhang2023/coverage/summary/')"
+
+
 # # run test
 # # head -6 data/Zhang2023/coverage_mappings.csv > data/Zhang2023/coverage_mappings_test.csv
 # # head -2 data/driver_genes/driver_gene_coords_for_coverage.bed > data/driver_genes/driver_gene_coords_for_coverage_test.bed

### 2024-03-06 infercnv/reports/gains_and_losses.Rmd
add chisq test, fix chr gain contingency plotting

@@
 
-```{r chr7_and_chr5}
+```{r chr7_and_others}
 events <-
@@
   dplyr::mutate(prop = n / dplyr::n()) %>%
-  # ignore events that happen in <10 cells
-  dplyr::filter(events != 'WT', n >= 10) %>%
+  # ignore events that happen in <= 25 cells
+  dplyr::filter(events != 'WT', n >= 25) %>%
   dplyr::mutate(
@@
       gsub('_gain', '+', .) %>%
-      gsub('_loss', '+', .),
+      gsub('_loss', '-', .),
     events = forcats::fct_infreq(events),
@@
     ))
+
+# plot
+p <-
   p_dat %>%
-  ggplot(aes(x = events, fill = cell_group, label = label)) +
-  geom_bar() +
-  geom_text(aes(y = n), hjust = 0, vjust = -0.1, angle = 45, size = 2) +
+  ggplot(aes(x = events)) +
+  geom_text(aes(y = n + 50,  label = label), 
+            vjust = 0.5, hjust = 0, angle = 90, size = 2) +
   scale_x_discrete(guide = guide_axis(angle = -90)) +
@@
   theme(legend.position = 'bottom')
-p_dat %>%
-  ggplot(aes(x = forcats::fct_infreq(events), fill = sample, label = label)) +
-  geom_bar() +
-  geom_text(aes(y = n), hjust = 0, vjust = -0.1, angle = 45, size = 2) +
-  scale_x_discrete(guide = guide_axis(angle = -90)) +
-  ylim(c(0, max(p_dat$n) * 1.1)) +
-  theme_classic() +
+p + 
+  geom_bar(aes(fill = sample))
+p + 
+  geom_bar(aes(fill = cell_group)) +
   theme(legend.position = 'none')
+```
+
+We also can run a Fisher exact test to test whether chr7+ and chr5+ gains co-occur more often than would be expected by chance, and also test all other co-occurences.
+
+```{r cont_tab}
+cont <-
+  sce@colData %>%
+  tibble::as_tibble() %>%
+  dplyr::mutate(chr7 = ifelse(chr7_gain == 1, 'chr7+', 'chr7_WT'),
+                chr5 = ifelse(chr5_gain == 1, 'chr5+', 'chr5_WT')) 
+table(cont$chr7, cont$chr5) %>%
+  knitr::kable()
+```
+
+```{r fisher_exact}
+chisq <- xfun::cache_rds({
+
+  chisq <-
+    colnames(sce@colData)[grepl('_gain$|_loss$', colnames(sce@colData))] %>%
+    purrr::set_names() %>%
+    purrr::map(function(event) {
+
+      if (1 %in% cont[, event, drop = T]) {
+
+        # run chi-squared test
+        chisq <- chisq.test(sce@colData[, 'chr7_gain', drop = T], 
+                            sce@colData[, event, drop = T])
+
+         # return relevant stats
+        tibble::tibble(
+          X_squared = chisq$statistic, 
+          p = chisq$p.value, 
+          df = chisq$parameter['df'])
+
+      }
+
+    }) %>%
+    dplyr::bind_rows(.id = 'event') %>%
+    # remove chr7+
+    dplyr::filter(event != 'chr7_gain') %>%
+    dplyr::arrange(p)
+
+  # return
+  chisq
+
+}, file = 'chisq.rds', rerun = params$rerun)
+
+chisq %>%
+  knitr::kable()
 ```

### 2024-03-06 scomatic/reports/targeted_mutation_calling.Rmd
add chisq test, fix chr gain contingency plotting

@@
   readr::read_csv(params$bam_mappings) %>%
-  dplyr::filter(!is.na(celltype))
+  dplyr::filter(!is.na(celltype)) %>%
+  # temporarily remove problematic BRI-401 endo bam
+  dplyr::filter(!(id == 'BRI_401' & celltype == 'Endothelial_cell'))
 
@@
     readr::read_tsv(col_names = c('pos', 'ref')) %>%
-    tidyr::separate_wider_delim(cols = pos, delim = stringr::regex(':|-'), names = c('chr', 'start', 'end')) %>%
+    tidyr::separate_wider_delim(cols = pos, delim = stringr::regex(':|-'), 
+                                names = c('chr', 'start', 'end')) %>%
     type.convert(as.is = T) %>%
@@
                       values_to = 'nt_n_reads', names_to = 'nt') %>%
-  # # ignore indels for now
-  # dplyr::filter(!nt %in% c('-', '_', 'INS', 'ins')) %>%
   # get number of ref and non-ref alleles at each position
@@
   dplyr::mutate(vaf = n_non_ref / total)
+```
 
+```{r plot}
 # plot positions
@@
       theme(axis.text.x = element_blank()) +
-      ggh4x::facet_grid2(celltype ~ mut_pos, scales = 'free_x') +
+      ggh4x::facet_grid2(celltype ~ mut_pos, scales = 'free') +
       scale_fill_manual(values = c(A_ref = 'darkgreen', A_alt = 'lightgreen',

### 2024-03-06 infercnv/src/03_knit_gains_and_losses.sh

@@ -9,13 +9,13 @@ cd $wd
 . ~/.bashrc
 mamba activate jupy
 
-# # knit by_fibroblast_cluster, < 0.5 | > 0.8
-# mkdir -p out/Zhang2023/by_fibroblast_cluster/summary/
-# Rscript -e "rmarkdown::render('reports/gains_and_losses.Rmd', params = list(sce = 'data/Zhang2023/sce/Fibroblast_sce.rds', cell_group_name = 'cluster_name', infercnv_dir = 'out/Zhang2023/by_stromal_cluster/', cache_dir = 'out/Zhang2023/by_fibroblast_cluster/summary/gains_and_losses_cache/', rerun = F), output_file = 'gains_and_losses.html', output_dir = 'out/Zhang2023/by_fibroblast_cluster/summary/')"
+# knit by_fibroblast_cluster, < 0.5 | > 0.8
+mkdir -p out/Zhang2023/by_fibroblast_cluster/summary/
+Rscript -e "rmarkdown::render('reports/gains_and_losses.Rmd', params = list(sce = 'data/Zhang2023/sce/Fibroblast_sce.rds', cell_group_name = 'cluster_name', infercnv_dir = 'out/Zhang2023/by_stromal_cluster/', cache_dir = 'out/Zhang2023/by_fibroblast_cluster/summary/gains_and_losses_cache/', rerun = F), output_file = 'gains_and_losses.html', output_dir = 'out/Zhang2023/by_fibroblast_cluster/summary/')"
 
-# knit by_fibroblast_cluster, < 0.2 | > 0.8
-mkdir -p out/Zhang2023/by_fibroblast_cluster_0.2/summary/
-Rscript -e "rmarkdown::render('reports/gains_and_losses.Rmd', params = list(sce = 'data/Zhang2023/sce/Fibroblast_sce.rds', cell_group_name = 'cluster_name', infercnv_dir = 'out/Zhang2023/by_stromal_cluster/', cache_dir = 'out/Zhang2023/by_fibroblast_cluster_0.2/summary/gains_and_losses_cache/', rerun = T, max_prop_for_wt = 0.2), output_file = 'gains_and_losses.html', output_dir = 'out/Zhang2023/by_fibroblast_cluster_0.2/summary/')"
+# # knit by_fibroblast_cluster, < 0.2 | > 0.8
+# mkdir -p out/Zhang2023/by_fibroblast_cluster_0.2/summary/
[…]

### 2024-03-06 scomatic/src/00_preprocess_drivers.R

@@ -151,6 +151,9 @@ hotspots <-
   dplyr::inner_join(lo) %>%
   dplyr::select(id = sampleID, chr, pos, everything())
 
+# # dndscv
+# dndsout <- dndscv(muts, max_muts_per_gene_per_sample = Inf, max_coding_muts_per_sample = Inf)
+
 # save
 hotspots %>%
   readr::write_tsv('data/driver_genes/lcm_wes_mutations.tsv')

### 2024-03-06 scomatic/src/02_run_scomatic_low_thresholds.sh

@@ -16,6 +16,7 @@ mamba activate jupy
 # run scomatic on drivers
 /software/team205/nextflow-23.04.1-all run nextflow/scomatic.nf \
     --projectDir $PWD \
+    --publish_celltype_bams true \
     --max_nM 1000000 \
     --max_NH 1000000 \
     --min_MQ 0 \
@@ -24,7 +25,6 @@ mamba activate jupy
     --min_cc 1 \
     --max_cell_types 1000000 \
     --output_dir $wd/out/Zhang2023/no_thresholds/ \
-    --cleanup false \
     -params-file $config_dir/GEX.json \
     -c config/Zhang2023.config \
     -c config/LSF.config \

### 2024-03-06 infercnv/reports/gains_and_losses.Rmd
change chisq to fisher exact in gains/losses statistical test

@@
 ```{r fisher_exact}
-chisq <- xfun::cache_rds({
+fisher <- xfun::cache_rds({
 
-  chisq <-
+  fisher <-
     colnames(sce@colData)[grepl('_gain$|_loss$', colnames(sce@colData))] %>%
@@
 
-      if (1 %in% cont[, event, drop = T]) {
+      event_vec <- sce@colData[, event, drop = T]
 
-        # run chi-squared test
-        chisq <- chisq.test(sce@colData[, 'chr7_gain', drop = T], 
-                            sce@colData[, event, drop = T])
+      if (1 %in% event_vec) {
+
+        # run fisher exact test
+        fisher <- fisher.test(sce$chr7_gain, event_vec)
 
@@
         tibble::tibble(
-          X_squared = chisq$statistic, 
-          p = chisq$p.value, 
-          df = chisq$parameter['df'])
+          OR = fisher$estimate, 
+          p = fisher$p.value)
+
+      } else {
+
+        # no event
+        tibble::tibble(OR = NA, p = NA)
 
@@
     dplyr::bind_rows(.id = 'event') %>%
-    # remove chr7+
+    # remove chr7+ (can't do chr7+ vs chr7+)
     dplyr::filter(event != 'chr7_gain') %>%
-    dplyr::arrange(p)
+    dplyr::arrange(-OR)
 
   # return
-  chisq
+  fisher
 
-}, file = 'chisq.rds', rerun = params$rerun)
+}, file = 'fisher.rds', rerun = params$rerun)
 
-chisq %>%
+fisher %>%
   knitr::kable()

### 2024-03-06 scomatic/reports/targeted_mutation_calling.Rmd
fix up bam mutation calling

@@
 
+```{r}
+# read targeted nanoseq 
+tnanoseq <-
+  '/lustre/scratch125/casm/team268im/lh22/synovium_nanoseq/ex_t_dndsout/dnds_nano_muts.tsv' %>%
+  readr::read_tsv()
+
+# read genes of interest
+target_genes <-
+  '/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/data/driver_genes/Sanger_TERT-v4_TE-95148282_hg19_highstringencyfilter_buccal_gene_list.tsv' %>%
+  readr::read_tsv()
+
+# run dndscv
+dnds <-
+  dndscv::dndscv(
+    unique(tnanoseq[, 1:5]),
+    max_muts_per_gene_per_sample = Inf,
+    max_coding_muts_per_sample = Inf,
+    gene_list = target_genes$gene,
+    outmats = T)
+
+# run sitednds
+sitednds <-
+  dndscv::sitednds(dnds)
+
+# get the significant genes
+signif_genes <-
+  dnds$sel_cv %>%
+  tibble::as_tibble() %>%
+  dplyr::filter(qglobal_cv < 0.1)
+
+# get the sites with positively selected mutations
+recursites <-
+  sitednds$recursites %>%
+  tibble::as_tibble()
+
+# convert to granges (1-based!)
+gr <- GRanges(
+  seqnames = paste0('chr', recursites$chr),
+  ranges = IRanges(start = recursites$pos, end = recursites$pos),
+  gene = recursites$gene,
+  pos_GRCh37 = recursites$pos)
+
+# get chain file
+chain <-
+  rtracklayer::import.chain('data/liftOver/hg19ToHg38.over.chain')
+
+# lift over
+lifted <-
+  rtracklayer::liftOver(gr, chain) %>%
+  unlist() %>%
+  tibble::as_tibble() %>%
+  dplyr::transmute(chr = gsub('chr', '', seqnames), 
+                   pos = start, gene, pos_GRCh37)
+
+# update positions
+recursites <-
+  recursites %>%
+  dplyr::rename(pos_GRCh37 = pos) %>%
+  dplyr::left_join(lifted, by = c('chr', 'gene', 'pos_GRCh37')) %>%
+  dplyr::select(chr, pos, gene, ref, mut, pos_GRCh37, everything())
+```
+
 # Pileup of BAMs from snRNA-seq data
@@
 
-          hotspots %>%
-            purrr::pmap(function(chr, pos, mut_pos, gene, ...) {
+          recursites %>%
+            # dplyr::filter(gene == 'TNFRSF14', aachange == 'W12*') %>% 
+            purrr::pmap(function(chr, pos, start20, stop20, gene, ...) {
+              # testing: # recursites %>% dplyr::filter(gene == 'TNFRSF14', aachange == 'W12*', pos == mut_pos) %>% as.list() %>% list2env(envir = globalenv())
+
+              # positions
+              positions <- (pos - 20):(pos + 20)
+
+              # get reads at each position of the window
               deepSNV::bam2R(
@@
                 chr = chr,
-                start = pos,
-                stop = pos,
+                # extend pileup window 20bp either side of mutated position
+                start = pos - 20,
+                stop = pos + 20,
                 q = min_phred,
@@
                 tibble::as_tibble() %>%
-                dplyr::mutate(chr = chr, pos = pos, mut_pos = mut_pos, gene = gene) %>%
+                dplyr::mutate(chr = chr, pos = positions, mut_pos = pos, gene = gene) %>%
                 dplyr::select(chr, pos, mut_pos, gene, dplyr::all_of(nucleotides))
+
             }) %>%
@@
 
-```{r plot}
+```{r plot, fig.width = 20}
 # plot positions
@@
         strandless_nt = paste0(strandless_nt, '_', is_ref) %>% 
-          forcats::fct_relevel(unique(c(paste0(topupper(nucleotides), '_ref'),
+          forcats::fct_relevel(unique(c(paste0(toupper(nucleotides), '_ref'),
                                         paste0(toupper(nucleotides), '_alt'))))) %>%

### 2024-03-06 scomatic/src/03_knit_driver_coverage.sh

@@ -9,40 +9,40 @@ cd $wd
 . ~/.bashrc
 mamba activate jupy
 
-# create mappings
-mappings=data/Zhang2023/mappings.csv
-bams_dir=/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/cellranger_output/
-ct_bams_dir=/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/out/Zhang2023/coverage/
-(
-  echo 'id,bam,celltype' ;
-  while read -r id ; do
-    # check if ct bams directory exists
-    if [ -d $ct_bams_dir/$id/ ] ; then
-      ct_bams=($(ls $ct_bams_dir/$id/celltype_bams/*.bam))
-      # check if scomatic successfully ran
-      if [ ${#ct_bams[*]} -gt 0 ] ; then
-        echo "$id,$bams_dir/${id/_/-}/possorted_genome_bam.bam,NA" ;
-        for file in ${ct_bams[@]} ; do
-          ct=$(basename $file | cut -d. -f2)
-          echo "$id,$file,$ct" ;
-        done
-      fi
-    fi
-  done < \
-  <(sed 1d $mappings | cut -d, -f2 | sort -u)
-) | cat > out/Zhang2023/coverage/mappings.csv
-
-# run
-/software/team205/nextflow-23.04.1-all run nextflow/driver_coverage.nf \
-    --mappings ${wd}out/Zhang2023/coverage/mappings.csv \
-    --drivers ${wd}data/driver_genes/driver_gene_coords_for_coverage.tsv \
-    --mutations ${wd}data/driver_genes/lcm_wes_mutations.tsv \
-    --out_dir ${wd}out/Zhang2023/coverage/ \
-    --min_MQ 0 \
-    -c /nfs/team205/kp9/nextflow/scomatic/LSF.config  \
-    --location local \
-    -w test/work/ \
-    -resume
+# # create mappings
+# mappings=data/Zhang2023/mappings.csv
+# bams_dir=/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/cellranger_output/
+# ct_bams_dir=/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/out/Zhang2023/coverage/
+# (
+#   echo 'id,bam,celltype' ;
+#   while read -r id ; do
+#     # check if ct bams directory exists
[…]

### 2024-03-07 scomatic/reports/targeted_mutation_calling.Rmd
add nextflow wrapper to targeted_mutation_calling by id/celltype

@@
 
-# Mutational calls in drivers from the LCM-WES data
-
-```{r mut_hotspots}
-# read in bam mappings, get celltype-split bams only
-mappings <-
-  readr::read_csv(params$bam_mappings) %>%
-  dplyr::filter(!is.na(celltype)) %>%
-  # temporarily remove problematic BRI-401 endo bam
-  dplyr::filter(!(id == 'BRI_401' & celltype == 'Endothelial_cell'))
-
-# read in drivers
-drivers <-
-  readr::read_tsv(params$drivers, col_names = c('chr', 'start', 'end', 'gene')) %>%
-  dplyr::mutate(chr = gsub('chr', '', chr))
-
-# read in mutations in drivers
-hotspots <-
-  readr::read_tsv(params$hotspots) %>%
-  dplyr::mutate(mut_pos = pos,
-                chr = gsub('chr', '', chr)) %>%
-  dplyr::group_by(chr, mut_pos, gene, ref_cod) %>%
-  # summarise each position
-  dplyr::summarise(n = dplyr::n(),
-                   id = paste(id, collapse = ','),
-                   mut_cod = paste(mut_cod, collapse = ','),
-                   impact = paste(impact, collapse = ',')) %>% 
-  dplyr::group_by(dplyr::across(everything())) %>%
-  # extend pileup window 20bp either side of mutated position
-  dplyr::reframe(pos = (mut_pos - 20):(mut_pos + 20))
-
-# save as a bed
-hotspots %>%
-  dplyr::transmute(chr = paste0('chr', chr), start = pos - 1, end = pos) %>%
-  readr::write_tsv(
-    'data/driver_genes/lcm_wes_mutations_20bp_window.bed',
-    col_names = F)
-
-# get ref allele at each position
-refs <- xfun::cache_rds({
-  refs <-
-    system('bedtools getfasta -fi /lustre/scratch125/casm/team268im/al28/bed_ref/GRCh38_full_analysis_set_plus_decoy_hla_genome.fa -bed data/driver_genes/lcm_wes_mutations_20bp_window.bed -tab', intern = T) %>%
-    paste(collapse = '\n') %>%
-    readr::read_tsv(col_names = c('pos', 'ref')) %>%
-    tidyr::separate_wider_delim(cols = pos, delim = stringr::regex(':|-'), 
-                                names = c('chr', 'start', 'end')) %>%
-    type.convert(as.is = T) %>%
-    dplyr::transmute(chr = gsub('chr', '', chr), pos = end, ref) %>%
-    dplyr::distinct()
-  refs
-}, file = 'refs.rds', rerun = params$rerun)
-```
-
-```{r}
+```{r load}
 # read targeted nanoseq 
@@
   readr::read_tsv()
+```
 
+```{r dndscv_and_liftover}
 # run dndscv
@@
             # dplyr::filter(gene == 'TNFRSF14', aachange == 'W12*') %>% 
-            purrr::pmap(function(chr, pos, start20, stop20, gene, ...) {
-              # testing: # recursites %>% dplyr::filter(gene == 'TNFRSF14', aachange == 'W12*', pos == mut_pos) %>% as.list() %>% list2env(envir = globalenv())
+            purrr::pmap(function(chr, pos, ref, mut, gene, ...) {
+              # testing: # recursites %>% dplyr::filter(gene == 'TNFRSF14', aachange == 'W12*') %>% as.list() %>% list2env(envir = globalenv())
 
@@
               # get reads at each position of the window
+              ct_bam %>%
                 deepSNV::bam2R(
-                ct_bam,
                   chr = chr,
@@
                   stop = pos + 20,
+                  # quality controls
                   q = min_phred,
@@
 ```
+
+# Mutational calls in drivers from the LCM-WES data
+
+```{r mut_hotspots, eval = F}
+# read in bam mappings, get celltype-split bams only
+mappings <-
+  readr::read_csv(params$bam_mappings) %>%
+  dplyr::filter(!is.na(celltype)) %>%
+  # temporarily remove problematic BRI-401 endo bam
+  dplyr::filter(!(id == 'BRI_401' & celltype == 'Endothelial_cell'))
+
+# read in drivers
+drivers <-
+  readr::read_tsv(params$drivers, col_names = c('chr', 'start', 'end', 'gene')) %>%
+  dplyr::mutate(chr = gsub('chr', '', chr))
+
+# read in mutations in drivers
+hotspots <-
+  readr::read_tsv(params$hotspots) %>%
+  dplyr::mutate(mut_pos = pos,
+                chr = gsub('chr', '', chr)) %>%
+  dplyr::group_by(chr, mut_pos, gene, ref_cod) %>%
+  # summarise each position
+  dplyr::summarise(n = dplyr::n(),
+                   id = paste(id, collapse = ','),
+                   mut_cod = paste(mut_cod, collapse = ','),
+                   impact = paste(impact, collapse = ',')) %>% 
+  dplyr::group_by(dplyr::across(everything())) %>%
+  # extend pileup window 20bp either side of mutated position
+  dplyr::reframe(pos = (mut_pos - 20):(mut_pos + 20))
+
+# save as a bed
+hotspots %>%
+  dplyr::transmute(chr = paste0('chr', chr), start = pos - 1, end = pos) %>%
+  readr::write_tsv(
+    'data/driver_genes/lcm_wes_mutations_20bp_window.bed',
+    col_names = F)
+
+# get ref allele at each position
+refs <- xfun::cache_rds({
+  refs <-
+    system('bedtools getfasta -fi /lustre/scratch125/casm/team268im/al28/bed_ref/GRCh38_full_analysis_set_plus_decoy_hla_genome.fa -bed data/driver_genes/lcm_wes_mutations_20bp_window.bed -tab', intern = T) %>%
+    paste(collapse = '\n') %>%
+    readr::read_tsv(col_names = c('pos', 'ref')) %>%
+    tidyr::separate_wider_delim(cols = pos, delim = stringr::regex(':|-'), 
+                                names = c('chr', 'start', 'end')) %>%
+    type.convert(as.is = T) %>%
+    dplyr::transmute(chr = gsub('chr', '', chr), pos = end, ref) %>%
+    dplyr::distinct()
+  refs
+}, file = 'refs.rds', rerun = params$rerun)
+```

### 2024-03-07 scomatic/src/03_knit_driver_coverage.sh

@@ -44,10 +44,25 @@ mamba activate jupy
 #     -w test/work/ \
 #     -resume
 
-# knit target_mutation_calling.Rmd
-mkdir out/Zhang2023/coverage/summary/
-Rscript -e "rmarkdown::render('reports/targeted_mutation_calling.Rmd', params = list(drivers = 'data/driver_genes/driver_gene_coords_for_coverage.bed', hotspots = 'data/driver_genes/lcm_wes_mutations.tsv', bam_mappings = 'out/Zhang2023/coverage/mappings.csv', rerun = F, cache_dir = 'out/Zhang2023/coverage/summary/targeted_mutation_calling_cache/'), output_file = 'targeted_mutation_calling.html', output_dir = 'out/Zhang2023/coverage/summary/')"
+# # knit target_mutation_calling.Rmd
+# mkdir out/Zhang2023/coverage/summary/
+# Rscript -e "rmarkdown::render('reports/targeted_mutation_calling.Rmd', params = list(drivers = 'data/driver_genes/driver_gene_coords_for_coverage.bed', hotspots = 'data/driver_genes/lcm_wes_mutations.tsv', bam_mappings = 'out/Zhang2023/coverage/mappings.csv', rerun = F, cache_dir = 'out/Zhang2023/coverage/summary/targeted_mutation_calling_cache/'), output_file = 'targeted_mutation_calling.html', output_dir = 'out/Zhang2023/coverage/summary/')"
 
+# run target_mutation_calling.nf on celltype bams only
+out_dir=${wd}out/Zhang2023/targeted_mutation_calling/
+mkdir -p $out_dir
+
+# get mappings of celltype-specific bams only
+cat ${wd}out/Zhang2023/coverage/mappings.csv | 
+awk -F',' -v OFS=',' '$3 != "NA" {print}' \
+> ${out_dir}/mappings.csv
+
+# run
+/software/team205/nextflow-23.04.1-all run nextflow/targeted_mutation_calling.nf \
+  --mappings ${out_dir}/mappings.csv \
+  --out_dir ${out_dir} \
+  -c /nfs/team205/kp9/nextflow/scomatic/LSF.config \
+  -w work/  #-resume
 
 # # run test
 # # head -6 data/Zhang2023/coverage_mappings.csv > data/Zhang2023/coverage_mappings_test.csv

### 2024-03-07 scomatic/reports/targeted_mutation_calling.Rmd
developing targeted mutation calling

@@
 
-```{r load}
-# read targeted nanoseq 
-tnanoseq <-
-  '/lustre/scratch125/casm/team268im/lh22/synovium_nanoseq/ex_t_dndsout/dnds_nano_muts.tsv' %>%
-  readr::read_tsv()
-
-# read genes of interest
-target_genes <-
-  '/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/data/driver_genes/Sanger_TERT-v4_TE-95148282_hg19_highstringencyfilter_buccal_gene_list.tsv' %>%
-  readr::read_tsv()
-```
-
-```{r dndscv_and_liftover}
-# run dndscv
-dnds <-
-  dndscv::dndscv(
-    unique(tnanoseq[, 1:5]),
-    max_muts_per_gene_per_sample = Inf,
-    max_coding_muts_per_sample = Inf,
-    gene_list = target_genes$gene,
-    outmats = T)
-
-# run sitednds
-sitednds <-
-  dndscv::sitednds(dnds)
-
-# get the significant genes
-signif_genes <-
-  dnds$sel_cv %>%
-  tibble::as_tibble() %>%
-  dplyr::filter(qglobal_cv < 0.1)
-
-# get the sites with positively selected mutations
-recursites <-
-  sitednds$recursites %>%
-  tibble::as_tibble()
-
-# convert to granges (1-based!)
-gr <- GRanges(
-  seqnames = paste0('chr', recursites$chr),
-  ranges = IRanges(start = recursites$pos, end = recursites$pos),
-  gene = recursites$gene,
-  pos_GRCh37 = recursites$pos)
-
-# get chain file
-chain <-
-  rtracklayer::import.chain('data/liftOver/hg19ToHg38.over.chain')
-
-# lift over
-lifted <-
-  rtracklayer::liftOver(gr, chain) %>%
-  unlist() %>%
-  tibble::as_tibble() %>%
-  dplyr::transmute(chr = gsub('chr', '', seqnames), 
-                   pos = start, gene, pos_GRCh37)
-
-# update positions
-recursites <-
-  recursites %>%
-  dplyr::rename(pos_GRCh37 = pos) %>%
-  dplyr::left_join(lifted, by = c('chr', 'gene', 'pos_GRCh37')) %>%
-  dplyr::select(chr, pos, gene, ref, mut, pos_GRCh37, everything())
-```
-
-# Pileup of BAMs from snRNA-seq data
-
-```{r pileup}
-# bam2r filtering arguments
-min_phred <- 30
-sam_flag <- 3844
-mapq <- 25
-nucleotides <- c("A", "T", "C", "G", "-", "INS", "a", "t", "c", "g", "_", "ins")
-
-# list of id - ct - bam
-bams <-
-  mappings %>%
-  {split(., .$id)} %>%
-  purrr::map(function(df) {split(df$bam, df$celltype)})
-
-# get bases at sites
-sites <- xfun::cache_rds({
-
-  sites <-
-    bams %>% 
-    purrr::map(function(id) {
-      id %>% 
-        purrr::map(function(ct_bam) {
-
-          recursites %>%
-            # dplyr::filter(gene == 'TNFRSF14', aachange == 'W12*') %>% 
-            purrr::pmap(function(chr, pos, ref, mut, gene, ...) {
-              # testing: # recursites %>% dplyr::filter(gene == 'TNFRSF14', aachange == 'W12*') %>% as.list() %>% list2env(envir = globalenv())
-
-              # positions
-              positions <- (pos - 20):(pos + 20)
-
-              # get reads at each position of the window
-              ct_bam %>%
-                deepSNV::bam2R(
-                  chr = chr,
-                  # extend pileup window 20bp either side of mutated position
-                  start = pos - 20,
-                  stop = pos + 20,
-                  # quality controls
-                  q = min_phred,
-                  mask = sam_flag,
-                  mq = mapq) %>%
-                tibble::as_tibble() %>%
-                dplyr::mutate(chr = chr, pos = positions, mut_pos = pos, gene = gene) %>%
-                dplyr::select(chr, pos, mut_pos, gene, dplyr::all_of(nucleotides))
-
-            }) %>%
-            dplyr::bind_rows() %>%
-            dplyr::rowwise() %>%
-            dplyr::mutate(
-              total = sum(dplyr::across(dplyr::all_of(nucleotides)))) %>%
-            dplyr::relocate(dplyr::all_of(nucleotides), .after = dplyr::last_col())
-
-        }) %>% dplyr::bind_rows(.id = 'celltype')
-    }) %>% dplyr::bind_rows(.id = 'id')
-
-  # return
-  sites
-
-}, file = 'sites.rds', rerun = params$rerun)
-
-# add refs, compare
-muts <-
-  sites %>%
-  dplyr::left_join(refs) %>%
-  tidyr::pivot_longer(cols = dplyr::all_of(nucleotides),
-                      values_to = 'nt_n_reads', names_to = 'nt') %>%
-  # get number of ref and non-ref alleles at each position
-  dplyr::mutate(n_ref = ifelse(ref == toupper(nt), nt_n_reads, 0),
-                n_non_ref = ifelse(ref != toupper(nt), nt_n_reads, 0)) %>%
-  dplyr::group_by(id, celltype, chr, pos, mut_pos, gene, total, ref) %>%
-  dplyr::summarise(n_ref = sum(n_ref), n_non_ref = sum(n_non_ref)) %>%
-  dplyr::mutate(vaf = n_non_ref / total)
+```{r load_files}
+list.files(pattern = '_reads.tsv') %>%
+  purrr::set_names(., gsub('_reads.tsv', '', .)) %>%
+  purrr::map(function(file) {
+    reads <- readr::read_tsv(reads)
+  })
 ```

### 2024-03-07 scomatic/src/03_knit_driver_coverage.sh

@@ -60,9 +60,10 @@ awk -F',' -v OFS=',' '$3 != "NA" {print}' \
 # run
 /software/team205/nextflow-23.04.1-all run nextflow/targeted_mutation_calling.nf \
   --mappings ${out_dir}/mappings.csv \
+  --window 1 \
   --out_dir ${out_dir} \
   -c /nfs/team205/kp9/nextflow/scomatic/LSF.config \
-  -w work/  #-resume
+  -w work/ 
 
 # # run test
 # # head -6 data/Zhang2023/coverage_mappings.csv > data/Zhang2023/coverage_mappings_test.csv

### 2024-03-11 scomatic/reports/targeted_mutation_calling.Rmd
fix targeted_mutation_calling and generate notebook of the results

@@
     fig_width: 8
-    keep_md: false
+    keep_md: true
     code_folding: show
@@
 params:
-  drivers: NULL
-  hotspots: NULL
-  bam_mappings: NULL
   cache_dir: './target_mutation_calling_cache/'
@@
 # wd <- ifelse(Sys.info()['nodename'] == "mib119104s", '~/Volumes/', '') |> paste0('/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/') ; setwd(wd)
-# rmarkdown::render('reports/driver_coverage.Rmd', output_file = 'driver_coverage.html', output_dir = 'reports/', params = list(rerun = F))
 
 # chunk options
-knitr::opts_knit$set(root.dir = '../')
 knitr::opts_chunk$set(
@@
 ```{r load_files}
+# all possible nucleotides
+nucleotides <- c("A", "T", "C", "G", "-", "INS", "a", "t", "c", "g", "_", "ins")
+
+# strandless nucleotides - colours and order
+nt_cols <-
+  c('INS' = '#e6ab02',
+    'DEL' = '#1c2d78',
+    'A' = '#1b9e77', 
+    'C' = '#d95f02', 
+    'G' = '#7570b3', 
+    'T' = '#e7298a',
+    'ref' = 'grey')
+
+# recursites (to add detected mut)
+recurmuts <-
+  readr::read_tsv('recursites.tsv') %>%
+  dplyr::transmute(
+    chr,
+    mut_pos = pos,
+    mut_pos_lab = paste0(pos, '\ mut: ', mut)
+  )
+
+# reads
+reads <-
   list.files(pattern = '_reads.tsv') %>%
@@
   purrr::map(function(file) {
-    reads <- readr::read_tsv(reads)
-  })
+    readr::read_tsv(file, show_col_types = F)
+  }) %>%
+  dplyr::bind_rows(.id = 'id') %>%
+  tidyr::pivot_longer(cols = dplyr::all_of(nucleotides)) %>%
+  dplyr::mutate(
+    # colour in alt alleles
+    status = dplyr::case_when(toupper(name) == ref ~ 'ref',
+                              name %in% c('-', '_') ~ 'DEL',
+                              TRUE ~ toupper(name)) %>%
+      forcats::fct_relevel(names(nt_cols))) %>%
+  # add detected mutations
+  dplyr::left_join(recurmuts, by = c('chr', 'mut_pos')) %>%
+  split(., .$gene)
 ```
 
-```{r plot, fig.width = 20}
-# plot positions
-unique(sites$gene) %>%
-  purrr::walk(function(i) {
+## Mutations
+
+First, we plot the rates of reference and non-reference alleles at mutation windows in all genes. Then, we do the same for substitutions only (ignoring indels).
+
+```{r plot}
+# function: create plot list
+plot_positions <- function(reads, include_indels = T) {
   p <-
-      sites %>%
-      dplyr::filter(gene == i) %>%
-      dplyr::left_join(refs, by = c('chr', 'pos')) %>%
-      tidyr::pivot_longer(cols = dplyr::all_of(nucleotides),
-                          values_to = 'nt_n_reads', names_to = 'nt') %>%
+    names(reads) %>%
+    purrr::set_names() %>%
+    purrr::map(function(gene) {
+
+      p_dat <-
+        reads[[gene]] %>%
+        # optionally exclude indels
+        dplyr::filter(include_indels == T |
+                !name %in% c('-', '_', 'INS', 'ins')) %>%
+        # only plot celltypes with any reads
+        dplyr::group_by(celltype) %>%
+        dplyr::filter(max(value) > 0) %>%
+        dplyr::ungroup()
+
+      if (nrow(p_dat) > 0) {
+
+        # get ref alleles for x axis labels
+        x_labs <-
+          p_dat %>%
+          dplyr::distinct(pos, ref) 
+        x_labs_vec <- x_labs$ref
+        names(x_labs_vec) <- x_labs$pos
+
+        g_title <- paste0(
+          gene, ' w/', 
+          ifelse(include_indels == T, '', 'o'), 
+          ' indels')
+
+        # get plot dims
+        g_p_dims <-
+          c(min(max(5, dplyr::n_distinct(p_dat$mut_pos) * 2 + 2), 50),
+            min(max(5, dplyr::n_distinct(p_dat$celltype) * 1.7 + 2), 50))
+
+        # get plot
+        g_p <-
+          p_dat %>%
           dplyr::mutate(
-        strandless_nt = 
-          ifelse(nt %in% c('-', '_'), 'DEL', toupper(nt)),
-        is_ref = ifelse(strandless_nt == ref, 'ref', 'alt'),
-        strandless_nt = paste0(strandless_nt, '_', is_ref) %>% 
-          forcats::fct_relevel(unique(c(paste0(toupper(nucleotides), '_ref'),
-                                        paste0(toupper(nucleotides), '_alt'))))) %>%
-      ggplot(aes(x = pos, y = nt_n_reads, fill = strandless_nt)) + 
-      geom_vline(aes(xintercept = mut_pos)) +
+            # make sure ordering of positions is maintained
+            x_pos = factor(as.character(pos), 
+                           levels = unique(as.character(sort(pos)))),
+            x_mut_pos = dplyr::case_when(pos == mut_pos ~ x_pos, TRUE ~ NA)) %>%
+          # plot
+          ggplot(aes(x = x_pos, y = value, fill = status)) +
           geom_col(position = 'stack') +
-      theme_bw() +
-      theme(axis.text.x = element_blank()) +
-      ggh4x::facet_grid2(celltype ~ mut_pos, scales = 'free') +
-      scale_fill_manual(values = c(A_ref = 'darkgreen', A_alt = 'lightgreen',
-                                   C_ref = 'darkblue', C_alt = 'lightblue',
-                                   G_ref = 'darkorange3', G_alt = 'peachpuff',
-                                   T_ref = 'darkred', T_alt = 'salmon',
-                                   INS_alt = 'black', DEL_alt = 'grey')) +
-      ggtitle(i)
-    print(p)
+          geom_vline(aes(xintercept = x_mut_pos)) +
+          theme_classic() +
+          scale_x_discrete(labels = x_labs_vec) +
+          scale_fill_manual(values = nt_cols) +
+          ggh4x::facet_grid2(celltype ~ mut_pos_lab, scales = 'free') +
+          labs(title = g_title, x = 'pos (ref shown)')
+
+        # return
+        list(plot = g_p, dims = g_p_dims)
+
+      }
     })
+
+  # return
+  p
+}
+
+# generate plot lists
+p_w_indels <- plot_positions(reads)
+p_wo_indels <- plot_positions(reads, F)
 ```
 
-# Mutational calls in drivers from the LCM-WES data
+```{r, results = 'asis'}
+chunks <- c()
 
-```{r mut_hotspots, eval = F}
-# read in bam mappings, get celltype-split bams only
-mappings <-
-  readr::read_csv(params$bam_mappings) %>%
-  dplyr::filter(!is.na(celltype)) %>%
-  # temporarily remove problematic BRI-401 endo bam
-  dplyr::filter(!(id == 'BRI_401' & celltype == 'Endothelial_cell'))
+for (gene in names(reads)) {
 
-# read in drivers
-drivers <-
-  readr::read_tsv(params$drivers, col_names = c('chr', 'start', 'end', 'gene')) %>%
-  dplyr::mutate(chr = gsub('chr', '', chr))
+  chunks <- c(chunks, paste0('\n### ', gene))
 
-# read in mutations in drivers
-hotspots <-
-  readr::read_tsv(params$hotspots) %>%
-  dplyr::mutate(mut_pos = pos,
-                chr = gsub('chr', '', chr)) %>%
-  dplyr::group_by(chr, mut_pos, gene, ref_cod) %>%
-  # summarise each position
-  dplyr::summarise(n = dplyr::n(),
-                   id = paste(id, collapse = ','),
-                   mut_cod = paste(mut_cod, collapse = ','),
-                   impact = paste(impact, collapse = ',')) %>% 
-  dplyr::group_by(dplyr::across(everything())) %>%
-  # extend pileup window 20bp either side of mutated position
-  dplyr::reframe(pos = (mut_pos - 20):(mut_pos + 20))
+  knit_plot <- c(
+    '',
+    paste0('```{r ', gene,'_w_indels, fig.dim = p_w_indels[["', gene, '"]]$dims, warning = F, message = F, echo = F}'),
+    paste0('print(p_w_indels[["', gene, '"]]$plot)'),
+    '```',
+    '',
+    paste0('```{r ', gene,'_wo_indels, fig.dim = p_wo_indels[["', gene, '"]]$dims, warning = F, message = F, echo = F}'),
+    paste0('print(p_wo_indels[["', gene, '"]]$plot)'),
+    '```',
+    ''
+  )
 
-# save as a bed
-hotspots %>%
-  dplyr::transmute(chr = paste0('chr', chr), start = pos - 1, end = pos) %>%
-  readr::write_tsv(
-    'data/driver_genes/lcm_wes_mutations_20bp_window.bed',
-    col_names = F)
+  knitted_plot <- knitr::knit_child(text = knit_plot, envir = environment(), quiet = T)
+  dev_off_if()
 
-# get ref allele at each position
-refs <- xfun::cache_rds({
-  refs <-
-    system('bedtools getfasta -fi /lustre/scratch125/casm/team268im/al28/bed_ref/GRCh38_full_analysis_set_plus_decoy_hla_genome.fa -bed data/driver_genes/lcm_wes_mutations_20bp_window.bed -tab', intern = T) %>%
-    paste(collapse = '\n') %>%
-    readr::read_tsv(col_names = c('pos', 'ref')) %>%
-    tidyr::separate_wider_delim(cols = pos, delim = stringr::regex(':|-'), 
-                                names = c('chr', 'start', 'end')) %>%
-    type.convert(as.is = T) %>%
-    dplyr::transmute(chr = gsub('chr', '', chr), pos = end, ref) %>%
-    dplyr::distinct()
-  refs
-}, file = 'refs.rds', rerun = params$rerun)
+  chunks <- c(chunks, knitted_plot)
+
+}
+
+cat(unlist(chunks), sep = '\n')
 ```
+
+## Mutations types, by patient across genes
+
+```{r muts_by_patient, fig.width = 10}
+reads %>%
+  dplyr::bind_rows() %>%
+  ggplot(aes(x = id, y = value, fill = status)) +
+  geom_col() +
+  scale_fill_manual(values = nt_cols) + 
+  theme_classic() +
+  scale_x_discrete(guide = guide_axis(angle = -90))
+```

### 2024-03-11 scomatic/src/03_knit_driver_coverage.sh

@@ -62,8 +62,9 @@ awk -F',' -v OFS=',' '$3 != "NA" {print}' \
   --mappings ${out_dir}/mappings.csv \
   --window 1 \
   --out_dir ${out_dir} \
-  -c /nfs/team205/kp9/nextflow/scomatic/LSF.config \
-  -w work/ 
+  -c config/LSF.config \
+  -w work/ \
+  -resume
 
 # # run test
 # # head -6 data/Zhang2023/coverage_mappings.csv > data/Zhang2023/coverage_mappings_test.csv

### 2024-03-11 scomatic/reports/targeted_mutation_calling.Rmd
extend windows for targeted_mutation_calling, fix plot dims

@@
 library(ggplot2)
-library(GenomicRanges)
 
@@
   split(., .$gene)
+
+# save reads
+reads %>% 
+  saveRDS('reads.rds')
 ```
@@
 # function: create plot list
-plot_positions <- function(reads, include_indels = T) {
+plot_positions <- function(reads, include_indels = T, include_refs = T) {
   p <-
@@
         # optionally exclude indels
-        dplyr::filter(include_indels == T |
-                !name %in% c('-', '_', 'INS', 'ins')) %>%
+        dplyr::filter(
+          include_indels == T | !status %in% c('DEL', 'INS'),
+          include_refs == T | status != 'ref') %>%
         # only plot celltypes with any reads
@@
         g_title <- paste0(
-          gene, ' w/', 
+          gene, ', w/', 
           ifelse(include_indels == T, '', 'o'), 
-          ' indels')
+          ' indels',
+          ifelse(include_refs == T, '', ', w/o refs'))
 
@@
         g_p_dims <-
-          c(min(max(5, dplyr::n_distinct(p_dat$mut_pos) * 2 + 2), 50),
-            min(max(5, dplyr::n_distinct(p_dat$celltype) * 1.7 + 2), 50))
+          c(min(max(7, (dplyr::n_distinct(p_dat$mut_pos) + 2) * 2), 50),
+            min(max(7, (dplyr::n_distinct(p_dat$celltype)) * 1.5), 50))
 
@@
 p_wo_indels <- plot_positions(reads, F)
+p_wo_indels_wo_refs <- plot_positions(reads, F, F)
 ```
@@
     '```',
+    '',
+    paste0('```{r ', gene,'_wo_indels_wo_refs, fig.dim = p_wo_indels_wo_refs[["', gene, '"]]$dims, warning = F, message = F, echo = F}'),
+    paste0('print(p_wo_indels_wo_refs[["', gene, '"]]$plot)'),
+    '```',
     ''
@@
 
-## Mutations types, by patient across genes
+## Mutation types, by patient across genes
 
-```{r muts_by_patient, fig.width = 10}
+### Mutation types, by patients across all genes
+
+```{r muts_by_patient, fig.width = 10, class.source = 'fold-hide'}
 reads %>%
@@
 ```
+
+### Mutation types, by patient across all genes, without indels
+
+```{r muts_by_patient_no_indels, fig.width = 10, class.source = 'fold-hide'}
+reads %>%
+  dplyr::bind_rows() %>%
+  dplyr::filter(!status %in% c('INS', 'DEL', 'ref')) %>%
+  ggplot(aes(x = id, y = value, fill = status)) +
+  geom_col() +
+  scale_fill_manual(values = nt_cols) + 
+  theme_classic() +
+  scale_x_discrete(guide = guide_axis(angle = -90))
+```

### 2024-03-11 scomatic/src/03_knit_driver_coverage.sh

@@ -60,7 +60,7 @@ awk -F',' -v OFS=',' '$3 != "NA" {print}' \
 # run
 /software/team205/nextflow-23.04.1-all run nextflow/targeted_mutation_calling.nf \
   --mappings ${out_dir}/mappings.csv \
-  --window 1 \
+  --window 5 \
   --out_dir ${out_dir} \
   -c config/LSF.config \
   -w work/ \

### 2024-03-11 infercnv/reports/gains_and_losses.Rmd
add uphyloplot repo

@@
 
-We also can observe the rate coincidence of chr7 and other gains. The LCM-WES data suggested a nested relationship in which there were chr5+ subclones within chr7+ populations. We look for this here.
+### Co-occurence of CNV events
+
+We also can observe the rate co-occurence of chr7+ and other CNVs. The LCM-WES data suggested a nested relationship in which there were chr5+ subclones within chr7+ populations. We look for this here.
 
@@
 
-We also can run a Fisher exact test to test whether chr7+ and chr5+ gains co-occur more often than would be expected by chance, and also test all other co-occurences.
+Here is a contingency table of the frequency of chr7+ versus chr5+ within cells.
 
@@
 ```
+
+We can run a Fisher exact test to test whether chr7+ and chr5+ gains co-occur more often than would be expected by chance, and also test all other co-occurent CNV relationships.
 
@@
 fisher %>%
+  dplyr::mutate(p = format(p, digits = 3)) %>%
   knitr::kable()

### 2024-03-11 scomatic/reports/targeted_mutation_calling.Rmd
add uphyloplot repo

@@
 # recursites (to add detected mut)
+recursites <-
+  readr::read_tsv('recursites.tsv') 
 recurmuts <-
-  readr::read_tsv('recursites.tsv') %>%
+  recursites %>%
   dplyr::transmute(
@@
   saveRDS('reads.rds')
+```
+
+## Recurrent mutations sites
+
+Recurrent mutations from the LCM-WES data were found using `dNdScv`. Mutated sites in the following genes will be interrogated.
+
+```{r plot_recurmuts, fig.width = 10}
+recursites %>%
+  dplyr::group_by(gene) %>%
+  dplyr::mutate(sum_freq = sum(freq)) %>%
+  ggplot(aes(x = reorder(gene, -sum_freq), y = freq, fill = qval)) +
+  geom_col(position = 'stack') +
+  scale_x_discrete(guide = guide_axis(angle = -90)) +
+  labs(x = 'gene') +
+  theme_classic()
 ```

### 2024-03-12 scomatic/reports/targeted_mutation_calling.Rmd
create separated targeted_mutation_calling step, move dndscv run to pre-processing

--- 2024-03-11 scomatic/reports/targeted_mutation_calling.Rmd
@@
 params:
+  recursites: NULL
   cache_dir: './target_mutation_calling_cache/'
@@
 # recursites (to add detected mut)
+recursites <-
+  readr::read_tsv(params$recursites) 
 recurmuts <-
-  readr::read_tsv('recursites.tsv') %>%
+  recursites %>%
   dplyr::transmute(
@@
   saveRDS('reads.rds')
+```
+
+## Recurrent mutations sites
+
+Recurrent mutations from the LCM-WES data were found using `dNdScv`. Mutated sites in the following genes will be interrogated.
+
+```{r plot_recurmuts, fig.width = 10, eval = F}
+recursites %>%
+  dplyr::group_by(gene) %>%
+  dplyr::mutate(sum_freq = sum(freq)) %>%
+  ggplot(aes(x = reorder(gene, -sum_freq), y = freq, fill = qval)) +
+  geom_col(position = 'stack') +
+  scale_x_discrete(guide = guide_axis(angle = -90)) +
+  labs(x = 'gene') +
+  theme_classic()
 ```

### 2024-03-12 scomatic/src/03_knit_driver_coverage.sh

@@ -9,62 +9,40 @@ cd $wd
 . ~/.bashrc
 mamba activate jupy
 
-# # create mappings
-# mappings=data/Zhang2023/mappings.csv
-# bams_dir=/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/cellranger_output/
-# ct_bams_dir=/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/out/Zhang2023/coverage/
-# (
-#   echo 'id,bam,celltype' ;
-#   while read -r id ; do
-#     # check if ct bams directory exists
-#     if [ -d $ct_bams_dir/$id/ ] ; then
-#       ct_bams=($(ls $ct_bams_dir/$id/celltype_bams/*.bam))
-#       # check if scomatic successfully ran
-#       if [ ${#ct_bams[*]} -gt 0 ] ; then
-#         echo "$id,$bams_dir/${id/_/-}/possorted_genome_bam.bam,NA" ;
-#         for file in ${ct_bams[@]} ; do
-#           ct=$(basename $file | cut -d. -f2)
-#           echo "$id,$file,$ct" ;
-#         done
-#       fi
-#     fi
-#   done < \
-#   <(sed 1d $mappings | cut -d, -f2 | sort -u)
-# ) | cat > out/Zhang2023/coverage/mappings.csv
-# 
-# # run
-# /software/team205/nextflow-23.04.1-all run nextflow/driver_coverage.nf \
-#     --mappings ${wd}out/Zhang2023/coverage/mappings.csv \
-#     --drivers ${wd}data/driver_genes/driver_gene_coords_for_coverage.tsv \
-#     --mutations ${wd}data/driver_genes/lcm_wes_mutations.tsv \
-#     --out_dir ${wd}out/Zhang2023/coverage/ \
-#     --min_MQ 0 \
-#     -c /nfs/team205/kp9/nextflow/scomatic/LSF.config  \
-#     --location local \
-#     -w test/work/ \
-#     -resume
-
-# # knit target_mutation_calling.Rmd
-# mkdir out/Zhang2023/coverage/summary/
[…]

### 2024-03-12 scomatic/src/04_get_targeted_mutation_sites.R

@@ -1,0 +1,132 @@
+#!/usr/bin/env Rscript
+
+#libraries
+library(magrittr)
+library(GenomicRanges)
+
+# directory
+wd <- 
+  ifelse(Sys.info()['nodename'] == "mib119104s", '~/Volumes/', '') |> 
+  paste0('/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/')
+setwd(wd)
+
+# reference files
+fasta <- '/lustre/scratch125/casm/team268im/al28/bed_ref/GRCh38_full_analysis_set_plus_decoy_hla_genome.fa'
+hg19_to_hg38 <- '/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/data/liftOver/hg19ToHg38.over.chain'
+
+# recursites from tnanoseq ----
+
+# files
+tnanoseq <- '/lustre/scratch125/casm/team268im/lh22/synovium_nanoseq/ex_t_dndsout/dnds_nano_muts.tsv'
+tgenes <- 'data/driver_genes/Sanger_TERT-v4_TE-95148282_hg19_highstringencyfilter_buccal_gene_list.tsv'
+
+# read targeted nanoseq 
+tnanoseq <- readr::read_tsv(tnanoseq)
+
+# read genes of interest
+target_genes <- readr::read_tsv(tgenes)
+
+# run dndscv
+dnds <-
+  dndscv::dndscv(
+    unique(tnanoseq[, 1:5]),
+    max_muts_per_gene_per_sample = Inf,
+    max_coding_muts_per_sample = Inf,
+    gene_list = target_genes$gene,
+    outmats = T)
+
+# run sitednds
+sitednds <-
+  dndscv::sitednds(dnds)
+
+# get the significant genes
+signif_genes <-
+  dnds$sel_cv %>%
+  tibble::as_tibble() %>%
+  dplyr::filter(qglobal_cv < 0.1)
+
+# get the sites with positively selected mutations
+recursites <-
+  sitednds$recursites %>%
+  tibble::as_tibble() %>%
+  # fix 1d table columns -> numeric columns
+  dplyr::mutate(dplyr::across(where(is.table), as.numeric))
+
+# convert to granges (1-based!)
+gr <- GRanges(
+  seqnames = paste0('chr', recursites$chr),
+  ranges = IRanges(start = recursites$pos, end = recursites$pos),
+  gene = recursites$gene,
[…]

### 2024-03-12 scomatic/src/05_run_targeted_mutation_calling.sh

@@ -1,0 +1,36 @@
+#!/bin/bash
+# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/ ; . ~/.bashrc ; mamba activate jupy ; /software/team205/bin/jsub lsf -q week -n targeted_mut_calling -c 1 -m 8g -l log "bash src/05_run_targeted_mutation_calling.sh" | bsub
+
+# wd
+wd=/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/
+out_dir=${wd}out/Zhang2023/targeted_mutation_calling/
+mkdir -p $out_dir
+cd $wd
+
+# conda activate an environment which has docopt installed so jsub works
+. ~/.bashrc
+mamba activate jupy
+
+# get mappings of celltype-specific bams only
+cat ${wd}out/Zhang2023/coverage/mappings.csv | 
+awk -F',' -v OFS=',' '$3 != "NA" {print}' \
+> ${out_dir}/mappings.csv
+
+# run target_mutation_calling.nf on celltype bams only
+/software/team205/nextflow-23.04.1-all run nextflow/targeted_mutation_calling.nf \
+  --mappings ${out_dir}/mappings.csv \
+  --window 5 \
+  --out_dir ${out_dir} \
+  -c config/LSF.config \
+  -w work/ \
+  -resume
+
+# run on TNFRSF14 nonsense mutations
+/software/team205/nextflow-23.04.1-all run nextflow/targeted_mutation_calling.nf \
+  --mappings ${out_dir}/mappings.csv \
+  --recursites $wd/data/driver_genes/TNFRSF14_withingene_sites_GRCh38.tsv \
+  --window 5 \
+  --out_dir ${out_dir}/TNFRSF14_nonsense/ \
+  -c config/LSF.config \
+  -w work/ \
+  -resume

### 2024-03-12 scomatic/src/withingenednds.R

@@ -1,0 +1,350 @@
+#' withingenednds
+#' 
+#' This function uses Poisson and Negative Binomial regression models at single-site level to study selection across different regions (coding and non-coding) within a gene.
+#' 
+#' @author Inigo Martincorena (Wellcome Sanger Institute)
+#' @details Martincorena I, et al. (2017) Universal patterns of selection in cancer and somatic tissues. Cell. 171(5):1029-1041.
+#' 
+#' @param mutations Data frame with all the mutations detected in the study (5-column input table as for dndscv: sampleID, chr, pos, ref, mut).
+#' @param gene Name of the gene of interest. This function is currently designed to work on a single gene, but combined analyses of multiple genes could be done using the sites output table generated by this function.
+#' @param covtable Table with all sites of interest in the gene. This should be a data frame with one row per site and the following columns: chr, pos, dc (duplex depth). Additional columns will not be used.
+#' @param dndsout dndscv output object for the dataset. This is mainly used for the MLEs of the substitution model. Running dndscv on all genes in the dataset is recommended unless the gene of interest is believed to have a different substitution model.
+#' @param genomeFile Path to a reference fasta file for the genome assembly.
+#' @param regionschr Optional data frame with user-defined regions of interest in the gene. This allows the user to define arbitrary regions within a gene (coding or non-coding) from which to calculate omega (selection or obs/exp) values (e.g. protein domains, splicing regulator regions, core promoters, etc). The table should contain the following columns: chr, start, end, wname (a unique name for the w parameter, e.g. wdomain1, wcorepromoter), impacts (e.g. Missense or Missense|Nonsense will restrict the w calculation with Missense or Missense and Nonsense mutations in the region, respectively), layered (1/0; using "0" removes other w parameters influencing the site, whereas using "1" models selection as relative to other w parameters active at these sites).
[…]

### 2024-03-13 infercnv/reports/gains_and_losses.Rmd
add GLM and plotting to gains_and_losses notebook, add metadata

@@
   infercnv_dir: NULL
+  metadata: NULL
   cache_dir: NULL
@@
   ggplot(aes(x = events)) +
-  geom_text(aes(y = n + 50,  label = label), 
+  geom_text(aes(y = n,  label = label), 
             vjust = 0.5, hjust = 0, angle = 90, size = 2) +
@@
 ```
+
+#### Fisher exact test
 
@@
 
+        # get contingencies
+        cont <-
+          table(sce@colData[, c('chr7_gain', event)])
+
          # return relevant stats
@@
           OR = fisher$estimate, 
-          p = fisher$p.value)
+          p = fisher$p.value,
+          sig = ifelse(fisher$p.value < 0.05, '*', ''),
+          n_WT = cont[1, 1],
+          n_chr7_gain = cont[2, 1],
+          n_event = cont[1, 2],
+          n_both = cont[2, 2])
 
@@
         # no event
-        tibble::tibble(OR = NA, p = NA)
-
+        tibble::tibble()
       }
@@
     # remove chr7+ (can't do chr7+ vs chr7+)
-    dplyr::filter(event != 'chr7_gain') %>%
+    dplyr::filter(event != 'chr7_gain', n_both > 0) %>%
     dplyr::arrange(-OR)
@@
   knitr::kable()
+```
+
+#### GLM
+
+We can also run a GLM.
+
+```{r glm}
+# read in metadata
+metadata <-
+  readr::read_tsv(params$metadata)
+
+# create glm
+# prop_chr7_gain ~ age (+ coverage) (per donor)
+for_glm <-
+  dplyr::inner_join(
+    tibble::as_tibble(sce@colData), 
+    metadata, 
+    by = c('sample' = 'id')) %>%
+  dplyr::filter(!is.na(age), !is.na(chr7_gain)) %>%
+  dplyr::group_by(sample, age, CTAP, sex, race, ethnicity) %>%
+  dplyr::summarise(n_chr7_gain = sum(chr7_gain),
+                   n_cells = dplyr::n(),
+                   ch7_gain_prop = n_chr7_gain / n_cells)
+
+for_glm %>%
+  {glm(.$n_chr7_gain ~ log(.$n_cells) + .$age)}
+```
+
+```{r plot_glm_vars}
+# n cells vs n chr7 gain
+for_glm %>%
+  ggplot(aes(x = n_cells, y = n_chr7_gain)) +
+  geom_point() +
+  theme_classic()
+
+# age vs prop chr7 gain
+for_glm %>%
+  ggplot(aes(x = age, y = ch7_gain_prop)) +
+  geom_point() +
+  theme_classic()
+
+# all categorical variables
+for_glm %>%
+  tidyr::pivot_longer(cols = c('CTAP', 'sex', 'race', 'ethnicity', 'subcluster')) %>%
+  dplyr::group_by(name, value) %>%
+  dplyr::mutate(med = median(ch7_gain_prop)) %>%
+  ggplot(aes(
+    x = tidytext::reorder_within(value, by = med, within = name), 
+    y = ch7_gain_prop)) +
+  geom_boxplot() +
+  geom_jitter(height = 0) +
+  facet_wrap(~ name, nrow = 1, scales = 'free') +
+  tidytext::scale_x_reordered(guide = guide_axis(angle = -45)) +
+  theme_classic() +
+  labs(x = '')
+
+# looking by cluster
+for_glm_w_cluster <-
+  dplyr::inner_join(
+    tibble::as_tibble(sce@colData), 
+    metadata, 
+    by = c('sample' = 'id')) %>%
+  dplyr::filter(!is.na(age), !is.na(chr7_gain)) %>%
+  dplyr::group_by(
+    sample, age, CTAP, sex, race, ethnicity, cluster_name) %>%
+  dplyr::summarise(n_chr7_gain = sum(chr7_gain),
+                   n_cells = dplyr::n(),
+                   ch7_gain_prop = n_chr7_gain / n_cells) 
+for_glm_w_cluster %>%
+  ggplot(aes(x = cluster_name, y = ch7_gain_prop)) +
+  geom_violin() +
+  geom_jitter(height = 0) +
+  theme_classic() +
+  scale_x_discrete(guide = guide_axis(angle = -45))
 ```

### 2024-03-13 infercnv/src/00_preprocess.R

@@ -12,6 +12,16 @@ dat_dir <- '/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal202
 cellranger_dir <- paste0(dat_dir, 'cellranger_output/')
 processed_dir <- paste0(dat_dir, 'processed_output/')
 
+# clinical metadata
+clinical <- 
+  readr::read_csv('data/Zhang2023/syn26710600/AMP-RA.SLE_clinical.csv') %>%
+  dplyr::rename(id2 = individualID) %>%
+  dplyr::mutate(age = as.numeric(ageEnrollment))
+readr::read_tsv('data/Zhang2023/syn26710600/CTAP_donor_mapping.tsv') %>%
+  dplyr::rename(id = donor, id2 = subject_id) %>%
+  dplyr::left_join(clinical) %>%
+  readr::write_tsv('data/Zhang2023/clinical_metadata.tsv')
+
 # annotations TSVs (cell celltype) ----
 readRDS(paste0(processed_dir, 'all_cells_reference.rds'))$meta_data %>%
   dplyr::select(cell, 

### 2024-03-13 infercnv/src/03_knit_gains_and_losses.sh

@@ -1,5 +1,5 @@
 #!/bin/bash
-# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/ ; . ~/.bashrc ; mamba activate jupy ; /software/team205/bin/jsub lsf -q week -n knit_gains_and_losses -c 1 -m 100g -l log "bash src/03_knit_gains_and_losses.sh" | bsub
+# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/ ; . ~/.bashrc ; mamba activate jupy ; /software/team205/bin/jsub lsf -q normal -n knit_gains_and_losses -c 1 -m 100g -l log "bash src/03_knit_gains_and_losses.sh" | bsub
 
 # wd
 wd=/lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/
@@ -11,7 +11,7 @@ mamba activate jupy
 
 # knit by_fibroblast_cluster, < 0.5 | > 0.8
 mkdir -p out/Zhang2023/by_fibroblast_cluster/summary/
-Rscript -e "rmarkdown::render('reports/gains_and_losses.Rmd', params = list(sce = 'data/Zhang2023/sce/Fibroblast_sce.rds', cell_group_name = 'cluster_name', infercnv_dir = 'out/Zhang2023/by_stromal_cluster/', cache_dir = 'out/Zhang2023/by_fibroblast_cluster/summary/gains_and_losses_cache/', rerun = F), output_file = 'gains_and_losses.html', output_dir = 'out/Zhang2023/by_fibroblast_cluster/summary/')"
+Rscript -e "rmarkdown::render('reports/gains_and_losses.Rmd', params = list(sce = 'data/Zhang2023/sce/Fibroblast_sce.rds', cell_group_name = 'cluster_name', metadata = 'data/Zhang2023/clinical_metadata.tsv', infercnv_dir = 'out/Zhang2023/by_stromal_cluster/', cache_dir = 'out/Zhang2023/by_fibroblast_cluster/summary/gains_and_losses_cache/', rerun = F), output_file = 'gains_and_losses.html', output_dir = 'out/Zhang2023/by_fibroblast_cluster/summary/')"
 
 # # knit by_fibroblast_cluster, < 0.2 | > 0.8
 # mkdir -p out/Zhang2023/by_fibroblast_cluster_0.2/summary/

### 2024-03-13 scomatic/src/04_get_targeted_mutation_sites.R
interrogate all nonsynonymous mutations in TNFRSF14

@@ -90,7 +90,7 @@ source('src/withingenednds.R')
 tnfrsf14 <-
   readr::read_tsv('data/driver_genes/TNFRSF14_withingene_sites.tsv') %>%
   # get all possible nonsense mutations
-  dplyr::filter(impact == 'Nonsense')
+  dplyr::filter(!is.na(impact), impact != 'Synonymous')
 
 # lift over GRCh37 -> GRCh38
 # convert to granges (1-based!)

### 2024-03-13 scomatic/src/05_run_targeted_mutation_calling.sh

@@ -16,21 +16,22 @@ cat ${wd}out/Zhang2023/coverage/mappings.csv |
 awk -F',' -v OFS=',' '$3 != "NA" {print}' \
 > ${out_dir}/mappings.csv
 
-# run target_mutation_calling.nf on celltype bams only
-/software/team205/nextflow-23.04.1-all run nextflow/targeted_mutation_calling.nf \
-  --mappings ${out_dir}/mappings.csv \
-  --window 5 \
-  --out_dir ${out_dir} \
-  -c config/LSF.config \
-  -w work/ \
-  -resume
+# # run target_mutation_calling.nf on celltype bams only
+# /software/team205/nextflow-23.04.1-all run nextflow/targeted_mutation_calling.nf \
+#   --mappings ${out_dir}/mappings.csv \
+#   --window 5 \
+#   --out_dir ${out_dir} \
+#   -c config/LSF.config \
+#   -w work/ \
+#   -resume
 
-# run on TNFRSF14 nonsense mutations
+# run on all impactful TNFRSF14 mutations
 /software/team205/nextflow-23.04.1-all run nextflow/targeted_mutation_calling.nf \
   --mappings ${out_dir}/mappings.csv \
   --recursites $wd/data/driver_genes/TNFRSF14_withingene_sites_GRCh38.tsv \
-  --window 5 \
-  --out_dir ${out_dir}/TNFRSF14_nonsense/ \
+  --window 1 \
+  --out_dir ${out_dir}/TNFRSF14/ \
   -c config/LSF.config \
-  -w work/ \
-  -resume
+  -w work/ 
+  
+  

### 2024-03-13 scomatic/src/06_plot_TNFRSF14_nonsynonymous_mutations.R

@@ -1,0 +1,150 @@
+#!/usr/bin/env Rscript
+
+# wd
+wd <- 
+  ifelse(Sys.info()['nodename'] == "mib119104s", '~/Volumes/', '') |> 
+  paste0('/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/')
+setwd(wd)
+
+#libraries
+library(magrittr)
+library(GenomicFeatures)
+
+# get recursites in TNFRSF14
+recursites <-
+  readr::read_tsv('data/driver_genes/TNFRSF14_withingene_sites_GRCh38.tsv')
+
+# get all reads
+reads <-
+  readRDS('out/Zhang2023/targeted_mutation_calling/TNFRSF14/summary/reads.rds')
+
+# get exon coords
+exons <- 
+  readRDS('data/dndscv/ref_cds.rds')$TNFRSF14 %>%
+  {tibble::tibble(
+    chr = as.numeric(.$chr),
+    start = .$intervals_cds[, 1],
+    end = .$intervals_cds[, 2],
+    gene = 'TNFRSF14') %>%
+  dplyr::mutate(feature = paste0('exon', dplyr::row_number()))}
+introns <-
+  gene_spans %>%
+  dplyr::group_by(chr, gene) %>%
+  dplyr::transmute(
+    new_end = start - 1, start = dplyr::lag(end) + 1) %>%
+  dplyr::filter(!is.na(start)) %>%
+  dplyr::transmute(
+    start, end = new_end,
+    feature = paste0('intron', dplyr::row_number()))
+
+# get all features
+features <-
+  dplyr::bind_rows(exons, introns) %>%
+  dplyr::arrange(start) %>%
+  dplyr::mutate(feature = forcats::fct_reorder(feature, dplyr::row_number())) 
+features_all_positions <-
+  features %>%
+  dplyr::mutate(pos = purrr::map2(start, end, ~ .x:.y)) %>%
+  tidyr::unnest(cols = 'pos') %>%
+  dplyr::select(-start, -end)
+
+# alts - colours and order
+nt_cols <-
+  c('INS' = '#e6ab02',
+    'DEL' = '#1c2d78',
+    'A' = '#1b9e77', 
+    'C' = '#d95f02', 
+    'G' = '#7570b3', 
+    'T' = '#e7298a',
+    'ref' = 'grey')
+
+# impacts - colours and order
+imp_cols <-
[…]

### 2024-07-02 infercnv/reports/paper_figures.Rmd
fix infercnv pipeline logic, begin plotting paper figures

--- 2024-03-13 infercnv/reports/gains_and_losses.Rmd
@@
 ---
-title: "Gains and losses in rheumatoid arthritis data from inferCNV"
+title: "Creating figures for the paper"
 author: "Alexandra Tidd"
@@
 ```{r setup, include = F, message = F, warning = F, class.source = 'fold-hide'}
-# wd <- ifelse(Sys.info()['nodename'] == "mib119104s", '~/Volumes/', '') |> paste0('/lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/') ; setwd(wd)
-# rmarkdown::render('reports/infercnv.Rmd', output_file = 'driver_coverage.html', output_dir = 'reports/', params = list(rerun = T))
+# module load R/4.4.0 ; Rscript -e "rmarkdown::render('reports/paper_figures.Rmd', output_dir = 'out/paper_figures/')" # nolint: line_length_linter.
 
@@
 library(ggplot2)
-library(gridExtra)
-library(Matrix)
 library(SummarizedExperiment)
 library(SingleCellExperiment)
-library(glmGamPoi)
-
-# dev.off() if there is a device
-dev_off_if <- function() {if (length(dev.list()) != 0) { dev.off() }}
 
 # chunk options
-knitr::opts_knit$set(root.dir = '../')
-knitr::opts_chunk$set(
-  warning = FALSE, 
-  dpi = 300, 
-  message = FALSE,
-  cache.path = params$cache_dir) 
+knitr::opts_knit$set(root.dir = "../")
+knitr::opts_chunk$set(warning = FALSE, dpi = 300, message = FALSE)
 ```
@@
 
-inferCNV was run on the scRNA-seq data to call copy number variants in synovial cells. No reference group was used, so the average profile of all cells was used in each patient as the baseline.
+inferCNV was run on the scRNA-seq data to call copy number variants in synovial 
+cells. No reference group was used, so the average profile of all cells was used 
+in each patient as the baseline.
 
-First, we load the CITE-seq object. We convert it to a SingleCellExperiment object. Then we add the CNV metadata from the inferCNV run to the object.
+First, we load the CITE-seq object. We convert it to a SingleCellExperiment 
+object. Then we add the CNV metadata from the inferCNV run to the object.
 
 ```{r load_sce, results = 'asis', class.source = 'fold-show'}
-# load gene coords
-coords <-
-  readr::read_tsv('data/hg38_gencode_v27.txt', col_names = c('name', 'chr', 'start', 'stop'))
-
-# load CTAP and metadata
-ctap <-
-  readr::read_tsv('data/Zhang2023/syn26710600/CTAP_donor_mapping.tsv') %>%
-  dplyr::select(sample = donor, CTAP)
-md <- 
-  readr::read_tsv('data/Zhang2023/syn26710600/SYNAPSE_METADATA_MANIFEST.tsv') %>%
-  dplyr::distinct()
-
-# load sce, add inferCNV 
-sce <- xfun::cache_rds({
-
-  # load sce
-  sce <- readRDS(params$sce)
-
-  # get cnv metadata
-  cnvs <- 
-    list.files(
-      params$infercnv_dir,
-      pattern = 'map_metadata_from_infercnv.txt',
-      recursive = T) %>%
-    purrr::set_names(., dirname(.)) %>%
-    purrr::map(function(file) {
-      file.path(params$infercnv_dir, file) %>% 
-        read.table(sep = '\t')
-      }) %>%
-    dplyr::bind_rows() %>%
-    dplyr::mutate(infercnv = T)
-
-  # left join cnvs to coldata, reorder, restore rownames
-  col_data <- merge(sce@colData, cnvs, by = 'row.names', all.x = T, sort = F)
-  rownames(col_data) <- col_data[, 1]
-  col_data <- 
-    col_data[colnames(sce), -1] %>%
-    DataFrame()
-
-  # add to sce
-  sce@colData <- col_data
-
-  # return
-  sce
-
-}, file = 'sce.rds', rerun = params$rerun)
-
-# remove NA's for inferCNV analysis
-sce <- sce[, !is.na(sce$proportion_cnv_chr1)]
-
- # assign cell group
-sce@colData$cell_group <-
-  sce@colData[, params$cell_group_name]
+# load sce with gains
+sce <-
+  "out/Zhang2023/by_fibroblast_cluster/summary/gains_and_losses_cache/" %>%
+  list.files(pattern = "sce_w_gains", full.names = TRUE) %>%
+  readRDS()
 ```
 
-The `inferCNV` output file `map_metadata_from_infercnv.txt` contains columns with different CNV measures at whole-chromosome level per cell, with rownames denoting cells.
-The columns have the following values...
+# Plot distribution of number of fibroblasts per patient to decide cut-off
 
-```{r echo = F, results = 'asis'}
-tibble::tribble(
-  ~name, ~`type - value`,
-  '`has_*`', 'binary - whether any CNV / loss / duplication is found within the given chr',
-  '`proportion_*`', 'proportion - proportion of genes that are part of any CNV / loss / duplication within the given chr',
-  '`proportion_scaled_*`', 'proportion - same as proportion_*, but taking into account whether genes that are duplicated / lost are a single copy or two copies, as a weight',
-  '`top_loss_*`', 'the top CNVs that affect the most genes'
-) %>%
+```{r plot_fibroblast_dist}
+n_fibroblasts <-
+  sce@colData %>%
+  tibble::as_tibble() %>%
+  dplyr::count(sample)
+n_fibroblasts %>%
+  ggplot(aes(x = n)) +
+  geom_histogram(binwidth = 20)
+```
+
+I will introduce a cut-off of 100 fibroblasts per patient.
+
+```{r cut_off}
+n_fibroblasts %>%
+  dplyr::summarise(`n >= 100` = sum(n >= 100),
+                   `n < 100` = sum(n < 100),
+                   n = dplyr::n()) %>%
   knitr::kable()
@@
 
-## HMM regional predictions
+This leaves 52 / 80 of the samples.
 
-We also load the CNV regions prediced by HMM. According to [their wiki](https://github.com/broadinstitute/infercnv/wiki/infercnv-i6-HMM-type#infercnv-i6-hmm)...
+# Plot distribution of chr7+ and chr5+ rates
 
-As well as aneuploidy, we want to look for smaller recurrent structural changes, including focal-/arm-level events.
+```{r}
+events <-
+  sce@colData %>%
+  tibble::as_tibble() %>%
+  dplyr::select(sample, cell, cell_group, chr7_gain, chr5_gain) %>%
+  tidyr::pivot_longer(-c("sample", "cell", "cell_group")) %>%
+  dplyr::group_by(sample, cell, cell_group) %>%
+  # remove cells in which no calls made
+  dplyr::filter(!all(is.na(value))) %>%
+  dplyr::mutate(event = dplyr::case_when(all(is.na(value) | value == 0) ~ "WT",
+                                         value == 1 ~ name, TRUE ~ NA)) %>%
+  dplyr::summarise(
+    events = paste(sort(unique(na.omit(event))), collapse = ","),
+    n_events = sum(value, na.rm = TRUE)) %>%
+  # get props and counts
+  dplyr::ungroup() %>%
+  dplyr::add_count(sample, name = "n_cells_per_sample") %>%
+  dplyr::add_count(sample, events, name = "n_cells_per_sample_x_event") %>%
+  dplyr::group_by(sample, n_cells_per_sample_x_event, n_cells_per_sample, events) %>%
+  dplyr::summarise(prop_cells_per_sample_x_event = n_cells_per_sample_x_event / n_cells_per_sample) %>%
+  dplyr::distinct() %>%
+  dplyr::mutate(sample = forcats::fct_reorder(sample, prop_cells_per_sample_x_event)) %>%
+  tidyr::pivot_longer(c("n_cells_per_sample_x_event", "prop_cells_per_sample_x_event"))
 
-> The inferCNV i6 HMM is a six-state CNV model that predicts the following CNV levels:
->
-> * State 1: 0x: complete loss
-> * State 2: 0.5x: loss of one copy
-> * State 3: 1x: neutral
-> * State 4: 1.5x: addition of one copy
-> * State 5: 2x: addition of two copies
-> * State 6: 3x: essentially a placeholder for >2x copies but modeled as 3x.
+# get ordering by total non-WT prop
 
-```{r get_hmm}
-cnv_regions <- xfun::cache_rds({
-
-  # get HMM predictions
-  cnv_regions <- 
-    list.files(
-      params$infercnv_dir,
-      pattern = 'HMM_CNV_predictions.HMMi6.leiden.hmm_mode-subclusters.Pnorm_0.5.pred_cnv_regions.dat',
-      recursive = T) %>%
-    purrr::set_names(., dirname(.)) %>%
-    purrr::map(function(file) {
-      file.path(params$infercnv_dir, file) %>% 
-        readr::read_tsv(show_col_types = F)
-      }) %>%
-    dplyr::bind_rows(.id = 'sample') %>%
-    # refactor chromosomes
-    dplyr::mutate(
-      chr = chr %>% {forcats::fct_reorder(., gsub('chr', '', .) %>% as.numeric())},
-      direction = dplyr::case_when(
-        state < 3 ~ 'loss',
-        state > 3 ~ 'dupli',
-        TRUE ~ 'neutral'),
-      ploidy = dplyr::case_when(
-        state == 1 ~ '0x',
-        state == 2 ~ '0.5x',
-        state == 3 ~ '1x',
-        state == 4 ~ '1.5x',
-        state == 5 ~ '2x',
-        state == 6 ~ '>2x',
-        TRUE ~ NA
-        ) %>% forcats::fct_reorder(state),
-      region = gsub('.*region_', '', cnv_name) %>% as.numeric(),
-      cell_group = cell_group_name %>% gsub('\\..*', '', .))
-
-  # calculate region midpoints for plotting
-  cnv_regions$midpoint <-
-    (cnv_regions$start + cnv_regions$end) / 2
-
-  # return
-  cnv_regions
-
-}, file = 'cnv_regions.rds', rerun = params$rerun)
-```
-
-```{r plot_cnv_regions_pred, fig.height = 20, fig.width = 10}
-cnv_regions %>%
-  ggplot(aes(x = midpoint, fill = ploidy, group = ploidy)) +
-  geom_density(data = cnv_regions %>% dplyr::filter(direction == 'dupli'),
-               aes(y = ..density..), 
-               position = position_stack(), 
-               colour = NA) +
-  geom_density(data = cnv_regions %>% dplyr::filter(direction == 'loss'),
-               aes(y = -..density..), 
-               position = position_stack(reverse = T), 
-               colour = NA) +
-  ggh4x::facet_grid2(chr ~ cell_group, scales = 'free_x', independent = 'x') +
-  theme_bw() +
-  theme(axis.text.x = element_blank()) +
-  scale_fill_manual(
-    values = c('0x' = '#00008B', '0.5x' = '#b4b0ff', '1' = 'white', 
-               '1.5x' = '#ffc2c2', '2x' = '#d64242', '>2x' = '#8B0000')) 
+pdf("test.pdf")
+events %>%
+  ggplot(aes(x = sample, y = value, fill = events)) +
+  geom_col() +
+  scale_x_discrete(guide = guide_axis(angle = -90)) +
+  facet_wrap(~ name, scale = "free_y", ncol = 1)
+dev.off()
 ```
@@
 
-```{r ct_comp, fig.width = 20, fig.height = 7}
+```{r ct_comp, fig.width = 12, fig.height = 7}
 p_dat <-
@@
     sample = forcats::fct_infreq(sample),
-    cell_group = forcats::fct_infreq(cell_group)) %>%
-  dplyr::add_count(sample, cell_group, name = 'n_sample_x_cell_group')
+    cell_group = forcats::fct_rev(forcats::fct_infreq(cell_group))) %>%
+  dplyr::add_count(sample, cell_group, name = 'n_sample_x_cell_group') %>%
+  dplyr::add_count(cell_group, name = 'n_cell_group')
 p_base <-
   p_dat %>%
-  ggplot(aes(fill = cell_group, colour = cell_group)) 
+  ggplot(aes(fill = cell_group, colour = cell_group)) +
+  scale_fill_manual(values = dittoSeq::dittoColors()) +
+  scale_colour_manual(values = dittoSeq::dittoColors())
 p1 <- 
@@
   theme_classic() +
-  theme(legend.position = 'none') +
-  scale_x_discrete(guide = guide_axis(angle = -90))
+  theme(legend.position = 'none',
+        axis.text.x = element_text(size = 7)) +
+  scale_x_discrete(guide = guide_axis(angle = -90)) +
+  labs(y = params$cell_group_name)
 p2 <- 
@@
   theme_void() +
-  theme(legend.position = 'none') 
+  theme(legend.position = 'none',
+        axis.text.y = element_text()) 
 p3 <-
@@
   geom_bar(aes(y = cell_group), position = 'stack') +
+  geom_text(
+    data = p_dat %>% dplyr::distinct(cell_group, n_cell_group),
+    aes(x = n_cell_group, y = cell_group, 
+        label = format(n_cell_group, nsmall = 1, big.mark = ",")), 
+    hjust = -0.1) +
   theme_void() +
-  theme(legend.position = 'none')
+  theme(legend.position = 'none') +
+  xlim(0, 1.2 * max(p_dat$n_cell_group))
 design <-
 'BB#
+BB#
 AAC'
+
+# plot
 patchwork::wrap_plots(A = p1, B = p2, C = p3, design = design)
@@
   coord_fixed()
+
+pal <-
+  c('T cell' = "#E6AB02",
+    'Stromal cell' = "#66A61E",
+    'Myeloid cell' = "#E7298A",
+    'B cell/plasma cell' = "#7570B3",
+    'Endothelial cell' = "#D95F02",
+    'NK' = "#1B9E77")
+dittoSeq::dittoDimPlot(sce, 'cell_type', size = 0.2) +
+  coord_fixed() +
+  scale_colour_manual(values = pal)
 ```
@@
 
-## Assigning whole chromosome-level events
+# Assigning whole chromosome-level events
 
@@
 
-### Co-occurence of CNV events
+```{r plot_heatmap, fig.width = 10, fig.height = 5}
+event_cols <-
+  c('WT' = 'grey',
+    'chr5_gain' = '#e7298a',
+    'chr5_gain,chr7_gain' = '#d95f02',
+    'chr7_gain' = '#e6ab02')
+
+p_dat <-
+  sce@colData %>%
+  tibble::as_tibble() %>%
+  dplyr::select(sample, cell, 
+                dplyr::ends_with('_gain'), 
+                dplyr::ends_with('_loss')) %>%
+  tidyr::pivot_longer(-c('sample', 'cell')) %>%
+  dplyr::filter(name %in% c('chr7_gain', 'chr5_gain')) %>%
+  dplyr::group_by(sample, cell) %>%
+  # remove cells in which no calls made
+  dplyr::filter(!all(is.na(value))) %>%
+  dplyr::mutate(
+    event = dplyr::case_when(
+      all(value == 0 | is.na(value)) ~ 'WT',
+      value == 1 ~ name,
+      TRUE ~ NA)) %>%
+  dplyr::summarise(
+    events = paste(sort(unique(na.omit(event))), collapse = ',')) %>%
+  dplyr::mutate(events = forcats::fct_relevel(events, names(event_cols))) %>%
+  # get n aneuploid for plot ordering
+  dplyr::group_by(sample) %>%
+  dplyr::mutate(prop_aneuploid = sum(events != 'WT') / dplyr::n()) %>%
+  # count
+  dplyr::group_by(sample, prop_aneuploid) %>%
+  dplyr::count(events) %>%
+  # for fill and stack
+  tidyr::crossing(tibble::tibble(plot = c('% cells', 'n cells')))
+
+# plot
+p_dat %>%
+  ggplot(aes(x = reorder(sample, -prop_aneuploid), y = n, fill = events)) +
+  geom_col(data = p_dat %>% dplyr::filter(plot == '% cells'),
+           position = 'fill') +
+  geom_col(data = p_dat %>% dplyr::filter(plot == 'n cells'),
+           position = 'stack') +
+  theme_classic() +
+  ggh4x::facet_grid2(plot ~ ., scales = 'free_y', independent = 'y') +
+  scale_x_discrete(guide = guide_axis(angle = -90)) +
+  scale_y_continuous(expand = c(0, 0)) +
+  scale_fill_manual(values = event_cols) +
+  theme(axis.title = element_blank(),
+        axis.text.x = element_text(size = 6))
+```
+
+## Co-occurence of CNV events
 
@@
 
-#### Fisher exact test
+### Fisher exact test
 
@@
     # remove chr7+ (can't do chr7+ vs chr7+)
-    dplyr::filter(event != 'chr7_gain', n_both > 0) %>%
-    dplyr::arrange(-OR)
+    dplyr::filter(event != 'chr7_gain', n_event > 0) %>%
+    dplyr::arrange(p)
 
@@
 
-}, file = 'fisher.rds', rerun = params$rerun)
+}, file = 'fisher.rds', rerun = T)
 
@@
 
-#### GLM
+### GLM
 
@@
 for_glm %>%
-  tidyr::pivot_longer(cols = c('CTAP', 'sex', 'race', 'ethnicity', 'subcluster')) %>%
+  tidyr::pivot_longer(cols = c('CTAP', 'sex', 'race', 'ethnicity')) %>%
   dplyr::group_by(name, value) %>%
@@
       dplyr::filter(rank(adj_pval) <= 10 | rank(lfc, ties.method = 'min') <= 10),
-    aes(label = name, ),
+    aes(label = name),
     min.segment.length = 0, nudge_x = 0.7, show.legend = F) +
@@
   scale_shape_manual(values = c('TRUE' = 55, 'FALSE' = 16), na.value = 16)
+```
+
+```{r plot_de_close}
+min_p_non_chr7 <-
+  de_res %>%
+  dplyr::filter(!is.na(direction)) %>%
+  dplyr::pull(adj_pval) %>% 
+  min()
+de_res %>%
+  ggplot(aes(x = lfc, y = -log10(adj_pval), colour = direction, 
+             size = chr == 'chr7', shape = chr == 'chr7')) +
+  geom_point() +
+  geom_hline(yintercept = -log10(min_p), linetype = 'dashed') +
+  geom_vline(xintercept = c(log2(1.5), log2(2)), linetype = 'dashed', colour = 'purple') +
+  geom_vline(xintercept = c(-1.5, 1.5), linetype = 'dashed') +
+  ggrepel::geom_text_repel(
+    data = 
+      de_res %>% 
+      dplyr::filter(!is.na(direction)) %>%
+      dplyr::group_by(direction) %>%
+      dplyr::filter(rank(adj_pval) <= 10 | rank(lfc, ties.method = 'min') <= 10),
+    aes(label = name),
+    min.segment.length = 0, nudge_x = 0.7, show.legend = F) +
+  theme_classic() +
+  scale_colour_manual(values = c('upregulated' = 'darkred',
+                                 'downregulated' = 'darkblue'), 
+                      na.value = 'darkgrey') +
+  scale_size_manual(values = c('TRUE' = 5, 'FALSE' = 4), na.value = 4) +
+  scale_shape_manual(values = c('TRUE' = 55, 'FALSE' = 16), na.value = 16) +
+  ylim(0, -log10(min_p_non_chr7) * 1.03)
 ```
@@
   knitr::kable()
+```
+
+# Figures for the paper
+
+```{r n_fibroblasts_dist}
+
 ```

### 2024-07-02 infercnv/src/01_run_infercnv.sh

@@ -1,16 +1,30 @@
 #!/bin/bash
-# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/ ; . ~/.bashrc ; mamba activate jupy ; ~/bin/jsub lsf -q week -n infercnv -c 1 -m 2g -l log "bash src/01_run_infercnv.sh" | bsub
+# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/ ; bsub -q week -M 2000 -R 'select[mem>=2000] span[hosts=1] rusage[mem=2000]' -J infercnv -o log "bash src/01_run_infercnv.sh"
 
-set -eo pipefail
+# get singularity module
+module load singularity
 
 # wd
 wd=/lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/
 cd $wd
 mkdir -p work/ out/ log/
 
-# conda activate an environment which has docopt installed so jsub works
-. ~/.bashrc
-mamba activate jupy
+# run infercnv on celltypes, running all individuals together
+# TODO: decide on cluster_by_groups 
+# arg definition from inferCNV - If observations are defined according to groups 
+# (ie. patients), each group of cells will be clustered separately. 
+# (default = FALSE, instead will use k_obs_groups setting)
+nextflow run nextflow/infercnv.nf \
+  --out_dir out/Zhang2023/by_celltype_run_all/ \
+  --mappings data/Zhang2023/stromal_mappings_run_all.csv \
+  --annotations data/Zhang2023/annotations.tsv \
+  --annotation_col "celltype" \
+  --analysis_mode "samples" \
+  --cluster_by_groups "FALSE" \
+  -c config/infercnv.config \
+  -c /lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/config/LSF.config \
+  -w work/ \
+  -resume
 
 # # run infercnv on celltypes
 # /software/team205//nextflow-23.04.1-all run nextflow/infercnv.nf \
@@ -24,7 +38,7 @@ mamba activate jupy
 #   -resume
     
 # # run infercnv on clusters
-# /software/team205/nextflow-23.04.1-all run nextflow/infercnv.nf \
+# nextflow run nextflow/infercnv.nf \
[…]

### 2024-07-02 infercnv/src/04_top_genes_per_celltype.R

@@ -1,0 +1,25 @@
+# libraries
+library(magrittr)
+library(Seurat)
+
+# dir
+wd <- 
+  ifelse(Sys.info()['nodename'] == "mib119104s", '~/Volumes/', '') |> 
+  paste0('/lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/') 
+setwd(wd)
+
+# read in sce, convert to seurat
+sce <-
+  readRDS('data/Zhang2023/sce/sce.rds')
+seu <- as.Seurat(sce)
+Idents(seu) <- seu$cell_type
+sce <- NULL
+
+# find top markers for each celltype
+markers <-
+  FindAllMarkers(seu)
+
+# save the markers
+markers %>%
+  tibble::as_tibble() %>%
+  readr::write_tsv('out/Zhang2023/by_celltype/summary/markers/celltype_markers.tsv')

### 2024-07-02 scomatic/src/06_plot_TNFRSF14_nonsynonymous_mutations.R

@@ -8,7 +8,9 @@ setwd(wd)
 
 #libraries
 library(magrittr)
-library(GenomicFeatures)
+library(ggplot2)
+
+# load ----
 
 # get recursites in TNFRSF14
 recursites <-
@@ -28,7 +30,7 @@ exons <-
     gene = 'TNFRSF14') %>%
   dplyr::mutate(feature = paste0('exon', dplyr::row_number()))}
 introns <-
-  gene_spans %>%
+  exons %>%
   dplyr::group_by(chr, gene) %>%
   dplyr::transmute(
     new_end = start - 1, start = dplyr::lag(end) + 1) %>%
@@ -48,6 +50,8 @@ features_all_positions <-
   tidyr::unnest(cols = 'pos') %>%
   dplyr::select(-start, -end)
 
+# colours and orderings ----
+
 # alts - colours and order
 nt_cols <-
   c('INS' = '#e6ab02',
@@ -66,6 +70,8 @@ imp_cols <-
     'Essential_Splice' = '#d95f02', 
     'Stop_loss' = '#7570b3')
 
+# generate plot data ----
+
 # plot mutations
 p_dat <-
   reads %>%
@@ -82,7 +88,18 @@ p_dat <-
   dplyr::filter(sum(value) > 0) %>%
   dplyr::left_join(features_all_positions %>% dplyr::select(-gene))
 
-pdf('test.pdf', height = 15, width = 20)
+# plot mutations, collapsed per patient
+p_dat_id_lvl <-
+  reads %>%
+  dplyr::bind_rows() %>%
+  dplyr::filter(pos == mut_pos) %>%
+  dplyr::left_join(recursites, relationship = 'many-to-many') %>%
+  dplyr::filter(name == mut, value > 0) %>%
+  dplyr::group_by(chr, pos, ref, status, aachange, celltype, gene, impact) %>%
+  dplyr::summarise(value = dplyr::n_distinct(id)) %>%
+  dplyr::left_join(features_all_positions %>% dplyr::select(-gene))
+
+pdf('out/Zhang2023/targeted_mutation_calling/TNFRSF14/summary/mutations.pdf', height = 15, width = 20)
 
 # exons and introns
 p_dat %>%
@@ -115,7 +132,10 @@ p_dat %>%
   geom_segment(aes(x = pos, xend = pos, y = 0, yend = value)) +
   geom_point() +
   ggrepel::geom_label_repel(
[…]

### 2024-07-08 infercnv/reports/paper_figures.Rmd
add schema, update to use qc'ed cells only, add more figures

--- 2024-03-11 infercnv/reports/gains_and_losses.Rmd
@@
 
-```{r setup, include = F, message = F, warning = F, class.source = 'fold-hide'}
-# wd <- ifelse(Sys.info()['nodename'] == "mib119104s", '~/Volumes/', '') |> paste0('/lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/') ; setwd(wd)
-# rmarkdown::render('reports/infercnv.Rmd', output_file = 'driver_coverage.html', output_dir = 'reports/', params = list(rerun = T))
+```{r setup, include = F, message = F, warning = F, class.source = "fold-hide"}
+# module load R/4.4.0 ; Rscript -e "rmarkdown::render("reports/paper_figures.Rmd", output_dir = "out/paper_figures/")" # nolint: line_length_linter.
 
@@
 
-inferCNV was run on the scRNA-seq data to call copy number variants in synovial cells. No reference group was used, so the average profile of all cells was used in each patient as the baseline.
+inferCNV was run on the scRNA-seq data to call copy number variants in synovial 
+cells. No reference group was used, so the average profile of all cells was used 
+in each patient as the baseline.
 
-First, we load the CITE-seq object. We convert it to a SingleCellExperiment object. Then we add the CNV metadata from the inferCNV run to the object.
+First, we load the CITE-seq data as a `SingleCellExperiment` object. The CNV 
+metadata from the `inferCNV` run has been aded to this object.
 
-```{r load_sce, results = 'asis', class.source = 'fold-show'}
-# load gene coords
-coords <-
-  readr::read_tsv('data/hg38_gencode_v27.txt', col_names = c('name', 'chr', 'start', 'stop'))
-
-# load CTAP and metadata
-ctap <-
-  readr::read_tsv('data/Zhang2023/syn26710600/CTAP_donor_mapping.tsv') %>%
-  dplyr::select(sample = donor, CTAP)
-md <- 
-  readr::read_tsv('data/Zhang2023/syn26710600/SYNAPSE_METADATA_MANIFEST.tsv') %>%
-  dplyr::distinct()
-
-# load sce, add inferCNV 
-sce <- xfun::cache_rds({
-
-  # load sce
-  sce <- readRDS(params$sce)
-
-  # get cnv metadata
-  cnvs <- 
-    list.files(
-      params$infercnv_dir,
-      pattern = 'map_metadata_from_infercnv.txt',
-      recursive = T) %>%
-    purrr::set_names(., dirname(.)) %>%
-    purrr::map(function(file) {
-      file.path(params$infercnv_dir, file) %>% 
-        read.table(sep = '\t')
-      }) %>%
-    dplyr::bind_rows() %>%
-    dplyr::mutate(infercnv = T)
-
-  # left join cnvs to coldata, reorder, restore rownames
-  col_data <- merge(sce@colData, cnvs, by = 'row.names', all.x = T, sort = F)
-  rownames(col_data) <- col_data[, 1]
-  col_data <- 
-    col_data[colnames(sce), -1] %>%
-    DataFrame()
-
-  # add to sce
-  sce@colData <- col_data
-
-  # return
-  sce
-
-}, file = 'sce.rds', rerun = params$rerun)
-
-# remove NA's for inferCNV analysis
-sce <- sce[, !is.na(sce$proportion_cnv_chr1)]
-
- # assign cell group
-sce@colData$cell_group <-
-  sce@colData[, params$cell_group_name]
+```{r load_sce, results = "asis", class.source = "fold-show"}
+# load sce with gains
+sce <-
+  "out/Zhang2023/by_fibroblast_cluster/summary/gains_and_losses_cache/" %>%
+  list.files(pattern = "sce_w_gains", full.names = TRUE) %>%
+  readRDS()
 ```
 
-The `inferCNV` output file `map_metadata_from_infercnv.txt` contains columns with different CNV measures at whole-chromosome level per cell, with rownames denoting cells.
-The columns have the following values...
+# Quantifying CNVs of interest in the samples' fibroblast populations
 
-```{r echo = F, results = 'asis'}
-tibble::tribble(
-  ~name, ~`type - value`,
-  '`has_*`', 'binary - whether any CNV / loss / duplication is found within the given chr',
-  '`proportion_*`', 'proportion - proportion of genes that are part of any CNV / loss / duplication within the given chr',
-  '`proportion_scaled_*`', 'proportion - same as proportion_*, but taking into account whether genes that are duplicated / lost are a single copy or two copies, as a weight',
-  '`top_loss_*`', 'the top CNVs that affect the most genes'
-) %>%
+We first prepare the data so that we can view the proportions of chr7 and chr5 
+gains in each patient's fibroblast population.
+
+```{r}
+events <-
+  sce@colData %>%
+  tibble::as_tibble() %>%
+  dplyr::select(sample, cell, cell_group, chr7_gain, chr5_gain) %>%
+  tidyr::pivot_longer(-c("sample", "cell", "cell_group")) %>%
+  dplyr::group_by(sample, cell, cell_group) %>%
+  # remove cells in which no calls made
+  dplyr::filter(!all(is.na(value))) %>%
+  dplyr::mutate(event = dplyr::case_when(all(is.na(value) | value == 0) ~ "WT",
+                                         value == 1 ~ name, TRUE ~ NA)) %>%
+  dplyr::summarise(
+    events = paste(sort(unique(na.omit(event))), collapse = ",") %>%
+      factor(levels = c("chr5_gain,chr7_gain", "chr7_gain", "chr5_gain", "WT")),
+    n_events = sum(value, na.rm = TRUE)) %>%
+  # get props and counts
+  dplyr::ungroup() %>%
+  dplyr::add_count(sample, name = "n_cells_per_sample") %>%
+  dplyr::add_count(sample, events, name = "n_cells_per_sample_x_event") %>%
+  dplyr::distinct(sample, events,
+                  n_cells_per_sample_x_event, n_cells_per_sample) %>%
+  dplyr::mutate(prop_cells_per_sample_x_event = n_cells_per_sample_x_event /
+                  n_cells_per_sample) %>%
+  # refactor samples by prop non-WT cells
+  dplyr::group_by(sample) %>%
+  dplyr::mutate(
+    n_wt_per_sample = dplyr::case_when(events == "WT" ~
+                                         n_cells_per_sample_x_event,
+                                       TRUE ~ 0),
+    prop_non_wt_per_sample = (n_cells_per_sample - sum(n_wt_per_sample)) /
+                  n_cells_per_sample) %>%
+  dplyr::ungroup() %>%
+  dplyr::mutate(sample = forcats::fct_reorder2(sample, n_cells_per_sample,
+                                               prop_non_wt_per_sample)) %>%
+  tidyr::pivot_longer(c("n_cells_per_sample_x_event",
+                        "prop_cells_per_sample_x_event"))
+```
+
+# Filtering out patients with few fibroblasts
+
+```{r}
+min_n_fibroblasts_per_sample <- 100
+```
+
+We set the minimum number of fibroblasts per patient to 
+`r min_n_fibroblasts_per_sample`. We can review how many patients pass the 
+cut-off to give us reliable results.
+
+```{r plot_fibroblast_dist}
+n_fibroblasts <-
+  sce@colData %>%
+  tibble::as_tibble() %>%
+  dplyr::count(sample)
+n_fibroblasts %>%
+  ggplot(aes(x = n)) +
+  geom_histogram(binwidth = 20) +
+  geom_vline(xintercept = min_n_fibroblasts_per_sample, linetype = "dashed", 
+             colour = "red") +
+  theme_classic()
+```
+
+```{r cut_off}
+enough_fibroblasts <-
+  dplyr::filter(n_fibroblasts, n >= min_n_fibroblasts_per_sample)$sample
+n_fibroblasts %>%
+  dplyr::summarise(`n >= 100` = sum(n >= 100),
+                   `n < 100` = sum(n < 100),
+                   n = dplyr::n()) %>%
   knitr::kable()
@@
 
-## HMM regional predictions
+This leaves `r length(enough_fibroblasts)` / `r nrow(n_fibroblasts)` of the 
+samples. We will filter the `SingleCellExperiment` object accordingly.
 
-We also load the CNV regions prediced by HMM. According to [their wiki](https://github.com/broadinstitute/infercnv/wiki/infercnv-i6-HMM-type#infercnv-i6-hmm)...
+We can also review the relationship between the number of cells in each 
+patient's fibroblast population and the proportions / counts of the CNVs of 
+interest (chr7 gain and chr5 gain).
 
-As well as aneuploidy, we want to look for smaller recurrent structural changes, including focal-/arm-level events.
-
-> The inferCNV i6 HMM is a six-state CNV model that predicts the following CNV levels:
->
-> * State 1: 0x: complete loss
-> * State 2: 0.5x: loss of one copy
-> * State 3: 1x: neutral
-> * State 4: 1.5x: addition of one copy
-> * State 5: 2x: addition of two copies
-> * State 6: 3x: essentially a placeholder for >2x copies but modeled as 3x.
-
-```{r get_hmm}
-cnv_regions <- xfun::cache_rds({
-
-  # get HMM predictions
-  cnv_regions <- 
-    list.files(
-      params$infercnv_dir,
-      pattern = 'HMM_CNV_predictions.HMMi6.leiden.hmm_mode-subclusters.Pnorm_0.5.pred_cnv_regions.dat',
-      recursive = T) %>%
-    purrr::set_names(., dirname(.)) %>%
-    purrr::map(function(file) {
-      file.path(params$infercnv_dir, file) %>% 
-        readr::read_tsv(show_col_types = F)
-      }) %>%
-    dplyr::bind_rows(.id = 'sample') %>%
-    # refactor chromosomes
-    dplyr::mutate(
-      chr = chr %>% {forcats::fct_reorder(., gsub('chr', '', .) %>% as.numeric())},
-      direction = dplyr::case_when(
-        state < 3 ~ 'loss',
-        state > 3 ~ 'dupli',
-        TRUE ~ 'neutral'),
-      ploidy = dplyr::case_when(
-        state == 1 ~ '0x',
-        state == 2 ~ '0.5x',
-        state == 3 ~ '1x',
-        state == 4 ~ '1.5x',
-        state == 5 ~ '2x',
-        state == 6 ~ '>2x',
-        TRUE ~ NA
-        ) %>% forcats::fct_reorder(state),
-      region = gsub('.*region_', '', cnv_name) %>% as.numeric(),
-      cell_group = cell_group_name %>% gsub('\\..*', '', .))
-
-  # calculate region midpoints for plotting
-  cnv_regions$midpoint <-
-    (cnv_regions$start + cnv_regions$end) / 2
-
-  # return
-  cnv_regions
-
-}, file = 'cnv_regions.rds', rerun = params$rerun)
+```{r plot_event_vs_fibroblast}
+events %>%
+  ggplot(aes(x = sample, y = value, fill = events)) +
+  geom_col() +
+  scale_x_discrete(guide = guide_axis(angle = -90)) +
+  facet_wrap(~ name, scale = "free_y", ncol = 1) +
+  theme_classic() +
+  scale_fill_brewer(palette = "Dark2") +
+  theme(legend.position = "bottom")
 ```
 
-```{r plot_cnv_regions_pred, fig.height = 20, fig.width = 10}
-cnv_regions %>%
-  ggplot(aes(x = midpoint, fill = ploidy, group = ploidy)) +
-  geom_density(data = cnv_regions %>% dplyr::filter(direction == 'dupli'),
-               aes(y = ..density..), 
-               position = position_stack(), 
-               colour = NA) +
-  geom_density(data = cnv_regions %>% dplyr::filter(direction == 'loss'),
-               aes(y = -..density..), 
-               position = position_stack(reverse = T), 
-               colour = NA) +
-  ggh4x::facet_grid2(chr ~ cell_group, scales = 'free_x', independent = 'x') +
-  theme_bw() +
-  theme(axis.text.x = element_blank()) +
-  scale_fill_manual(
-    values = c('0x' = '#00008B', '0.5x' = '#b4b0ff', '1' = 'white', 
-               '1.5x' = '#ffc2c2', '2x' = '#d64242', '>2x' = '#8B0000')) 
+We can now apply the cut-off.
+
+```{r apply_cutoff}
+sce <- sce[, sce$sample %in% enough_fibroblasts]
+events <- events %>% dplyr::filter(sample %in% enough_fibroblasts)
 ```
+
+After applying the cut-off, we look for sample-size effects once again.
+
+```{r plot_event_vs_fibroblast_after_cutoff}
+events %>%
+  ggplot(aes(x = sample, y = value, fill = events)) +
+  geom_col() +
+  scale_x_discrete(guide = guide_axis(angle = -90)) +
+  facet_wrap(~ name, scale = "free_y", ncol = 1) +
+  theme_classic() +
+  scale_fill_brewer(palette = "Dark2") +
+  theme(legend.position = "bottom")
+```
+
+We can see that there is still a strong relationship between the size of the
+fibroblast populations by patient and the proportion of those cells that are
+called as having a chr7 or chr5 gain. This is most likely related to with the
+fact that we set `ref_group_names = NULL`. This compounds the effect of the
+number of cells on the ability to detect CNVs (the false negative rate), 
+because...
+
+1. fewer cells makes it less likely to capture these somewhat rare events
+
+2. fewer cells makes the reference profile noisier and less representative of 
+the 'baseline' expression levels across chromosomes, further limiting CNV 
+calling
+
+Keeping this in mind, we will proceed.
 
@@
     sample = forcats::fct_infreq(sample),
-    cell_group = forcats::fct_infreq(cell_group)) %>%
-  dplyr::add_count(sample, cell_group, name = 'n_sample_x_cell_group')
+    cell_group = forcats::fct_rev(forcats::fct_infreq(cell_group))) %>%
+  dplyr::add_count(sample, cell_group, name = "n_sample_x_cell_group") %>%
+  dplyr::add_count(cell_group, name = "n_cell_group")
 p_base <-
   p_dat %>%
-  ggplot(aes(fill = cell_group, colour = cell_group)) 
+  ggplot(aes(fill = cell_group, colour = cell_group)) +
+  scale_fill_manual(values = dittoSeq::dittoColors()) +
+  scale_colour_manual(values = dittoSeq::dittoColors())
 p1 <- 
@@
   theme_classic() +
-  theme(legend.position = 'none') +
-  scale_x_discrete(guide = guide_axis(angle = -90))
+  theme(legend.position = "none",
+        axis.text.x = element_text(size = 7)) +
+  scale_x_discrete(guide = guide_axis(angle = -90)) +
+  labs(y = params$cell_group_name)
 p2 <- 
   p_base +
-  geom_bar(aes(x = sample), position = 'stack') +
+  geom_bar(aes(x = sample), position = "stack") +
   theme_void() +
-  theme(legend.position = 'none') 
+  theme(legend.position = "none",
+        axis.text.y = element_text()) 
 p3 <-
   p_base +
-  geom_bar(aes(y = cell_group), position = 'stack') +
+  geom_bar(aes(y = cell_group), position = "stack") +
+  geom_text(
+    data = p_dat %>% dplyr::distinct(cell_group, n_cell_group),
+    aes(x = n_cell_group, y = cell_group, 
+        label = format(n_cell_group, nsmall = 1, big.mark = ",")), 
+    hjust = -0.1) +
   theme_void() +
-  theme(legend.position = 'none')
+  theme(legend.position = "none") +
+  xlim(0, 1.2 * max(p_dat$n_cell_group))
 design <-
-'BB#
-AAC'
+"BB#
+BB#
+AAC"
+
+# plot
 patchwork::wrap_plots(A = p1, B = p2, C = p3, design = design)
@@
   coord_fixed()
+
+pal <-
+  c("T cell" = "#E6AB02",
+    "Stromal cell" = "#66A61E",
+    "Myeloid cell" = "#E7298A",
+    "B cell/plasma cell" = "#7570B3",
+    "Endothelial cell" = "#D95F02",
+    "NK" = "#1B9E77")
+dittoSeq::dittoDimPlot(sce, "cell_type", size = 0.2) +
+  coord_fixed() +
+  scale_colour_manual(values = pal)
 ```
@@
 
-```{r plot_chr_gains, fig.width = 10, fig.height = 4, results = 'hide'}
-paste0('chr', 1:22) %>% 
+```{r plot_chr_gains, fig.width = 10, fig.height = 4, results = "hide"}
+paste0("chr", 1:22) %>% 
   purrr::map(function(chr) {
     events <-
-      paste('proportion', c('dupli', 'loss'), chr, sep = '_')
-    reducedDim(sce, 'UMAP_uwot') %>%
+      paste("proportion", c("dupli", "loss"), chr, sep = "_")
+    reducedDim(sce, "UMAP_uwot") %>%
       cbind(sce@colData[, events]) %>%
@@
 
-```{r plot_gains_and_losses_by_ct, results = 'hide'}
+```{r plot_gains_and_losses_by_ct, results = "hide"}
 # at single-cell level
@@
   purrr::map(function(cgn) {
-    c('dupli', 'loss') %>%
+    c("dupli", "loss") %>%
       purrr::map(function(direction) {
@@
           p_mat <-
-            p_dat[, grepl(paste0('proportion_', direction), colnames(p_dat))]
+            p_dat[, grepl(paste0("proportion_", direction), colnames(p_dat))]
           p_annot <- 
-            p_dat[, c('sample'), drop = F] %>%
+            p_dat[, c("sample"), drop = F] %>%
             as.data.frame()
@@
             legend = F,
-            main = paste0(params$cell_group_name, ': ', cgn, '\ndirection: ', direction)) %>%
+            main = paste0(params$cell_group_name, ": ", cgn, "\ndirection: ", direction)) %>%
             print()
@@
   dplyr::summarise(
-    dplyr::across(dplyr::starts_with('proportion_'), ~ mean(.x, na.rm = T))) %>%
+    dplyr::across(dplyr::starts_with("proportion_"), ~ mean(.x, na.rm = T))) %>%
   dplyr::ungroup() %>%
-  dplyr::mutate(sample_x_cg = paste0(sample, '_', cell_group)) %>%
-  tibble::column_to_rownames('sample_x_cg')
+  dplyr::mutate(sample_x_cg = paste0(sample, "_", cell_group)) %>%
+  tibble::column_to_rownames("sample_x_cg")
 
 # split gain and loss
-c('dupli', 'loss') %>% 
+c("dupli", "loss") %>% 
   purrr::map(function(direction) {
     p_mat <-
-      p_dat[, grepl(paste0('proportion_', direction), colnames(p_dat))]
+      p_dat[, grepl(paste0("proportion_", direction), colnames(p_dat))]
     p_annot <- 
-      p_dat[, c('sample', 'cell_group')] %>%
+      p_dat[, c("sample", "cell_group")] %>%
       as.data.frame()
@@
       annotation_row = p_annot,
-      main = paste0('direction: ', direction)) %>%
+      main = paste0("direction: ", direction)) %>%
       print()
@@
             axis.title.x = element_blank()) +
-      ylab('count'),
+      ylab("count"),
     plot_func(p_dat, position_stack(reverse = T)) +
-      ylab('proportion')) %>%
+      ylab("proportion")) %>%
     patchwork::wrap_plots(ncol = 1) 
@@
 # plot, across all and by patient
-plot_prop_chr7_gain(p_dat, 'chr7 gain - all cells')
+plot_prop_chr7_gain(p_dat, "chr7 gain - all cells")
 p_dat %>%
@@
     plot_prop_chr7_gain(
-      paste0('chr7 gains - ', id, ' (', unique(df$n_cells), ' cells)')) %>%
+      paste0("chr7 gains - ", id, " (", unique(df$n_cells), " cells)")) %>%
       print()
@@
 
-```{r assign_gains, class.source = 'fold-show'}
+```{r assign_gains, class.source = "fold-show"}
 sce <- xfun::cache_rds({
@@
   # apply cut-offs for whole chr gains and losses
-  paste0('chr', 1:22) %>%
+  paste0("chr", 1:22) %>%
     purrr::walk(function(chr) {
@@
         sce@colData[, paste0("proportion_dupli_", chr)]
-      sce@colData[, paste0(chr, '_gain')] <<-
+      sce@colData[, paste0(chr, "_gain")] <<-
         ifelse(chr_dupli > params$min_prop_for_cnv, 1, 
@@
         sce@colData[, paste0("proportion_loss_", chr)]
-      sce@colData[, paste0(chr, '_loss')] <<-
+      sce@colData[, paste0(chr, "_loss")] <<-
         ifelse(chr_loss > params$min_prop_for_cnv, 1, 
@@
 
-}, file = 'sce_w_gains.rds', rerun = params$rerun)
+}, file = "sce_w_gains.rds", rerun = params$rerun)
 ```
 
-### Co-occurence of CNV events
+```{r plot_heatmap, fig.width = 10, fig.height = 5}
+event_cols <-
+  c("WT" = "grey",
+    "chr5_gain" = "#e7298a",
+    "chr5_gain,chr7_gain" = "#d95f02",
+    "chr7_gain" = "#e6ab02")
+
+p_dat <-
+  sce@colData %>%
+  tibble::as_tibble() %>%
+  dplyr::select(sample, cell, 
+                dplyr::ends_with("_gain"), 
+                dplyr::ends_with("_loss")) %>%
+  tidyr::pivot_longer(-c("sample", "cell")) %>%
+  dplyr::filter(name %in% c("chr7_gain", "chr5_gain")) %>%
+  dplyr::group_by(sample, cell) %>%
+  # remove cells in which no calls made
+  dplyr::filter(!all(is.na(value))) %>%
+  dplyr::mutate(
+    event = dplyr::case_when(
+      all(value == 0 | is.na(value)) ~ "WT",
+      value == 1 ~ name,
+      TRUE ~ NA)) %>%
+  dplyr::summarise(
+    events = paste(sort(unique(na.omit(event))), collapse = ",")) %>%
+  dplyr::mutate(events = forcats::fct_relevel(events, names(event_cols))) %>%
+  # get n aneuploid for plot ordering
+  dplyr::group_by(sample) %>%
+  dplyr::mutate(prop_aneuploid = sum(events != "WT") / dplyr::n()) %>%
+  # count
+  dplyr::group_by(sample, prop_aneuploid) %>%
+  dplyr::count(events) %>%
+  # for fill and stack
+  tidyr::crossing(tibble::tibble(plot = c("% cells", "n cells")))
+
+# plot
+p_dat %>%
+  ggplot(aes(x = reorder(sample, -prop_aneuploid), y = n, fill = events)) +
+  geom_col(data = p_dat %>% dplyr::filter(plot == "% cells"),
+           position = "fill") +
+  geom_col(data = p_dat %>% dplyr::filter(plot == "n cells"),
+           position = "stack") +
+  theme_classic() +
+  ggh4x::facet_grid2(plot ~ ., scales = "free_y", independent = "y") +
+  scale_x_discrete(guide = guide_axis(angle = -90)) +
+  scale_y_continuous(expand = c(0, 0)) +
+  scale_fill_manual(values = event_cols) +
+  theme(axis.title = element_blank(),
+        axis.text.x = element_text(size = 6))
+```
+
+## Co-occurence of CNV events
 
@@
   dplyr::select(sample, cell, cell_group, 
-                dplyr::ends_with('_gain'), 
-                dplyr::ends_with('_loss')) %>%
-  tidyr::pivot_longer(-c('sample', 'cell', 'cell_group')) %>%
+                dplyr::ends_with("_gain"), 
+                dplyr::ends_with("_loss")) %>%
+  tidyr::pivot_longer(-c("sample", "cell", "cell_group")) %>%
   dplyr::group_by(sample, cell, cell_group) %>%
@@
     event = dplyr::case_when(
-      all(value == 0 | is.na(value)) ~ 'WT',
+      all(value == 0 | is.na(value)) ~ "WT",
       value == 1 ~ name,
@@
   dplyr::summarise(
-    events = paste(sort(unique(na.omit(event))), collapse = ','),
+    events = paste(sort(unique(na.omit(event))), collapse = ","),
     n_events = sum(value, na.rm = T))
@@
   # ignore events that happen in <= 25 cells
-  dplyr::filter(events != 'WT', n >= 25) %>%
+  dplyr::filter(events != "WT", n >= 25) %>%
   dplyr::mutate(
     events = events %>% 
-      gsub('chr', '', .) %>% 
-      gsub('_gain', '+', .) %>%
-      gsub('_loss', '-', .),
+      gsub("chr", "", .) %>% 
+      gsub("_gain", "+", .) %>%
+      gsub("_loss", "-", .),
     events = forcats::fct_infreq(events),
@@
       n < 10 ~ label1,
-      TRUE ~ paste0(label1, ' (', round(prop*100, 2), '%)')
+      TRUE ~ paste0(label1, " (", round(prop*100, 2), "%)")
     ))
@@
   theme_classic() +
-  theme(legend.position = 'bottom')
+  theme(legend.position = "bottom")
 p + 
@@
   geom_bar(aes(fill = cell_group)) +
-  theme(legend.position = 'none')
+  theme(legend.position = "none")
 ```
+
+### Fisher exact test
 
@@
   tibble::as_tibble() %>%
-  dplyr::mutate(chr7 = ifelse(chr7_gain == 1, 'chr7+', 'chr7_WT'),
-                chr5 = ifelse(chr5_gain == 1, 'chr5+', 'chr5_WT')) 
+  dplyr::mutate(chr7 = ifelse(chr7_gain == 1, "chr7+", "chr7_WT"),
+                chr5 = ifelse(chr5_gain == 1, "chr5+", "chr5_WT")) 
 table(cont$chr7, cont$chr5) %>%
@@
   fisher <-
-    colnames(sce@colData)[grepl('_gain$|_loss$', colnames(sce@colData))] %>%
+    colnames(sce@colData)[grepl("_gain$|_loss$", colnames(sce@colData))] %>%
     purrr::set_names() %>%
@@
 
+        # get contingencies
+        cont <-
+          table(sce@colData[, c("chr7_gain", event)])
+
          # return relevant stats
@@
           OR = fisher$estimate, 
-          p = fisher$p.value)
+          p = fisher$p.value,
+          sig = ifelse(fisher$p.value < 0.05, "*", ""),
+          n_WT = cont[1, 1],
+          n_chr7_gain = cont[2, 1],
+          n_event = cont[1, 2],
+          n_both = cont[2, 2])
 
@@
     }) %>%
-    dplyr::bind_rows(.id = 'event') %>%
+    dplyr::bind_rows(.id = "event") %>%
     # remove chr7+ (can't do chr7+ vs chr7+)
-    dplyr::filter(event != 'chr7_gain') %>%
-    dplyr::arrange(-OR)
+    dplyr::filter(event != "chr7_gain", n_event > 0) %>%
+    dplyr::arrange(p)
 
@@
 
-}, file = 'fisher.rds', rerun = params$rerun)
+}, file = "fisher.rds", rerun = T)
 
@@
   knitr::kable()
+```
+
+### GLM
+
+We can also run a GLM.
+
+```{r glm}
+# read in metadata
+metadata <-
+  readr::read_tsv(params$metadata)
+
+# create glm
+# prop_chr7_gain ~ age (+ coverage) (per donor)
+for_glm <-
+  dplyr::inner_join(
+    tibble::as_tibble(sce@colData), 
+    metadata, 
+    by = c("sample" = "id")) %>%
+  dplyr::filter(!is.na(age), !is.na(chr7_gain)) %>%
+  dplyr::group_by(sample, age, CTAP, sex, race, ethnicity) %>%
+  dplyr::summarise(n_chr7_gain = sum(chr7_gain),
+                   n_cells = dplyr::n(),
+                   ch7_gain_prop = n_chr7_gain / n_cells)
+
+for_glm %>%
+  {glm(.$n_chr7_gain ~ log(.$n_cells) + .$age)}
+```
+
+```{r plot_glm_vars}
+# n cells vs n chr7 gain
+for_glm %>%
+  ggplot(aes(x = n_cells, y = n_chr7_gain)) +
+  geom_point() +
+  theme_classic()
+
+# age vs prop chr7 gain
+for_glm %>%
+  ggplot(aes(x = age, y = ch7_gain_prop)) +
+  geom_point() +
+  theme_classic()
+
+# all categorical variables
+for_glm %>%
+  tidyr::pivot_longer(cols = c("CTAP", "sex", "race", "ethnicity")) %>%
+  dplyr::group_by(name, value) %>%
+  dplyr::mutate(med = median(ch7_gain_prop)) %>%
+  ggplot(aes(
+    x = tidytext::reorder_within(value, by = med, within = name), 
+    y = ch7_gain_prop)) +
+  geom_boxplot() +
+  geom_jitter(height = 0) +
+  facet_wrap(~ name, nrow = 1, scales = "free") +
+  tidytext::scale_x_reordered(guide = guide_axis(angle = -45)) +
+  theme_classic() +
+  labs(x = "")
+
+# looking by cluster
+for_glm_w_cluster <-
+  dplyr::inner_join(
+    tibble::as_tibble(sce@colData), 
+    metadata, 
+    by = c("sample" = "id")) %>%
+  dplyr::filter(!is.na(age), !is.na(chr7_gain)) %>%
+  dplyr::group_by(
+    sample, age, CTAP, sex, race, ethnicity, cluster_name) %>%
+  dplyr::summarise(n_chr7_gain = sum(chr7_gain),
+                   n_cells = dplyr::n(),
+                   ch7_gain_prop = n_chr7_gain / n_cells) 
+for_glm_w_cluster %>%
+  ggplot(aes(x = cluster_name, y = ch7_gain_prop)) +
+  geom_violin() +
+  geom_jitter(height = 0) +
+  theme_classic() +
+  scale_x_discrete(guide = guide_axis(angle = -45))
 ```
@@
 
-```{r chr7_gain_prop, results = 'asis'}
+```{r chr7_gain_prop, results = "asis"}
 dat <-
@@
     chr7_gain_status = dplyr::case_when(
-      is.na(chr7_gain_value) ~ 'unknown/intermediate',
-      chr7_gain_value == 0 ~ 'WT',
-      chr7_gain_value == 1 ~ 'gain')
+      is.na(chr7_gain_value) ~ "unknown/intermediate",
+      chr7_gain_value == 0 ~ "WT",
+      chr7_gain_value == 1 ~ "gain")
   )
 
-cat('Counts and proportions of chr7 gain, all cells')
+cat("Counts and proportions of chr7 gain, all cells")
 dat %>%
@@
 
-cat('Counts and proportions of chr7 gain, by cell group')
+cat("Counts and proportions of chr7 gain, by cell group")
 dat %>%
@@
 
-```{r pseudobulk_chr7_gain, class.source = 'fold-show'}
+```{r pseudobulk_chr7_gain, class.source = "fold-show"}
 reduced_sce <- xfun::cache_rds({
@@
 
-}, file = 'reduced_sce.rds', rerun = params$rerun)
+}, file = "reduced_sce.rds", rerun = params$rerun)
 ```
@@
   fit
-}, file = 'fit.rds', rerun = params$rerun)
+}, file = "fit.rds", rerun = params$rerun)
 
@@
       direction = dplyr::case_when(
-        lfc < -1.5 & adj_pval < min_p ~ 'downregulated',
-        lfc > 1.5 & adj_pval < min_p ~ 'upregulated',
+        lfc < -1.5 & adj_pval < min_p ~ "downregulated",
+        lfc > 1.5 & adj_pval < min_p ~ "upregulated",
         TRUE ~ NA)) %>%
@@
 
-}, file = 'de_res.rds', rerun = params$rerun)
+}, file = "de_res.rds", rerun = params$rerun)
 ```
@@
   ggplot(aes(x = lfc, y = -log10(adj_pval), colour = direction, 
-             size = chr == 'chr7', shape = chr == 'chr7')) +
+             size = chr == "chr7", shape = chr == "chr7")) +
   geom_point() +
-  geom_hline(yintercept = -log10(min_p), linetype = 'dashed') +
-  geom_vline(xintercept = c(log2(1.5), log2(2)), linetype = 'dashed', colour = 'purple') +
-  geom_vline(xintercept = c(-1.5, 1.5), linetype = 'dashed') +
+  geom_hline(yintercept = -log10(min_p), linetype = "dashed") +
+  geom_vline(xintercept = c(log2(1.5), log2(2)), linetype = "dashed", colour = "purple") +
+  geom_vline(xintercept = c(-1.5, 1.5), linetype = "dashed") +
   ggrepel::geom_text_repel(
@@
       dplyr::group_by(direction) %>%
-      dplyr::filter(rank(adj_pval) <= 10 | rank(lfc, ties.method = 'min') <= 10),
-    aes(label = name, ),
+      dplyr::filter(rank(adj_pval) <= 10 | rank(lfc, ties.method = "min") <= 10),
+    aes(label = name),
     min.segment.length = 0, nudge_x = 0.7, show.legend = F) +
   theme_classic() +
-  scale_colour_manual(values = c('upregulated' = 'darkred',
-                                 'downregulated' = 'darkblue'), 
-                      na.value = 'darkgrey') +
-  scale_size_manual(values = c('TRUE' = 3, 'FALSE' = 2), na.value = 2) +
-  scale_shape_manual(values = c('TRUE' = 55, 'FALSE' = 16), na.value = 16)
+  scale_colour_manual(values = c("upregulated" = "darkred",
+                                 "downregulated" = "darkblue"), 
+                      na.value = "darkgrey") +
+  scale_size_manual(values = c("TRUE" = 3, "FALSE" = 2), na.value = 2) +
+  scale_shape_manual(values = c("TRUE" = 55, "FALSE" = 16), na.value = 16)
+```
+
+```{r plot_de_close}
+min_p_non_chr7 <-
+  de_res %>%
+  dplyr::filter(!is.na(direction)) %>%
+  dplyr::pull(adj_pval) %>% 
+  min()
+de_res %>%
+  ggplot(aes(x = lfc, y = -log10(adj_pval), colour = direction, 
+             size = chr == "chr7", shape = chr == "chr7")) +
+  geom_point() +
+  geom_hline(yintercept = -log10(min_p), linetype = "dashed") +
+  geom_vline(xintercept = c(log2(1.5), log2(2)), linetype = "dashed", colour = "purple") +
+  geom_vline(xintercept = c(-1.5, 1.5), linetype = "dashed") +
+  ggrepel::geom_text_repel(
+    data = 
+      de_res %>% 
+      dplyr::filter(!is.na(direction)) %>%
+      dplyr::group_by(direction) %>%
+      dplyr::filter(rank(adj_pval) <= 10 | rank(lfc, ties.method = "min") <= 10),
+    aes(label = name),
+    min.segment.length = 0, nudge_x = 0.7, show.legend = F) +
+  theme_classic() +
+  scale_colour_manual(values = c("upregulated" = "darkred",
+                                 "downregulated" = "darkblue"), 
+                      na.value = "darkgrey") +
+  scale_size_manual(values = c("TRUE" = 5, "FALSE" = 4), na.value = 4) +
+  scale_shape_manual(values = c("TRUE" = 55, "FALSE" = 16), na.value = 16) +
+  ylim(0, -log10(min_p_non_chr7) * 1.03)
 ```
@@
 
+# Figures for the paper
+
+```{r n_fibroblasts_dist}
+
+```
+
 # Cell group markers
@@
   seu
-}, file = 'seu.rds', rerun = params$rerun)
+}, file = "seu.rds", rerun = params$rerun)
 ```
@@
   markers
-}, file = 'markers.rds', rerun = params$rerun)
+}, file = "markers.rds", rerun = params$rerun)
 ```
@@
       patchwork::plot_annotation(
-        title = paste0(params$cell_group_name, ': ', cg)) %>%
+        title = paste0(params$cell_group_name, ": ", cg)) %>%
       print()

### 2024-07-08 infercnv/src/00_preprocess_sce.R

@@ -8,35 +8,30 @@ library(SummarizedExperiment)
 library(SingleCellExperiment)
 
 # dirs
-wd <- 
-  ifelse(Sys.info()['nodename'] == "mib119104s", '~/Volumes/', '') |>
-  paste0('/lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/') 
-setwd(wd)
-data_dir <- '/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/processed_output/'
-out_dir <- 'data/Zhang2023/sce/'
-dir.create(out_dir)
+data_dir <- "/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/processed_output/"
+out_dir <- "out/Zhang2023/sce/"
 
 # function: convert Zhang 2023 object to sce
 get_sce <- function(obj, assays) {
-  
+
   # get cell order from metadata
   cell_order <- obj$meta_data$cell
-  
+
   # filter and reorder to cells that pass qc from counts
   assays_pass <-
     assays %>%
     purrr::map(~ .x[rownames(assays$logcounts), cell_order])
-  
+
   # populate sce
-  sce <- 
+  sce <-
     SingleCellExperiment::SingleCellExperiment(
       assays = assays_pass,
       colData = obj$meta_data
     )
-  
+
   # name, reorder, and add umap embeddings
-  reducedDim(sce, 'UMAP_uwot') <- obj$umap$embedding
-  
+  reducedDim(sce, "UMAP_uwot") <- obj$umap$embedding
+
   # return
   sce
 }
@@ -44,46 +39,42 @@ get_sce <- function(obj, assays) {
 # all cells ----
 
 # load object
-obj <- 
-  readRDS(file.path(data_dir, 'all_cells_reference.rds'))
+obj <- readRDS(file.path(data_dir, "all_cells_reference.rds"))
 
-# load rna data
-assays <-
-  list(
-    counts = 
-      readRDS(file.path(data_dir, 'raw_mRNA_count_matrix.rds')),
-    logcounts = 
-      readRDS(file.path(data_dir, 'qc_mRNA_314011cells_log_normalized_matrix.rds')))
+# load rna data (raw + normalised)
+counts_rds <- file.path(data_dir, "raw_mRNA_count_matrix.rds")
[…]

### 2024-07-08 infercnv/src/01_run_infercnv.sh

@@ -4,33 +4,41 @@
 # get singularity module
 module load singularity
 
-# wd
-wd=/lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/
-cd $wd
-mkdir -p work/ out/ log/
-
 # run infercnv on celltypes, running all individuals together
-# TODO: decide on cluster_by_groups 
+# TODO: decide on cluster_by_groups value
 # arg definition from inferCNV - If observations are defined according to groups 
 # (ie. patients), each group of cells will be clustered separately. 
 # (default = FALSE, instead will use k_obs_groups setting)
-nextflow run nextflow/infercnv.nf \
-  --out_dir out/Zhang2023/by_celltype_run_all/ \
-  --mappings data/Zhang2023/stromal_mappings_run_all.csv \
-  --annotations data/Zhang2023/annotations.tsv \
-  --annotation_col "celltype" \
-  --analysis_mode "samples" \
-  --cluster_by_groups "FALSE" \
+# run infercnv on fibroblast clusters, running all individuals together
+
+# run infercnv on all fibroblasts, running all individuals together
+mkdir out/Zhang2023/all_fibroblasts/
+
+# generate annotations, filter to only fibroblasts
+cat data/Zhang2023/annotations.tsv |
+awk -F'\t' -v OFS="\t" \
+  '{if ((NR == 1) || (($2 == "Stromal cell") && $3 ~ /^F-/)) {print $1,$2,$3,"all"} }' \
+> out/Zhang2023/all_fibroblasts/annotations.tsv
+
+# generate mappings for all
+echo -e "id,raw_counts_matrix\nall,out/Zhang2023/sce/counts.rds" \
+> out/Zhang2023/all_fibroblasts/mappings.csv
+
+nextflow run nf-infercnv \
+  --out_dir out/Zhang2023/all_fibroblasts/ \
+  --mappings out/Zhang2023/all_fibroblasts/mappings.csv \
+  --annotations out/Zhang2023/all_fibroblasts/annotations.tsv \
+  --annotation_col all \
+  --analysis_mode subclusters \
+  --cluster_by_groups FALSE \
   -c config/infercnv.config \
[…]

### 2024-07-09 infercnv/reports/gains_and_losses.Rmd
fix annotations col

--- 2024-03-11 infercnv/reports/gains_and_losses.Rmd
@@
 author: "Alexandra Tidd"
-date: "`r format(Sys.time(), '%d %B, %Y')`"
+date: "`r format(Sys.time(), "%d %B, %Y")`"
 output:
@@
 
-```{r setup, include = F, message = F, warning = F, class.source = 'fold-hide'}
-# wd <- ifelse(Sys.info()['nodename'] == "mib119104s", '~/Volumes/', '') |> paste0('/lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/') ; setwd(wd)
-# rmarkdown::render('reports/infercnv.Rmd', output_file = 'driver_coverage.html', output_dir = 'reports/', params = list(rerun = T))
+```{r setup, include = F, message = F, warning = F, class.source = "fold-hide"}
+# rmarkdown::render("reports/gains_and_losses.Rmd", output_file = "gains_and_losses.html", output_dir = "reports/", params = list(rerun = T))
 
@@
 # chunk options
-knitr::opts_knit$set(root.dir = '../')
-knitr::opts_chunk$set(
-  warning = FALSE, 
-  dpi = 300, 
-  message = FALSE,
+knitr::opts_knit$set(root.dir = "../")
+knitr::opts_chunk$set(warning = FALSE, dpi = 300, message = FALSE,
                       cache.path = params$cache_dir)
@@
 
-```{r load_sce, results = 'asis', class.source = 'fold-show'}
+```{r load_sce, results = "asis", class.source = "fold-show"}
 # load gene coords
 coords <-
-  readr::read_tsv('data/hg38_gencode_v27.txt', col_names = c('name', 'chr', 'start', 'stop'))
+  "data/hg38_gencode_v27.txt" %>%
+  readr::read_tsv(col_names = c("name", "chr", "start", "stop"))
 
@@
 ctap <-
-  readr::read_tsv('data/Zhang2023/syn26710600/CTAP_donor_mapping.tsv') %>%
+  "data/Zhang2023/syn26710600/CTAP_donor_mapping.tsv" %>%
+  readr::read_tsv() %>%
   dplyr::select(sample = donor, CTAP)
 md <-
-  readr::read_tsv('data/Zhang2023/syn26710600/SYNAPSE_METADATA_MANIFEST.tsv') %>%
+  "data/Zhang2023/syn26710600/SYNAPSE_METADATA_MANIFEST.tsv" %>%
+  readr::read_tsv() %>%
   dplyr::distinct()
@@
   cnvs <-
-    list.files(
-      params$infercnv_dir,
-      pattern = 'map_metadata_from_infercnv.txt',
-      recursive = T) %>%
+    list.files(params$infercnv_dir, pattern = "map_metadata_from_infercnv.txt",
+               recursive = TRUE) %>%
     purrr::set_names(., dirname(.)) %>%
@@
       file.path(params$infercnv_dir, file) %>%
-        read.table(sep = '\t')
+        read.table(sep = "\t")
     }) %>%
     dplyr::bind_rows() %>%
-    dplyr::mutate(infercnv = T)
+    dplyr::mutate(infercnv = TRUE)
 
   # left join cnvs to coldata, reorder, restore rownames
-  col_data <- merge(sce@colData, cnvs, by = 'row.names', all.x = T, sort = F)
+  col_data <- merge(sce@colData, cnvs, by = "row.names", all.x = TRUE,
+                    sort = FALSE)
   rownames(col_data) <- col_data[, 1]
@@
 
-}, file = 'sce.rds', rerun = params$rerun)
+}, file = "sce.rds", rerun = params$rerun)
 
@@
 
-```{r echo = F, results = 'asis'}
+```{r echo = F, results = "asis"}
 tibble::tribble(
   ~name, ~`type - value`,
-  '`has_*`', 'binary - whether any CNV / loss / duplication is found within the given chr',
-  '`proportion_*`', 'proportion - proportion of genes that are part of any CNV / loss / duplication within the given chr',
-  '`proportion_scaled_*`', 'proportion - same as proportion_*, but taking into account whether genes that are duplicated / lost are a single copy or two copies, as a weight',
-  '`top_loss_*`', 'the top CNVs that affect the most genes'
+  "`has_*`", "binary - whether any CNV / loss / duplication is found within the given chr",
+  "`proportion_*`", "proportion - proportion of genes that are part of any CNV / loss / duplication within the given chr",
+  "`proportion_scaled_*`", "proportion - same as proportion_*, but taking into account whether genes that are duplicated / lost are a single copy or two copies, as a weight",
+  "`top_loss_*`", "the top CNVs that affect the most genes"
 ) %>%
@@
       params$infercnv_dir,
-      pattern = 'HMM_CNV_predictions.HMMi6.leiden.hmm_mode-subclusters.Pnorm_0.5.pred_cnv_regions.dat',
+      pattern = "HMM_CNV_predictions.HMMi6.leiden.hmm_mode-subclusters.Pnorm_0.5.pred_cnv_regions.dat",
       recursive = T) %>%
@@
       }) %>%
-    dplyr::bind_rows(.id = 'sample') %>%
+    dplyr::bind_rows(.id = "sample") %>%
     # refactor chromosomes
     dplyr::mutate(
-      chr = chr %>% {forcats::fct_reorder(., gsub('chr', '', .) %>% as.numeric())},
+      chr = chr %>% {forcats::fct_reorder(., gsub("chr", "", .) %>% as.numeric())},
       direction = dplyr::case_when(
-        state < 3 ~ 'loss',
-        state > 3 ~ 'dupli',
-        TRUE ~ 'neutral'),
+        state < 3 ~ "loss",
+        state > 3 ~ "dupli",
+        TRUE ~ "neutral"),
       ploidy = dplyr::case_when(
-        state == 1 ~ '0x',
-        state == 2 ~ '0.5x',
-        state == 3 ~ '1x',
-        state == 4 ~ '1.5x',
-        state == 5 ~ '2x',
-        state == 6 ~ '>2x',
+        state == 1 ~ "0x",
+        state == 2 ~ "0.5x",
+        state == 3 ~ "1x",
+        state == 4 ~ "1.5x",
+        state == 5 ~ "2x",
+        state == 6 ~ ">2x",
         TRUE ~ NA
         ) %>% forcats::fct_reorder(state),
-      region = gsub('.*region_', '', cnv_name) %>% as.numeric(),
-      cell_group = cell_group_name %>% gsub('\\..*', '', .))
+      region = gsub(".*region_", "", cnv_name) %>% as.numeric(),
+      cell_group = cell_group_name %>% gsub("\\..*", "", .))
 
@@
 
-}, file = 'cnv_regions.rds', rerun = params$rerun)
+}, file = "cnv_regions.rds", rerun = params$rerun)
 ```
@@
   ggplot(aes(x = midpoint, fill = ploidy, group = ploidy)) +
-  geom_density(data = cnv_regions %>% dplyr::filter(direction == 'dupli'),
+  geom_density(data = cnv_regions %>% dplyr::filter(direction == "dupli"),
                aes(y = ..density..), 
@@
                colour = NA) +
-  geom_density(data = cnv_regions %>% dplyr::filter(direction == 'loss'),
+  geom_density(data = cnv_regions %>% dplyr::filter(direction == "loss"),
                aes(y = -..density..), 
@@
                colour = NA) +
-  ggh4x::facet_grid2(chr ~ cell_group, scales = 'free_x', independent = 'x') +
+  ggh4x::facet_grid2(chr ~ cell_group, scales = "free_x", independent = "x") +
   theme_bw() +
@@
   scale_fill_manual(
-    values = c('0x' = '#00008B', '0.5x' = '#b4b0ff', '1' = 'white', 
-               '1.5x' = '#ffc2c2', '2x' = '#d64242', '>2x' = '#8B0000')) 
+    values = c("0x" = "#00008B", "0.5x" = "#b4b0ff", "1" = "white", 
+               "1.5x" = "#ffc2c2", "2x" = "#d64242", ">2x" = "#8B0000")) 
 ```
@@
             axis.title.x = element_blank()) +
-      ylab('count'),
+      ylab("proportion"),
     plot_func(p_dat, position_stack(reverse = T)) +
-      ylab('proportion')) %>%
+      ylab("count")) %>%
     patchwork::wrap_plots(ncol = 1) 
@@
 
-## Assigning whole chromosome-level events
+# Assigning whole chromosome-level events
 
-Based on the observed distribution of `proportion_dupli_chr7` values and comparing this with the `inferCNV` heatmaps, whole chromosome gains are defined as `proportion_dupli_chr*` > `r params$min_prop_for_cnv`.
-Diploidy is defined as `proportion_dupli_chr*` < `r params$max_prop_for_wt`.
+Based on the observed distribution of `proportion_dupli_chr7` values and 
+comparing this with the `inferCNV` heatmaps, whole chromosome gains are defined 
+as `proportion_dupli_chr*` > `r params$min_prop_for_cnv`. Diploidy is defined as 
+`proportion_dupli_chr*` < `r params$max_prop_for_wt`.
 
-```{r assign_gains, class.source = 'fold-show'}
+```{r assign_gains, class.source = "fold-show"}
 sce <- xfun::cache_rds({
@@
         sce@colData[, paste0("proportion_loss_", chr)]
-      sce@colData[, paste0(chr, '_loss')] <<-
+      sce@colData[, paste0(chr, "_loss")] <<-
         ifelse(chr_loss > params$min_prop_for_cnv, 1,

### 2024-07-09 infercnv/reports/paper_figures.Rmd
fix annotations col

@@
 ```{r setup, include = F, message = F, warning = F, class.source = "fold-hide"}
-# module load R/4.4.0 ; Rscript -e "rmarkdown::render("reports/paper_figures.Rmd", output_dir = "out/paper_figures/")" # nolint: line_length_linter.
+# module load R/4.4.0 ; Rscript -e 'rmarkdown::render("reports/paper_figures.Rmd", output_dir = "out/paper_figures/")' # nolint: line_length_linter.
 
@@
 
-1. fewer cells makes it less likely to capture these somewhat rare events
+1. having fewer cells makes it less likely to capture these somewhat rare events
 
-2. fewer cells makes the reference profile noisier and less representative of 
-the 'baseline' expression levels across chromosomes, further limiting CNV 
-calling
+2. having fewer cells makes the reference profile noisier and less
+representative of the 'baseline' expression levels across chromosomes, further
+limiting CNV calling
 
@@
 
-## Cell type compositions
+# Plot 1 - chr7, chr5, and chr3 barplots by sample
 
-```{r ct_comp, fig.width = 12, fig.height = 7}
-p_dat <-
+```{r}
+p <-
   sce@colData %>%
   tibble::as_tibble() %>%
-  dplyr::mutate(
-    sample = forcats::fct_infreq(sample),
-    cell_group = forcats::fct_rev(forcats::fct_infreq(cell_group))) %>%
-  dplyr::add_count(sample, cell_group, name = "n_sample_x_cell_group") %>%
-  dplyr::add_count(cell_group, name = "n_cell_group")
-p_base <-
-  p_dat %>%
-  ggplot(aes(fill = cell_group, colour = cell_group)) +
-  scale_fill_manual(values = dittoSeq::dittoColors()) +
-  scale_colour_manual(values = dittoSeq::dittoColors())
-p1 <- 
-  p_base +
-  geom_point(aes(x = sample, y = cell_group,
-                 alpha = n_sample_x_cell_group, size = n_sample_x_cell_group)) +
+  dplyr::add_count(sample) %>%
+  tidyr::pivot_longer(cols = c("chr7_gain", "chr5_gain", "chr3_gain")) %>%
+  dplyr::filter(!is.na(value)) %>%
+  dplyr::mutate(value = as.character(value)) %>%
+  ggplot(aes(x = reorder(sample, -n), fill = value)) +
+  scale_x_discrete(guide = guide_axis(angle = -90)) +
   theme_classic() +
-  theme(legend.position = "none",
-        axis.text.x = element_text(size = 7)) +
-  scale_x_discrete(guide = guide_axis(angle = -90)) +
-  labs(y = params$cell_group_name)
-p2 <- 
-  p_base +
-  geom_bar(aes(x = sample), position = "stack") +
-  theme_void() +
-  theme(legend.position = "none",
-        axis.text.y = element_text()) 
-p3 <-
-  p_base +
-  geom_bar(aes(y = cell_group), position = "stack") +
-  geom_text(
-    data = p_dat %>% dplyr::distinct(cell_group, n_cell_group),
-    aes(x = n_cell_group, y = cell_group, 
-        label = format(n_cell_group, nsmall = 1, big.mark = ",")), 
-    hjust = -0.1) +
-  theme_void() +
-  theme(legend.position = "none") +
-  xlim(0, 1.2 * max(p_dat$n_cell_group))
-design <-
-"BB#
-BB#
-AAC"
-
-# plot
-patchwork::wrap_plots(A = p1, B = p2, C = p3, design = design)
+  scale_fill_brewer(palette = "Dark2") +
+  facet_grid(name ~ .)
+p + geom_bar() + labs(x = "sample", y = "n cells")
+p + geom_bar(position = "fill") + labs(x = "sample", y = "proportion of cells")
 ```
 
-## UMAPs
+# Plot 2 - chr7, chr5, and chr3 barplots by celltype
 
-First, we plot UMAPs by cell grouping.
-
-```{r plot_umaps}
-dittoSeq::dittoDimPlot(sce, params$cell_group_name, size = 0.2) +
-  coord_fixed()
-
-pal <-
-  c("T cell" = "#E6AB02",
-    "Stromal cell" = "#66A61E",
-    "Myeloid cell" = "#E7298A",
-    "B cell/plasma cell" = "#7570B3",
-    "Endothelial cell" = "#D95F02",
-    "NK" = "#1B9E77")
-dittoSeq::dittoDimPlot(sce, "cell_type", size = 0.2) +
-  coord_fixed() +
-  scale_colour_manual(values = pal)
+```{r}
+cnvs <-
+  "out/Zhang2023/by_celltype/summary/gains_and_losses_cache/" %>%
+  list.files(pattern = "cnvs", full.names = TRUE) %>%
+  readRDS()
+p <-
+  cnvs %>%
+  dplyr::filter(event == "dupli", measure == "proportion",
+                (value < 0.5 | value > 0.8),
+                chr %in% c("chr7", "chr5", "chr3")) %>%
+  dplyr::mutate(
+    value = dplyr::case_when(value <= 0.5 ~ 0, value >= 0.8 ~ 1) %>% 
+      as.character()) %>%
+  ggplot(aes(x = cell_type, fill = value)) +
+  theme_classic() +
+  facet_grid(chr ~ .) +
+  scale_fill_brewer(palette = "Dark2") +
+  scale_x_discrete(guide = guide_axis(angle = -45))
+p + geom_bar()
+p + geom_bar(position = "fill") + ylab("proportion")
 ```
 
-Then, we visualise gains and loss proportions for each chromosome on the UMAP.
+# Plot 3 - number of transcripts per cell vs CNVs
 
-```{r plot_chr_gains, fig.width = 10, fig.height = 4, results = "hide"}
-paste0("chr", 1:22) %>% 
-  purrr::map(function(chr) {
-    events <-
-      paste("proportion", c("dupli", "loss"), chr, sep = "_")
-    reducedDim(sce, "UMAP_uwot") %>%
-      cbind(sce@colData[, events]) %>%
-      tibble::as_tibble() %>%
-      tidyr::pivot_longer(3:4) %>%
-      dplyr::arrange(value) %>%
-      ggplot(aes(x = UMAP1, y = UMAP2, colour = value)) + 
-      geom_point(size = 0.1) +
-      coord_fixed() +
-      theme_classic() +
-      scale_color_viridis_c() +
-      facet_grid(~ name) %>%
-      print()
-  })
+Next, we look at the distribution of the the number of transcripts in cells in 
+which chr7 gain has been detected versus those without. If the distribution is 
+greatly skewed in one or the other, this would suggest that this technical 
+variate is strongly affecting the performance of `inferCNV` between cells.
+
+```{r}
+colSums(sce@assays@data$counts) %>%
+  tibble::enframe(name = "cell", value = "n_transcripts") %>%
+  dplyr::inner_join(sce@colData %>% tibble::as_tibble()) %>%
+  dplyr::filter(!is.na(chr7_gain)) %>%
+  dplyr::mutate(chr7_gain = as.character(chr7_gain)) %>%
+  ggplot(aes(x = n_transcripts, fill = chr7_gain)) +
+  geom_histogram(alpha = 0.5) +
+  theme_classic()
 ```
 
-## Heatmaps
+The distributions appear to be sensible.
 
-Now, we will plot gain and loss proportions by chromosome.
-
-First, we plot single cell-level CNVs by direction and by celltype.
-
-```{r plot_gains_and_losses_by_ct, results = "hide"}
-# at single-cell level
-sce@colData[, params$cell_group_name] %>%
-  unique() %>%
-  purrr::map(function(cgn) {
-    c("dupli", "loss") %>%
-      purrr::map(function(direction) {
-        p_dat <-
-          sce@colData[sce@colData[, params$cell_group_name] == cgn, ]
-        if (nrow(p_dat) > 0) {
-          p_mat <-
-            p_dat[, grepl(paste0("proportion_", direction), colnames(p_dat))]
-          p_annot <- 
-            p_dat[, c("sample"), drop = F] %>%
-            as.data.frame()
-          pheatmap::pheatmap(
-            p_mat, 
-            show_rownames = F, 
-            annotation_row = p_annot,
-            # hclust error: size cannot be NA nor exceed 65536
-            cluster_rows = ((nrow(p_mat) < 65536) == T),
-            legend = F,
-            main = paste0(params$cell_group_name, ": ", cgn, "\ndirection: ", direction)) %>%
-            print()
-        }
-      })
-  })
-```
-
-We also plot sample-x-celltype level events (mean-summarised).
-
-```{r plot_gains_and_losses_by_id, fig.width = 12}
-# mean proportion loss / gain per chromosome per celltype per id
-p_dat <-
-  sce@colData %>%
-  tibble::as_tibble() %>%
-  dplyr::mutate(cell_group = get(params$cell_group_name)) %>%
-  dplyr::group_by(sample, cell_group) %>%
-  dplyr::summarise(
-    dplyr::across(dplyr::starts_with("proportion_"), ~ mean(.x, na.rm = T))) %>%
-  dplyr::ungroup() %>%
-  dplyr::mutate(sample_x_cg = paste0(sample, "_", cell_group)) %>%
-  tibble::column_to_rownames("sample_x_cg")
-
-# split gain and loss
-c("dupli", "loss") %>% 
-  purrr::map(function(direction) {
-    p_mat <-
-      p_dat[, grepl(paste0("proportion_", direction), colnames(p_dat))]
-    p_annot <- 
-      p_dat[, c("sample", "cell_group")] %>%
-      as.data.frame()
-    pheatmap::pheatmap(
-      p_mat, 
-      show_rownames = F, 
-      annotation_row = p_annot,
-      main = paste0("direction: ", direction)) %>%
-      print()
-  })
-```
-
-## chr7 gain
-
-We are most interested in the chr7 gain events in fibroblasts. We visualise their distribution across the dataset.
-
-```{r chr7_prop_dist}
-# prep data
-p_dat <-
-  sce@colData %>%
-  tibble::as_tibble() %>%
-  dplyr::group_by(sample) %>%
-  dplyr::mutate(n_cells = dplyr::n())
-
-# define functions
-plot_prop_chr7_gain <- function(p_dat, title) {
-  plot_func <- function(p_dat, position_arg) {
-    p_dat %>%
-      ggplot() +
-      geom_bar(aes(x = cell_group, 
-                   fill = proportion_dupli_chr7, 
-                   group = proportion_dupli_chr7),
-               position = position_arg) +
-      scale_x_discrete(guide = guide_axis(angle = -45)) +
-      theme_classic() +
-      scale_fill_viridis_b(n.breaks = 9, limits = c(0, 1)) 
-  }
-
-  # plot both count (stack) and proportion (fill), annotate n cells
-  list(
-    plot_func(p_dat, position_fill(reverse = T)) +
-      ggtitle(title) +
-      theme(axis.text.x = element_blank(),
-            axis.title.x = element_blank()) +
-      ylab("count"),
-    plot_func(p_dat, position_stack(reverse = T)) +
-      ylab("proportion")) %>%
-    patchwork::wrap_plots(ncol = 1) 
-}
-
-# plot, across all and by patient
-plot_prop_chr7_gain(p_dat, "chr7 gain - all cells")
-p_dat %>%
-  {split(., .$sample)} %>%
-  purrr::walk2(names(.), ., function(id, df) {
-    df %>%
-    plot_prop_chr7_gain(
-      paste0("chr7 gains - ", id, " (", unique(df$n_cells), " cells)")) %>%
-      print()
-  })
-```
-
-# Assigning whole chromosome-level events
-
-Based on the observed distribution of `proportion_dupli_chr7` values and comparing this with the `inferCNV` heatmaps, whole chromosome gains are defined as `proportion_dupli_chr*` > `r params$min_prop_for_cnv`.
-Diploidy is defined as `proportion_dupli_chr*` < `r params$max_prop_for_wt`.
-
-```{r assign_gains, class.source = "fold-show"}
-sce <- xfun::cache_rds({
-
-  # apply cut-offs for whole chr gains and losses
-  paste0("chr", 1:22) %>%
-    purrr::walk(function(chr) {
-      chr_dupli <-
-        sce@colData[, paste0("proportion_dupli_", chr)]
-      sce@colData[, paste0(chr, "_gain")] <<-
-        ifelse(chr_dupli > params$min_prop_for_cnv, 1, 
-               ifelse(chr_dupli < params$max_prop_for_wt, 0, 
-                      NA))
-      chr_loss <-
-        sce@colData[, paste0("proportion_loss_", chr)]
-      sce@colData[, paste0(chr, "_loss")] <<-
-        ifelse(chr_loss > params$min_prop_for_cnv, 1, 
-               ifelse(chr_loss < params$max_prop_for_wt, 0, 
-                      NA))
-    })
-
-  # return
-  sce
-
-}, file = "sce_w_gains.rds", rerun = params$rerun)
-```
-
-```{r plot_heatmap, fig.width = 10, fig.height = 5}
-event_cols <-
-  c("WT" = "grey",
-    "chr5_gain" = "#e7298a",
-    "chr5_gain,chr7_gain" = "#d95f02",
-    "chr7_gain" = "#e6ab02")
-
-p_dat <-
-  sce@colData %>%
-  tibble::as_tibble() %>%
-  dplyr::select(sample, cell, 
-                dplyr::ends_with("_gain"), 
-                dplyr::ends_with("_loss")) %>%
-  tidyr::pivot_longer(-c("sample", "cell")) %>%
-  dplyr::filter(name %in% c("chr7_gain", "chr5_gain")) %>%
-  dplyr::group_by(sample, cell) %>%
-  # remove cells in which no calls made
-  dplyr::filter(!all(is.na(value))) %>%
-  dplyr::mutate(
-    event = dplyr::case_when(
-      all(value == 0 | is.na(value)) ~ "WT",
-      value == 1 ~ name,
-      TRUE ~ NA)) %>%
-  dplyr::summarise(
-    events = paste(sort(unique(na.omit(event))), collapse = ",")) %>%
-  dplyr::mutate(events = forcats::fct_relevel(events, names(event_cols))) %>%
-  # get n aneuploid for plot ordering
-  dplyr::group_by(sample) %>%
-  dplyr::mutate(prop_aneuploid = sum(events != "WT") / dplyr::n()) %>%
-  # count
-  dplyr::group_by(sample, prop_aneuploid) %>%
-  dplyr::count(events) %>%
-  # for fill and stack
-  tidyr::crossing(tibble::tibble(plot = c("% cells", "n cells")))
-
-# plot
-p_dat %>%
-  ggplot(aes(x = reorder(sample, -prop_aneuploid), y = n, fill = events)) +
-  geom_col(data = p_dat %>% dplyr::filter(plot == "% cells"),
-           position = "fill") +
-  geom_col(data = p_dat %>% dplyr::filter(plot == "n cells"),
-           position = "stack") +
-  theme_classic() +
-  ggh4x::facet_grid2(plot ~ ., scales = "free_y", independent = "y") +
-  scale_x_discrete(guide = guide_axis(angle = -90)) +
-  scale_y_continuous(expand = c(0, 0)) +
-  scale_fill_manual(values = event_cols) +
-  theme(axis.title = element_blank(),
-        axis.text.x = element_text(size = 6))
-```
-
-## Co-occurence of CNV events
-
-We also can observe the rate co-occurence of chr7+ and other CNVs. The LCM-WES data suggested a nested relationship in which there were chr5+ subclones within chr7+ populations. We look for this here.
-
-```{r chr7_and_others}
-events <-
-  sce@colData %>%
-  tibble::as_tibble() %>%
-  dplyr::select(sample, cell, cell_group, 
-                dplyr::ends_with("_gain"), 
-                dplyr::ends_with("_loss")) %>%
-  tidyr::pivot_longer(-c("sample", "cell", "cell_group")) %>%
-  dplyr::group_by(sample, cell, cell_group) %>%
-  # remove cells in which no calls made
-  dplyr::filter(!all(is.na(value))) %>%
-  dplyr::mutate(
-    event = dplyr::case_when(
-      all(value == 0 | is.na(value)) ~ "WT",
-      value == 1 ~ name,
-      TRUE ~ NA)) %>%
-  dplyr::summarise(
-    events = paste(sort(unique(na.omit(event))), collapse = ","),
-    n_events = sum(value, na.rm = T))
-
-p_dat <-
-  events %>%
-  dplyr::ungroup() %>%
-  dplyr::add_count(events) %>%
-  dplyr::mutate(prop = n / dplyr::n()) %>%
-  # ignore events that happen in <= 25 cells
-  dplyr::filter(events != "WT", n >= 25) %>%
-  dplyr::mutate(
-    events = events %>% 
-      gsub("chr", "", .) %>% 
-      gsub("_gain", "+", .) %>%
-      gsub("_loss", "-", .),
-    events = forcats::fct_infreq(events),
-    label1 = scales::label_comma()(n), 
-    label = dplyr::case_when(
-      n < 10 ~ label1,
-      TRUE ~ paste0(label1, " (", round(prop*100, 2), "%)")
-    ))
-
-# plot
-p <-
-  p_dat %>%
-  ggplot(aes(x = events)) +
-  geom_text(aes(y = n,  label = label), 
-            vjust = 0.5, hjust = 0, angle = 90, size = 2) +
-  scale_x_discrete(guide = guide_axis(angle = -90)) +
-  ylim(c(0, max(p_dat$n) * 1.1)) +
-  theme_classic() +
-  theme(legend.position = "bottom")
-p + 
-  geom_bar(aes(fill = sample))
-p + 
-  geom_bar(aes(fill = cell_group)) +
-  theme(legend.position = "none")
-```
-
-### Fisher exact test
-
-Here is a contingency table of the frequency of chr7+ versus chr5+ within cells.
-
-```{r cont_tab}
-cont <-
-  sce@colData %>%
-  tibble::as_tibble() %>%
-  dplyr::mutate(chr7 = ifelse(chr7_gain == 1, "chr7+", "chr7_WT"),
-                chr5 = ifelse(chr5_gain == 1, "chr5+", "chr5_WT")) 
-table(cont$chr7, cont$chr5) %>%
-  knitr::kable()
-```
-
-We can run a Fisher exact test to test whether chr7+ and chr5+ gains co-occur more often than would be expected by chance, and also test all other co-occurent CNV relationships.
-
-```{r fisher_exact}
-fisher <- xfun::cache_rds({
-
-  fisher <-
-    colnames(sce@colData)[grepl("_gain$|_loss$", colnames(sce@colData))] %>%
-    purrr::set_names() %>%
-    purrr::map(function(event) {
-
-      event_vec <- sce@colData[, event, drop = T]
-
-      if (1 %in% event_vec) {
-
-        # run fisher exact test
-        fisher <- fisher.test(sce$chr7_gain, event_vec)
-
-        # get contingencies
-        cont <-
-          table(sce@colData[, c("chr7_gain", event)])
-
-         # return relevant stats
-        tibble::tibble(
-          OR = fisher$estimate, 
-          p = fisher$p.value,
-          sig = ifelse(fisher$p.value < 0.05, "*", ""),
-          n_WT = cont[1, 1],
-          n_chr7_gain = cont[2, 1],
-          n_event = cont[1, 2],
-          n_both = cont[2, 2])
-
-      } else {
-
-        # no event
-        tibble::tibble()
-      }
-
-    }) %>%
-    dplyr::bind_rows(.id = "event") %>%
-    # remove chr7+ (can't do chr7+ vs chr7+)
-    dplyr::filter(event != "chr7_gain", n_event > 0) %>%
-    dplyr::arrange(p)
-
-  # return
-  fisher
-
-}, file = "fisher.rds", rerun = T)
-
-fisher %>%
-  dplyr::mutate(p = format(p, digits = 3)) %>%
-  knitr::kable()
-```
-
-### GLM
-
-We can also run a GLM.
-
-```{r glm}
-# read in metadata
-metadata <-
-  readr::read_tsv(params$metadata)
-
-# create glm
-# prop_chr7_gain ~ age (+ coverage) (per donor)
-for_glm <-
-  dplyr::inner_join(
-    tibble::as_tibble(sce@colData), 
-    metadata, 
-    by = c("sample" = "id")) %>%
-  dplyr::filter(!is.na(age), !is.na(chr7_gain)) %>%
-  dplyr::group_by(sample, age, CTAP, sex, race, ethnicity) %>%
-  dplyr::summarise(n_chr7_gain = sum(chr7_gain),
-                   n_cells = dplyr::n(),
-                   ch7_gain_prop = n_chr7_gain / n_cells)
-
-for_glm %>%
-  {glm(.$n_chr7_gain ~ log(.$n_cells) + .$age)}
-```
-
-```{r plot_glm_vars}
-# n cells vs n chr7 gain
-for_glm %>%
-  ggplot(aes(x = n_cells, y = n_chr7_gain)) +
-  geom_point() +
-  theme_classic()
-
-# age vs prop chr7 gain
-for_glm %>%
-  ggplot(aes(x = age, y = ch7_gain_prop)) +
-  geom_point() +
-  theme_classic()
-
-# all categorical variables
-for_glm %>%
-  tidyr::pivot_longer(cols = c("CTAP", "sex", "race", "ethnicity")) %>%
-  dplyr::group_by(name, value) %>%
-  dplyr::mutate(med = median(ch7_gain_prop)) %>%
-  ggplot(aes(
-    x = tidytext::reorder_within(value, by = med, within = name), 
-    y = ch7_gain_prop)) +
-  geom_boxplot() +
-  geom_jitter(height = 0) +
-  facet_wrap(~ name, nrow = 1, scales = "free") +
-  tidytext::scale_x_reordered(guide = guide_axis(angle = -45)) +
-  theme_classic() +
-  labs(x = "")
-
-# looking by cluster
-for_glm_w_cluster <-
-  dplyr::inner_join(
-    tibble::as_tibble(sce@colData), 
-    metadata, 
-    by = c("sample" = "id")) %>%
-  dplyr::filter(!is.na(age), !is.na(chr7_gain)) %>%
-  dplyr::group_by(
-    sample, age, CTAP, sex, race, ethnicity, cluster_name) %>%
-  dplyr::summarise(n_chr7_gain = sum(chr7_gain),
-                   n_cells = dplyr::n(),
-                   ch7_gain_prop = n_chr7_gain / n_cells) 
-for_glm_w_cluster %>%
-  ggplot(aes(x = cluster_name, y = ch7_gain_prop)) +
-  geom_violin() +
-  geom_jitter(height = 0) +
-  theme_classic() +
-  scale_x_discrete(guide = guide_axis(angle = -45))
-```
-
-# Pseudobulking and DGEA
-
-We pseudobulk chr7+ cells and perform differential expression analysis.
-In order to do this, we use the `glmGamPoi` package in R.
-This analysis follows a [tutorial](https://bioconductor.org/packages/devel/bioc/vignettes/glmGamPoi/inst/doc/pseudobulk.html) and the [glmGamPoi GitHub](https://github.com/const-ae/glmGamPoi).
-
-We will pseudobulk by patient, cell group, and chr7 ploidy to produce a reduced object.
-
-```{r chr7_gain_prop, results = "asis"}
-dat <-
-  tibble::tibble(chr7_gain_value = sce$chr7_gain,
-               cell_group = sce$cell_group) %>%
-  dplyr::mutate(
-    chr7_gain_status = dplyr::case_when(
-      is.na(chr7_gain_value) ~ "unknown/intermediate",
-      chr7_gain_value == 0 ~ "WT",
-      chr7_gain_value == 1 ~ "gain")
-  )
-
-cat("Counts and proportions of chr7 gain, all cells")
-dat %>%
-  dplyr::mutate(n_cells = dplyr::n()) %>%
-  dplyr::group_by(chr7_gain_status, n_cells) %>%
-  dplyr::summarise(n = dplyr::n(),
-                   `%` = 100 * dplyr::n() / unique(n_cells)) %>%
-  knitr::kable()
-
-cat("Counts and proportions of chr7 gain, by cell group")
-dat %>%
-  dplyr::group_by(cell_group) %>%
-  dplyr::mutate(n_cells = dplyr::n()) %>%
-  dplyr::group_by(cell_group, chr7_gain_status, n_cells) %>%
-  dplyr::summarise(n = dplyr::n(),
-                   `%` = 100 * dplyr::n() / unique(n_cells)) %>%
-  knitr::kable()
-```
-
-```{r pseudobulk_chr7_gain, class.source = "fold-show"}
-reduced_sce <- xfun::cache_rds({
-
-  # pseudobulk the sce object by patient, cell group and chr7 ploidy
-  # (exclude intermediate chr7 values)
-  reduced_sce <-
-    glmGamPoi::pseudobulk(
-      sce[, !is.na(sce$sample) & 
-            !is.na(sce$chr7_gain) &
-            !is.na(sce$cell_group)], 
-      group_by = vars(sample, cell_group, chr7_gain)
-    )
-
-  # return
-  reduced_sce
-
-}, file = "reduced_sce.rds", rerun = params$rerun)
-```
-
-With the reduced data, we can conduct differential expression analysis, treating the data like bulk RNA-seq data.
-
-We will identify which genes in stromal cells are changed most by chr7 gain. We do this by fitting one GLM model for each gene, with the chr7 status, cell group and patient as factors.
-
-```{r deseq, message = T}
-# fit one GLM model for each gene
-fit <- xfun::cache_rds({
-  fit <-
-    glmGamPoi::glm_gp(
-      reduced_sce,
-      design = ~ chr7_gain + sample + cell_group,
-      verbose = T
-    )
-  fit
-}, file = "fit.rds", rerun = params$rerun)
-
-# show model output
-summary(fit)
-```
-
-Next, we test for differential gene expression.
-
-```{r test_ge}
-# minimum accepted adj_pval
-min_p <- 0.05
-
-# get differential expression
-de_res <- xfun::cache_rds({
-
-  # test differences
-  de_res <-
-    glmGamPoi::test_de(
-      fit, 
-      contrast = 
-        cond(chr7_gain = 1) -
-        cond(chr7_gain = 0)
-    )
-
-  # define significant up/downregulation and position
-  de_res <-
-    de_res %>% 
-    tibble::as_tibble() %>%
-    dplyr::mutate(
-      direction = dplyr::case_when(
-        lfc < -1.5 & adj_pval < min_p ~ "downregulated",
-        lfc > 1.5 & adj_pval < min_p ~ "upregulated",
-        TRUE ~ NA)) %>%
-    dplyr::left_join(coords)
-
-  # return
-  de_res
-
-}, file = "de_res.rds", rerun = params$rerun)
-```
-
-```{r plot_de}
-de_res %>%
-  ggplot(aes(x = lfc, y = -log10(adj_pval), colour = direction, 
-             size = chr == "chr7", shape = chr == "chr7")) +
-  geom_point() +
-  geom_hline(yintercept = -log10(min_p), linetype = "dashed") +
-  geom_vline(xintercept = c(log2(1.5), log2(2)), linetype = "dashed", colour = "purple") +
-  geom_vline(xintercept = c(-1.5, 1.5), linetype = "dashed") +
-  ggrepel::geom_text_repel(
-    data = 
-      de_res %>% 
-      dplyr::filter(!is.na(direction)) %>%
-      dplyr::group_by(direction) %>%
-      dplyr::filter(rank(adj_pval) <= 10 | rank(lfc, ties.method = "min") <= 10),
-    aes(label = name),
-    min.segment.length = 0, nudge_x = 0.7, show.legend = F) +
-  theme_classic() +
-  scale_colour_manual(values = c("upregulated" = "darkred",
-                                 "downregulated" = "darkblue"), 
-                      na.value = "darkgrey") +
-  scale_size_manual(values = c("TRUE" = 3, "FALSE" = 2), na.value = 2) +
-  scale_shape_manual(values = c("TRUE" = 55, "FALSE" = 16), na.value = 16)
-```
-
-```{r plot_de_close}
-min_p_non_chr7 <-
-  de_res %>%
-  dplyr::filter(!is.na(direction)) %>%
-  dplyr::pull(adj_pval) %>% 
-  min()
-de_res %>%
-  ggplot(aes(x = lfc, y = -log10(adj_pval), colour = direction, 
-             size = chr == "chr7", shape = chr == "chr7")) +
-  geom_point() +
-  geom_hline(yintercept = -log10(min_p), linetype = "dashed") +
-  geom_vline(xintercept = c(log2(1.5), log2(2)), linetype = "dashed", colour = "purple") +
-  geom_vline(xintercept = c(-1.5, 1.5), linetype = "dashed") +
-  ggrepel::geom_text_repel(
-    data = 
-      de_res %>% 
-      dplyr::filter(!is.na(direction)) %>%
-      dplyr::group_by(direction) %>%
-      dplyr::filter(rank(adj_pval) <= 10 | rank(lfc, ties.method = "min") <= 10),
-    aes(label = name),
-    min.segment.length = 0, nudge_x = 0.7, show.legend = F) +
-  theme_classic() +
-  scale_colour_manual(values = c("upregulated" = "darkred",
-                                 "downregulated" = "darkblue"), 
-                      na.value = "darkgrey") +
-  scale_size_manual(values = c("TRUE" = 5, "FALSE" = 4), na.value = 4) +
-  scale_shape_manual(values = c("TRUE" = 55, "FALSE" = 16), na.value = 16) +
-  ylim(0, -log10(min_p_non_chr7) * 1.03)
-```
-
-These are the most strongly upregulated genes:
-
-```{r get_downreg_degs}
-de_res %>%
-  tibble::as_tibble() %>% 
-  dplyr::filter(adj_pval < 0.01) %>%
-  dplyr::arrange(-lfc, adj_pval) %>%
-  head(10) %>%
-  knitr::kable()
-```
-
-These are the most strongly downregulated genes:
-
-```{r get_upreg_degs}
-de_res %>%
-  tibble::as_tibble() %>% 
-  dplyr::filter(adj_pval < 0.01) %>%
-  dplyr::arrange(lfc, adj_pval) %>%
-  head(10) %>%
-  knitr::kable()
-```
-
-# Figures for the paper
-
-```{r n_fibroblasts_dist}
-
-```
-
-# Cell group markers
-
-We convert the `SingleCellExperiment` object to a `Seurat` object in order to find the top markers of each cell group.
-
-```{r exit}
-knitr::knit_exit()
-```
-
-```{r sce_to_seu}
-
-seu <- xfun::cache_rds({
-  seu <- Seurat::as.Seurat(sce)
-  Seurat::Idents(seu) <- seu$cell_group
-  seu <- Seurat::ScaleData(seu)
-  seu
-}, file = "seu.rds", rerun = params$rerun)
-```
-
-Now, we find all differentially expressed genes for each cell group.
-
-```{r markers}
-markers <- xfun::cache_rds({
-  markers <-
-    Seurat::FindAllMarkers(seu)
-  markers
-}, file = "markers.rds", rerun = params$rerun)
-```
-
-```{r inspect_markers, fig.height = 10, fig.width = 10}
-top2 <-
-  markers %>%
-  tibble::as_tibble() %>% 
-  dplyr::filter(p_val_adj < 0.01, avg_log2FC > 1.5) %>% 
-  dplyr::group_by(cluster) %>% 
-  dplyr::slice_max(order_by = avg_log2FC, n = 2)
-unique(top2$cluster) %>%
-  purrr::walk(function(cg) {
-    genes <- 
-      top2 %>% 
-      dplyr::filter(cluster == cg) %>% 
-      dplyr::pull(gene)
-    list(
-      Seurat::VlnPlot(seu, features = genes),
-      Seurat::FeaturePlot(seu, features = genes, label = T) + coord_fixed()) %>%
-      patchwork::wrap_plots(nrow = 2) +
-      patchwork::plot_annotation(
-        title = paste0(params$cell_group_name, ": ", cg)) %>%
-      print()
-  })
-```
-
-```{r marker_heatmap, fig.height = 10, fig.width = 10}
-top10 <-
-  markers %>%
-  dplyr::group_by(cluster) %>%
-  dplyr::filter(avg_log2FC > 1) %>%
-  dplyr::slice_max(order_by = avg_log2FC, n = 100) %>%
-  dplyr::ungroup()
-Seurat::DoHeatmap(seu, features = top10$gene) + 
-  Seurat::NoLegend()
-```
+# Plot 3 - inferCNV plot with all individuals together

### 2024-07-09 infercnv/src/01_run_infercnv.sh

@@ -17,7 +17,7 @@ mkdir out/Zhang2023/all_fibroblasts/
 # generate annotations, filter to only fibroblasts
 cat data/Zhang2023/annotations.tsv |
 awk -F'\t' -v OFS="\t" \
-  '{if ((NR == 1) || (($2 == "Stromal cell") && $3 ~ /^F-/)) {print $1,$2,$3,"all"} }' \
+  '{if (NR == 1) {print $1,$2,$3,"id"} ; if(($2 == "Stromal cell") && $3 ~ /^F-/) {print $1,$2,$3,"all"} }' \
 > out/Zhang2023/all_fibroblasts/annotations.tsv
 
 # generate mappings for all
@@ -28,7 +28,7 @@ nextflow run nf-infercnv \
   --out_dir out/Zhang2023/all_fibroblasts/ \
   --mappings out/Zhang2023/all_fibroblasts/mappings.csv \
   --annotations out/Zhang2023/all_fibroblasts/annotations.tsv \
-  --annotation_col all \
+  --annotation_col id \
   --analysis_mode subclusters \
   --cluster_by_groups FALSE \
   -c config/infercnv.config \

### 2024-07-30 infercnv/reports/paper_figures.Rmd
add params, deal with NULL-> NA conversions
fix channel issues

@@
 
-# Plot 3 - inferCNV plot with all individuals together
+# Plot 4 - inferCNV plot with all individuals together
+
+I am currently trying to run this. I have run into some issues with `inferCNV`
+dependencies breaking on farm22, so this has been on hold for the last 2 weeks.
+
+# Plot 5 - Mutation hotspots
+
+We load Luke's object. The `nano_muts` are the targeted mutations. The 
+`exome_muts` is a list of significant drivers in each cell type (dNdScv output).
+
+```{r}
+# load the metadata
+sm <- readr::read_tsv("/lustre/scratch125/casm/team268im/lh22/Rheumatoid_Arthritis_Paper/Data/MetaData/sample_meta.tsv")
+
+# load the muts and convert to list
+load("/lustre/scratch125/casm/team268im/lh22/Rheumatoid_Arthritis_Paper/Figures/Figure_2/LCM_NanoSeq_Overlap/GOI/pileup_muts.rda")
+muts <- list(exome = exome_muts,
+             LCM = LCM_muts,
+             targeted = nano_muts,
+             pileup = pileup_muts)
+
+# genes of interest
+goi <- "NF1"
+
+# get annotations from dndscv output
+mut_annots <-
+  muts$exome %>%
+  tibble::as_tibble() %>%
+  dplyr::select(chr, pos, gene, ref, mut, type, tri, impact, aachange)
+
+muts$targeted %>%
+  tibble::as_tibble() %>%
+  dplyr::select()
+```

### 2024-07-30 infercnv/src/01_run_infercnv.sh

@@ -11,8 +11,54 @@ module load singularity
 # (default = FALSE, instead will use k_obs_groups setting)
 # run infercnv on fibroblast clusters, running all individuals together
 
+# run infercnv by celltype
+mkdir -p out/Zhang2023/by_celltype_test/
+
+# generate annotations, differentiate mural cells and fibroblasts within stromal compartment
+cat data/Zhang2023/annotations.tsv |
+awk -F'\t' '{
+    if (NR == 1) {
+      print $1, $2, "id"
+    } else if ($2 == "Stromal cell") {
+        if ($3 ~ /^F-/) {
+            print $1, "Fibroblast", $4
+        } else if ($3 ~ /^M-/) {
+            print $1, "Mural cell", $4
+        } 
+    } else {
+        print $1, $2, $4
+    }
+}' OFS='\t' \
+> out/Zhang2023/by_celltype_test/annotations.tsv
+
+# generate mappings for all
+cat data/Zhang2023/mappings.csv |
+awk -F, '{ if (NR==1) { print } ; if (NR>1) {print $1",out/Zhang2023/sce/counts.rds"}}' \
+> out/Zhang2023/by_celltype_test/mappings.csv
+
+# run infercnv
+nextflow run nf-infercnv \
+  --out_dir out/Zhang2023/by_celltype_test/ \
+  --mappings out/Zhang2023/by_celltype_test/mappings.csv \
+  --annotations out/Zhang2023/by_celltype_test/annotations.tsv \
+  --annotation_col celltype \
+  --analysis_mode subclusters \
+  --cluster_by_groups FALSE \
+  --cutoff 0.1 \
+  --window_length 151 \
+  --noise_logistic FALSE \
+  --denoise TRUE \
+  --HMM TRUE \
+  --HMM_transition_prob 0.000001 \
+  --HMM_report_by subcluster \
+  --sd_amplifier 0.65 \
+  --useRaster FALSE \
+  -c config/infercnv.config \
+  -resume \
+  -N at31@sanger.ac.uk
+
 # run infercnv on all fibroblasts, running all individuals together
-mkdir out/Zhang2023/all_fibroblasts/
+mkdir -p out/Zhang2023/all_fibroblasts/
 
 # generate annotations, filter to only fibroblasts
 cat data/Zhang2023/annotations.tsv |
[…]

### 2024-07-30 infercnv/reports/paper_figures.Rmd
add lolliplotting
change ref files from channel to file, split infercnv run script

@@
 ```{r}
+lh22_dir <- "/lustre/scratch125/casm/team268im/lh22/Rheumatoid_Arthritis_Paper/"
+
 # load the metadata
-sm <- readr::read_tsv("/lustre/scratch125/casm/team268im/lh22/Rheumatoid_Arthritis_Paper/Data/MetaData/sample_meta.tsv")
+sm <- readr::read_tsv(file.path(lh22_dir, "Data/MetaData/sample_meta.tsv"))
 
 # load the muts and convert to list
-load("/lustre/scratch125/casm/team268im/lh22/Rheumatoid_Arthritis_Paper/Figures/Figure_2/LCM_NanoSeq_Overlap/GOI/pileup_muts.rda")
+load(file.path(lh22_dir, "Figures/Figure_2/LCM_NanoSeq_Overlap/GOI",
+               "pileup_muts.rda"))
 muts <- list(exome = exome_muts,
@@
              pileup = pileup_muts)
+
+# lollipopPlot
 

### 2024-07-30 infercnv/src/01_run_infercnv_all_fibroblasts.sh

@@ -1,8 +1,5 @@
 #!/bin/bash
-# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/ ; bsub -q week -M 2000 -R 'select[mem>=2000] span[hosts=1] rusage[mem=2000]' -J infercnv -o log "bash src/01_run_infercnv.sh"
-
-# get singularity module
-module load singularity
+# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/ ; bsub -q week -M 2000 -R 'select[mem>=2000] span[hosts=1] rusage[mem=2000]' -J infercnv_all -o log "bash src/01_run_infercnv.sh"
 
 # run infercnv on celltypes, running all individuals together
 # TODO: decide on cluster_by_groups value
@@ -11,51 +8,8 @@ module load singularity
 # (default = FALSE, instead will use k_obs_groups setting)
 # run infercnv on fibroblast clusters, running all individuals together
 
-# run infercnv by celltype
-mkdir -p out/Zhang2023/by_celltype_test/
-
-# generate annotations, differentiate mural cells and fibroblasts within stromal compartment
-cat data/Zhang2023/annotations.tsv |
-awk -F'\t' '{
-    if (NR == 1) {
-      print $1, $2, "id"
-    } else if ($2 == "Stromal cell") {
-        if ($3 ~ /^F-/) {
-            print $1, "Fibroblast", $4
-        } else if ($3 ~ /^M-/) {
-            print $1, "Mural cell", $4
-        } 
-    } else {
-        print $1, $2, $4
-    }
-}' OFS='\t' \
-> out/Zhang2023/by_celltype_test/annotations.tsv
-
-# generate mappings for all
-cat data/Zhang2023/mappings.csv |
-awk -F, '{ if (NR==1) { print } ; if (NR>1) {print $1",out/Zhang2023/sce/counts.rds"}}' \
-> out/Zhang2023/by_celltype_test/mappings.csv
-
-# run infercnv
-nextflow run nf-infercnv \
-  --out_dir out/Zhang2023/by_celltype_test/ \
-  --mappings out/Zhang2023/by_celltype_test/mappings.csv \
-  --annotations out/Zhang2023/by_celltype_test/annotations.tsv \
-  --annotation_col celltype \
-  --analysis_mode subclusters \
-  --cluster_by_groups FALSE \
-  --cutoff 0.1 \
-  --window_length 151 \
[…]

### 2024-07-30 infercnv/src/02_run_infercnv_by_celltype.sh

@@ -1,0 +1,51 @@
+#!/bin/bash
+# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/ ; bsub -q week -M 2000 -R 'select[mem>=2000] span[hosts=1] rusage[mem=2000]' -J infercnv_by_celltype -o log "bash src/01_run_infercnv.sh"
+
+# get singularity module
+module load singularity
+
+# run infercnv by celltype
+mkdir -p out/Zhang2023/by_celltype_test/
+
+# generate annotations, differentiate mural cells and fibroblasts within stromal compartment
+cat data/Zhang2023/annotations.tsv |
+awk -F'\t' '{
+    if (NR == 1) {
+      print $1, $2, "id"
+    } else if ($2 == "Stromal cell") {
+        if ($3 ~ /^F-/) {
+            print $1, "Fibroblast", $4
+        } else if ($3 ~ /^M-/) {
+            print $1, "Mural cell", $4
+        } 
+    } else {
+        print $1, $2, $4
+    }
+}' OFS='\t' \
+> out/Zhang2023/by_celltype_test/annotations.tsv
+
+# generate mappings for all
+cat data/Zhang2023/mappings.csv |
+awk -F, '{ if (NR==1) { print } ; if (NR>1) {print $1",out/Zhang2023/sce/counts.rds"}}' \
+> out/Zhang2023/by_celltype_test/mappings.csv
+
+# run infercnv
+nextflow run nf-infercnv \
+  --out_dir out/Zhang2023/by_celltype_test/ \
+  --mappings out/Zhang2023/by_celltype_test/mappings.csv \
+  --annotations out/Zhang2023/by_celltype_test/annotations.tsv \
+  --annotation_col celltype \
+  --analysis_mode subclusters \
+  --cluster_by_groups FALSE \
+  --cutoff 0.1 \
+  --window_length 151 \
+  --noise_logistic FALSE \
+  --denoise TRUE \
+  --HMM TRUE \
+  --HMM_transition_prob 0.000001 \
+  --HMM_report_by subcluster \
+  --sd_amplifier 0.65 \
+  --useRaster FALSE \
+  -c config/infercnv.config \
+  -resume \
+  -N at31@sanger.ac.uk

### 2024-08-04 infercnv/reports/paper_figures.Rmd
reorder scripts, trying pdf to help plot.new() issue

--- 2024-07-09 infercnv/reports/paper_figures.Rmd
@@
 
-# Plot 3 - inferCNV plot with all individuals together
+# Plot 4 - inferCNV plot with all individuals together
+
+I am currently trying to run this. I have run into some issues with `inferCNV`
+dependencies breaking on farm22, so this has been on hold for the last 2 weeks.
+
+# Plot 5 - Mutation hotspots
+
+We load Luke's object. The `nano_muts` are the targeted mutations. The 
+`exome_muts` is a list of significant drivers in each cell type (dNdScv output).
+
+```{r}
+# add library path
+.libPaths(c("/lustre/scratch125/casm/team268im/at31/.R/lib/4.2.2/", .libPaths()))
+
+# libraries
+library(GenomicRanges)
+library(trackViewer)
+library(TxDb.Hsapiens.UCSC.hg19.knownGene)  # package for gene model
+library(org.Hs.eg.db)                       # package for gene name
+
+# dirs
+lh22_dir <- "/lustre/scratch125/casm/team268im/lh22/Rheumatoid_Arthritis_Paper/"
+
+# load the metadata
+sm <- readr::read_tsv(file.path(lh22_dir, "Data/MetaData/sample_meta.tsv"))
+
+# load the muts
+load(file.path(lh22_dir, "Figures/Figure_2/LCM_NanoSeq_Overlap/GOI",
+               "pileup_muts.rda"))
+# muts <- list(exome = exome_muts,
+#              LCM = LCM_muts,
+#              targeted = nano_muts,
+#              pileup = pileup_muts)
+
+# get NF1 mutations, convert to granges
+
+# get mutations, convert to granges
+gr_ls <-
+  nano_muts %>%
+  tibble::as_tibble() %>%
+  dplyr::add_count(mut_site) %>%
+  {
+    GenomicRanges::GRanges(
+      seqnames = paste0("chr", .$chr),
+      ranges = IRanges::IRanges(start = .$pos,
+                                width = 1,
+                                names = .$mut_site),
+      strand = ifelse(.$strand == 1, "+", "-"),
+      gene = .$gene, score = .$n)
+  } %>%
+  {split(., .$gene)}
+
+# get transcripts in the range
+trs <-
+  geneModelFromTxdb(TxDb.Hsapiens.UCSC.hg19.knownGene,
+                    org.Hs.eg.db, gr = gr_ls$NF1)
+
+entrezID <- get("NF1", org.Hs.egSYMBOL2EG)
+theTrack <- geneTrack(entrezID, TxDb.Hsapiens.UCSC.hg19.knownGene)[[1]]
+viewerStyle <- trackViewerStyle()
+setTrackViewerStyleParam(viewerStyle, "margin", c(.1, .05, .02, .02))
+trackList <- trackList(trs)
+vp <- viewTracks(trackList,
+                 gr = gr, viewerStyle = viewerStyle,
+                 autoOptimizeStyle = TRUE)
+
+pdf("test.pdf")
+vp
+dev.off()
+
+# select transcripts of interest
+features <- c(range(trs[[1]]$dat), range(trs[[2]]$dat))
+names(features) <- c(trs[[1]]$name, trs[[2]]$name)
+features$fill <- c("lightblue", "mistyrose")
+
+# lollipop
+pdf("test.pdf", width = 30)
+lolliplot(gr_ls$NF1, features,
+          ranges = GRanges("chr17", IRanges(29421945, 29704695, names = "NF1")), 
+          type = "circle")
+dev.off()
+
+# get genes from biomart
+txdb <- TxDb.Hsapiens.UCSC.hg19.knownGene
+broads <- GenomicFeatures::genes(txdb)
+seqlevels(broads) <- gsub("chr", "", seqlevels(broads))
+
+# get gene annotations from broads
+feats <- broads[seqnames(broads) %in% seqlevels(gr)]
+findOverlaps(gr, broads)
+
+feats <-
+  GenomicRanges::GRanges("12", IRanges::IRanges(start = 6702320, end = 6702330),
+                         strand = "+", gene = "NF1")
+
+trackViewer::lolliplot(SNP.gr = head(gr, 1), features = feats)
+```

### 2024-08-04 infercnv/src/01_run_infercnv_all_fibroblasts.sh

@@ -1,5 +1,5 @@
 #!/bin/bash
-# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/ ; bsub -q week -M 2000 -R 'select[mem>=2000] span[hosts=1] rusage[mem=2000]' -J infercnv_all -o log "bash src/01_run_infercnv.sh"
+# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/ ; bsub -q week -M 2000 -R 'select[mem>=2000] span[hosts=1] rusage[mem=2000]' -J infercnv_all -o log "bash src/01_run_infercnv_all_fibroblasts.sh"
 
 # run infercnv on celltypes, running all individuals together
 # TODO: decide on cluster_by_groups value
@@ -12,7 +12,8 @@
 module load singularity
 
 # run infercnv on all fibroblasts, running all individuals together
-mkdir -p out/Zhang2023/all_fibroblasts/
+# must partition work subdirectories so concomitant jobs don't cause lock issues
+mkdir -p out/Zhang2023/all_fibroblasts/ work/all_fibroblasts/
 
 # generate annotations, filter to only fibroblasts
 cat data/Zhang2023/annotations.tsv |
@@ -29,11 +30,12 @@ echo -e "id,raw_counts_matrix\nall,out/Zhang2023/sce/counts.rds" \
 # (/lustre/scratch126/casm/team268im/mp34/analysis/synovium_scRNA/infercnv_analysis.R)
 nextflow run nf-infercnv \
   --out_dir out/Zhang2023/all_fibroblasts/ \
+  -w work/all_fibroblasts/ \
   --mappings out/Zhang2023/all_fibroblasts/mappings.csv \
   --annotations out/Zhang2023/all_fibroblasts/annotations.tsv \
   --annotation_col cluster \
   --analysis_mode subclusters \
-  --cluster_by_groups FALSE \
+  --cluster_by_groups TRUE \
   --cutoff 0.1 \
   --window_length 151 \
   --noise_logistic FALSE \
@@ -43,6 +45,7 @@ nextflow run nf-infercnv \
   --HMM_report_by subcluster \
   --sd_amplifier 0.65 \
   --useRaster FALSE \
+  --output_format pdf \
   -c config/infercnv.config \
   -resume \
   -N at31@sanger.ac.uk

### 2024-08-04 infercnv/src/02_run_infercnv_by_celltype.sh

@@ -1,11 +1,11 @@
 #!/bin/bash
-# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/ ; bsub -q week -M 2000 -R 'select[mem>=2000] span[hosts=1] rusage[mem=2000]' -J infercnv_by_celltype -o log "bash src/01_run_infercnv.sh"
+# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/ ; bsub -q week -M 2000 -R 'select[mem>=2000] span[hosts=1] rusage[mem=2000]' -J infercnv_by_celltype -o log "bash src/02_run_infercnv_by_celltype.sh"
 
 # get singularity module
 module load singularity
 
 # run infercnv by celltype
-mkdir -p out/Zhang2023/by_celltype_test/
+mkdir -p out/Zhang2023/by_celltype_test/ 
 
 # generate annotations, differentiate mural cells and fibroblasts within stromal compartment
 cat data/Zhang2023/annotations.tsv |
@@ -36,7 +36,7 @@ nextflow run nf-infercnv \
   --annotations out/Zhang2023/by_celltype_test/annotations.tsv \
   --annotation_col celltype \
   --analysis_mode subclusters \
-  --cluster_by_groups FALSE \
+  --cluster_by_groups TRUE \
   --cutoff 0.1 \
   --window_length 151 \
   --noise_logistic FALSE \

### 2024-08-07 infercnv/reports/paper_figures.Rmd
trying different lolliplot methods

--- 2024-07-30 infercnv/reports/paper_figures.Rmd
@@
 ```
+
+# Datasets
+
+## LCM NanoSeq
 
@@
 ```{r}
+# add library path
+.libPaths(c("/lustre/scratch125/casm/team268im/at31/.R/lib/4.2.2/", .libPaths()))
+
+# libraries
+library(magrittr)
+library(GenomicRanges)
+library(trackViewer)
+library(TxDb.Hsapiens.UCSC.hg19.knownGene)  # package for gene model
+library(org.Hs.eg.db)                       # package for gene name
+library(biomaRt)
+library(httr)                               # package for REST API
+
+# dirs
 lh22_dir <- "/lustre/scratch125/casm/team268im/lh22/Rheumatoid_Arthritis_Paper/"
@@
 
-# load the muts and convert to list
+# load the muts (targeted mutations in the nano_muts object)
 load(file.path(lh22_dir, "Figures/Figure_2/LCM_NanoSeq_Overlap/GOI",
                "pileup_muts.rda"))
-muts <- list(exome = exome_muts,
-             LCM = LCM_muts,
-             targeted = nano_muts,
-             pileup = pileup_muts)
+nano_muts <-
+  nano_muts %>%
+  tibble::as_tibble()
 
-# lollipopPlot
+# get mutations, convert to granges
+gr_muts_ls <-
+  nano_muts %>%
+  tibble::as_tibble() %>%
+  dplyr::add_count(mut_site) %>%
+  {
+    GenomicRanges::GRanges(
+      seqnames = paste0("chr", .$chr),
+      ranges = IRanges::IRanges(start = .$pos,
+                                width = 1,
+                                names = .$mut_site),
+      strand = ifelse(.$strand == 1, "+", "-"),
+      gene = .$gene, score = .$n)
+  } %>%
+  {split(., .$gene)}
 
-# genes of interest
-goi <- "NF1"
+# get the refcds
+data("refcds_hg19", package = "dndscv")
+names(RefCDS) <- RefCDS %>% purrr::map(~ .$gene_name)
 
-# get annotations from dndscv output
-mut_annots <-
-  muts$exome %>%
+# write urls for proteinpaint
+
+# write custom input file for proteinpaint
+nano_muts %>%
+  dplyr::filter(gene == "NF1") %>%
+  dplyr::transmute(
+    aachange, 
+    aapos = as.numeric(stringr::str_sub(aachange, 2, -2)),
+    impact = dplyr::case_when(impact == "Missense" ~ "M",
+                              impact == "Nonsense" ~ "N",
+                              impact == "Synonymous" ~ "S",
+                              impact == "no-SNV" & type == "del" ~ "D",
+                              impact == "no-SNV" & type == "ins" ~ "I",
+                              TRUE ~ NA),
+    donor) %>%
+  dplyr::filter(!is.na(aapos)) %>%
+  readr::write_csv("protein_paint.csv")
+
+# write custom input file for mutationmapper
+nano_muts %>%
+  dplyr::filter(gene == "NF1") %>%
+  dplyr::transmute(
+    Chromosome = chr, Start_Position = pos - 1, End_Position = pos,
+    Reference_Allele = ref, Variant_Allele = mut, Hugo_Symbol = gene
+  ) %>%
+  readr::write_tsv("mutation_mapper.tsv")
+
+# get the gene
+g <- "NF1"
+
+# retrieve refcds element for gene
+x <- RefCDS[[g]]
+
+# get the position of each base along the exons
+# RefCDS is 1-based - must convert to 0-based for calculations, then
+# convert back to 1-based for plotting
+exons <-
+  tibble::tibble(start = x$intervals_cds[, 1] - 1,
+                 end = x$intervals_cds[, 2]) %>%
+  dplyr::mutate(dplyr::across(everything(), 
+                              ~ .x - (x$intervals_cds[1, 1] - 1),
+                              .names = "rel_{.col}"),
+                exon = dplyr::row_number(),
+                exon_length = end - start,
+                cds_end = cumsum(exon_length),
+                cds_start = dplyr::lag(cds_end, default = 0),
+                # convert starts to 1-based for plotting
+                dplyr::across(dplyr::contains("start"),
+                              ~ .x + 1))
+
+# expand to get every position
+exons_exp <-
+  exons %>%
+  dplyr::group_by(rel_start, rel_end, exon, exon_length) %>%
+  dplyr::reframe(pos = start:end, cds_pos = cds_start:cds_end)
+
+# load interpro entries
+interpro2go <-
+  readLines("data/interpro/interpro2go.txt") %>%
   tibble::as_tibble() %>%
-  dplyr::select(chr, pos, gene, ref, mut, type, tri, impact, aachange)
+  dplyr::filter(!grepl("^!", value)) %>%
+  tidyr::separate_wider_delim("value", delim = " ",
+                              names = c("interpro_accession", "rest"),
+                              too_many = "merge") %>%
+  tidyr::separate(rest, sep = " > GO:| ; GO:",
+                  into = c("domain_name", "GO_name", "GO_id")) %>%
+  dplyr::mutate(interpro_accession = gsub("InterPro:", "", interpro_accession))
+interpro <-
+  readr::read_tsv("data/interpro/entry.list") %>%
+  dplyr::rename(interpro_accession = ENTRY_AC)
 
-muts$targeted %>%
-  tibble::as_tibble() %>%
-  dplyr::select()
+# retrieve protein domain annotations
+library(ensembldb)
+library(EnsDb.Hsapiens.v75)
+edb <- EnsDb.Hsapiens.v75
+domains <-
+  transcripts(edb, filter = GeneNameFilter(g),
+              columns = c("protein_id", "tx_biotype", "uniprot_id",
+                          "protein_domain_id", "protein_domain_source",
+                          "prot_dom_start", "prot_dom_end",
+                          "interpro_accession"))
+domains_exp <-
+  txs %>%
+  dplyr::as_tibble() %>%
+  dplyr::left_join(interpro2go) %>%
+  dplyr::filter(tx_biotype == "protein_coding",
+                !is.na(interpro_accession), !is.na(domain_name)) %>%
+  dplyr::mutate(dom_start = start + prot_dom_start,
+                dom_end = start + prot_dom_end) %>%
+  dplyr::group_by(tx_id, protein_id, protein_domain_id, protein_domain_source,
+                  prot_dom_start, prot_dom_end, interpro_accession,
+                  domain_name, GO_name) %>%
+  dplyr::reframe(pos = dom_start:dom_end)
+
+# combine all exon and domain annotations
+annots <-
+  exons_exp %>%
+  dplyr::full_join(domains_exp)
+
+# add exon annots to muts
+impact_palette <-
+  c("Synonymous" = "#b3b3b3",
+    "Missense" = "#5f9ea0",
+    "Nonsense" = "#68218a")
+p_dat <-
+  nano_muts %>%
+  dplyr::group_by(chr, pos, impact) %>%
+  dplyr::count() %>%
+  dplyr::right_join(annots) %>%
+  dplyr::mutate(alternate = ifelse(exon %% 2 == 0, "e", "o"),
+                impact = factor(impact, levels = names(impact_palette)))
+
+# plot
+pdf("test.pdf", width = 30)
+# mutations
+p_dat %>%
+  ggplot(aes(x = cds_pos, y = n, fill = impact, colour = impact)) +
+  geom_col(width = 1) +
+  geom_point() +
+  theme_classic() +
+  scale_fill_manual(values = impact_palette,
+                    na.value = "white", na.translate = FALSE) +
+  scale_colour_manual(values = impact_palette,
+                      na.value = "white", na.translate = FALSE)
+# exon bounds
+p_dat %>%
+  dplyr::group_by(exon, alternate) %>%
+  dplyr::summarise(xmin = min(cds_pos), xmax = max(cds_pos)) %>%
+  ggplot(aes(xmin = xmin, xmax = xmax, ymin = -1, ymax = 1,
+             fill = alternate)) +
+  geom_hline(yintercept = 0, colour = "black") +
+  geom_rect(colour = "black") +
+  theme_void() +
+  theme(axis.text.x = element_text()) +
+  scale_fill_manual(values = c("e" = "#e3b3b2",
+                               "o" = "#cd6f6e"))
+# protein domains
+p_dat %>%
+  dplyr::filter(!is.na(protein_domain_id)) %>%
+  dplyr::group_by(tx_id, domain_name, GO_name, protein_domain_id) %>%
+  dplyr::summarise(xmin = min(cds_pos), xmax = max(cds_pos)) %>%
+  ggplot(aes(xmin = xmin, xmax = xmax, ymin = -1, ymax = 1,
+             fill = GO_name)) +
+  geom_hline(yintercept = 0, colour = "black") +
+  geom_rect(colour = "black") +
+  theme_void() +
+  scale_fill_discrete(na.value = "white", na.translate = FALSE) +
+  facet_grid(tx_id ~ .)
+
+dev.off()
+
+# get gene coords (GRCh37)
+ensembl <- useEnsembl(biomart = "ensembl", dataset = "hsapiens_gene_ensembl",
+                      GRCh = 37)
+bm <-
+  getBM(attributes = c("chromosome_name", "start_position",
+                       "end_position", "strand", "hgnc_symbol"),
+        filters = c("hgnc_symbol"),
+        values = list(names(gr_muts_ls)),
+        mart = ensembl)
+gr_genes <-
+  GenomicRanges::GRanges(
+    seqnames = paste0("chr", bm$chromosome_name),
+    ranges = IRanges::IRanges(start = bm$start_position,
+                              end = bm$end_position,
+                              names = bm$hgnc_symbol),
+    strand = ifelse(bm$strand == 1, "+", "-"),
+    gene = bm$hgnc_symbol)
+gr_genes_ls <- split(gr_genes, gr_genes$gene)
+
+# get transcripts per gene
+nf1_trs <-
+  geneModelFromTxdb(TxDb.Hsapiens.UCSC.hg19.knownGene,
+                    org.Hs.eg.db, gr = gr_genes_ls$NF1) %>%
+  purrr::map(~ .$dat) %>%
+  {Reduce(c, .)} %>%
+  subset(symbol == "NF1" & feature == "CDS")
+nf1_trs_fill <-
+  tibble::tibble(transcript = unique(nf1_trs$transcript),
+                 fill = rainbow(dplyr::n_distinct(nf1_trs$transcript))) %>%
+  tibble::column_to_rownames("transcript")
+nf1_trs$fill <-  nf1_trs_fill[nf1_trs$transcript, ]
+nf1_trs$featureLayerID <- nf1_trs$transcript
+names(nf1_trs) <- nf1_trs$transcript
+
+# get protein domains
+APIurl <- "https://www.ebi.ac.uk/proteins/api/" # base URL of the API
+taxid <- "9606" # human tax ID
+gene <- "NF1" # target gene
+orgDB <- "org.Hs.eg.db" # org database to get the uniprot accession id
+eid <- mget(gene, get(sub(".db", "SYMBOL2EG", orgDB)))[[1]]
+chr <- mget(eid, get(sub(".db", "CHR", orgDB)))[[1]]
+accession <- unlist(lapply(eid, function(.ele){
+  mget(.ele, get(sub(".db", "UNIPROT", orgDB)))
+  }))
+featureURL <- paste0(APIurl,
+                      "features?offset=0&size=-1&reviewed=true",
+                      "&types=DNA_BIND%2CMOTIF%2CDOMAIN",
+                      "&taxid=", taxid,
+                      "&accession=", paste(accession, collapse = "%2C"))
+response <- GET(featureURL)
+content <- httr::content(response)[[1]]
+acc <- content$accession
+sequence <- content$sequence
+gr <- GRanges(chr, IRanges(1, nchar(sequence)))
+domains <- do.call(rbind, content$features)
+domains <- GRanges(chr, IRanges(as.numeric(domains[, "begin"]),
+                                as.numeric(domains[, "end"]),
+                                names = domains[, "description"]))
+names(domains)[1] <- "DNA_BIND" ## this is hard coding.
+domains$fill <- 1 + seq_along(domains)
+domains$height <- 0.04
+## GET variations. This part can be replaced by user-defined data.
+variationURL <- paste0(APIurl,
+                        "variation?offset=0&size=-1",
+                        "&sourcetype=uniprot&dbtype=dbSNP",
+                        "&taxid=", taxid,
+                        "&accession=", acc)
+response <- GET(variationURL)
+content <- httr::content(response)[[1]]
+keep <- sapply(content$features, function(.ele) length(.ele$evidences)>2 && # filter the data by at least 2 evidences
+                  !grepl("Unclassified", .ele$clinicalSignificances)) # filter the data by classified clinical significances.
+nkeep <- c("wildType", "alternativeSequence", "begin", "end",
+            "somaticStatus", "consequenceType", "score")
+content$features <- lapply(content$features[keep], function(.ele){
+  .ele$score <- length(.ele$evidences)
+  unlist(.ele[nkeep]) 
+})
+variation <- do.call(rbind, content$features)
+variation <-
+  GRanges(chr,
+          IRanges(as.numeric(variation[, "begin"]),
+                  width = 1,
+                  names = paste0(variation[, "wildType"],
+                                  variation[, "begin"],
+                                  variation[, "alternativeSequence"])),
+          score = as.numeric(variation[, "score"]),
+          color = as.numeric(factor(variation[, "consequenceType"]))+1)
+variation$label.parameter.gp <- gpar(cex=.5)
+seqlevels(variation) <- paste0("chr", seqlevels(variation))
+seqlevels(domains) <- paste0("chr", seqlevels(domains))
+
+domains %>%
+  tibble::as_tibble()
+
+pdf("test.pdf")
+lolliplot(gr_muts_ls$NF1, domains, nf1_trs)
+dev.off()
+
+# plot transcripts per gene
+pdf("test.pdf", width = 20)
+lolliplot(gr_muts_ls$NF1, nf1_trs)
+dev.off()
 ```
+
+```{r}
+load(file="/lustre/scratch125/casm/team268im/lh22/Rheumatoid_Arthritis_Paper/Figures/Figure_1/GOI.rda")
+goi <- list(exome_genes = exome_genes, goi_exome = goi_exome, 
+            goi_targeted = goi_targeted, targeted_genes = targeted_genes)
+
+nanoseqVarsPath <- "nano_muts.tsv"
+```

### 2024-08-07 infercnv/src/nb15_nanoseqLollipopGenePlots_targ_nano_grch37_031023.Rmd
trying different lolliplot methods

---
title: "nanoseqLollipopGenePlots"
author: "Matt Neville"
date: '2022-07-12'
output: html_document
---

# 1. Prep
## Paths/Libraries
```{r setup}
#Input files on the farm
dndsPath <- "~/Mounts/peter/liver/analysis/targeted_liver_panel/"
refCDSpath <- "~/Mounts/peter/liver/code/dNdScv/RefCDS_GRCh37_vF.Rdat"
lof_gof_path <- "~/Mounts/peter/liver/data/NanoSeq/lof_gof38.tsv"
nanoseqVarsPath <- "~/Mounts/peter/liver//analysis/targeted_liver_panel/combined_muts_annot.tsv"

#Local output
# outputPath <- "/Volumes/GoogleDrive/My Drive/Sperm/lolliplot/"
outputPath <- "/Users/nb15/Mounts/peter/liver/analysis/targeted_liver_panel/lolliplots/coverage/"

library(Biostrings)
library(GenomicRanges)
library(MetBrewer)
library(egg)
library(plyr)
library(tidyverse)
```

## Load Datasets
```{r loadDatasets}

## load dnds hg37 refCDS object from dNdS -> to get coding sequence intervals
load(refCDSpath)

# Katrina's clinvar annotated LOF/GOF list
lof_gof <- read_tsv(lof_gof_path, col_types = cols()) %>%
  dplyr::rename(mut = alt)

## dataframe that holds list number in RefCDS object for each gene; so you can easily extract your gene of interest cds intervals
genes <- as.data.frame(matrix(nrow=length(RefCDS), ncol=2))
colnames(genes) <- c("gene_name","row_in_refcds")
for( i in 1:length(RefCDS)){
  genes$gene_name[i] <- RefCDS[[i]]$gene_name
  genes$row_in_refcds[i] <- i
}

## load dnds object to annotate muts and colour by impact
annotmuts <- readRDS(paste0(dndsPath, "dnds_all_dc.rds"))$annotmuts

## merge nanoseq_data with dnds annotations and clinvar annotations
nanoseq_data <- read_tsv(nanoseqVarsPath, col_types = cols()) %>% 
  dplyr::select(sampleID, chr, pos, ref, mut, duplex_cov) %>% 
  mutate(sampleID = as.character(sampleID)) %>%
  mutate(chr = gsub("chr","",chr)) %>%
  left_join(annotmuts,by=c("sampleID","chr","pos","ref","mut")) %>%  
  left_join(lof_gof,by=c("chr","pos","ref","mut")) %>%  
  mutate(label = if_else(fxn == "altered_function", "G", "")) %>%
  mutate(label = if_else(fxn == "LOF", paste0(label,"L"), label)) %>%
  # rename cols to match nanoseq_data colnames
  dplyr::rename(CHROM = chr,
         POS = pos,
         REF = ref,
         ALT = mut) %>% 
  #Change label based on clinvar Katrina annotation
  # mutate(label = if_else(fxn == "altered_function", "G", "")) %>% 
  # mutate(label = if_else(fxn == "LOF", paste0(label,"L"), label)) %>% 
  mutate(impact = replace_na(impact, "Non-coding")) %>% 
  mutate(impact = str_replace(impact, "no-SNV", "Indel"))  %>% 
  mutate(name = if_else(impact == "Indel", 
                        #Formating for indels
                        replace_na(str_remove(ntchange, ".*-"), ".") %>% 
                          str_replace("frshift", "FS") %>% 
                          str_replace("inframe", "IF"), 
                        #Name for SNVs
                        replace_na(aachange, ".")))

## define genes of interest by dNdS sig genes or by custom list or put in gene names (example below)
#goi <- readRDS(paste0(dndsPath, "spermTargdNdS.RDS"))$sel_cv %>% 
  # dplyr::filter(qglobal_cv < 0.05 | qallsubs_cv < 0.05) %>% 
#  pull(gene_name)

#goi <- c("FOXO1","KLF15","CYP2E1","ACVR2A")
goi <- c("A1CF","TNRC6B","FOXO1","GPAM","ACVR2A","TF","CHD4")

```

# 2. Coverage
## Calculate cov
Bash code on farm to generate the total coverage across all samples in cohort
```{bash burdens, eval = F}

#Go to targeted directory, create folders
cd /lustre/scratch126/casm/team154pc/nb15/liver/data/NanoSeq/plate_9_DN942040D_exome_1/Analysis
mkdir -p cov cov/cut cov/source cov/jobLogs

#Unzip  and rename all nanoseq coverage files to single folder then
#Format files to remove trinuc context
for sample in `ls variants_subs`;
  do echo $sample
  cp variants_subs/$sample/results.cov.bed cov/source/$sample.bed
  sed 's/;/\t/g' cov/source/$sample.bed | cut -f 1-3,6 > cov/cut/$sample.bed
done

#Make summary files with coverage for all indivs at all positions
cd /lustre/scratch126/casm/team154pc/nb15/liver/data/NanoSeq/plate_9_DN942040D_exome_1/Analysis/cov/cut
bsub -G team154-grp -o ../jobLogs/sumCov.%J.out -e ../jobLogs/sumCov.%J.err -q normal -R 'select[mem>=1000] rusage[mem=1000] span[hosts=1]' -M 1000 'module load bedtools; bedtools unionbedg -i $(ls *.bed) -header -names $(ls *.bed) > ../summaryCov.bed'

#Get sum of coverage at each site
cd /lustre/scratch126/casm/team154pc/nb15/liver/data/NanoSeq/plate_9_DN942040D_exome_1/Analysis/cov

#These 2 lines need to be in a bash script so could for instance type 'nano getTotalCov.sh', paste these 2 lines in and then make it executable with 'chmod +x getTotalCov.sh'

#!/bin/bash
awk 'OFS="\t" {sum=0; for (col=4; col<=NF; col++) sum += $col; print $1, $2, $3, sum;}' summaryCov.bed | tail -n +2 > totalCov.bed

bsub -G team154-grp -q normal -o jobLogs/sumCov.%J.out -e jobLogs/sumCov.%J.err -R 'select[mem>=1000] rusage[mem=1000] span[hosts=1]' -M 1000 ./getTotalCov.sh

```

## Read into R
```{r cov}
#Coverage summaries
totTargCov <- read_tsv("~/Mounts/peter/liver/data/NanoSeq/plate_9_DN942040D_exome_1/Analysis/cov/totalCov.bed", col_types = "c-dd", col_names = c("chr", "pos", "cov"))

```

# 3. Plot
## Plotting functions
```{r cov}
# jitterLables and reAdjustLabels adapted from lolliplots package
jitterLables <- function(coor, xscale, lineW, weight=1.2){
    if(weight==1.2) {
      stopifnot("Please sort your inputs by start position"= 
                  order(coor)==1:length(coor))
    }
    if(weight<0.5) return(coor)
    stopifnot(length(xscale)==2)
    pos <- (coor-1)/(xscale[2]-1)
    # pos <- convertX(unit(coor01, "native"), "npc", valueOnly=TRUE)
    pos.diff <- diff(c(0, pos, 1))
    idx <- which(pos.diff < weight*lineW)
    if(length(idx)<1){
        return(coor)
    }
    if(all(idx %in% c(1, length(pos)+1))){
        return(coor)
    }
    idx.diff <- diff(c(-1, idx))
    idx.grp <- rle(idx.diff)
    idx.grp$values[idx.grp$values==1] <- length(pos) + 1:sum(idx.grp$values==1)
    idx.grp <- inverse.rle(idx.grp)
    idx.grp.w <- which(idx.grp>length(pos))-1
    idx.grp.w <- idx.grp.w[idx.grp.w>0]
    idx.grp[idx.grp.w] <- idx.grp[idx.grp.w+1]
    idx.grp <- split(idx, idx.grp)
    flag <- as.numeric(names(idx.grp))>length(pos)
    idx.grp.mul <- lapply(idx.grp[flag], function(.ele){
        c(.ele[1]-1, .ele)
    })
    idx.grp.sin <- lapply(idx.grp[!flag], function(.ele){
        lapply(as.list(.ele), function(.ele){c(.ele-1, .ele)})
    })
    idx.grp.sin <- unlist(idx.grp.sin, recursive = FALSE)
    idx.grp <- c(idx.grp.mul, idx.grp.sin)

    adj.pos <- lapply(idx.grp, function(.ele){
        .ele <- .ele[.ele>0 & .ele<=length(pos)]
        this.pos <- pos[.ele]
        names(this.pos) <- .ele
        if(length(this.pos)%%2==1){
            center <- ceiling(length(this.pos)/2)
        }else{
            center <- length(this.pos)/2 + .5
        }
        if(length(this.pos)>5){ ## too much, how to jitter?
            this.pos <- this.pos + 
                ((1:length(this.pos))-center) * (weight-.1) * 
                lineW/ceiling(log(length(this.pos), 5))
        }else{
            this.pos <- this.pos + 
                ((1:length(this.pos))-center) * (weight-.1) * lineW
        }
        this.pos
    })
    names(adj.pos) <- NULL
    adj.pos <- unlist(adj.pos)
    coor[as.numeric(names(adj.pos))] <- adj.pos*diff(xscale)+xscale[1]

    Recall(coor, xscale=xscale, lineW=lineW, weight=weight-0.2)
}

reAdjustLabels <- function(coor, lineW, xmax){
  # resort
  coor <- sort(coor)
  bins <- ceiling(1/lineW)
  pos <- (coor-1)/(xmax-1)
  # pos <- convertX(unit(coor, "native"), "npc", valueOnly=TRUE)
  pos.bin <- cut(pos, c(-Inf, (0:bins)*lineW, Inf), labels=0:(bins+1), right=FALSE)

  ## split the coors by into clusters
  ## give the clusters with more idx more spaces if there are spaces between clusters
  tbl <- table(pos.bin)
  if(all(tbl<2)) return(coor)
  tbl.len <- length(tbl)
  if(tbl.len<3) return(coor)
  loops <- 1000
  loop <- 1
  while(any(tbl==0) && any(tbl>1) && loop < loops){
    tbl.bk <- tbl
    for(i in order(tbl.bk, decreasing=TRUE)){
      if(tbl[i]>1 && tbl.bk[i]==tbl[i]){
        if(i==1){
          if(tbl[2]<tbl[1]){
            half <- sum(tbl[1:2])/2
            tbl[2] <- ceiling(half)
            tbl[1] <- floor(half)
          }
        }else{
          if(i==tbl.len){
            if(tbl[tbl.len]>tbl[tbl.len-1]){
              half <- sum(tbl[(tbl.len-1):tbl.len])/2
              tbl[tbl.len-1] <- ceiling(half)
              tbl[tbl.len] <- floor(half)
            }
          }else{
            if(tbl[i-1]<tbl[i+1]){
              ## i-1 and i should be balanced
              half <- sum(tbl[(i-1):i])/2
              tbl[i-1] <- floor(half)
              tbl[i] <- ceiling(half)
            }else{
              half <- sum(tbl[i:(i+1)])/2
              tbl[i] <- floor(half)
              tbl[i+1] <- ceiling(half)
            }
          }
        }
      }
    }
    loop <- loop + 1
  }
  coef <- unlist(lapply(tbl, function(.ele){
    if(.ele==0) return(0)
    .ele <- seq(from=0, to=1, length.out=.ele+1)
    (.ele[-length(.ele)] + .ele[-1])/2
  }))
  coef <- coef[coef!=0]
  coor <- (rep(as.numeric(names(tbl)), tbl) - 1 + coef) * lineW
  # coor <- convertX(unit(coor, "npc"), "native", valueOnly=TRUE)
  coor = ((xmax-1) * coor) + 1
  coor
}

# Covert genomic bp to coding sequence coordinate
getCDSindex <- function(intervals) {
  strand <- intervals$strand[1]
  allBP <- tibble()
  for(exon in 1:length(intervals$length)) {
    exonBP <- tibble(pos = seq(intervals$start[exon], intervals$end[exon]))
    allBP <- bind_rows(allBP, exonBP) 
  }
  # Forward strand standard
  if(strand == 1) {
  allBP <- allBP %>% 
      mutate(cdsPos = seq(1:sum(intervals$length)))    
  } else {
    #Reverse strand flip cds coords so all genes show left to right
    allBP <- allBP %>% 
      mutate(cdsPos = seq(from = sum(intervals$length), to = 1))  
  }
  return(allBP)
}

#For a mutation outside the CDS (splice or non-coding) set its CDS position as .5 off the nearest exon border
getCDSindexNonCDS <- function(cdsCoordsIndex, vars, strand) {
  nonCDSvars <- vars %>% filter(impact %in% c("Essential_Splice", "Non-coding") | (impact == "Indel" & is.na(ntchange)))
  closestCDS <- cdsCoordsIndex %>% dplyr::slice(rep(1:n(), each = dim(nonCDSvars)[1])) %>% 
    mutate(nonCDSpos = rep(nonCDSvars$POS, times = dim(cdsCoordsIndex)[1])) %>% 
    mutate(gap = nonCDSpos - pos) %>% 
    mutate(absGap = abs(gap)) %>% 
    group_by(nonCDSpos) %>% filter(absGap == min(absGap)) %>% 
    ungroup() %>% 
    #flip direction if reverse strand
    mutate(gap = gap * strand) %>% 
    #Set offset based on strand direction
    mutate(cdsPos = if_else(gap < 0, cdsPos - 0.5, cdsPos + 0.5)) %>% 
    select(POS = nonCDSpos, cdsPos)
  return(closestCDS)
}
```

## Lolliplot
Loops through selected genes, outputs file
```{r lolliplot}

#Plot Settings
#Set height of coverage y-axis
yMax = max(totTargCov$cov)+500

#True means drop non-coding (intronic) mutations, False means keep (will show at exon borders)
filterNonCoding = T
#Controls how spaced variant points will be. If you change pdf output size, you may need to adjust this
pointWidth = 0.017

## loop over genes
for(g in 1:length(goi)){
g_CDS <- RefCDS[[which(genes$gene_name==goi[g])]]

intervals_cds0 <- tibble(chr = g_CDS$chr,
                        strand = g_CDS$strand,
                        start = g_CDS$intervals_cds[,1],
                        end = g_CDS$intervals_cds[,2]) %>% 
  arrange(start) %>% 
  mutate(length = (end + 1) - start)

cdsCoordsIndex <- getCDSindex(intervals_cds0)

intervals_cds <- intervals_cds0 %>% 
  left_join(cdsCoordsIndex %>% select(start = pos, cdsStart = cdsPos), by = "start") %>% 
  left_join(cdsCoordsIndex %>% select(end = pos, cdsEnd = cdsPos), by = "end")

# Coverage
geneCov <- cdsCoordsIndex %>% 
  left_join(totTargCov %>% select(-chr), by = "pos") %>% 
  mutate(cov = replace_na(cov, 0))

# subset nanoseq_data to variants in the gene
# Can choose to filter non-coding variants or not
gene0 <- nanoseq_data %>% filter(CHROM == intervals_cds$chr[1] & POS >= min(intervals_cds %>% pull(start)) & POS <= max(intervals_cds %>% pull(end)))

gene <- gene0 %>% 
  left_join((cdsCoordsIndex %>% dplyr::rename(POS = pos) %>% bind_rows(getCDSindexNonCDS(cdsCoordsIndex, gene0, strand = intervals_cds$strand[1]))), by = "POS") %>% 
  arrange(cdsPos) %>% 
  drop_na(cdsPos)

  if(dim(gene)[1] == 0) {
    xMax <- max(c(intervals_cds$cdsEnd,intervals_cds$cdsStart)) + max(c(intervals_cds$cdsEnd,intervals_cds$cdsStart))*0.05
    xMin <- 1 - xMax*0.01
    pCov <- ggplot(geneCov, aes(x = cdsPos, y = cov)) + 
      #Coverage area
      geom_area(fill = '#ccece6') + ylab('Duplex_Cov') + 
      scale_x_continuous(expand = c(0, 0), limits = c(xMin, xMax)) +
      coord_cartesian(clip = "off") +
      scale_y_continuous(limits = c(0, yMax), expand = c(0, 0)) +
      ggtitle(goi[g]) +
      theme(
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA),
        panel.grid = element_blank(),
        panel.border = element_blank(),
        plot.margin = unit(c(0.5, 0.5, 0, 0.5), "cm"),
        panel.spacing = unit(c(0, 0, 0, 0), "cm"),
        axis.text.y = element_text(margin = margin(0,0.3,0,0.2, "cm")),
        axis.ticks = element_blank(),
        axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.line = element_blank(),
        plot.title = element_text(hjust = 0.5),
        axis.ticks.length.x = unit(0, "null"),
        legend.position = c(0.5, 0.85),
        legend.direction = "horizontal",
        legend.title = element_blank(),
        legend.key = element_rect(fill = NA)
      )

    pCDS <- ggplot(intervals_cds, aes(xmin = cdsStart, xmax = cdsEnd, ymin = 0, ymax = 1)) +
      geom_rect(fill = "#d0d1e6", colour = "black") +
      theme_void() + 
      xlab("CDS") +
      scale_x_continuous(expand = c(0, 0), limits = c(xMin,xMax)) +
      scale_y_continuous(expand = c(0, 0)) +
      theme(plot.margin = unit(c(0, 0.5, 0.5, 0.5), "cm"))

    ggarrange(pCov, pCDS, heights = c(15,1), nrow = 2, ncol = 1)

    filename <- paste0(outputPath, goi[g],"_lollipop.pdf")
    pdf(file = filename, height = 3, width = 11, onefile=FALSE)
    ggarrange(pCov, pCDS, heights = c(15,1), nrow = 2, ncol = 1)
    dev.off()
  }else {
  #Set colours for variant csq
  cols <- met.brewer("Archambault",5)
  gene$col <- NA
  gene$col[which(gene$impact=="Non-coding")] <- "gray100"
  gene$col[which(gene$impact=="Synonymous")] <- "gray76"
  gene$col[which(gene$impact=="Missense")] <- "#abd9e9"
  gene$col[which(gene$impact=="Nonsense")] <- "#fdae61"
  gene$col[which(gene$impact=="Indel")] <- cols[3]
  gene$col[which(gene$impact=="Essential_Splice")] <- "#fee090"
  gene$col[which(gene$impact=="Stop_loss")] <- cols[5]

  # Set label position with lolliplot functions
  lab.pos <- jitterLables(gene %>% arrange(cdsPos) %>% pull(cdsPos), 
                          xscale=c(1, max(cdsCoordsIndex %>% pull(cdsPos))), 
                          lineW=pointWidth)

  lab.pos <- reAdjustLabels(lab.pos, lineW=pointWidth, xmax = max(cdsCoordsIndex %>% pull(cdsPos)))

  variants <- gene %>% 
    mutate(labPos = lab.pos) %>% mutate(aachange = replace_na(aachange, ".")) %>%
    mutate(label = replace_na(label, " "))

  xMax <- max(c(intervals_cds$cdsEnd,intervals_cds$cdsStart)) + max(c(intervals_cds$cdsEnd,intervals_cds$cdsStart))*0.05

  xMin <- 1 - xMax*0.01

  pCov <- ggplot(geneCov, aes(x = cdsPos, y = cov)) + 
    #Coverage area
    geom_area(fill = '#ccece6') + ylab('Duplex_Cov') + 
    #Variant stems
    geom_segment(data = variants, aes(x=cdsPos, xend=cdsPos, y=0, yend= (0.125 * yMax))) +
    geom_segment(data = variants, aes(x=cdsPos, xend=labPos, y=(0.125 * yMax), yend=(0.25 * yMax))) +
    geom_segment(data = variants, aes(x=labPos, xend=labPos, y=(0.25 * yMax), yend=(0.325 * yMax))) +
    #Variant points
    geom_point(data = variants, aes(x = labPos, y = (0.325 * yMax), fill=col), shape=21,  color="black", size=5) +
    #Label aa change
    geom_text(data = variants, aes(label = name, x = labPos, y = (0.5 * yMax)), angle = 90) + 
    #Label clinvar
    geom_text(data = variants, aes(label = label, x = labPos, y = (0.325 * yMax)), size=3.5) +
    scale_fill_identity(guide = "legend", breaks = variants$col, labels = variants$impact) +
    scale_x_continuous(expand = c(0, 0), limits = c(xMin, xMax)) +
    coord_cartesian(clip = "off") +
    scale_y_continuous(limits = c(0, yMax), expand = c(0, 0)) +
    ggtitle(goi[g]) +
    theme(
      panel.background = element_rect(fill = "transparent", colour = NA),
      plot.background = element_rect(fill = "transparent", colour = NA),
      panel.grid = element_blank(),
      panel.border = element_blank(),
      plot.margin = unit(c(0.5, 0.5, 0, 0.5), "cm"),
      panel.spacing = unit(c(0, 0, 0, 0), "cm"),
      axis.text.y = element_text(margin = margin(0,0.3,0,0.2, "cm")),
      axis.ticks = element_blank(),
      axis.text.x = element_blank(),
      axis.title.x = element_blank(),
      axis.line = element_blank(),
      plot.title = element_text(hjust = 0.5),
      axis.ticks.length.x = unit(0, "null"),
      legend.position = c(0.5, 0.85),
      legend.direction = "horizontal",
      legend.title = element_blank(),
      legend.key = element_rect(fill = NA)
    )

  pCDS <- ggplot(intervals_cds, aes(xmin = cdsStart, xmax = cdsEnd, ymin = 0, ymax = 1)) +
    geom_rect(fill = "#d0d1e6", colour = "black") +
    theme_void() + 
    xlab("CDS") +
    scale_x_continuous(expand = c(0, 0), limits = c(xMin,xMax)) +
    scale_y_continuous(expand = c(0, 0)) +
    theme(plot.margin = unit(c(0, 0.5, 0.5, 0.5), "cm"))

  ggarrange(pCov, pCDS, heights = c(15,1), nrow = 2, ncol = 1)

  filename <- paste0(outputPath, goi[g],"_lollipop.pdf")
  pdf(file = filename, height = 3, width = 11, onefile=FALSE)

  ggarrange(pCov, pCDS, heights = c(15,1), nrow = 2, ncol = 1)

  dev.off()
  }
}

```

### 2024-08-07 infercnv/src/01_run_infercnv_all_fibroblasts.sh

@@ -1,5 +1,5 @@
 #!/bin/bash
-# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/ ; bsub -q week -M 2000 -R 'select[mem>=2000] span[hosts=1] rusage[mem=2000]' -J infercnv_all -o log "bash src/01_run_infercnv_all_fibroblasts.sh"
+# cd /lustre/scratch125/casm/team268im/at31/RA_som_mut/infercnv/ ; bsub -q week -M 10000 -R 'select[mem>=10000] span[hosts=1] rusage[mem=10000]' -J infercnv_all -o log "bash src/01_run_infercnv_all_fibroblasts.sh"
 
 # run infercnv on celltypes, running all individuals together
 # TODO: decide on cluster_by_groups value
@@ -34,7 +34,7 @@ nextflow run nf-infercnv \
   --mappings out/Zhang2023/all_fibroblasts/mappings.csv \
   --annotations out/Zhang2023/all_fibroblasts/annotations.tsv \
   --annotation_col cluster \
-  --analysis_mode subclusters \
+  --analysis_mode samples \
   --cluster_by_groups TRUE \
   --cutoff 0.1 \
   --window_length 151 \

### 2024-08-07 infercnv/src/gene_coverage_histogram_plotting_coding_farm_TwinsUK.R

@@ -1,0 +1,616 @@
+library("tidyverse")
+library("dndscv")
+library("scales")
+library("patchwork")
+
+args <- commandArgs(TRUE)
+
+gene_num <- as.numeric(args[1])
+datasets <- c("plate_014_SQPP-1106-C_targeted_buccal_v3","plate_017_SQPP-1640-T_targeted_buccal_v2","plate_018_SQPP-1728-D_targeted_buccal_v2","plate_019_SQPP-1734-B_targeted_buccal_v2",
+              "plate_020_SQPP-1737-E_targeted_buccal_v2","plate_021_SQPP-2297-F_targeted_buccal","plate_022_SQPP-2906-H_targeted_buccal","plate_024_SQPP_4022-L_targeted_buccal",
+              "plate_029_SQPP-13328-P_targeted_buccal","plate_035_SQPP-18868-M_targeted_buccal","plate_036_SQPP-18870-G_targeted_buccal","plate_038_SQPP-19909-I_targeted_buccal",
+              "plate_040_SQPP-20086-J_targeted_buccal","plate_042_SQPP-22755-I_targeted_buccal")
+
+twins_uk_sample_list <- read.table("/lustre/scratch126/casm/team268im/al28/TwinsUK_Manuscript/TwinsUKSampleList.csv", sep = ",", stringsAsFactors = F, header = F)
+twins_uk_sample_list <- twins_uk_sample_list$V1
+
+exclude_sample_list <- read.table("/lustre/scratch126/casm/team268im/al28/TwinsUK_Manuscript/SAMPLES2EXCLUDE.2023-11-15.tsv", sep = "\t", stringsAsFactors=F, header = T)
+twins_uk_sample_list <- twins_uk_sample_list[which(!(twins_uk_sample_list) %in% rownames(exclude_sample_list))]
+twins_uk_sample_list <- twins_uk_sample_list[which(!(grepl("Unknown",twins_uk_sample_list)))]
+
+buccal_genes <- read.table("/lustre/scratch126/casm/team268im/al28/TwinsUK_Manuscript/Sanger_TERT-v4_TE-95148282_hg19_highstringencyfilter_buccal_gene_list.tsv", sep = "\t", stringsAsFactors = F, header = T)
+gene <- buccal_genes$gene[gene_num]
+
[…]

### 2024-08-08 paper/reports/paper_figures.Rmd
move paper prep to separate subdir, integrate Matt's plotting functions

--- 2024-08-07 infercnv/reports/paper_figures.Rmd
@@
 ## LCM NanoSeq
+
+```{r}
+# dirs
+lh22_dir <- "/lustre/scratch125/casm/team268im/lh22/Rheumatoid_Arthritis_Paper/"
+nb15_dir <- "/lustre/scratch126/casm/team154pc/nb15/liver/"
+
+# functions and libraries
+source("bin/lolliplot.R")
+library(magrittr)
+library(ggplot2)
+
+# load the annotated mutations from nanoseq output
+load(file.path(lh22_dir, "Figures/Figure_2/LCM_NanoSeq_Overlap/GOI",
+               "pileup_muts.rda"))
+nano_muts <- nano_muts %>% tibble::as_tibble()
+
+# run dNdScv
+dndscv_in <- nano_muts %>% dplyr::select(sampleID, chr, pos, ref, mut)
+dndscv_out <- dndscv::dndscv(dndscv_in)
+
+# load refcds, add gene names
+data("refcds_hg19", package = "dndscv")
+names(RefCDS) <- RefCDS %>% purrr::map(~ .$gene_name)
+
+# combine annotated mutations with dnds output
+nanoseq_data <-
+  nano_muts %>%
+  dplyr::mutate(impact = dplyr::case_when(impact == "no-SNV" ~ "Indel",
+                                          is.na(impact) ~ "Non-coding",
+                                          TRUE ~ impact)) %>%
+  dplyr::select(sampleID, chr, pos, ref, mut, duplex_cov, impact, donor) %>%
+  dplyr::left_join(dndscv_out$annotmuts) %>%
+  # rename cols
+  dplyr::rename(CHROM = chr, POS = pos, REF = ref, ALT = mut)
+
+# plotting params
+filterNonCoding <- TRUE
+pointWidth <- 0.017
+
+# define genes of interest
+g <- c("NF1")
+
+# get gene info
+g_CDS <- RefCDS[[g]]
+
+#  get gene intervals
+intervals_cds0 <- tibble::tibble(
+  chr = g_CDS$chr,
+  strand = g_CDS$strand,
+  start = g_CDS$intervals_cds[, 1],
+  end = g_CDS$intervals_cds[, 2]
+) %>%
+  dplyr::arrange(start) %>%
+  dplyr::mutate(length = (end + 1) - start)
+
+# get genomic position to CDS position index
+cds_coords_index <- getCDSindex(intervals_cds0)
+
+# convert CDS intervals
+intervals_cds <-
+  intervals_cds0 %>%
+  dplyr::left_join(cds_coords_index %>%
+                     dplyr::select(start = pos, cdsStart = cdsPos),
+                   by = "start") %>%
+  dplyr::left_join(cds_coords_index %>%
+                     dplyr::select(end = pos, cdsEnd = cdsPos),
+                   by = "end")
+
+# subset nanoseq_data to variants in the gene
+gene0 <-
+  nanoseq_data %>%
+  dplyr::filter(CHROM == unique(intervals_cds$chr),
+                POS >= min(intervals_cds$start),
+                POS <= max(intervals_cds$end))
+
+# get coords outside the CDS
+noncds <-
+  cds_coords_index %>%
+  dplyr::rename(POS = pos) %>%
+  dplyr::bind_rows(getCDSindexNonCDS(cds_coords_index, gene0,
+                                     strand = unique(intervals_cds$strand)))
+
+gene <-
+  gene0 %>%
+  dplyr::left_join(noncds, by = "POS") %>%
+  dplyr::arrange(cdsPos) %>%
+  tidyr::drop_na(cdsPos)
+
+# set impact colours
+impact_palette <-
+  c("Non-coding" = "gray100",
+    "Synonymous" = "gray76",
+    "Missense" = "#abd9e9",
+    "Nonsense" = "#fdae61",
+    "Indel" = "pink",
+    "Essential_Splice" = "#fee090",
+    "Stop_loss" = "#bb0f12")
+gene <-
+  gene %>%
+  dplyr::mutate(impact = factor(impact, levels = names(impact_palette)),
+                col = impact_palette[impact])
+
+# set label positions
+lab_pos <-
+  jitterLabels(sort(gene$cdsPos),
+               xscale = c(1, max(cds_coords_index$cdsPos)),
+               lineW = pointWidth) %>%
+  reAdjustLabels(lineW = pointWidth, xmax = max(cds_coords_index$cdsPos))
+variants <-
+  gene %>%
+  dplyr::mutate(labPos = lab_pos,
+                aachange = tidyr::replace_na(aachange, "."))
+
+# xmax and xmin
+x_max <-
+  max(c(intervals_cds$cdsEnd, intervals_cds$cdsStart)) +
+  max(c(intervals_cds$cdsEnd, intervals_cds$cdsStart)) * 0.05
+x_min <- 1 - x_max * 0.01
+
+# plot variants
+p_var <-
+  variants %>%
+  dplyr::group_by(CHROM, POS, REF, ALT) %>%
+  dplyr::mutate(n_donors = dplyr::n_distinct(donor)) %>%
+  ggplot() +
+  # variant stems
+  geom_segment(aes(x = cdsPos, xend = cdsPos, y = 0,
+                   yend = (0.125 * max(lab_pos)))) +
+  geom_segment(aes(x = cdsPos, xend = labPos, y = (0.125 * max(lab_pos)),
+                   yend = (0.25 * max(lab_pos)))) +
+  geom_segment(aes(x = labPos, xend = labPos, y = (0.25 * max(lab_pos)),
+                   yend = (0.325 * max(lab_pos)))) +
+  # variant points
+  geom_point(aes(x = labPos, y = 0.325 * max(lab_pos), fill = impact,
+                 size = n_donors),
+             shape = 21, colour = "black", size = 5) +
+  # label aa change
+  # geom_text(aes(label = name, x = labPos, y = (0.5 * max(lab_pos))),
+  #           angle = 90) +
+  theme_void() +
+  scale_fill_manual(guide = "legend", values = impact_palette) +
+  scale_x_continuous(expand = c(0, 0), limits = c(x_min, x_max)) +
+  coord_cartesian(clip = "off") +
+  scale_y_continuous(limits = c(0, max(lab_pos)), expand = c(0, 0)) +
+  ggtitle(g)
+
+# plot CDS
+p_cds <-
+  intervals_cds %>%
+  ggplot(aes(xmin = cdsStart, xmax = cdsEnd, ymin = 0, ymax = 1)) +
+  geom_rect(fill = "#d0d1e6", colour = "black") +
+  theme_void() + 
+  xlab("CDS") +
+  scale_x_continuous(expand = c(0, 0), limits = c(x_min, x_max)) +
+  scale_y_continuous(expand = c(0, 0)) +
+  theme(plot.margin = unit(c(0, 0.5, 0.5, 0.5), "cm"))
+
+pdf("test.pdf", width = 20)
+cowplot::plot_grid(p_var, p_cds, ncol = 1, rel_heights = c(15, 1))
+dev.off()
+
+```
 

### 2024-08-08 infercnv/bin/lolliplot.R

@@ -1,0 +1,174 @@
+# jitterLabels and reAdjustLabels adapted from lolliplots package
+jitterLabels <- function(coor, xscale, lineW, weight = 1.2) {
+  if (weight == 1.2) {
+    stopifnot(
+      "Please sort your inputs by start position" =
+        order(coor) == 1:length(coor)
+    )
+  }
+  if (weight < 0.5) {
+    return(coor)
+  }
+  stopifnot(length(xscale) == 2)
+  pos <- (coor - 1) / (xscale[2] - 1)
+  # pos <- convertX(unit(coor01, "native"), "npc", valueOnly=TRUE)
+  pos.diff <- diff(c(0, pos, 1))
+  idx <- which(pos.diff < weight * lineW)
+  if (length(idx) < 1) {
+    return(coor)
+  }
+  if (all(idx %in% c(1, length(pos) + 1))) {
+    return(coor)
+  }
+  idx.diff <- diff(c(-1, idx))
+  idx.grp <- rle(idx.diff)
+  idx.grp$values[idx.grp$values == 1] <- length(pos) + 1:sum(idx.grp$values == 1)
+  idx.grp <- inverse.rle(idx.grp)
+  idx.grp.w <- which(idx.grp > length(pos)) - 1
+  idx.grp.w <- idx.grp.w[idx.grp.w > 0]
+  idx.grp[idx.grp.w] <- idx.grp[idx.grp.w + 1]
+  idx.grp <- split(idx, idx.grp)
+  flag <- as.numeric(names(idx.grp)) > length(pos)
+  idx.grp.mul <- lapply(idx.grp[flag], function(.ele) {
+    c(.ele[1] - 1, .ele)
+  })
+  idx.grp.sin <- lapply(idx.grp[!flag], function(.ele) {
+    lapply(as.list(.ele), function(.ele) {
+      c(.ele - 1, .ele)
+    })
+  })
+  idx.grp.sin <- unlist(idx.grp.sin, recursive = FALSE)
+  idx.grp <- c(idx.grp.mul, idx.grp.sin)
+
+  adj.pos <- lapply(idx.grp, function(.ele) {
+    .ele <- .ele[.ele > 0 & .ele <= length(pos)]
+    this.pos <- pos[.ele]
+    names(this.pos) <- .ele
+    if (length(this.pos) %% 2 == 1) {
+      center <- ceiling(length(this.pos) / 2)
+    } else {
[…]

### 2024-08-20 infercnv/reports/paper_figures.Rmd
using chatgpt to write new lolliplot function

@@
 
-# libraries
+# libraries and functions
+source("bin/lolliplot.R")
 library(magrittr)
@@
 
+# dirs
+lh22_dir <- "/lustre/scratch125/casm/team268im/lh22/Rheumatoid_Arthritis_Paper/"
+nb15_dir <- "/lustre/scratch126/casm/team154pc/nb15/liver/"
+
 # chunk options
@@
 knitr::opts_chunk$set(warning = FALSE, dpi = 300, message = FALSE)
+
+# plotting params
+filterNonCoding <- TRUE
+pointWidth <- 0.017
+mut_impact_palette <-
+  c("Non-coding" = "gray100",
+    "Synonymous" = "gray76",
+    "Missense" = "#abd9e9",
+    "Nonsense" = "#fdae61",
+    "Indel" = "pink",
+    "Essential_Splice" = "#fee090",
+    "Stop_loss" = "#bb0f12")
 ```
 
-# Datasets
+First, we load the NanoSeq output and run dNdScv to annotate the mutations and 
+to get a list of significantly positively selected genes.
 
-## LCM NanoSeq
+```{r load_nanoseq_muts}
+# load the annotated mutations from nanoseq output (`nano_muts`)
+load(file.path(lh22_dir, "Figures/Figure_2/LCM_NanoSeq_Overlap/GOI",
+               "pileup_muts.rda"))
+nano_muts <- nano_muts %>% tibble::as_tibble()
+
+# run dNdScv
+dndscv_in <- nano_muts %>% dplyr::select(sampleID, chr, pos, ref, mut)
+dndscv_out <- dndscv::dndscv(dndscv_in)
+
+# combine annotated mutations with dnds output
+nanoseq_data <-
+  nano_muts %>%
+  dplyr::mutate(
+    impact = dplyr::case_when(impact == "no-SNV" ~ "Indel",
+                              is.na(impact) ~ "Non-coding",
+                              TRUE ~ impact),
+    name = dplyr::case_when(impact == "Indel" ~
+      # names for indels
+      ntchange %>%
+        tidyr::replace_na(".") %>%
+        stringr::str_remove(".*-") %>%
+        stringr::str_replace("frshift", "FS") %>%
+        stringr::str_replace("inframe", "IF"),
+      # names for snvs
+      TRUE ~ tidyr::replace_na(aachange, "."))) %>%
+  dplyr::select(sampleID, chr, pos, ref, mut, duplex_cov, impact, donor, name) %>%
+  dplyr::left_join(dndscv_out$annotmuts) %>%
+  # rename cols
+  dplyr::rename(CHROM = chr, POS = pos, REF = ref, ALT = mut)
+```
+
+Now, we get coding sequence annotations for these genes of interest, including 
+the TSS, the bounds of exons, and the protein domains.
+
+```{r get_representative_domains, eval = F}
+# TODO: dplyr::filter(protein_representative == TRUE)
+accession <- "P21359"
+url <-
+  paste0(
+    "https://www.ebi.ac.uk/interpro/api/entry/all/protein/UniProt/",
+    accession,
+    "/?extra_fields=hierarchy%2Cshort_name&amp;format=tsv&amp;page_size=200")
+response <- httr::GET(url)
+content <- httr::content(response)
+domains <-
+  content$results %>%
+  purrr::map(function(domain) {
+    md <-
+      domain$metadata %>%
+      purrr::compact() %>%
+      tibble::as_tibble() %>%
+      dplyr::rename_with(~ paste0("metadata_", .), .cols = everything())
+    prot <-
+      domain$proteins[[1]] %>%
+      purrr::compact() %>%
+      tibble::as_tibble() %>%
+      tidyr::unnest_wider(entry_protein_locations) %>%
+      dplyr::mutate(fragments = fragments[[1]]) %>%
+      tidyr::unnest_wider(fragments, names_sep = "_") %>%
+      dplyr::rename_with(~ paste0("protein_", .), .cols = everything())
+    # extra <-
+    #   domain$extra_fields %>%
+    #   purrr::compact() %>%
+    #   tibble::as_tibble() %>%
+    #   tidyr::unnest_wider(hierarchy, names_sep = "_")
+    dplyr::bind_cols(md, prot)
+  }) %>%
+  dplyr::bind_rows()
+```
+
+```{r get_gene_info}
+# load refcds, add gene names
+data("refcds_hg19", package = "dndscv")
+names(RefCDS) <- RefCDS %>% purrr::map(~ .$gene_name)
+
+#  get gene info
+goi <-
+  list("NF1", "TNFRSF14") %>%
+  purrr::set_names() %>%
+  purrr::map(function(gene) {
+
+    # get gene info
+    g <- RefCDS[[gene]]
+
+    # get cds intervals
+    intervals_cds0 <- tibble::tibble(
+      chr = g$chr,
+      strand = g$strand,
+      start = g$intervals_cds[, 1],
+      end = g$intervals_cds[, 2]
+    ) %>%
+      dplyr::arrange(start) %>%
+      dplyr::mutate(length = (end + 1) - start)
+
+    # get genomic position to CDS position index
+    index <- getCDSindex(intervals_cds0)
+
+    # convert CDS intervals
+    intervals_cds <-
+      intervals_cds0 %>%
+      dplyr::left_join(dplyr::select(index, start = pos, cdsStart = cdsPos),
+                       by = "start") %>%
+      dplyr::left_join(dplyr::select(index, end = pos, cdsEnd = cdsPos),
+                       by = "end")
+
+    # get protein domains from InterPro REST API
+    domains <- getProteinDomains(gene)
+
+    # subset nanoseq_data to variants in the gene
+    gene_cds_muts <-
+      nanoseq_data %>%
+      dplyr::filter(CHROM == unique(intervals_cds$chr),
+                    POS >= min(intervals_cds$start),
+                    POS <= max(intervals_cds$end))
+
+    # get coords outside the CDS
+    gene_non_cds_muts <-
+      index %>%
+      dplyr::rename(POS = pos) %>%
+      dplyr::bind_rows(getCDSindexNonCDS(index, gene_cds_muts,
+                                         strand = unique(intervals_cds$strand)))
+
+    # get all muts in the gene
+    gene_muts <-
+      gene_cds_muts %>%
+      dplyr::left_join(gene_non_cds_muts, by = "POS") %>%
+      dplyr::arrange(cdsPos) %>%
+      tidyr::drop_na(cdsPos) %>%
+      # count number of patients with the mutation
+      dplyr::group_by(CHROM, POS, REF, ALT) %>%
+      dplyr::mutate(n_donors = dplyr::n_distinct(pid)) %>%
+      dplyr::ungroup()
+
+    # get label positions
+    lab_pos <-
+      jitterLabels(sort(gene_muts$cdsPos),
+                   xscale = c(1, max(intervals_cds$cdsEnd)),
+                   lineW = pointWidth) %>%
+      reAdjustLabels(lineW = pointWidth, xmax = max(intervals_cds$cdsEnd))
+
+    # prep data for plotting
+    p_dat <-
+      gene_muts %>%
+      dplyr::mutate(lab_pos = lab_pos,
+                    impact = factor(impact, levels = names(mut_impact_palette)),
+                    aachange = tidyr::replace_na(aachange, "."))
+
+    # get plotting limits
+    x_max <-
+      max(c(intervals_cds$cdsEnd, intervals_cds$cdsStart)) +
+      max(c(intervals_cds$cdsEnd, intervals_cds$cdsStart)) * 0.02
+    x_min <- 1 - x_max * 0.01
+    if (min(p_dat$lab_pos < x_min)) x_min <- 1 - x_max * 0.02
+    legend_pos <- c(0.5, 0.95)
+    y_max_count <- max(p_dat$n_donors) + 2
+    if (y_max_count > 10) y_max_count <- y_max_count + 2
+
+    # plot
+    p <-
+      ggplot(p_dat, aes(x = cdsPos, y = n_donors)) +
+      # variant stems
+      geom_segment(aes(x = cdsPos, xend = cdsPos, y = 0, yend = (0.2 * n_donors))) +
+      geom_segment(aes(x = cdsPos, xend = lab_pos, y = 0.2 * n_donors, yend = 0.3 * n_donors)) +
+      geom_segment(aes(x = lab_pos, xend = lab_pos, y = 0.3 * n_donors, yend = n_donors)) +
+      #Variant points
+      geom_point(aes(x = lab_pos, y = n_donors, fill=impact), shape=21,  color="black", size=5) +
+      #Label aa change
+      geom_text(aes(label = name, x = lab_pos, y = (n_donors + y_max_count/8)), angle = 90) + 
+      ggtitle(gene) +
+      # label clinvar
+      # geom_text(aes(label = label, x = lab_pos, y = (n_donors)), size = 3.5) +
+      scale_fill_manual(guide = "legend", values = mut_impact_palette) +
+      scale_x_continuous(expand = c(0, 0), limits = c(x_min, x_max)) +
+      ylab("Sperm samples with mutation") +
+      coord_cartesian(clip = "off") +
+      scale_y_continuous(limits = c(0, y_max_count), expand = c(0, 0),
+                         sec.axis = sec_axis(~ . * (200000/y_max_count),
+                         name = "Duplex Coverage")) +
+      theme(
+        panel.background = element_rect(fill = "transparent", colour = NA),
+        plot.background = element_rect(fill = "transparent", colour = NA),
+        panel.grid = element_blank(),
+        panel.border = element_blank(),
+        plot.margin = unit(c(0.5, 0.5, 0, 0.5), "cm"),
+        panel.spacing = unit(c(0, 0, 0, 0), "cm"),
+        axis.text.y = element_text(margin = margin(0,0.3,0,0.2, "cm")),
+        axis.ticks = element_blank(),
+        axis.text.x = element_blank(),
+        axis.title.x = element_blank(),
+        axis.line = element_blank(),
+        plot.title = element_text(hjust = 0.5),
+        axis.ticks.length.x = unit(0, "null"),
+        legend.position = legend_pos,
+        legend.direction = "horizontal",
+        legend.title = element_blank(),
+        legend.key = element_rect(fill = NA))
+
+    # plot mutations
+    p_muts <-
+      p_dat %>%
+      dplyr::group_by(CHROM, POS, REF, ALT) %>%
+      dplyr::mutate(n_donors = dplyr::n_distinct(donor)) %>%
+      ggplot() +
+      # variant stems
+      geom_segment(aes(x = cdsPos, xend = cdsPos, y = 0,
+                       yend = (0.125 * max(lab_pos)))) +
+      geom_segment(aes(x = cdsPos, xend = labPos, y = (0.125 * max(lab_pos)),
+                       yend = (0.25 * max(lab_pos)))) +
+      geom_segment(aes(x = labPos, xend = labPos, y = (0.25 * max(lab_pos)),
+                       yend = (0.325 * max(lab_pos)))) +
+      # variant points
+      geom_point(aes(x = labPos, y = 0.325 * max(lab_pos), fill = impact,
+                     size = n_donors),
+                shape = 21, colour = "black", size = 5) +
+      # label aa change
+      # geom_text(aes(label = name, x = labPos, y = (0.5 * max(lab_pos))),
+      #           angle = 90) +
+      theme_void() +
+      scale_fill_manual(guide = "legend", values = mut_impact_palette) +
+      scale_x_continuous(expand = c(0, 0), limits = c(x_min, x_max)) +
+      coord_cartesian(clip = "off") +
+      scale_y_continuous(limits = c(0, max(lab_pos)), expand = c(0, 0)) +
+      ggtitle(g)
+
+    # plot CDS
+    p_cds <-
+      intervals_cds %>%
+      ggplot(aes(xmin = cdsStart, xmax = cdsEnd, ymin = 0, ymax = 1)) +
+      geom_rect(fill = "#d0d1e6", colour = "black") +
+      theme_void() + 
+      xlab("CDS") +
+      scale_x_continuous(expand = c(0, 0), limits = c(x_min, x_max)) +
+      scale_y_continuous(expand = c(0, 0)) +
+      theme(plot.margin = unit(c(0, 0.5, 0.5, 0.5), "cm"))
+
+    pdf("test.pdf", width = 20)
+    print(p)
+    cowplot::plot_grid(p_muts, p_cds, ncol = 1, rel_heights = c(15, 1))
+    dev.off()
+
+    # return all
+    list(gene_muts = gene_muts,
+         intervals_cds = intervals_cds,
+         domains = domains,
+         index = index)
+  })
+```
+
+```{r plot_lolliplot}
+# set label positions
+lab_pos <-
+  jitterLabels(sort(gene$cdsPos),
+               xscale = c(1, max(cds_coords_index$cdsPos)),
+               lineW = pointWidth) %>%
+  reAdjustLabels(lineW = pointWidth, xmax = max(cds_coords_index$cdsPos))
+variants <-
+  gene %>%
+  dplyr::mutate(labPos = lab_pos,
+                aachange = tidyr::replace_na(aachange, "."))
+
+# xmax and xmin
+x_max <-
+  max(c(intervals_cds$cdsEnd, intervals_cds$cdsStart)) +
+  max(c(intervals_cds$cdsEnd, intervals_cds$cdsStart)) * 0.05
+x_min <- 1 - x_max * 0.01
+
+# plot variants
+p_var <-
+  variants %>%
+  dplyr::group_by(CHROM, POS, REF, ALT) %>%
+  dplyr::mutate(n_donors = dplyr::n_distinct(donor)) %>%
+  ggplot() +
+  # variant stems
+  geom_segment(aes(x = cdsPos, xend = cdsPos, y = 0,
+                   yend = (0.125 * max(lab_pos)))) +
+  geom_segment(aes(x = cdsPos, xend = labPos, y = (0.125 * max(lab_pos)),
+                   yend = (0.25 * max(lab_pos)))) +
+  geom_segment(aes(x = labPos, xend = labPos, y = (0.25 * max(lab_pos)),
+                   yend = (0.325 * max(lab_pos)))) +
+  # variant points
+  geom_point(aes(x = labPos, y = 0.325 * max(lab_pos), fill = impact,
+                 size = n_donors),
+             shape = 21, colour = "black", size = 5) +
+  # label aa change
+  # geom_text(aes(label = name, x = labPos, y = (0.5 * max(lab_pos))),
+  #           angle = 90) +
+  theme_void() +
+  scale_fill_manual(guide = "legend", values = mut_impact_palette) +
+  scale_x_continuous(expand = c(0, 0), limits = c(x_min, x_max)) +
+  coord_cartesian(clip = "off") +
+  scale_y_continuous(limits = c(0, max(lab_pos)), expand = c(0, 0)) +
+  ggtitle(g)
+
+# plot CDS
+p_cds <-
+  intervals_cds %>%
+  ggplot(aes(xmin = cdsStart, xmax = cdsEnd, ymin = 0, ymax = 1)) +
+  geom_rect(fill = "#d0d1e6", colour = "black") +
+  theme_void() + 
+  xlab("CDS") +
+  scale_x_continuous(expand = c(0, 0), limits = c(x_min, x_max)) +
+  scale_y_continuous(expand = c(0, 0)) +
+  theme(plot.margin = unit(c(0, 0.5, 0.5, 0.5), "cm"))
+
+pdf("test.pdf", width = 20)
+cowplot::plot_grid(p_var, p_cds, ncol = 1, rel_heights = c(15, 1))
+dev.off()
+
+```
 

### 2024-08-20 infercnv/src/chatgpt_lolliplot.R

@@ -1,0 +1,188 @@
+# Load the libraries
+library(httr)
+library(jsonlite)
+library(ggplot2)
+library(cowplot)
+library(magrittr)
+
+# Function to fetch exon and intron data from Ensembl (GRCh37) for all transcripts of a gene
+get_all_transcripts_data_grch37 <- function(gene_name) {
+  ensembl_url <-
+    paste0("https://grch37.rest.ensembl.org/lookup/symbol/homo_sapiens/",
+           gene_name, "?expand=1")
+  response <- GET(ensembl_url, content_type("application/json"))
+
+  if (status_code(response) != 200) {
+    stop("Failed to retrieve data from Ensembl API")
+  }
+
+  gene_data <- fromJSON(content(response, as = "text", encoding = "UTF-8"))
+
+  all_transcripts <- list()
+
+  names(gene_data$Transcript$Exon) <- gene_data$Transcript$id
+  for (transcript in names(gene_data$Transcript$Exon)) {
+    exons <- gene_data$Transcript$Exon[[transcript]]
+
+    if (nrow(exons) > 1) {
+      introns <- data.frame(
+        start = exons[-1, "end"] + 1,
+        end = exons[-length(exons$end), "start"] - 1,
+        Type = "Intron"
+      )
+    }
+
+    exons$Type <- "Exon"
+    exons <- exons[, c("start", "end", "Type")]
+    introns <- introns[, c("start", "end", "Type")]
+
+    transcript_data <- rbind(exons, introns)
+    transcript_data$transcript_id <- transcript
+
+    all_transcripts[[transcript]] <- transcript_data
+  }
+
+  out <- do.call(rbind, all_transcripts)
+
+  # Differentiate canonical transcript
+  canonical_transcript_id <-
+    gsub("\\..*", "", gene_data$canonical_transcript)
+  out$canonical <- 
+    out$transcript_id == canonical_transcript_id
+  canonical <- out %>%
+    dplyr::filter(canonical == TRUE) %>%
+    dplyr::mutate(Type = "Canonical")
+  out <- dplyr::bind_rows(out, canonical) %>%
+    dplyr::mutate(
+      Type = factor(Type, levels = c("Canonical", "Intron", "Exon")),
+      size = dplyr::case_match(Type, "Canonical" ~ 1, "Exon" ~ 0.9,
+                               "Intron" ~ 0.5))
+
+  # Add biotype information
+  out <-
+    gene_data$Transcript %>%
[…]
